---
title: "Spike-at-Zero Modelling with mfp2"
subtitle: 'Package version `r as.character(utils::packageVersion("mfp2"))`'
author: "Edwin Kipruto and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Spike-at-Zero Modelling with mfp2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-size: 15px;
  line-height: 1.5;
}

/* Justify narrative paragraphs */
p {
  text-align: justify;
  text-justify: inter-word;
}

/* Justify normal prose, but not code, tables, headings, or captions */
p, li {
  text-align: justify;
  line-height: 1.5;
}

h1, h2, h3, h4 {
  margin-top: 1.4em;
  margin-bottom: 0.6em;
  color: #1f77b4;
}

#TOC a {
  color: inherit;
  text-decoration: none;
}

pre {
  font-size: 13px;
  line-height: 1.4;
  overflow-x: auto;
}

code {
  font-size: 90%;
}

img {
  max-width: 100%;
  height: auto;
}

.figure {
  margin-top: 1.2em;
  margin-bottom: 1.2em;
}

.caption {
  font-size: 90%;
  color: #555555;
  text-align: left;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  dev = "png",
  dpi = 100
)

oldwidth <- getOption("width")
options(width = 100)

library(mfp2)
library(ggplot2)
library(patchwork)
library(survival)
```

# Introduction

Covariates in biomedical, epidemiological, environmental, and pharmacological studies may have a semi-continuous distribution. In this setting, a proportion of observations is concentrated at zero, while the remaining observations are continuously distributed over positive values. This distributional pattern is often described as a *spike at zero* (SAZ) or *semi-continuous variable* (Robertson et al., 1994; Royston and Sauerbrei, 2008; Becher et al., 2012; Lorenz et al., 2017). We use SAZ throughout this vignette.

In `mfp2`, the structural-zero component is defined by `x <= 0`, not only by
exact observed zeros. Any negative values are recoded to zero before the SAZ
eligibility check and component modelling, while positive values form the
continuous component.

Typical examples include alcohol consumption, tobacco exposure, and laboratory measurements restricted by a limit of detection or quantification. Here, zero may denote no alcohol intake, no tobacco exposure, or a value below the measurement threshold, while positive values describe the amount consumed, the intensity or duration of exposure, or the measured concentration above that limit (Becher et al., 2012; Lorenz et al., 2017).

Several simple strategies are commonly used for SAZ variables, but each
has limitations. Categorisation uses the zero or unexposed group as the
reference category. This is easy to interpret, but it leads to loss of
information, depends on arbitrary cut-points, and represents the
exposure-response relationship as a step function. Another strategy is
to include the SAZ variable as an ordinary continuous covariate, either
untransformed or after applying a transformation. This uses the
numerical scale of the variable, but it does not model the spike
explicitly. It treats zero as part of the same continuous
covariate-response curve as the positive values, which may be
inappropriate when the zero group is substantively distinct. For
fractional-polynomial transformations, this approach may also require
adding an arbitrary constant or shifting the variable before
transformation, because many FP powers are not defined at zero (Royston
and Sauerbrei, 2008; Becher et al., 2012; Lorenz et al., 2017).

A more sensible strategy is to represent the SAZ covariate by two
components: a binary component distinguishing zero from positive values,
and a continuous component describing how the outcome changes over the
positive range. Fractional polynomials can then be used to estimate a
functional form for the positive component without forcing the zero
values onto the same smooth curve.

The SAZ implementation in `mfp2` follows the modified FP-spike proposal
of Becher et al. (2012). Under this approach, the covariate is
represented by two components: a binary spike component distinguishing
zero from positive values, and an FP component fitted only to the
positive values. This avoids adding an arbitrary constant to the zero
values before FP transformation. Because SAZ is embedded in the
multivariable FP backfitting algorithm, more than one SAZ covariate can
be considered in the same model, together with ordinary continuous,
binary, and categorical covariates. During model selection, `mfp2`
decides separately for each requested SAZ covariate whether the final
model should retain both components, the binary component only, the
positive FP component only, or neither component.

Lorenz et al. (2017) compared several strategies for modelling SAZ
covariates. The table below summarises these options in the context of
`mfp2`, with the FP-spike approach following the modified proposal of
Becher et al. (2012).

| Strategy | Model idea | Main advantage | Main limitation |
|---|---|---|---|
| Categorisation | Treat zero as the reference category and split positive values into exposure categories. | Simple to communicate and useful as an exploratory analysis. | Cutpoints are arbitrary; the fitted effect is a step function; power is lost. |
| Linear | Include the original variable as an ordinary linear term. | Uses the continuous measurement directly. | Ignores the spike and assumes linearity on the model scale. |
| Standard FP | Apply ordinary FP modelling to the variable after making it positive. | Uses the continuous scale and can capture nonlinear effects. | Ignores the spike; any constant added to make zeros transformable may affect the result. |
| Linear plus spike indicator | Include a binary spike component and a linear term for the positive values. | Separates the zero group from the linear trend among positive values. | Still assumes linearity over the positive range and no simplification of the model components. |
| FP-spike | Include a binary spike component and use FP selection for the positive values. | Allows a separate contrast for the zero group and a nonlinear functional form over the positive range, with possible component simplification. | Requires adequate information in both components and adds model-selection uncertainty. |

SAZ modelling is most useful when `x = 0` identifies a meaningful group, such as no exposure or a value below a measurement threshold. The positive values should also be able to carry additional continuous and possibly nonlinear covariate-response information. In addition, both the zero component and the positive-value component should contain enough observations for reliable estimation of the model parameters.

SAZ should not be used simply because a continuous predictor contains a few zero values. If zero is only a boundary value, a coding artifact, or a consequence of measurement precision, ordinary preprocessing, shifting, scaling, or data cleaning may be more appropriate. Conversely, if the covariate is already binary, SAZ is not needed because the variable already represents a two-level effect.

# Relationship to `zero` and `catzero` handling

`mfp2()` provides three related options for covariates containing nonpositive
values. They should not be treated as interchangeable.

| Requested handling | Positive-value function | Binary nonpositive-component indicator | Component selection |
|---|---:|---:|---:|
| `zero` | Yes | No | No |
| `catzero` | Yes | Yes | No; both are retained while the term remains selected |
| `spike` | Yes | Yes | Yes; SAZ chooses both components or either component alone |

In the formula interface, request these options with
`fp(x, zero = TRUE)`, `fp(x, catzero = TRUE)`, or
`fp(x, spike = TRUE)`. In the matrix interface, use `zero_vars`,
`catzero_vars`, or `spike_vars`.

For all three options, original values less than or equal to zero belong to the
nonpositive component. The detailed sections below focus on `spike`, for which
both component eligibility and component selection are data dependent.

# Quick start: a worked example

This section presents a SAZ analysis using artificial data with a known data-generating structure. We simulate a covariate with a zero-group effect and a nonlinear positive-value effect. We then check the component sizes, fit the model, inspect the printed output, plot the fitted function, and make predictions. The artificial data set has a zero-group effect and a nonlinear positive-value effect, so the expected SAZ decision is to retain both components.

## Simulate data

```{r quick-start-data}
set.seed(2101)

## Sample size for the artificial data set.
n <- 1000

## Generate a semi-continuous SAZ covariate.
## About 30% of observations are exactly zero; the rest are positive.
is_zero <- rbinom(n, size = 1, prob = 0.30) == 1
x_pos <- rgamma(n, shape = 2.5, rate = 0.60) + 0.10
x_saz <- ifelse(is_zero, 0, x_pos)

## Define the true SAZ contribution.
## The zero group has its own effect, and the positive values have
## a nonlinear log-shaped relationship with the outcome.
f_saz <- numeric(n)
f_saz[x_saz == 0] <- 1.00
f_saz[x_saz > 0] <- 0.70 * log(x_saz[x_saz > 0])

## Generate a Gaussian outcome from the true SAZ contribution plus noise.
y <- f_saz + rnorm(n, sd = 0.75)

## Create a data frame.
dat_both <- data.frame(y = y, x_saz = x_saz)
```

## Check component sizes

Before fitting an SAZ model, inspect the number of observations in the
zero and positive components. Both need at least 10% of the sample under
the default `min_saz_component_prop = 0.10`. This example shows that the SAZ variable has 31% zeros and 69% positive values.

```{r saz-helper-functions}
saz_component_summary <- function(data, vars) {
  out <- lapply(vars, function(v) {
    x <- data[[v]]
    positive <- x > 0
    data.frame(
      variable = v,
      n_zero = sum(!positive),
      prop_zero = mean(!positive),
      n_positive = sum(positive),
      prop_positive = mean(positive),
      distinct_positive = length(unique(x[positive])),
      row.names = NULL
    )
  })
  do.call(rbind, out)
}

saz_component_summary(dat_both, "x_saz")
```

## Specify and fit the model

In the formula interface, SAZ modelling is requested inside `fp()` using
`spike = TRUE`. In the matrix interface, the corresponding argument is
`spike_vars`, which specifies the variables to evaluate as SAZ candidates.

```{r quick-start-fit}
fit_both <- mfp2(
  y ~ fp(x_saz, select = 0.05, alpha = 0.05, spike = TRUE),
  data = dat_both,
  family = "gaussian",
  criterion = "pvalue",
  min_saz_component_prop = 0.10,
  verbose = FALSE
)

## Print the selected model and SAZ component decision.
fit_both
```

The `selected` column indicates whether the variable was retained by the
model-selection procedure at the specified `select` level. In this example,
the `saz_decision` column shows that `cont + binary` was selected. This means
that the final model retains the two-component SAZ representation: a binary
zero-component indicator and a positive-value FP function. This is consistent
with the data-generating structure, which included both a zero-group effect
and a positive-value effect.

The binary zero-component indicator contributes to the fitted linear predictor
for observations with `x_saz = 0`. The selected positive-value FP function
contributes to the fitted linear predictor for observations with `x_saz > 0`,
allowing the fitted outcome to vary over the positive range of `x_saz`. The
selected FP power is `0`, matching the power used to simulate the positive-value
part of the covariate-response function. Therefore, the selected positive-value
function is logarithmic. Thus, the zero group is represented separately, while
observations with positive values follow a fitted log function.

### Variable-specific FP settings

Parameters supplied inside `fp()` take precedence over the corresponding
global defaults for that variable. For example, in

```r
mfp2(
  y ~ fp(exposure, spike = TRUE, df = 2) + age,
  data = dat,
  family = "gaussian",
  df = 4
)
```

the variable `exposure` is restricted to a maximum FP1 form because `df = 2`
is specified inside `fp()`. The global setting `df = 4` does not change the
maximum FP degree for `exposure`. Instead, it applies to `age`, which has no
variable-specific `fp()` settings in the formula.


## Plot the fitted SAZ function

The fitted SAZ function can be displayed with the standard `plot()` method for
an `mfp2` object. For this example, the selected SAZ decision is
`cont + binary`, so the plot shows two fitted components: the fitted value for
the zero group at `x_saz = 0` and the fitted positive-value function over the
range where `x_saz > 0`.

```{r quick-start-plot, fig.cap="Fitted spike-at-zero function for the two-component example."}
p_both <- plot(
  fit_both,
  terms = "x_saz",
  partial_only = TRUE,
  linewidth = 1
)

p_both[[1]]
```

The horizontal axis is the original scale of `x_saz`. The vertical axis is the
fitted contribution of the selected SAZ term on the linear predictor scale. In
this Gaussian example, that scale is the mean outcome scale. For logistic,
Poisson, and Cox models, the corresponding scales would be the log-odds,
log-mean, and log-hazard scales, respectively.

In this fitted model, the zero group is represented separately by the binary
zero-component indicator. Observations with positive values follow the selected
positive-value FP function. Because the selected FP power is `0`, the fitted
positive-value function is logarithmic.

Confidence intervals in the plot are conditional on the final selected model.
They reflect uncertainty in the estimated coefficients for this selected
two-component SAZ model, but not the additional uncertainty introduced by
selecting the FP power, or the SAZ component structure.

## Prediction

Prediction uses the transformation metadata stored in the fitted object,
including shifts, scales, selected powers, centering constants, and SAZ
component decisions. This ensures that new observations are transformed in the
same way as the data used to fit the model. Supply the SAZ variable on its
original raw scale; do not manually create the binary zero-component indicator
or the positive-value FP terms before calling `predict()`.

```{r quick-start-predict, eval=FALSE}
newdat <- data.frame(
  x_saz = c(0, 0.5, 1, 2, 4, 8)
)

predict(fit_both, newdata = newdat, type = "link")
```

The prediction at `x_saz = 0` uses the fitted zero-component contribution.
Predictions for `x_saz > 0` use the selected positive-value FP function. With
`type = "link"`, predictions are returned on the linear predictor scale. In
this Gaussian model with the identity link, this is also the fitted mean
outcome scale.

In a multivariable analysis, the formula can include additional covariates in
the usual way. They are omitted from the quick-start fit only to keep the
output focused on the SAZ decision.

```{r multivariable-saz-formula, eval=FALSE}
y ~ fp(x_saz, spike = TRUE) +   fp(age, df = 4) + fp(bmi, df = 4) + sex
```

For prediction from a multivariable model, `newdata` must contain values for
all variables included in the fitted model.

# The two-stage SAZ algorithm

The SAZ procedure has two stages. Stage 1 chooses the positive-component
function or drops the current SAZ variable. Stage 2 then decides which of three
Stage 2 candidate representations should be retained in the final model.

## Stage 1: choose the positive-component function or drop the SAZ variable

Stage 1 evaluates the current SAZ variable using the full two-component
representation. Suppressing other covariates, this is

\[
\eta_i = \alpha + \gamma Z_i + f^+(x_i).
\]

In a multivariable fit, any other currently retained covariate terms are held
fixed. The purpose of Stage 1 is to decide whether the SAZ variable should be
retained at all and, if so, which functional form should be used for the
positive component.

For an active SAZ variable, the non-null candidate representations include the
binary zero-component term together with a linear, FP, or ACD-related
positive-component function, depending on `df` and `acdx`. The null candidate
omits the current variable entirely.

With `criterion = "pvalue"`, Stage 1 uses the usual MFP closed-test
function-selection procedure for the positive-component function. The difference
from ordinary MFP selection is that the binary zero-component term is included
while the positive-component function is being selected. For Gaussian models,
setting `ftest = TRUE` uses F-tests rather than chi-square likelihood-ratio
tests in the p-value based procedure.

Assuming that FP2 is the most complex allowed function for the positive
component, the first comparison is between the best FP2 positive-component
function, together with the binary zero-component term, and the null model that
omits the current variable entirely. If this comparison is not significant at
the chosen `select` level, the SAZ variable is dropped and Stage 2 is not
needed.

If the SAZ variable is retained, the procedure then checks whether a simpler
positive-component function is sufficient. The best FP2 representation with the
binary zero-component term is compared with the corresponding linear
positive-component representation with the binary zero-component term. If this
comparison is not significant, the positive-component function is simplified to
linear. If it is significant, the procedure then compares the best FP2
representation with the best FP1 representation, again keeping the binary
zero-component term in both representations. If FP2 is not significantly better
than FP1, the FP1 form is selected; otherwise, the FP2 form is retained.

With `criterion = "aic"` or `criterion = "bic"`, the same candidate
representations are compared using the selected information criterion rather
than the closed-test sequence. Under the FP2 assumption above, these are the
null representation, the linear positive-component function plus the binary
zero-component term, the best FP1 positive-component function plus the binary
zero-component term, and the best FP2 positive-component function plus the
binary zero-component term. The candidate with the smallest AIC or BIC is
selected. If the selected candidate is the null representation, the SAZ variable
is dropped and Stage 2 is not needed. Otherwise, the selected two-component
representation is passed to Stage 2.

## Stage 2: choose the retained SAZ representation

Stage 2 is run only if Stage 1 retains the SAZ variable. The selected
positive-component function \(f^+(x_i)\) from Stage 1 is treated as fixed for
this component-selection step. Stage 2 then compares three Stage 2 candidate
representations:

\[
M_{BP}: \quad \eta_i = \alpha + \gamma Z_i + f^+(x_i),
\]

\[
M_P: \quad \eta_i = \alpha + f^+(x_i),
\]

\[
M_B: \quad \eta_i = \alpha + \gamma Z_i.
\]

Here \(B\) denotes the binary zero-component term \(\gamma Z_i\), and \(P\)
denotes the positive-component function \(f^+(x_i)\). Thus, \(M_{BP}\) retains
both components, \(M_P\) retains only the positive-component function, and
\(M_B\) retains only the binary zero-component term.

The package prints these final representations using the labels `cont + binary`,
`continuous only`, and `binary only`. In the notation here, `continuous only`
means that \(M_P\) is retained: the positive-component function remains in the
model and the binary zero-component term is dropped.

With `criterion = "pvalue"`, the full representation \(M_{BP}\) is compared with
each reduced representation. The comparison of \(M_P\) with \(M_{BP}\) assesses
whether the binary zero-component term is needed. This is because \(M_P\) is
obtained from \(M_{BP}\) by removing \(\gamma Z_i\). If this comparison is
significant, removing the binary zero-component term makes the model
significantly worse, so the binary zero-component term is retained. If it is not
significant, the binary zero-component term can be removed.

The comparison of \(M_B\) with \(M_{BP}\) assesses whether the
positive-component function is needed. This is because \(M_B\) is obtained from
\(M_{BP}\) by removing \(f^+(x_i)\). If this comparison is significant, removing
the positive-component function makes the model significantly worse, so the
positive-component function is retained. If it is not significant, the
positive-component function can be removed.

The final decision follows directly from these two comparisons:

| Evidence from Stage 2 tests | Final representation | Package label |
|---|---|---|
| Binary zero-component term needed and positive-component function needed | \(M_{BP}\) | `cont + binary` |
| Binary zero-component term not needed, positive-component function needed | \(M_P\) | `continuous only` |
| Binary zero-component term needed, positive-component function not needed | \(M_B\) | `binary only` |
| Neither component clearly needed relative to \(M_{BP}\) | choose between \(M_P\) and \(M_B\) by the larger log-likelihood | `continuous only` or `binary only` |

The last case means that both reduced representations are adequate relative to
the full two-component representation. In that case, `mfp2` does not retain
\(M_{BP}\); it chooses between \(M_P\) and \(M_B\), selecting the reduced
representation with the larger log-likelihood.

With `criterion = "aic"` or `criterion = "bic"`, Stage 2 does not use hypothesis
tests. Instead, \(M_{BP}\), \(M_P\), and \(M_B\) are compared directly, and the
representation with the smallest selected information criterion is retained.

# Model formulation

Let \(x\) be a semi-continuous covariate. In an active SAZ term, non-positive values are assigned to the zero component. Define the binary zero-component indicator as

\[
Z_i = I(x_i \le 0).
\]

The positive component is represented by a function fitted only to observations with \(x_i > 0\). To make this explicit, let \(\phi_j(x_i; p)\) denote the \(j\)th selected basis function for the positive values of \(x\). The corresponding positive-component basis function is

\[
\phi_j^+(x_i; p) =
\begin{cases}
0, & x_i \le 0, \\
\phi_j(x_i; p), & x_i > 0.
\end{cases}
\]

The selected positive-component function is then written as

\[
f^+(x_i)
=
\sum_{j=1}^{J} \beta_j \phi_j^+(x_i; p),
\]

where \(J\) is the number of retained basis functions and \(\beta_j\) are the corresponding regression coefficients. For an FP1 positive component, \(J = 1\), while for an FP2 positive component, \(J = 2\). Thus, \(f^+(x_i)\) denotes the positive-component contribution to the linear predictor.

Suppressing other covariates for notational clarity, the full two-component SAZ representation is

\[
M_{BP}: \quad
\eta_i = \alpha + \gamma Z_i + f^+(x_i),
\]

where \(\eta_i\) is the model linear predictor and \(\alpha\) is the intercept. The interpretation of \(\eta_i\) depends on the regression family:

- Gaussian: contribution to the mean outcome;
- binomial: contribution to log-odds;
- Poisson: contribution to log mean;
- Cox: contribution to log hazard.

The coefficient \(\gamma\) is the regression coefficient for the binary zero-component indicator \(Z_i\). Because \(Z_i = 1\) when \(x_i \le 0\) and \(Z_i = 0\) when \(x_i > 0\), the term \(\gamma Z_i\) contributes \(\gamma\) to the linear predictor for observations in the zero component and contributes nothing for observations with positive values.

The positive-component function \(f^+(x_i)\) behaves in the opposite way. It is zero for observations with \(x_i \le 0\), and for observations with \(x_i > 0\) it contributes the selected linear or fractional-polynomial function of the positive values. Therefore, the zero component and the positive component are represented by separate terms:

\[
\gamma Z_i
\quad \textrm{and} \quad
f^+(x_i).
\]

This prevents zero observations from being forced onto the function fitted to the positive values.

# More examples

A useful way to illustrate SAZ modelling is to use data with a known
data-generating mechanism. The examples below continue with one SAZ candidate,
`x_saz`, so that the printed `mfp2` output remains easy to inspect. The table
summarises the examples used in this vignette.

| Example | Distributional feature of `x_saz` | True data-generating role | Intended modelling lesson |
|---|---|---|---|
| Both components | about 30% zeros | zero-group effect and nonlinear positive-value effect | SAZ may retain both components |
| Binary only | about 40% zeros | zero-group effect, but no additional positive-value effect | SAZ may retain the binary zero-component indicator only |
| Positive-value only | about 30% zeros | nonlinear effect over the positive range, with no additional zero-group effect | SAZ may retain the positive-value FP component only |
| Too few zeros | about 3% zeros | no intended SAZ effect | SAZ handling may be reset when there are too few zeros |
| Too few positives | about 94% zeros | no intended SAZ effect | SAZ handling may be reset when there are too few positive observations |

The examples use a Gaussian outcome to keep the code and interpretation
straightforward. The same SAZ ideas apply to binomial, Poisson, and Cox models;
only the linear predictor scale changes.

## Binary zero-component indicator only

In this example, the data-generating mechanism includes a zero-group effect but
no additional effect of the positive values. Thus, observations with `x_saz = 0`
have a different fitted contribution from observations with `x_saz > 0`, but
among observations with positive values, the magnitude of `x_saz` does not
contribute further to the outcome.

```{r artificial-binary-only}
set.seed(2102)

## Sample size for the artificial data set.
n <- 1000

## Generate a semi-continuous SAZ covariate.
## About 40% of observations are exactly zero; the rest are positive.
is_zero <- rbinom(n, size = 1, prob = 0.40) == 1
x_pos <- rgamma(n, shape = 2.5, rate = 0.60) + 0.10
x_saz <- ifelse(is_zero, 0, x_pos)

## Define the true SAZ contribution.
## Only the zero group has a separate effect. Among observations with
## positive values, the actual value of x_saz has no effect on the outcome.
f_saz <- numeric(n)
f_saz[x_saz == 0] <- 1.50

## Generate a Gaussian outcome from the zero-group effect plus noise.
y <- f_saz + rnorm(n, sd = 0.75)

## Keep only the outcome and the SAZ covariate.
dat_binary_only <- data.frame(y = y, x_saz = x_saz)

## Check that both the zero and positive components are large enough
## for SAZ modelling under the default threshold.
saz_component_summary(dat_binary_only, "x_saz")

## Request SAZ modelling for x_saz.
## Because the data-generating mechanism contains only a zero-group effect,
## the expected result is selection of the binary zero-component indicator only.
fit_binary_only <- mfp2(
  y ~ fp(x_saz, spike = TRUE),
  data = dat_binary_only,
  family = "gaussian",
  criterion = "pvalue",
  select = 0.05,
  alpha = 0.05,
  min_saz_component_prop = 0.10,
  verbose = FALSE
)

## Print the selected model and SAZ component decision.
fit_binary_only
```

In this example, the final SAZ decision is `binary`. This is identical to the
data-generating structure, in which `x_saz` affects the outcome only through
membership in the zero group and not through variation among the positive
values. The model therefore retains only the binary zero-component indicator
and does not retain a positive-value FP function. Thus, `x_saz` contributes one
fitted level for observations with `x_saz = 0` and another fitted level for
observations with `x_saz > 0`. Among observations with positive values, the
fitted contribution does not depend on the magnitude of `x_saz`.


## Positive-component function only

In this example, the data-generating structure includes a nonlinear effect over
the positive range of `x_saz`, but no additional zero-component effect. The
expected SAZ decision is therefore to retain the positive-component function
and not retain the binary zero-component term.

```{r artificial-positive-only}
set.seed(2103)

## Sample size for the artificial data set.
## A slightly larger sample and lower noise are used here to make the
## positive-component-only decision stable in this teaching example.
n <- 1500

## Generate a semi-continuous SAZ covariate.
## About 30% of observations are exactly zero; the rest are positive.
is_zero <- rbinom(n, size = 1, prob = 0.30) == 1
x_pos <- runif(n, min = 0.05, max = 10)
x_saz <- ifelse(is_zero, 0, x_pos)

## Define the true SAZ contribution.
## There is no separate zero-component effect. The outcome changes only over
## the positive range of x_saz, following a square-root relationship.
f_saz <- numeric(n)
f_saz[x_saz > 0] <- 1.20 * sqrt(x_saz[x_saz > 0])

## Generate a Gaussian outcome from the positive-component function plus noise.
y <- f_saz + rnorm(n, sd = 0.35)

## Keep only the outcome and the SAZ covariate.
dat_positive_only <- data.frame(y = y, x_saz = x_saz)

## Check that both the zero and positive components are large enough
## for SAZ modelling under the default threshold.
saz_component_summary(dat_positive_only, "x_saz")

## Request SAZ modelling for x_saz.
## Because the data-generating structure contains a positive-component function
## but no additional zero-component effect, the expected result is selection of
## the positive-component function only.
fit_positive_only <- mfp2(
  y ~ fp(x_saz, spike = TRUE),
  data = dat_positive_only,
  family = "gaussian",
  criterion = "pvalue",
  select = 0.05,
  alpha = 0.05,
  min_saz_component_prop = 0.10,
  verbose = FALSE
)

## Print the selected model and SAZ component decision.
fit_positive_only
```

In this example, the final SAZ decision is `continuous only`. In the notation
used above, this corresponds to the Stage 2 representation \(M_P\): the
positive-component function is retained and the binary zero-component term is
dropped. This matches the data-generating structure, in which the outcome
changes over the positive range of `x_saz`, but there is no additional
zero-component effect.

Observations with `x_saz = 0` remain in the model, but no separate coefficient
is estimated for membership in the zero component. With the selected power
`0.5`, the retained positive-component basis function can be written as

\[
\phi_1^+(x_{\mathrm{saz}}; 0.5) =
\begin{cases}
0, & x_{\mathrm{saz}} = 0, \\
x_{\mathrm{saz}}^{0.5}, & x_{\mathrm{saz}} > 0.
\end{cases}
\]

The retained positive-component function is therefore

\[
f^+(x_{\mathrm{saz}})
=
\beta_1 \phi_1^+(x_{\mathrm{saz}}; 0.5).
\]

Thus, in this one-variable Gaussian model,

\[
\eta
=
\beta_0 + f^+(x_{\mathrm{saz}}).
\]

For observations with `x_saz = 0`, \(f^+(x_{\mathrm{saz}}) = 0\), so the fitted
linear predictor is \(\beta_0\). For observations with `x_saz > 0`, the fitted
linear predictor is

\[
\beta_0 + \beta_1 x_{\mathrm{saz}}^{0.5}.
\]

The selected FP power is `0.5`, matching the square-root function used to
simulate the positive-component part of the covariate-response function.

```{r positive-only-plot, fig.cap="Fitted spike-at-zero function for the positive-component-only example."}
p_positive_only <- plot(
  fit_positive_only,
  terms = "x_saz",
  partial_only = TRUE,
  linewidth = 1
)

p_positive_only[[1]]
```
# Information-criterion based SAZ selection

The same SAZ model can be fitted with AIC or BIC instead of the p-value
based procedure. In that case, `mfp2` compares the candidate SAZ models
by the selected information criterion. BIC penalises complexity more
strongly than AIC and may therefore select a simpler functional form or
component structure.

```{r aic-bic, eval=FALSE}
## Fit the SAZ model using AIC instead of p-value based closed testing.
fit_both_aic <- mfp2(
  y ~ fp(x_saz, spike = TRUE),
  data = dat_both,
  family = "gaussian",
  criterion = "aic",
  min_saz_component_prop = 0.10,
  verbose = FALSE
)

## Fit the SAZ model using BIC.
fit_both_bic <- mfp2(
  y ~ fp(x_saz, spike = TRUE),
  data = dat_both,
  family = "gaussian",
  criterion = "bic",
  min_saz_component_prop = 0.10,
  verbose = FALSE
)

## Print the fitted models.
fit_both_aic
fit_both_bic
```

# Eligibility of SAZ

SAZ modelling is intended for semi-continuous covariates with two empirically meaningful components:

1. a zero component, and  
2. a positive continuous component.

The method is useful only when both components contain enough observations to support separate modelling. If one component is too small, the SAZ decomposition is unstable: the zero indicator may be estimated from too few zero observations, or the positive-value function may be estimated from too few positive observations.

For this reason, `mfp2()` applies eligibility checks before fitting the SAZ procedure. A requested spike variable is retained as a SAZ candidate only if it passes these checks.

## Component size requirements

Let

\[
p_0 = n^{-1}\sum_i I(x_i \le 0)
\]

be the proportion of observations in the zero component, and let

\[
p_+ = n^{-1}\sum_i I(x_i > 0)
\]

be the proportion of observations in the positive component. A requested SAZ variable remains eligible only if

\[
p_0 \ge m \quad \textrm{and} \quad p_+ \ge m,
\]

where \(m\) is `min_saz_component_prop`.

The default value is

```r
min_saz_component_prop = 0.10
```

so at least 10 percent of observations must be in each component. With the default threshold, a variable with 3 percent zeros and 97 percent positives is not eligible for SAZ, and neither is a variable with 94 percent zeros and 6 percent positives.

This rule is symmetric: SAZ requires enough information in both parts of the covariate distribution.

Two common reasons for automatic spike reset are:

1. **Too few zero observations.**  
   The variable is essentially continuous, and the zero component is too sparse to estimate as a separate binary effect.

2. **Too few positive observations.**  
   The variable is dominated by zeros, and the positive-value function has too little empirical support.

A requested spike variable is also reset if the variable is binary. A binary variable is already represented by two levels, so adding a separate zero-component indicator would be redundant or non-identifiable.

If a requested spike variable fails the SAZ eligibility checks, `mfp2()` resets the spike flag before df and shift/scale preprocessing. The variable then keeps only non-spike handling that was explicitly requested by the user. With `spike = TRUE` alone, no zero-component indicator and no spike-implied zero handling are retained. Consequently, if the variable is still eligible for nonlinear FP modelling, automatic shifting is applied as for an ordinary FP covariate, so original zeros are moved to positive shifted values before FP transformation. If the variable is restricted to a linear term (`df = 1`), its shift is set to zero.


## Examples of automatic spike reset

The following examples show the two directions in which the component-size rule can fail. In both cases, SAZ is requested by the user, but the request is reset because one component is below `min_saz_component_prop`.

These examples are designed to demonstrate eligibility behaviour, not to demonstrate a selected SAZ effect. The simulated outcome is therefore generated independently of the SAZ candidate.

### Too few zero observations

In this example, only about 3 percent of observations are exactly zero. With the default eligibility threshold of 10 percent, the zero component is too small for SAZ modelling.

The requested spike handling is therefore reset before fitting. The variable is not decomposed into a binary zero-component term and a positive-component function; instead, it follows the applicable non-spike preprocessing and model-selection path.

```{r artificial-too-few-zeros}
set.seed(2105)

## Sample size for the artificial data set.
n <- 1000

## Generate a semi-continuous covariate with too few zeros for SAZ modelling.
## Only about 3% of observations are exactly zero, which is below the default
## minimum component proportion of 10%.
is_zero <- rbinom(n, size = 1, prob = 0.03) == 1
x_pos <- rgamma(n, shape = 2.5, rate = 0.60) + 0.10
x_saz <- ifelse(is_zero, 0, x_pos)

## Generate a Gaussian outcome with no true association with x_saz.
## This example is designed to demonstrate automatic reset due to too few zeros,
## not to demonstrate a selected SAZ effect.
y <- rnorm(n, sd = 0.75)

## Keep only the outcome and the SAZ candidate.
dat_few_zeros <- data.frame(y = y, x_saz = x_saz)

## Check the zero and positive component sizes before fitting the model.
## The zero component should be below min_saz_component_prop = 0.10.
saz_component_summary(dat_few_zeros, "x_saz")

## Request SAZ modelling for x_saz.
## Because the zero component is too small, mfp2 should reset the spike request
## under the default component-size rule.
fit_few_zeros <- mfp2(
  y ~ fp(x_saz, spike = TRUE),
  data = dat_few_zeros,
  family = "gaussian",
  criterion = "pvalue",
  select = 0.05,
  alpha = 0.05,
  min_saz_component_prop = 0.10,
  verbose = FALSE
)

## Print the selected model. The output should indicate that spike handling
## was not retained for x_saz.
fit_few_zeros
```

When only a very small fraction of observations are zero, the zero component is not well represented. Estimating a separate binary zero-component effect would be driven by too few observations and may be unstable.

Under the default rule, `mfp2()` therefore resets the spike request. In the printed model summary, the `spike_decision` column shows `not SAZ` for `x_saz`. This means that `x_saz` is not treated as a retained SAZ covariate: no binary zero-component term \(\gamma Z_i\) is estimated, and no positive-component function \(f^+(x_i)\) is fitted as part of a SAZ decomposition. The variable instead proceeds through the non-spike preprocessing and selection rules. Because no explicit `zero` or `catzero` handling was requested in this example, the reset variable follows ordinary non-spike preprocessing: since nonlinear FP modelling is allowed, the original zero values are shifted to positive values before FP transformations are evaluated.

### Too few positive observations

In this example, about 94 percent of observations are zero, leaving only about 6 percent positive observations. With the default eligibility threshold of 10 percent, the positive component is too small for SAZ modelling.

The problem here is the opposite of the previous example. There are many zeros, but too few positive values to support a reliable positive-value FP function.

```{r artificial-too-few-positives}
set.seed(2106)

## Sample size for the artificial data set.
n <- 1000

## Generate a semi-continuous covariate with too few positive values
## for SAZ modelling. About 94% of observations are zero, leaving only
## about 6% positive observations, which is below the default minimum
## component proportion of 10%.
is_zero <- rbinom(n, size = 1, prob = 0.94) == 1
x_pos <- rgamma(n, shape = 2.5, rate = 0.60) + 0.10
x_saz <- ifelse(is_zero, 0, x_pos)

## Generate a Gaussian outcome with no true association with x_saz.
## This example is designed to demonstrate automatic reset due to too few
## positive observations, not to demonstrate a selected SAZ effect.
y <- rnorm(n, sd = 0.75)

## Keep only the outcome and the SAZ candidate.
dat_few_positives <- data.frame(y = y, x_saz = x_saz)

## Check the zero and positive component sizes before fitting the model.
## The positive component should be below min_saz_component_prop = 0.10.
saz_component_summary(dat_few_positives, "x_saz")

## Request SAZ modelling for x_saz.
## Because the positive component is too small, mfp2 should reset the spike
## request under the default component-size rule.
fit_few_positives <- mfp2(
  y ~ fp(x_saz, spike = TRUE),
  data = dat_few_positives,
  family = "gaussian",
  criterion = "pvalue",
  select = 0.05,
  alpha = 0.05,
  min_saz_component_prop = 0.10,
  verbose = FALSE
)

## Print the selected model. The output should indicate that spike handling
## was not used for x_saz because the positive component is too small.
fit_few_positives
```

When almost all observations are zero, the positive-value function is estimated from too few observations. This is especially problematic for fractional polynomial selection, because the shape of the positive component would be determined by a sparse subset of the data.

The default reset protects the analysis from over-interpreting a weakly supported positive range. As in the previous example, the spike request is not retained, and the variable is handled using its non-spike specification.

## Positive-component support and FP complexity

Passing the component-size requirement is necessary for SAZ, but it does not automatically mean that a complex positive-value function can be estimated.

For ordinary MFP modelling, the maximum fractional polynomial degrees of freedom are adjusted when a variable has few distinct values. SAZ applies the same principle to the positive component of a spike variable rather than to the full variable including zeros.

This distinction matters because the repeated zero values do not provide information about the shape of the positive-value function. The positive component must have enough distinct positive values to support the requested FP complexity.

For a retained SAZ variable, `mfp2()` limits the positive-component complexity according to the number of distinct positive values:

- **1--3 values:** linear only (`df = 1`)
- **4--5 values:** FP1 at most (`df = min(df, 2)`)
- **6 or more values:** requested `df` retained

Thus, a spike variable with many zeros and only a few distinct positive values may still be eligible for SAZ, but the positive component will be restricted to a simpler functional form.

This rule avoids fitting nonlinear FP functions to a positive component with too little support. The zero observations are still handled by the binary zero-component term, while the positive-component function is only made as flexible as the positive data can support.

# References

Becher, H., Lorenz, E., Royston, P. and Sauerbrei, W. (2012).
Analysing covariates with spike at zero: a modified FP procedure and
conceptual issues. *Biometrical Journal*, 54(5), 686--700.

Lorenz, E., Jenkner, C., Sauerbrei, W. and Becher, H. (2017). Modeling
variables with a spike at zero: examples and practical recommendations.
*American Journal of Epidemiology*, 185(8), 650--660.

Royston, P. and Altman, D. G. (1994). Regression using fractional
polynomials of continuous covariates: parsimonious parametric modelling.
*Applied Statistics*, 43, 429--467.

Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A
Pragmatic Approach to Regression Analysis Based on Fractional
Polynomials for Modelling Continuous Variables*. John Wiley & Sons.

Royston, P., Sauerbrei, W. and Becher, H. (2010). Modelling continuous
exposures with a spike at zero: a new procedure based on fractional
polynomials. *Statistics in Medicine*, 29, 1219--1227.

```{r cleanup, include=FALSE}
options(width = oldwidth)
```
