---
title: "ACD Transformations with mfp2"
author: "Edwin Kipruto, Michael Kammer, Patrick Royston, and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{ACD Transformations with mfp2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-size: 15px;
  line-height: 1.6;
}

/* Colour section headings in the main text only */
h1, h2, h3, h4 {
  margin-top: 1.4em;
  margin-bottom: 0.6em;
  color: #1f77b4;
}

/* Keep TOC links unchanged */
#TOC a {
  color: inherit;
  text-decoration: none;
}

/* Style ordinary links in the document body */
body > p a,
body > div:not(#TOC) a,
.main-container a {
  color: #1f77b4;
  text-decoration: none;
}

body > p a:hover,
body > div:not(#TOC) a:hover,
.main-container a:hover {
  color: #0b4f8a;
  text-decoration: underline;
}

pre {
  font-size: 13px;
  line-height: 1.4;
  overflow-x: auto;
}

code {
  font-size: 90%;
}

img {
  max-width: 100%;
  height: auto;
}

.figure {
  margin-top: 1.2em;
  margin-bottom: 1.2em;
}

.caption {
  font-size: 90%;
  color: #555555;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center"
)

oldwidth <- getOption("width")
options(width = 100)

library(mfp2)
library(ggplot2)
library(patchwork)
library(survival)
library(formatR)
```

# Overview

Royston (2014) proposed the approximate cumulative distribution (ACD)
transformation as a smooth covariate-rank transformation for use in regression
models with sigmoid dose-response functions. The transformation was motivated by
the need to represent relationships that are singly or doubly asymptotic, or
that contain a sigmoid-shaped component.

Standard flexible functions, such as fractional polynomials (FPs) and restricted
cubic splines, can be useful for many nonlinear relationships. However, they may
struggle to represent curves with one or two asymptotes. The ACD transformation
addresses this problem by first compressing the tails of the observed covariate
distribution through a scaled-rank transformation and then approximating the
empirical cumulative distribution function (ECDF) by a smooth function.

The `mfp2` package implements ACD transformations for continuous covariates. In
`mfp2`, ACD modelling corresponds to the MFPA extension of the standard MFP
algorithm for variables selected by the user. This extension allows the original
covariate `x` and its ACD transformation to be considered together within a
fractional polynomial model-selection framework:

$$
A_x = \operatorname{ACD}(x).
$$

In the multivariable setting, the ACD extension follows the same backfitting
principle as MFP. For variables selected for ACD assessment, the usual function
selection procedure (FSP) is replaced by a function-selection procedure with ACD
transformation (FSPA). Variables not selected for ACD assessment are handled by
the standard MFP procedure.

The ACD transformation can be used in both interfaces of `mfp2()`:

- in the matrix interface, by specifying the variable name in the `acdx`
  argument;
- in the formula interface, by setting `acdx = TRUE` inside `fp()`.

The variable representing the ACD transformation of a covariate `x` is named
`A_x` internally in `mfp2`.

# Definition of the ACD transformation

Let the observed values of a continuous covariate `X` be denoted by

$$
x_1, \ldots, x_n.
$$

The ACD transformation proceeds in three steps.

## Inverse normal transformation of ranks

First, the rank of each observation is transformed to an approximately standard
normal scale:

$$
z_i =
\Phi^{-1}
\left(
\frac{\operatorname{rank}(x_i) - 0.5}{n}
\right),
$$

where

$$
\Phi^{-1}
$$

is the inverse standard normal cumulative distribution function. If all `x_i`
values are distinct, the transformed values `z_i` resemble expected standard
normal order statistics.

## Power-linear approximation

Second, a one-term fractional polynomial regression is fitted to approximate
`z_i` as a smooth function of `x_i`:

$$
\hat{z}_i =
\hat{\beta}_0 +
\hat{\beta}_1 (x_i + \text{shift})^p.
$$

Here, `p` is the best-fitting FP1 power. The power `p` is selected from the usual
FP power set:

$$
S = \{-2, -1, -0.5, 0, 0.5, 1, 2, 3\}.
$$

Ordinary least squares is used to estimate

$$
\hat{\beta}_0
\quad \text{and} \quad
\hat{\beta}_1.
$$

A power of 0 corresponds to the natural logarithm. If all observed `x_i` values
are positive, the shift is zero. If some values are nonpositive, a shift is
used to ensure that all transformed values are positive:

$$
x_i + \text{shift} > 0.
$$

## Back-transformation to the interval `(0, 1)`

Third, the fitted values are transformed back to the interval `(0, 1)` using the
standard normal cumulative distribution function:

$$
\operatorname{ACD}(x_i)
=
a_i
=
\Phi(\hat{z}_i)
=
\Phi
\left[
\hat{\beta}_0 +
\hat{\beta}_1 (x_i + \text{shift})^p
\right].
$$

The result is a smooth approximation to the ECDF of `X`. Importantly, after this
smoothing step, the transformed value $\operatorname{ACD}(x_i)$ is a smooth 
function of the original covariate value $x_i$ and no longer depends directly on
its observed rank.

# Interpretation

The ACD transformation maps `X` to an approximately uniform scale on `(0, 1)`.
When a regression model includes a linear effect of `ACD(X)`, the expected
outcome changes smoothly from its value at the lower end of the distribution of
`X` to its value at the upper end.

For example, in a model of the form

$$
E(Y) = \beta_0 + \beta_1 \operatorname{ACD}(X),
$$

the ACD transformation maps the observed distribution of `X` approximately onto
the interval `(0, 1)`. At the lower end of the observed distribution,
$\operatorname{ACD}(X) \approx 0$,

so that

$$
E(Y \mid \operatorname{ACD}(X) \approx 0)
\approx
\beta_0 + \beta_1 \times 0
=
\beta_0.
$$

At the upper end of the observed distribution, $\operatorname{ACD}(X) \approx 1$,

so that

$$
E(Y \mid \operatorname{ACD}(X) \approx 1)
\approx
\beta_0 + \beta_1 \times 1
=
\beta_0 + \beta_1.
$$

Therefore, the difference between the predicted values at the upper and lower
ends of the observed distribution is

$$
(\beta_0 + \beta_1) - \beta_0 = \beta_1.
$$

Thus, $\beta_1$ represents the range of predicted values across the observed
distribution of `X`.

Because the mapping from `X` to `ACD(X)` is S-shaped, it changes slowly in the
tails of the distribution and more rapidly in the central range. Consequently, a
linear effect of `ACD(X)` typically produces a sigmoid-shaped relationship
between `X` and `Y`.

This type of relationship may be difficult to represent adequately using
standard FP functions alone, particularly when the curve is singly or doubly
asymptotic.

The ACD transformation may also reduce the influence of extreme covariate values.
By construction, `ACD(X)` has an approximately uniform distribution on `(0, 1)`.
Extreme values on the original scale of `X` are therefore compressed into the
tails of the ACD scale.

# ACD in the fractional polynomial family

Royston and Sauerbrei (2016) extended the MFP approach by considering both the
original covariate `x` and its ACD transformation simultaneously:

$$
A_x = \operatorname{ACD}(x).
$$

To reduce instability and overfitting, the most complex model is restricted to 
one-term FP functions of $x$ and $A_x$. Specifically, the most complex
ACD-extended model has the form

$$
\beta_1 x^{p_1}
+
\beta_2 A_x^{p_2}.
$$

This model is denoted

$$
FP1(p_1, p_2).
$$

The powers `p1` and `p2` are selected from the usual FP power set:

$$
S = \{-2, -1, -0.5, 0, 0.5, 1, 2, 3\},
$$

where a power of 0 denotes the natural logarithm. All 64 ($8 \times 8 = 64$) 
combinations of the two powers are considered during model selection.

This family provides a flexible way to model sigmoid-like effects while
remaining within the general framework of fractional polynomial model selection.
It also permits standard FP1 or linear functions when the ACD component is not
needed.

# Function-selection procedure for ACD (FSPA)

When a variable is selected for ACD assessment, `mfp2()` uses the FSPA. The
procedure starts from the most complex candidate ACD-extended model,

$$
FP1(p_1, p_2),
$$

and compares it with a set of simpler nested models, as described in the
[closed test procedure](#FSPA). The aim is to select the simplest function that
is adequately supported by the data. The procedure considers six nested model 
families as shown below:

```{r acd-model-table, echo=FALSE, results='asis'}
acd_models <- data.frame(
  Model = c("M1", "M2", "M3", "M4", "M5", "M6"),
  Notation = c(
    "$FP1(p_1, p_2)$",
    "$FP1(p_1, \\cdot)$",
    "$FP1(\\cdot, p_2)$",
    "$FP1(1, \\cdot)$",
    "$FP1(\\cdot, 1)$",
    "$FP1(\\cdot, \\cdot)$"
  ),
  Function = c(
    "$\\beta_1 x^{p_1} + \\beta_2 A_x^{p_2}$",
    "$\\beta_1 x^{p_1}$",
    "$\\beta_2 A_x^{p_2}$",
    "$\\beta_1 x$",
    "$\\beta_2 A_x$",
    "Null model"
  ),
  Interpretation = c(
    "Most complex allowed ACD-extended function",
    "Standard FP1 function of x",
    "FP1 function of ACD(x)",
    "Linear function of x",
    "Linear function of ACD(x)",
    "x omitted"
  ),
  check.names = FALSE
)

knitr::kable(
  acd_models,
  align = c("l", "l", "l", "l"),
  escape = FALSE,
  caption = "Six nested model families considered by the FSPA."
)
```
The dot $\cdot$ indicates that the corresponding term is omitted.

## Closed test procedure {#FSA}

Selection among the six sub-models is performed using a closed test procedure.
The first test is performed at the significance level specified by `select`,
whereas the remaining tests are performed at the significance level specified by
`alpha`.

The sequence of tests is as follows:

1. Compare model M1 with model M6 on 4 degrees of freedom. If the test is not
   significant, omit `x`. Otherwise, continue to Step 2.

2. Compare model M1 with model M4 on 3 degrees of freedom. If the test is not
   significant, select model M4 and stop. Otherwise, continue to Step 3.

3. Compare model M1 with model M2 on 2 degrees of freedom. If the test is not
   significant, select model M2 and stop. Otherwise, continue to Step 4.

4. Compare model M1 with model M3 on 2 degrees of freedom. If the test is
   significant, model M1 cannot be simplified; select model M1 and stop.
   Otherwise, continue to Step 5.

5. Compare model M3 with model M5 on 1 degree of freedom. If the test is
   significant, select model M3. Otherwise, select model M5.

The result is the selection of one of the six models. The ordering of these tests
is intentional. The FSPA favours a linear or standard FP1 function when such a
function provides an adequate fit. ACD-based models are selected only when the
simpler standard functions do not fit sufficiently well relative to the full
ACD-extended model.

This follows the general philosophy of MFP: the selected model should be as
simple as possible, and additional complexity should be supported by improved
fit.

## Information criteria

As an alternative to p-value-based selection, the candidate models M1 to M6 can
be compared using AIC or BIC. When `criterion = "aic"` or `criterion = "bic"`,
the model with the smallest information criterion is selected.

# Simulation example

The following simulation creates a sigmoid relationship between a continuous
covariate `x` and a continuous outcome `y`.

```{r simulate-acd}
set.seed(54)

n <- 500
x <- matrix(rnorm(n), ncol = 1, dimnames = list(NULL, "x"))

sigmoid <- function(x) {
  1 / (1 + exp(-1.7 * x))
}

y <- as.numeric(20 * sigmoid(x) + rnorm(n, mean = 0, sd = 0.5))
```

The true relationship between `x` and `y` is sigmoid-shaped. This provides a
simple setting in which the ACD extension is expected to be useful.

# Fitting an ACD model

## Matrix interface

We first fit the model using the matrix interface. The variable `x` is specified
in the `acdx` argument, indicating that the ACD extension should be considered
for this covariate.

```{r fit-acd-matrix}
fit_acd_matrix <- mfp2(
  x, y,
  acdx = "x",
  verbose = FALSE
)

fit_acd_matrix$fp_terms
```

## Formula interface

The same model can be fitted using the formula interface by setting
`acdx = TRUE` inside `fp()`.

```{r fit-acd-formula}
dat <- data.frame(y = y, x = as.numeric(x))

fit_acd_formula <- mfp2(
  y ~ fp(x, acdx = TRUE),
  data = dat,
  verbose = FALSE
)

fit_acd_formula$fp_terms
```

# Comparison with standard MFP

For comparison, we fit a standard MFP model without the ACD extension.

```{r fit-standard-mfp}
fit_standard <- mfp2(
  y ~ fp(x, acdx = FALSE),
  data = dat,
  verbose = FALSE
)

fit_standard$fp_terms
```

A standard FP function may select a flexible FP2 model, but it may still be less
suitable for representing a clearly sigmoid-shaped relationship. The ACD
extension is designed for situations where such shapes are scientifically
plausible and supported by the data.

# Plotting the fitted functions

The fitted functions can be inspected using `fracplot()`.

```{r plot-acd, fig.cap="ACD model compared with a standard MFP model.", fig.width=8, fig.height=4}
p1 <- fracplot(fit_acd_matrix, terms = "x")[[1]] +
  ggplot2::ylab("Partial predictor + residuals") +
  ggplot2::ggtitle("ACD model")

p2 <- fracplot(fit_standard, terms = "x")[[1]] +
  ggplot2::ylab("Partial predictor + residuals") +
  ggplot2::ggtitle("Standard MFP model")

patchwork::wrap_plots(p1, p2, ncol = 2)
```

In this simulated example, the ACD model is expected to follow the sigmoid shape
more naturally than the standard MFP model.

# Practical remarks

ACD transformations are useful when a sigmoid-shaped relationship is
scientifically plausible. They may also reduce the influence of extreme
covariate values, because the observed covariate distribution is mapped to an
approximately uniform scale on `(0, 1)`.

The ACD extension should be regarded as an additional modelling option within the
MFP framework, not as a replacement for standard MFP modelling.

# References

Royston, P. (2014). A smooth covariate rank transformation for use in regression
models with a sigmoid dose-response function. *Stata Journal*, 14, 329--341.

Royston, P. and Sauerbrei, W. (2016). mfpa: Extension of mfp using the ACD
covariate transformation for enhanced parametric multivariable modeling.
*Stata Journal*, 16, 72--87.

Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A
Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for
Modelling Continuous Variables*. John Wiley & Sons.

```{r cleanup, include=FALSE}
options(width = oldwidth)
```