---
title: "Multivariable Fractional Polynomials with Extensions"
author:
- Edwin Kipruto
- Michael Kammer
- Patrick Royston
- Willi Sauerbrei
date: "`r format(Sys.time(), '%B %d, %Y')`"
bibliography: assets/mfp_refs.bib
link-citations: true
output:
  html_document:
    self_contained: false
    fig_caption: yes
    toc: yes
    toc_depth: 3
    number_sections: yes
    df_print: paged
  word_document:
    fig_caption: yes
    toc: yes
    toc_depth: 3
    number_sections: yes
  pdf_document:
    fig_caption: yes
    toc: yes
    toc_depth: 3
editor_options:
  markdown:
    wrap: sentence
vignette: >
  %\VignetteIndexEntry{Multivariable Fractional Polynomials with Extensions}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
   \usepackage[utf8]{inputenc}
---

```{r include=FALSE}
# the code in this chunk enables us to truncate the print output for each
# chunk using the `out.lines` option
# save the built-in output hook
hook_output <- knitr::knit_hooks$get("output")

# set a new output hook to truncate text output
knitr::knit_hooks$set(output = function(x, options) {
  if (!is.null(n <- options$out.lines)) {
    x <- xfun::split_lines(x)
    if (length(x) > n) {
        
      # truncate the output
      x <- c(head(x, n), "....\n")
    }
    x <- paste(x, collapse = "\n")
  }
  hook_output(x, options)
})
```

<!-- justify text in the entire document-->
<style>
body {
  text-align: justify;
}

figcaption {
  text-align: justify;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introduction to mfp2 package

`mfp2` is a package that selects the MFP model. In addition, it has the ability to model a sigmoid relationship between `x` and an outcome variable `y` using the ACD transformation proposed by Royston (2016). The package offers three options for variable and function selection: p-value, Akaike information criterion (AIC), and Bayesian information criterion (BIC). Furthermore, it provides functions for prediction and plotting. Currently, the package implements linear, logistic, Poisson, and Cox regression models. However, the package is designed in such a way that it can easily incorporate other generalized linear models or parametric survival models.

The main function, `mfp2()`, implements both MFP and MFP with ACD transformation.
It offers two interfaces for input data. The first interface allows direct input of the predictor matrix `x` and the outcome vector `y`. The second interface uses a formula object in conjunction with a data.frame, similar to the `glm()` function
with slight modifications. Both interfaces are equivalent in terms of functionality.


The authors of `mfp2` are Edwin Kipruto, Michael Kammer, Patrick Royston, and Willi Sauerbrei, with contribution from Gregory Steiner and Georg Heinze. The R package is maintained by Edwin Kipruto, while the STATA version of `mfp` is maintained by Patrick Royston. 

This vignette describes basic usage of `mfp2` in R. There are additional vignettes available that will further enhance your understanding on the mfp2 package.

* ["Introduction to Multivariable Fractional Polynomial"](https://github.com/EdwinKipruto/mfp2/tree/main/inst/docs/mfp.html) provides an overview of multivariable fractional polynomials.


## Estimation algorithm
The estimation algorithm employed in `mfp2` sequentially processes the predictors using a back-fitting approach. It calculates the p-values of each predictor 
using the likelihood ratio test, assuming linearity. Subsequently, the 
predictors are arranged based on these p-values. By default, the predictors are arranged in order of decreasing statistical significance. This ordering aims to prioritize modeling relatively important variables before less important ones. 
This approach may help mitigate potential challenges in model fitting arising
from collinearity or more generally, the presence of "concurvity" among the predictors (stata??). Although alternative options for predictor ordering are available, we prefer the default option.

If a predictor contains nonpositive values, the program by default shifts the location of the predictor x to ensure positivity. In addition, it scales the 
shifted predictor before the first cycle of the algorithm. For more information on shifting and scaling, please refer to section xx.

At the initial cycle, the best-fitting FP function for the first variable (after ordering) is determined, with all the other variables assumed to be linear. The functional form (but not the estimated regression
coefficients) is kept, and the process is repeated for the other variables. The
first iteration concludes when all the variables have been processed in this way.
The next cycle is similar, except that the functional forms from the initial cycle are retained for all variables except the one currently being processed


A variable whose functional form is prespecified to be linear is tested
for exclusion within the above procedure when its nominal p-value is less than 1
or argument `keep = FALSE`; otherwise, it is included.
Updating of FP functions and candidate variables continues until the functions
and variables included in the overall model do not change (convergence). 
Convergence is usually achieved within 1–5 cycles.


## Installation
To install the `mfp2` package, enter the following command in the R console:
```{r, eval=FALSE}
install.packages("mfp2")
```


## Quick Start
The purpose of this section is to provide users with a comprehensive 
understanding of the `mfp2` package. We will provide a concise overview of its 
key functions and resulting outputs. We will delve into each function in detail, highlighting their specific use. This will provide users with a deeper understanding of the package's functionality. The package includes a built-in dataset that is specifically designed for analysis within the `mfp2` framework. 

To begin, let's load the `mfp2` package:
```{r}
library(mfp2)
```

## Linear Regression
The default `family` in `mfp2` package is `Gaussian`, which fits a Gaussian 
linear model. In this section, we will demonstrate how to fit this model. We will use the prostate cancer data (Stamey et al., 1989) included
in our package. The dataset contains seven predictors (six continuous variables
and one binary variable) and a continuous outcome variable (log prostate-specific antigen (lpsa)) of 97 patients with prostate cancer. Our aim is to determine whether non-linear functional relationships exist between the predictors and
the outcome variable.

Load the `prostate` dataset from the `mfp2` package and display the first few 
rows of the dataset  

```{r out.lines = 10}
data("prostate")
head(prostate)
x <- as.matrix(prostate[,2:8])
y <- as.numeric(prostate$lpsa)
```
The command loads a dataframe from the R data archive since the `mfp2` package 
is already loaded. We create a matrix `x` and a numeric vector `y` from the dataframe. 

### Fitting MFP Models Using Default and Formula Interface

The default interface of `mfp2()` requires a matrix of predictors `x` and a 
numeric vector of response `y` for continuous outcomes. If you have one predictor, make sure you convert it into a matrix with a single column.

We demonstrate how to fit the Gaussian linear model using the default interface 
of `mfp2()` with default parameters.

```{r, eval=FALSE}
fit <- mfp2(x, y)
```

The fit is an object of class `mfp2` that inherits from `glm` and `lm`. This 
means that `mfp2` inherits properties and methods from `glm` and `lm`, which 
allows the `mfp2` to utilize methods and functions specific to `glm` and `lm`. 
Various methods are defined for extracting components from the `mfp2` object, 
such as coef, print, summary, fracplot, and predict. These methods allow users 
to access different components of the mfp2 object, including coefficient 
estimates, plotting functionalities, and predictions.

We demonstrate how to fit the same model using the formula interface. In this 
case, we will use the `fp()` function. 

```{r, eval=FALSE}
fit <- mfp2(lpsa ~ fp(age) + svi + fp(pgg45) + fp(cavol) + fp(weight) + fp(bph) + fp(cp), data = prostate)
```

The main distinction between the `mfp2()` and `glm()` functions in R is the 
inclusion of the `fp()` function within the formula. The presence of the `fp()` 
function in the formula indicates that the variables included within it should 
undergo fractional polynomial (FP) transformation, provided that the degree of 
freedom (df) is not equal to 1. A df of 1 indicates a linear relationship,
which does not require transformation. Note that df is an argument in the 
`fp()` function. For more details on the fp() function, please refer to section
xx

The variable `svi` is a binary variable and is therefore not passed to the `fp()` function. This is because binary or factor variables do not undergo FP transformation. If a binary variable is mistakenly passed to the `fp()` function,
the program will automatically set the df to 1, treating the variable as linear. However, passing a factor variable to the `fp()` function will result in an error. For more details on handling factor variables, refer to section XX.

### Shifting and Scaling of Predictors
Fractional polynomials are defined only for positive variables due to the use of
logarithms and other powers. Thus, `mfp2()`function estimates shifting factors
for each variables to ensure positivity. The function `find_shift_factor()`,
used internally by `mfp2`, automatically estimates shifting factors for continuous variables. The formula used to estimate the shifting factor for a variable, 
say `x1`, is given by:
\[
shift_{x1} = \gamma - min(x1)
\]
where $min(x1)$ is the smallest observed value of `x1`, while $\gamma$ is the 
minimum increment between successive ordered sample values of `x1`, excluding 0 
(Royston and Sauerbrei,2008). The new variable $x1' = x1 + shift_{x1}$ will then be used by `mfp2()` in estimating the FP powers. 

For example, to estimate shifting factors for predictor matrix `x` from prostate
data in R, you can run the following code:

```{r}
# minimum values for each predictor
apply(x, 2, min)

# shifting values for each predictor
apply(x, 2, find_shift_factor)
```
We see that among the continuous variables, only the variable `pgg45` is shifted
by a factor of 1, which is attributed to its minimum value being 0. Even though
the variable `svi` also has a minimum value of 0, it is not shifted by the 
program because its a binary variable. The user can manually set the shifting 
factors for each variable in `mfp2()` function. 

If the values of the variables are too large or too small, it is important to 
scale the variables to reduce the chances of numerical underflow or overflow 
which can lead to inaccuracies and difficulties in estimating the model. Scaling
can be done automatically or by directly specifying the scaling values for
each variables so that the magnitude of the `x` values are not too extreme. 
By default scaling factors are estimated by the program as follows.

After adjusting the location of `x` so that its minimum value is positive, 
creating `x'` automatic scaling will divide each value of `x'` by 
$10^p$ where the exponent `p` is given by 
\[
p = sign(k) \times floor(|k|) \quad \text{where} \quad k = log_{10} (max(x')- min(x'))
\]

The `mfp2()` function uses this formula to scale `x` matrix, and the scaling 
process is implemented through the `find_scale_factor()` function. The following
R code demonstrates the estimation of scaling factors for `x`. From the output
below, we see that the variables `age`, `cavol`, and `cp` have scaling factors 
of 10 each, while the variables `pgg45` and `weight` have scaling factors of 100
each. Each variable will be divided by its corresponding scaling factor. A 
scaling factor of 1 implies no scaling.

```{r}
# shift x 
shift <- apply(x, 2, find_shift_factor)
xnew <- sweep(x, 2, shift, "+")

# scaling factors
apply(xnew, 2, find_scale_factor)

```

To manually enter shifting and scaling factors, the `mfp2()` function provides 
the `shift` and `scale` arguments. In the default usage of `mfp2()`, a vector of
shifting or scaling factors, with a length equal to the number of predictors, can
be provided. In the formula interface, shifting or scaling factors 
can be directly specified within the` fp()` function. Below is an example to 
illustrate this:


```{r, eval = FALSE}
# Default interface
mfp2(x,y, shift = c(0, 0, 1, 0, 0, 0, 0), scale = c(10, 1, 100, 10, 100, 1, 10))

# Formula interface
mfp2(lpsa ~ fp(age, shift = 0, scale = 10) + svi + fp(pgg45, shift = 1, scale = 100) + fp(cavol, shift = 0, scale = 10) + fp(weight, shift = 0, scale = 100) + fp(bph, shift = 0, scale = 1) + fp(cp, shift = 0, scale = 10),
            data = prostate)
```

In the default interface, each variable in the `x` matrix is assigned a shifting 
and scaling factor based on their respective positions. For instance, the first 
variable in the column of `x`, which is `age`, is assigned a shifting factor of 0 
and a scaling factor of 10. The second variable, `svi`, is assigned a shifting 
factor of 0 and a scaling factor of 1, and so on.

### Setting degrees of freedom for each variable {#df_vars}

The degrees of freedom (df) for each predictor (excluding the intercept) are 
twice the degrees of freedom of the FP. For instance, if the maximum allowed 
complexity of variable $x_1$ is a second-degree FP (FP2), then the degrees of 
freedom assigned to this variable should be 4. The default `df` is 4 for each 
predictor.

After assigning the default degrees of freedom to each variable, the program 
proceeds and overrides the default value based on the number of unique values 
for a given variable. The rules for overriding the default degrees of freedom 
are as follows:

* If a variable has 2-3 distinct values, it is assigned `df` = 1 (linear).\
* If a variable has 4-5 distinct values, it is assigned `df` = min(2, default).\
* If a variable has 6 or more distinct values, it is assigned `df` = default.

These rules ensure that the appropriate `df` are assigned to variables. For 
instance, it is not sensible to fit an FP2 function to a variable with only 3 distinct values. 

The following code illustrates how to set different `df` for  each variable. In 
the default interface, the `df` of the binary variable `svi` is explicitly set 
to 1, while In the formula interface, there is no need to specify the `df`, as the
program automatically assigns `df = 1` for binary variables.

If the user attempts to enter `df = 4` for `svi` in the default interface, the 
program will reset the `df` to 1 and issue a warning.


```{r eval = FALSE}
# Default Interface
mfp2(x,y, df = c(4, 1, 4, 4, 4, 4, 4))

# Formula Interface
mfp2(lpsa ~ fp(age, df = 4) + svi + fp(pgg45, df = 4) + fp(cavol, df = 4) + fp(weight, df = 4) +
       fp(bph, df = 4) + fp(cp, df = 4),data = prostate)

```

If the user does not explicitly assign `df` to variables, the program will automatically assign a `df = 4` to each variable. It is important to note that 
even if a continuous variable is not passed to the `fp`()` function in the 
formula interface, the `df` of that variable will still be set to the default 
value and will later be adjusted based on the unique values. The following three examples are equivalent:

```{r eval = FALSE}
# Default Interface
mfp2(x,y)

# Formula Interface
mfp2(lpsa ~ fp(age) + svi + fp(pgg45) + fp(cavol) + fp(weight) +
       fp(bph) + fp(cp), data = prostate)

# Formula Interface but `cp` not passed to the fp() function
mfp2(lpsa ~ fp(age) + svi + fp(pgg45) + fp(cavol) + fp(weight) +
       fp(bph) + cp, data = prostate)
```

### Tuning parameters for MFP
The two key components of MFP are variable selection with backward elimination
(BE) and function selection for continuous variables through the function 
selection procedure (FSP). These components require two nominal significance 
levels: $\alpha_1$  for variable selection with BE and $\alpha_2$  for comparing
the fit of the functions within the FSP. The choice of these significance levels
strongly influences the complexity and stability of the final model. While it 
is possible to use the same nominal significance level for both components, they 
can also differ based on the aims of the analysis.

The `mfp2()` function has an argument called `criterion`, which allows the user 
to specify the criteria for variable and function selection. The default 
criterion is `pvalue`, which enables the user to set the two nominal significance
levels. Please refer to section [section 1.4.4.1](#Nominal_significance_levels) 
for instructions on setting the nominal significance levels.


Information criteria, such as the Akaike information criterion (AIC) and the
Bayesian information criterion (BIC), have been proposed for selecting models
fitted on the same data. The `mfp2` package offers an alternative approach to 
variable and function selection by utilizing information criteria. In the MFP 
framework, each predictor is evaluated univariately while accounting for the
other predictors within an overarching back-fitting algorithm. This algorithm 
iteratively assesses each predictor and selects the model (null, linear, FP1, 
FP2, etc.) with the minimum AIC or BIC for each variable.

The "null" model refers to a model without the predictor of interest. A "linear" 
model assumes a linear relationship with the outcome variable, while an FP1 model 
assumes a non-linear relationship using the FP1 function. The `criterion` argument
allows users to specify whether they want to use AIC or BIC criteria for both
variable and function selection. If AIC or BIC is selected as the criterion, the
nominal significance levels set through the `select` and `alpha` arguments will 
be ignored. For details on using AIC and BIC in variable and function selection,
please refer to [section 1.4.4.2](#information_criteria).

#### Nominal significance levels {#Nominal_significance_levels}

The `mfp2()` function has the arguments `select` and `alpha` for setting the 
nominal significance level for variable selection by BE and for testing between
FP models of different degrees, respectively. It is important to note that when
using these arguments, you should ensure that the `criterion` is set to `"pvalue"` 
to correctly use the specified significance levels.

For variable selection, a significance level can be set using the `select` 
argument. A value of 1 (`select = 1`) for all variables forces them all into the model. Setting the nominal significance level to be 1 for a given variable forces
it into the model, leaving others to be selected or not. A variable is dropped if
its removal leads to a nonsignificant increase in deviance

On the other hand, the `alpha` argument is used to determine the complexity of 
the selected FP function. A value of 1 (`alpha = 1`) will choose the most 
complex FP function permitted for a given variable. For example, if FP2 is the
most complex function allowed for variable x1, setting alpha = 1 will select the FP2 function.

The rules for setting `select` and `alpha` are the same as those for setting `df` 
([see section 1.4.3](#df_vars)). The following R codes shows how to set equal
nominal significance levels for variable and function selection 
$(\alpha_1 = alpha_2 = 0.05)$ for each variable and produces identical results.
Setting different nominal significance levels is straightforward. Simply replace
the value 0.05 with the desired significance level of your choice.


```{r eval = FALSE}
# Default Interface
mfp2(x,y, select = rep(0.05, ncol(x)), alpha = rep(0.05, ncol(x)))

# Formula Interface
mfp2(lpsa ~ fp(age, select = 0.05) + svi + fp(pgg45, select = 0.05) + fp(cavol, select = 0.05) + 
       fp(weight, select = 0.05) + fp(bph, select = 0.05) + fp(cp, select = 0.05), 
     select = 0.05, alpha = 0.05, data = prostate)

```

In the formula interface, binary variables such as `svi` that are not passed in
the `fp()` function utilize the global `select` argument. In our example, the
global parameter is set to `select = 0.05`. If several binary variables exist in
the model, the global parameters will be used for all of them. However, if 
specific parameters need to be set for individual binary variables, the user can
use the `fp()`function. In summary, if a variable is not passed through the `fp`
function, it will utilize the global parameters.

Suppose we want to force the variables "age" and "svi" in the model. To achieve
this, we have two options:

* Set `select = 1` specifically for these variables: By setting `select = 1` for `age` and `svi` in the `mfp2()` function, we force these variables to be included
in the model.\

* Use the `keep` argument: Alternatively, we can use the `keep` argument, which resets the nominal significance levels for `age` and `svi` to 1 if they are 
different from 1. This ensures that these variables are retained in the model.

```{r eval = FALSE}
#------Default Interface
# Set select to 1 for age and svi
mfp2(x,y, select = c(1, 1, 0.05, 0.05, 0.05, 0.05, 0.05), alpha = rep(0.05, ncol(x)))

# use keep argument
mfp2(x,y, select = c(0.05, 0.05, 0.05, 0.05, 0.05, 0.05, 0.05), alpha = rep(0.05, ncol(x)), 
     keep = c("age", "svi"))

#-------Formula Interface
# use fp() function and set select to 1 for age and svi
mfp2(lpsa ~ fp(age, select = 1) + fp(svi, df = 1, select = 1) + fp(pgg45, select = 0.05) + 
       fp(cavol, select = 0.05) + fp(weight, select = 0.05) + fp(bph, select = 0.05) + fp(cp, select = 0.05), 
     select = 0.05, alpha = 0.05, data = prostate)

# use keep argument
mfp2(lpsa ~ fp(age, select = 0.05) + svi + fp(pgg45, select = 0.05) + fp(cavol, select = 0.05) + 
       fp(weight, select = 0.05) + fp(bph, select = 0.05) + fp(cp, select = 0.05), 
     select = 0.05, alpha = 0.05, keep = c("age", "svi"),  data = prostate)
```

#### Information criterion {#information_criteria}
Instead of using nominal significance levels ($\alpha_1, \alpha_2$) for variable and function selection, 
an alternative approach is to directly utilize the AIC or BIC. This can be 
achieved by setting the `criterion` argument to either "aic" or "bic". 
Additionally, if there is a need to force certain variables, such as "age" and
"svi", into the model, the `keep` argument can be used.

The following R code demonstrates how to implement this approach using both the
default and formula interfaces, as well as how to force specific variables into
the model:

```{r, eval=FALSE}
# Default Interface
mfp2(x,y,criterion = "aic", keep = c("age", "svi"))

# Formula Interface
mfp2(lpsa ~ fp(age) + fp(svi, df = 1) + fp(pgg45) + fp(cavol) + fp(weight) + fp(bph) +  fp(cp),
     criterion = "aic", keep = c("age", "svi"), data = prostate)

```

### Model comparison tests
The FSP in MFP is used to compare various models, such as FP2 vs NULL, 
FP2 vs Linear, and FP2 vs FP1, when the `criterion = "pvalue"` (
see section xx for details).

When comparing Gaussian models, the mfp2() function provides two options for 
calculating p-values: the F-test and the Chi-square test. The Chi-square test is
the default option. For other model families like Cox or logistic regression 
models, the Chi-square test is used. For more detailed information, please refer
to page 23 of the paper available at this link: https://www.stata.com/manuals/rfp.pdf

To use the F-test in R, we can set the `ftest` argument to TRUE (`ftest = TRUE`),
as demonstrated in the example below:

```{r, eval=FALSE}
# Default Interface
mfp2(x,y,criterion = "pvalue", ftest = TRUE)

# Formula Interface
mfp2(lpsa ~ fp(age) + svi + fp(pgg45) + fp(cavol) + fp(weight) + fp(bph) +  fp(cp),
     criterion = "pvalue", ftest = TRUE, data = prostate)

```

It is important to note that the p-values reported by the `mfp` program in `STATA`
for Gaussian models are based solely on the F-test. To facilitate result comparison,
the user must ensure that the `ftest = TRUE` in R. 

### Fraction Polynomial Powers


### Explanation of output from model-selection algorithm
## Logistic Regression

## Poisson regression

## Survival data
We illustrate two of the analyses performed by Sauerbrei and Royston (1999). We use
brcancer.dta, which contains prognostic factors data from the German Breast Cancer Study
Group of patients with node-positive breast cancer. The response variable is recurrence-free survival
time (rectime), and the censoring variable is censrec. There are 686 patients with 299 events. We
use Cox regression to predict the log hazard of recurrence from prognostic factors of which five are
continuous (x1, x3, x5, x6, x7) and three are binary (x2, x4a, x4b). Hormonal therapy (hormon) is
known to reduce recurrence rates and is forced into the model. We use mfp to build a model from the
initial set of eight predictors by using the backfitting model-selection algorithm. We set the nominal
p-value for variable and FP selection to 0.05 for all variables except hormon, which it is set to 1:

