---
title: "Introduction to the mfp2 Package"
subtitle: 'Package version `r as.character(utils::packageVersion("mfp2"))`'
author: "Edwin Kipruto, Michael Kammer, Patrick Royston, and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Introduction to the mfp2 Package}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-size: 15px;
  line-height: 1.6;
}

/* Colour section headings in the main text only */
h1, h2, h3, h4 {
  margin-top: 1.4em;
  margin-bottom: 0.6em;
  color: #1f77b4;
}

/* Keep TOC links unchanged */
#TOC a {
  color: inherit;
  text-decoration: none;
}

/* Style ordinary links in the document body */
body > p a,
body > div:not(#TOC) a,
.main-container a {
  color: #1f77b4;
  text-decoration: none;
}

body > p a:hover,
body > div:not(#TOC) a:hover,
.main-container a:hover {
  color: #0b4f8a;
  text-decoration: underline;
}

pre {
  font-size: 13px;
  line-height: 1.4;
  overflow-x: auto;
}

code {
  font-size: 90%;
}

img {
  max-width: 100%;
  height: auto;
}

.figure {
  margin-top: 1.2em;
  margin-bottom: 1.2em;
}

.caption {
  font-size: 90%;
  color: #555555;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  dev = "png",
  dpi = 96
)

oldwidth <- getOption("width")
options(width = 100)

library(mfp2)
library(ggplot2)
library(patchwork)
library(survival)
library(formatR)
```

# Introduction to the `mfp2` package

`mfp2` is a package that implements the multivariable fractional
polynomial (MFP) approach to model selection. The MFP approach combines
variable selection with the selection of suitable functional forms for
continuous covariates.

Compared with the original `mfp` package in R, `mfp2` provides a broader
and more flexible implementation of fractional polynomial modelling. In
particular, while `mfp` is restricted to FP1 and FP2 functions, `mfp2`
allows fractional polynomial functions of arbitrary degree, denoted FPm,
where \(m = 1, \ldots, M\). The package also extends model selection
beyond p-value based procedures by allowing selection using Akaike
information criterion (AIC) and Bayesian information criterion (BIC).
Additional functionality includes support for sigmoid relationships
through the approximate cumulative distribution (ACD) transformation,
spike-at-zero (SAZ) modelling for semi-continuous covariates, and
interaction modelling between categorical grouping variables and
continuous covariates through `mfpi()`. These extensions are implemented
with attention to computational efficiency, including compiled code for
performance-critical transformation steps, and are accompanied by
documented functions and vignettes that illustrate the main modelling
workflows.

The ACD transformation was proposed by Royston (2014) and implemented in
the Stata procedure MFPA (Royston and Sauerbrei, 2016). It enables the
modelling of sigmoid relationships between a continuous covariate `x`
and an outcome variable `y`. In `mfp2`, ACD modelling is available
through the main `mfp2()` function.

The `mfp2()` function also implements SAZ modelling for semi-continuous
covariates with a point mass at zero. SAZ modelling is useful when zero
values may represent a distinct subgroup, while positive values are
modelled using an FP or linear function. In `mfp2`, an SAZ covariate is
initially represented by two components: a binary indicator for the zero
part and an FP or linear function for the positive part of the covariate.
After this full SAZ representation has been considered, the algorithm
evaluates whether both components are needed, whether only the FP or
linear positive-part component should be retained, or whether only the
binary zero indicator should be retained. If the covariate is not
selected, both components are dropped. For more details on the SAZ
algorithm, see the `mfp2()` documentation and the dedicated SAZ vignette
available within the package.

Beyond `mfp2()`, the package provides `mfpi()`, a dedicated function for
multivariable fractional polynomial interaction (MFPI) analysis. The
`mfpi()` function investigates interactions between a grouping variable,
such as treatment, and continuous covariates using FP functions. This
avoids the need to categorize continuous variables or assume linearity
when assessing whether the effect of a grouping variable varies across
the range of a continuous covariate (Royston and Sauerbrei, 2004).
Thus, within the `mfp2` package, `mfp2()` is the main function for MFP,
MFPA/ACD, and SAZ modelling, whereas `mfpi()` is the dedicated function
for interaction analysis. For more details on MFPI, see the `mfpi()`
documentation and the dedicated MFPI vignette available within the
package.

Currently, the package implements Gaussian, binomial, Poisson, and Cox
regression models. Binomial family objects may use links other than the default
logit; for example, `binomial(link = "probit")` fits a binomial probit model.
Future development may extend this implementation to
other likelihood-based regression models, such as Gamma and inverse
Gaussian GLMs, negative binomial regression, and parametric survival
models.

The `mfp2()` function offers two interfaces for input data. The first
interface, the matrix interface, allows direct input of the covariate
matrix `x` and the outcome `y`. The second interface, the formula
interface, uses a formula object together with a data frame, similar to
the `glm()` function with slight modifications. Both interfaces use the
same MFP fitting machinery, but the formula interface additionally constructs
the model matrix and can expand nominal factors. The matrix interface requires
all predictors, including categorical indicators, to be supplied as numeric
matrix columns.

The package also includes several example datasets. We will illustrate
working with `mfp2` using the prostate dataset. Details about the
dataset are found in the book by Royston and Sauerbrei (2008) and on the
[MFP website](https://mfp.imbi.uni-freiburg.de/).

The current authors of `mfp2` are Edwin Kipruto, Michael Kammer, Patrick
Royston, and Willi Sauerbrei. We acknowledge the comments and
suggestions provided by Georg Heinze and Gregory Steiner, which helped
us improve our package. The R package is maintained by Edwin Kipruto,
while the Stata MFP and MFPA programs are maintained by Patrick Royston.
Parts of the text and some recommendations, for example Section 2.3.3,
have been taken from the Stata programs MFP and MFPA.

# Estimation algorithm

The estimation algorithm used in `mfp2` processes covariates
sequentially using a backfitting approach. First, assuming linear
effects for all covariates, the algorithm calculates the p-value for
each covariate using likelihood-ratio tests. The covariates are then
ordered according to these p-values. By default, they are arranged in
decreasing order of statistical significance (i.e. ascending order of
p-values), so that covariates with stronger evidence of association are
considered before those with weaker evidence. This ordering may help
reduce difficulties in model fitting arising from collinearity or, more
generally, from concurvity among covariates (StataCorp, 2023).
Alternative ordering options are available in the `mfp2` package. The
option `xorder = "descending"` considers covariates in decreasing order
of p-values (i.e. increasing order of statistical significance), whereas
`xorder = "original"` preserves the order of covariates in the input
data. We generally recommend the default option, `xorder = "ascending"`,
although subject-matter knowledge may sometimes justify a different
ordering.

For a nonlinear FP candidate, the program estimates a shift when positivity
is required and scales the shifted covariate before the first cycle. Shifts are
forced to zero for linear, binary, `zero`, `catzero`, and retained `spike`
terms. After selection, the final basis is expressed as a function of the
shifted-but-unscaled predictor, such as $\phi(x + s)$; the coefficient must be
interpreted with that basis and is not generally a coefficient of the raw
untransformed $x$. Further details on shifting
and scaling are provided in the section on [shifting and scaling of
covariates](#shifting-scaling).

During the initial cycle, the algorithm determines the best-fitting FP
function for the first covariate in the ordered list, adjusting for all
other covariates and initially assuming linear effects for continuous
covariates. The selected powers for this covariate are retained, and the
same procedure is then applied sequentially to the remaining covariates.
The first cycle ends after all covariates have been processed. At this
stage, some covariates may have been eliminated, while the selected
continuous covariates may have been assigned linear or FP functions up to
the maximum degree allowed by their specified degrees of freedom.

The model obtained at the end of the first cycle provides the starting
point for the next cycle. Subsequent cycles proceed in the same way as
the first cycle, except that the selected powers from the previous cycle
are retained for all covariates other than the one currently being
evaluated.

Updating of covariate inclusion and FP functions continues over
successive cycles until the set of selected covariates and their
corresponding functional forms remain unchanged from one cycle to the
next. At this point, the MFP algorithm is considered to have converged.
In practice, convergence usually occurs within two to four cycles.
Non-convergence, which may arise from oscillation between two or more
models, is rare. If non-convergence occurs, changing the nominal
significance levels for variable selection or function selection may
help to resolve the issue.

## Installation

To install and load the `mfp2` package, enter the following command in
the R console:

```{r, eval=FALSE}
# install the package
install.packages("mfp2")

# load the package
library(mfp2)
```

# Linear regression

The default `family` in `mfp2` package is `family = "gaussian"`, which
fits a Gaussian linear model. We will use the prostate cancer data
(Stamey et al., 1989) included in our package to demonstrate how to fit
this model. The dataset contains seven covariates (six continuous
(`age`, `pgg45`, `cavol`, `weight`, `bph`, and `cp`) and one binary
(`svi`)) and a continuous outcome variable log prostate-specific antigen
(lpsa) of 97 male patients with prostate cancer. Our aim is to determine
variables with influence on the outcome and whether non-linear
functional relationships exist between the covariates and the outcome
variable.

Load the `prostate` dataset from the `mfp2` package and display the
first six rows

```{r out.lines = 10}
# Load prostate data
data("prostate")
head(prostate, 6)

# create covariate matrix x and numeric vector y 
x <- as.matrix(prostate[,2:8])
y <- as.numeric(prostate$lpsa)
```

The command `data("prostate")` loads a data frame from the `mfp2`
package. We create a matrix `x` and a numeric vector `y` from the data
frame to be used by the matrix interface of the `mfp2()` function.

# Fitting MFP models using the matrix and formula interfaces

The default interface of `mfp2()` requires a matrix of covariates `x`
and a numeric vector of response `y` for continuous outcomes. If you
have one covariate, make sure you convert it into a matrix with a single
column.

We demonstrate how to fit the Gaussian linear model using the default
parameters. The formula interface uses the `fp()` function to state that
continuous variables should be modelled using fractional polynomials
(with default parameters). The two approaches lead to the same results
as demonstrated by calling `summary()` with the fitted models as input.

```{r}
# default interface
fit_default <- mfp2(x, y, verbose = FALSE)
summary(fit_default)

# formula interface
fit_formula <- mfp2(lpsa ~ fp(age) + svi + fp(pgg45) + fp(cavol) + fp(weight) + fp(bph) + fp(cp),
                    data = prostate, 
                    verbose = FALSE)
summary(fit_formula)
```

The main distinction between the formulas used in `mfp2()` and `glm()`
functions in **R** is the inclusion of the `fp()` function within the
formula. The presence of the `fp()` function in the formula indicates
that the variables included within it should undergo FP transformation,
provided that the degree of freedom (df) is not equal to 1. A df of 1
indicates a linear relationship, which does not require transformation.
Note that `df` is an argument of the `fp()` function.

The variable `svi` is a binary variable and is therefore not passed to
the `fp()` function. This is because binary or factor variables should
not undergo FP transformation. If a binary variable is passed to the
`fp()` function, the program will automatically set the `df` to 1,
treating the variable as linear. However, passing a factor variable to
the `fp()` function will result in an error. For more details on how
`mfp2` handles categorical variables, refer to the section on [handling
categorical variables](#handling-categorical).

# Shifting and scaling of covariates {#shifting-scaling}

Some fractional-polynomial functions, including logarithms, negative powers,
and fractional powers, require positive input values. When necessary, `mfp2()`
automatically adds a shift before estimating the fractional-polynomial
function.

Variables that are already positive usually receive a shift of zero. Binary
and linear terms are not shifted. Variables fitted with structural-zero
handling also receive a shift of zero so that zero keeps its intended meaning.

`mfp2()` can also scale continuous variables to avoid working with values that
are extremely large or small. Scaling improves numerical stability during FP
power selection. A scale of 1 means that no scaling is needed.

Shifting and scaling are performed automatically. Users do not need to
calculate these values before fitting the model.

The values used by a fitted model are stored in its `transformations`
component:

```{r inspect-shift-scale}
fit_default$transformations[, c("shift", "scale"), drop = FALSE]
```

This table reports the shift and scale used for each model term. For example, a
shift of 1 means that `mfp2()` added 1 before evaluating the selected FP
function. A scale of 10 means that the shifted values were divided by 10 during
FP power selection.

The scale is used for numerical stability during power selection and is removed
before the final model is fitted. A fitted shift remains part of the selected
function. Prediction data should always be supplied on the original variable
scale; `predict()` applies the fitted preprocessing automatically.

## Manually specifying shifts and scales

Most users should allow `mfp2()` to select shifts and scales automatically.
Manual values can be supplied when a transformation is prespecified or when
several analyses must use exactly the same preprocessing.

With the matrix interface, supply one value for each covariate:

```{r manual-shift-scale-matrix, eval = FALSE}
fit_manual <- mfp2(
  x,
  y,
  shift = c(
    age = 0,
    svi = 0,
    pgg45 = 1,
    cavol = 0,
    weight = 0,
    bph = 0,
    cp = 0
  ),
  scale = c(
    age = 10,
    svi = 1,
    pgg45 = 100,
    cavol = 10,
    weight = 100,
    bph = 1,
    cp = 10
  )
)
```

With the formula interface, specify the values in each `fp()` term:

```{r manual-shift-scale-formula, eval = FALSE}
fit_manual <- mfp2(
  lpsa ~
    fp(age, shift = 0, scale = 10) +
    svi +
    fp(pgg45, shift = 1, scale = 100) +
    fp(cavol, shift = 0, scale = 10) +
    fp(weight, shift = 0, scale = 100) +
    fp(bph, shift = 0, scale = 1) +
    fp(cp, shift = 0, scale = 10),
  data = prostate
)
```

Set `shift = 0` to prevent automatic shifting and `scale = 1` to prevent
automatic scaling for a variable. Manually supplied values must be suitable for
the FP powers being considered. In particular, the shifted values must satisfy
the domain requirements of logarithmic, negative, and fractional powers.

## Centering and prediction on the original predictor scale

Shifting and scaling are preprocessing operations used while candidate FP or ACD
functions are estimated. Centering is applied later, after the functional form
has been selected. With `center = TRUE`, each ordinary selected transformed continuous term is
centered on its mean in the fitted sample. For zero-handled FP columns, the mean
is computed from the nonzero transformed values and structural-zero rows remain
zero after centering. Binary terms are centered by subtracting their lower
observed value.

The fitted object records the shift, scale, and centering values. Consequently,
new observations passed to `predict.mfp2()` should be supplied on the original
predictor scale. The fitted preprocessing operations are reapplied internally;
users should not manually shift, scale, transform, or center `newdata`.

## Subsetting and analysis populations {#subsetting-analysis-populations}

The `subset` argument has a deliberately different role from subsetting the data
before calling `mfp2()`. When shift or scale values are estimated automatically,
`mfp2()` estimates them from the full input data first. It then applies `subset`
and performs model selection and final fitting on the retained observations.
This permits several subgroup models to use the same covariate transformations.
For example, models for women and men may be fitted separately while using shift
and scale values estimated from the same combined data set.

In the formula interface, a subset expression is evaluated once using ordinary
formula lookup: names are resolved first from columns of `data` and then from the
formula environment. Thus all of the following forms are valid:

```{r subset-formula-examples, eval=FALSE}
keep_rows <- seq_len(nrow(prostate)) %% 2L == 0L

mfp2(lpsa ~ fp(age) + fp(cavol), data = prostate, subset = age >= 60)
mfp2(lpsa ~ fp(age) + fp(cavol), data = prostate, subset = keep_rows)
mfp2(lpsa ~ fp(age) + fp(cavol), data = prostate, subset = c(10, 3, 8, 15, 20))
```

Numeric row positions must be unique. Their supplied order is retained rather
than sorted. In the matrix interface, `subset` is evaluated in the calling
environment because there is no formula data mask. Matrix column names are not
available as standalone variables, so write an explicit matrix expression when
needed:

```{r subset-matrix-example, eval=FALSE}
mfp2(x, y, subset = x[, "age"] >= 60)
```

Observation-level `weights`, external `offset` values, and matrix-interface
`strata` must remain aligned with the original rows, even when a subset is used.
The argument should not be used to delete missing observations or to define
cross-validation folds. Prepare those analysis data explicitly before fitting.

### Factors and grouped terms after subsetting

Subsetting may remove categorical levels. In formula fits, unused levels are
dropped before the fitted design matrix is constructed. When default contrasts
are used, `model.matrix()` regenerates the appropriate lower-dimensional
contrast basis. A factor with fewer than two retained levels cannot be fitted.

Factor-specific custom contrasts are preserved when all fitted levels remain.
If subsetting removes levels from a factor with custom contrasts, `mfp2()` stops
rather than guessing a reduced contrast basis, because no unique reconstruction
is generally available.

The matrix interface receives only numeric columns and therefore cannot rebuild
factor levels or contrasts. Its supplied columns remain unchanged. A mapped
`term_groups` block is rejected when subsetting reduces its estimable dimension,
because the original categorical representation cannot be reconstructed safely.

# Setting degrees of freedom for each variable {#df_vars}

The degrees of freedom (df) for each covariate (excluding the intercept)
are twice the degrees of freedom of the FP. For instance, if the maximum
allowed complexity of variable $x1$ is a second-degree FP (FP2), then
the degrees of freedom assigned to this variable should be 4. To find
the best fitting function, FSP selects power terms $p1$ and $p2$ and the
corresponding regression coefficients $\beta_1$ and $\beta_2$. The
default `df` is 4 ($m = 2$) for each covariate.

After assigning the default degrees of freedom to each variable, the
program overrides the default value if a variable has only a very small
number of unique values as such variables may not be handled as other
continuous variable. As proposed in the Stata program the rules for
overriding the default degrees of freedom are as follows:

- If a variable has 2-3 distinct values, it is assigned `df` = 1
  (linear).\
- If a variable has 4-5 distinct values, it is assigned `df` = min(2,
  default).\
- If a variable has 6 or more distinct values, it is assigned `df` =
  default.

These three rules apply to formula fits and to matrix fits that use a scalar
`df`. A matrix fit may instead supply an explicit `df` value for every column.
In that case, variables with 2--3 distinct values are still forced to `df = 1`,
but supplied values for variables with 4 or more distinct values are retained.
This interface-specific distinction matters when a per-column `df` vector is
used.

These rules ensure that the appropriate `df` are assigned to variables.
For instance, it is not sensible to fit an FP2 function to a variable
with only 3 distinct values (StataCorp, 2023).

The following code illustrates how to set different `df` for each
variable. In the default interface, the `df` of the binary variable
`svi` is explicitly set to 1 (linear), while in the formula interface,
there is no need to specify the `df`, as the program automatically
assigns `df = 1` for binary variables.

If the user attempts to enter `df = 4` for `svi` in the default
interface, the program will reset the `df` to 1 and issue a warning.

```{r eval = FALSE}
# Default Interface
mfp2(x,y, df = c(4, 1, 4, 4, 4, 4, 4))

# Formula Interface
mfp2(lpsa ~ fp(age, df = 4) + svi + fp(pgg45, df = 4) + fp(cavol, df = 4) + 
       fp(weight, df = 4) + fp(bph, df = 4) + fp(cp, df = 4),
     data = prostate)
```

In addition, if the user does not explicitly assign `df` to variables,
the program will automatically assign `df = 4` to each variable,
corresponding to $m
= 2$ as the default for MFP modelling. It is important to note that even
if a continuous variable is not passed to the `fp()` function in the
formula interface, the `df` of that variable will still be set to the
default value and will later be adjusted based on the unique values. The
following examples are all equivalent:

```{r eval = FALSE}
# Default Interface
mfp2(x,y)

# Formula Interface, all continuous variables passed to the fp() function
mfp2(lpsa ~ fp(age) + svi + fp(pgg45) + fp(cavol) + fp(weight) + fp(bph) + fp(cp), 
     data = prostate)

# Formula Interface, `cp` not passed to the fp() function
mfp2(lpsa ~ fp(age) + svi + fp(pgg45) + fp(cavol) + fp(weight) + fp(bph) + cp, 
     data = prostate)

# Formula Interface, all covariates not passed to the fp() function
mfp2(lpsa ~ age + svi + pgg45 + cavol + weight + bph + cp, 
     data = prostate)
```

# Tuning parameters for MFP

The two key components of MFP are variable selection with backward
elimination (BE) and function selection for continuous variables through
the function selection procedure (FSP). The `mfp2()` function has an
argument called `criterion`, which allows the user to specify the
criterion applied for variable and function selection.

The default criterion is `pvalue`, which enables the user to set two
nominal significance levels: $\alpha_1$ for variable selection with BE
and $\alpha_2$ for comparing the fit of the functions (linear, best FP1,
best FP2 and so on) within the FSP. The choice of these significance
levels strongly influences the complexity and stability of the final
model. Please refer to the section on [nominal significance
levels](#nominal-significance-levels) for instructions on setting the
nominal significance levels.

The `mfp2` package also supports alternative approaches to variable and
function selection based on information criteria. Within the MFP
framework, each covariate is evaluated in turn while adjusting for the
other covariates through the backfitting algorithm. For each covariate,
the algorithm compares candidate models, such as the null, linear, FP1,
and FP2 models, and selects the model with the smallest AIC or BIC,
depending on whether `criterion = "aic"` or `criterion = "bic"` is
specified; see the section on [information
criteria](#information-criteria).

The "null" model refers to a model without the covariate of interest. A
"linear" model assumes a linear relationship with the outcome variable,
while an FP1 model assumes a non-linear relationship using the FP1
function. The `criterion` argument allows users to specify whether they
want to use AIC or BIC criteria for both variable and function
selection. If AIC or BIC is selected as the criterion, the nominal
significance levels set through the `select` and `alpha` arguments will
be ignored. For details on using AIC and BIC in variable and function
selection, see the section on [information
criteria](#information-criteria).

## Nominal significance levels {#nominal-significance-levels}

The `mfp2()` function has the arguments `select` and `alpha` for setting
the nominal significance level for variable selection by BE and for
testing between FP models of different degrees, respectively. When using
these arguments, you should ensure that the `criterion` is set to
`"pvalue"` to correctly use the specified significance levels.

For variable selection, a significance level can be set using the
`select` argument. A value of 1 (`select = 1`) for all variables forces
them all into the model. Setting the nominal significance level to be 1
for a given variable forces it into the model, leaving others to be
selected or not. A variable is dropped if its removal leads to a
nonsignificant increase in deviance. Default is `select = 0.05` for each
variable.

On the other hand, the `alpha` argument is used to determine the
complexity of the selected FP function. A value of 1 (`alpha = 1`) will
choose the most complex FP function permitted for a given variable. For
example, if FP2 is the most complex function allowed for a continuous
variable, setting `alpha = 1` will select the best FP2 function. Default
is `alpha = 0.05` for each continuous variable.

The rules for setting `select` and `alpha` are the same as those for
setting `df`. The following R codes shows how to set equal nominal
significance levels for variable and function selection
$(\alpha_1 = \alpha_2 = 0.05)$ for each variable and produces identical
results. Setting different nominal significance levels is
straightforward. Simply replace the value 0.05 with the desired
significance level of your choice.

```{r eval = FALSE}
# Default Interface
mfp2(x,y, select = rep(0.05, ncol(x)), alpha = rep(0.05, ncol(x)))

# Formula Interface
mfp2(lpsa ~ fp(age, select = 0.05) + svi + fp(pgg45, select = 0.05) + 
       fp(cavol, select = 0.05) + fp(weight, select = 0.05) + 
       fp(bph, select = 0.05) + fp(cp, select = 0.05), 
     select = 0.05, alpha = 0.05, 
     data = prostate)
```

In the formula interface, binary variables such as `svi` that are not
passed in the `fp()` function utilize the global `select` argument. In
our example, the global parameter is set to `select = 0.05`. If several
binary variables exist in the model, the global parameters will be used
for all of them. However, if specific parameters need to be set for
individual binary variables, the user can use the `fp()`function. In
summary, if a variable is not passed through the `fp` function, it will
utilize the global parameters.

Suppose we want to force the variables `"age"` and `"svi"` in the model.
To achieve this, we have two options:

- Set `select = 1` for `age` and `svi` in the `mfp2()` function.\
- Alternatively, we can use the `keep` argument, which resets the
  nominal significance levels for `age` and `svi` to 1 if they are
  different from 1. This ensures that these variables are retained in
  the model.

```{r eval = FALSE}
# default Interface
# Set select to 1 for variables "age" and "svi"
mfp2(x,y, 
     select = c(1, 1, 0.05, 0.05, 0.05, 0.05, 0.05),
     alpha = rep(0.05, ncol(x)))

# use keep argument
mfp2(x,y,
     select = rep(0.05, ncol(x)), 
     alpha = rep(0.05, ncol(x)), 
     keep = c("age", "svi"))

# formula Interface
# use fp() function and set select to 1 for "age" and "svi"
mfp2(lpsa ~ fp(age, select = 1) + fp(svi, df = 1, select = 1) + fp(pgg45, select = 0.05) + 
       fp(cavol, select = 0.05) + fp(weight, select = 0.05) + fp(bph, select = 0.05) + fp(cp, select = 0.05), 
     select = 0.05, alpha = 0.05, data = prostate)

# use keep argument
mfp2(lpsa ~ fp(age, select = 0.05) + svi + fp(pgg45, select = 0.05) + fp(cavol, select = 0.05) + 
       fp(weight, select = 0.05) + fp(bph, select = 0.05) + fp(cp, select = 0.05), 
     select = 0.05, alpha = 0.05, 
     keep = c("age", "svi"), 
     data = prostate)
```

## Information criteria {#information-criteria}

Instead of using nominal significance levels for variable and function
selection, an alternative approach is to directly use adapted versions
of AIC or BIC, defined below.

$$\begin{aligned}
AIC & = -2 \log(L) + 2k \\
BIC & = -2 \log(L) + \log(n) \times k,
\end{aligned}$$

where $\log(L)$ is the maximum log-likelihood of the fitted model, which
measures how well the model fits the data. The parameter $k$ corresponds
to the number of estimated parameters in the model (regression estimates
and FP powers) and $n$ is the sample size; for Cox models, $n$ is the
number of observed events. AIC and BIC both account for model fit and
model complexity, with lower values indicating better-fitting MFP
models.

For instance, let's assume we want to select the best model for a
variable of interest ($z$) with fixed adjustment variables $x1$ (with
power $p = 3$ as the best FP1 function) and $x2$ (linear function was
pre-specified). Then, we can compare the AIC and BIC of different
models, such as FP2, FP1, linear, and null models for $z$. The
adjustment models all have the same number of parameters (2 for $x1$, 1
for $x2$, and perhaps an intercept if it exists). For $z$, we have 4, 2,
and 1 parameter(s) for FP2, FP1, and linear models, respectively, for a
model without an intercept. In a model without an intercept, the total
number of parameters, including adjustment variables, for variable z are
7 (4 + 3) for FP2, 5 (2 + 3) for FP1, and 4 (1 + 3) for a linear model.
These total number of parameters for each model type is used in
calculating the AIC and BIC. The model with the smallest AIC or BIC is
then selected for $z$.

In the `mfp2` package, this can be done by setting the `criterion`
argument in `mfp2()` to either `"aic"` or `"bic"`. If certain variables,
such as `age` and `svi`, should be forced into the model, the `keep`
argument can be used.

The argument `force_max_fp` can be used to force selection of the most
complex permitted FP function for a covariate. For example, if the
maximum complexity allowed for a covariate $z$ is FP2, setting
`force_max_fp = TRUE` for $z$ forces the best FP2 function to be
selected when `criterion = "aic"` or `criterion = "bic"`. This is
analogous to setting `select = 1` and `alpha = 1` when
`criterion = "pvalue"`.

The following R code demonstrates how to implement this approach using
both the default and formula interfaces, as well as how to force
specific variables into the model:

```{r, eval=FALSE}
# Default Interface
mfp2(x, y, criterion = "aic", keep = c("age", "svi"))

# Formula Interface
mfp2(lpsa ~ fp(age) + fp(svi, df = 1) + fp(pgg45) + fp(cavol) + fp(weight) + fp(bph) +  fp(cp),
     criterion = "aic", keep = c("age", "svi"), data = prostate)
```

# Model comparison tests

The function selection procedure (FSP) in `mfp2()` compares a sequence of
candidate models for each variable under consideration. For example, when the
most complex permitted FP function is FP2 and `criterion = "pvalue"`, the FSP
compares the best-fitting FP2 model with the null model, the linear model, and
the best-fitting FP1 model.  For each candidate model, the deviance, defined as
$-2\log(L)$, is computed. The corresponding deviance differences and p-values
are then used to guide model selection.

For Gaussian models, `mfp2()` provides two options for calculating p-values: the
F-test and the Chi-square test. The Chi-square test is used by default. For
non-Gaussian models, including logistic, Poisson, and Cox regression models, the
Chi-square test is used. Further details are available in the
[MFP Stata manual](https://www.stata.com/manuals/rfp.pdf).

To use the F-test for Gaussian models, set `ftest = TRUE`, as shown below. To use
the Chi-square test, set `ftest = FALSE`.

```{r, eval=FALSE}
# Default Interface
mfp2(x, y,
     criterion = "pvalue", 
     ftest = TRUE)

# Formula Interface
mfp2(lpsa ~ fp(age) + svi + fp(pgg45) + fp(cavol) + fp(weight) + fp(bph) +  fp(cp),
     criterion = "pvalue", 
     ftest = TRUE, 
     data = prostate)
```

Please note that, for Gaussian models, the p-values reported by the `mfp`
program in Stata are based on the F-test. Therefore, if results from `mfp2` are
to be compared with those from Stata, `ftest = TRUE` should be specified in
**R**, since the default in `mfp2` is `ftest = FALSE`.

The older `mfp` package in **R** uses the Chi-square test for all model types.
Consequently, for Gaussian models, its results may differ from those obtained
from the Stata `mfp` program. In summary, `mfp2` can reproduce the Gaussian-model
results from Stata when `ftest = TRUE`, and can reproduce the behaviour of the
older **R** `mfp` package when `ftest = FALSE`.

# Subject-matter knowledge may require changes in fractional polynomial powers

By default, `mfp2()` uses the predefined set of powers

$$
S = \{-2, -1, -0.5, 0, 0.5, 1, 2, 3\}
$$

for each continuous covariate, as proposed by Royston and Altman (1994).
However, in some situations, users may wish to restrict or modify the set of
candidate powers on the basis of subject-matter knowledge. This can be done
using the `powers` argument, which takes a named list of power vectors.

More extreme powers may also be considered when they are scientifically
plausible or when they may improve model fit. For example, if one of the
boundary FP2 functions, such as $(-2, -2)$ or $(3, 3)$, is selected, it may be
reasonable to examine whether a more extreme function, such as $(-3, -3)$ or
$(4, 4)$, provides a better model fit.

The following example illustrates how different power sets can be assigned to
continuous covariates. For the `age` covariate, two powers, 1 and 0.5, are
evaluated. This gives three possible FP2 functions, with power combinations
$(0.5, 0.5)$, $(0.5, 1)$, and $(1, 1)$, as well as the corresponding FP1
functions. The linear function, corresponding to power 1, is also allowed.

The `cavol` variable is assigned the powers 1, 0, and 1. The remaining
continuous variables, namely `pgg45`, `weight`, `bph`, and `cp`, are not
assigned user-defined power sets. Therefore, the default power set $S$ is used
for these variables, and all possible fractional polynomial functions up to the
degree specified by `df` are considered.

When using the `fp()` function in the formula interface, the supplied powers must
be given as a vector rather than a list, because they apply only to the
individual variable inside that `fp()` term.

```{r, eval=FALSE}
# Create a named list of user-defined power sets for selected covariates
custom_powers <- list( age = c(1, 0.5), cavol = c(1, 0, 1))

# Default interface
mfp2(x, y, criterion = "pvalue", powers = custom_powers)

# Formula interface
mfp2(lpsa ~ fp(age, powers = c(1, 0.5)) + svi + fp(pgg45) + fp(weight) +
       fp(cavol, powers = c(1, 0, 1)) + fp(bph) + fp(cp),
  data = prostate
)
```

## Candidate FP powers

Candidate FP powers can be supplied either through the top-level `powers`
argument or term by term inside `fp()` or `fp2()` in the formula interface. The
values define a set of candidate base powers, not a final selected FP power
vector. Duplicate candidate powers are removed. For example, `powers = c(4, 4)`
is treated as the single candidate set `powers = 4`; repeated selected powers
such as FP2(4, 4) are generated internally when an FP2 model is fitted.

In the formula interface, powers supplied inside `fp()` or `fp2()` take
precedence over powers supplied through the top-level `powers` argument for the
same variable. This avoids ambiguity when a variable-specific formula term and a
global `powers` list both specify candidate powers.

# Explanation of output from the model-selection algorithm

Using the prostate example, we briefly illustrate how the MFP model-selection
algorithm works. The algorithm uses a backfitting procedure. It starts from a
full multivariable model that includes all candidate variables, with continuous
variables initially treated as linear. The algorithm then visits the variables
one at a time. At each visit, the variable currently under consideration is
assessed while adjusting for the current forms of the other variables in the
model.

For continuous variables, the function selection procedure (FSP) is used to
assess both whether the variable should remain in the model and, if retained,
which functional form should be used. The first comparison in the FSP evaluates
the best-fitting FP2 function against the null model, that is, a model excluding
the variable under consideration. If this comparison is not significant at the
specified `select` level, the variable is removed at that visit. If the variable
is retained, subsequent comparisons determine whether a linear, FP1, or FP2
function is selected.

For non-continuous variables, such as binary variables or dummy variables created
from categorical covariates, no FP function selection is performed. These
variables are treated as linear terms with `df = 1`. At each visit, the model
including the variable is compared with the null model that excludes it. If the
test is significant at the specified `select` level, the variable is retained;
otherwise, it is removed at that visit. Thus, for non-continuous variables, the
procedure only assesses variable inclusion, whereas for continuous variables it
assesses both variable inclusion and functional form.

At the end of the first cycle, some variables may have been removed, and some
continuous variables may have been assigned linear, FP1, or FP2 functions. These
selected variables and functions then form the starting model for the next
cycle.

The selected variables and functions from the first cycle form the starting
model for the second cycle. In this cycle, all candidate variables are visited
again, including variables that may have been removed in the previous cycle.
This is an important feature of the backfitting algorithm: the evidence for a
variable, and the best functional form for a continuous variable, may change
after other variables have been removed or transformed. The results from the
second cycle then form the starting model for the third cycle. The process
continues until the selected variables and their functional forms remain
unchanged from one cycle to the next. In most applications, convergence occurs
within three or four cycles.

The order in which variables are visited is important. Misspecifying the
functional form of a variable with a strong effect may have a greater impact on
the model than misspecifying a variable with a weak effect. Therefore, by
default, variables are ordered according to their p-values from the initial full
linear model, so that variables with stronger evidence of association are
considered first.

To explain the output of the MFP algorithm, we use the default interface of
`mfp2()` to fit an MFP model with the default settings. All continuous variables
are assigned `df = 4`, so that FP2 is the most complex permitted function. The
criterion is set to `"pvalue"`, and the nominal significance levels for variable
selection and function selection are both set to 0.05. For this Gaussian model,
we set `ftest = TRUE`, so that p-values are calculated using the F-test.

```{r out.lines = 103}
fit <- mfp2(x, y, criterion = "pvalue", select = 0.05, alpha = 0.05, 
            ftest = TRUE)
```

The **R** output first displays the degrees of freedom used for each variable,
shown as the initial degrees of freedom. The program correctly identifies `svi`
as a binary variable and assigns it `df = 1`. The output also shows the visiting
order of the variables. In this example, `cavol` has the smallest p-value in the
initial full linear model and is therefore evaluated first, whereas `age` has
the largest p-value and is evaluated last.

The first cycle begins with `cavol`. The algorithm compares the best-fitting FP2
function for `cavol`, with powers $(-0.5, 1)$, against a null model that
excludes `cavol`. This comparison is made while adjusting for the other six
variables (`svi` to `age`), which are still represented by linear functions at
this stage. The test is highly significant $(p = 0.0000)$, indicating that
`cavol` should remain in the model. The best FP2 function is then compared with
a linear function, and this test is also significant $(p = 0.0011)$, suggesting
evidence of nonlinearity. Finally, the best FP2 function is compared with the
best FP1 function, with power 0. This comparison is not significant
$(p = 0.2226)$. Therefore, at this stage of the backfitting procedure,
`cavol` is represented by an FP1 function with power 0, that is, by
$\log(\texttt{cavol})$.

The next variable visited is `svi`, which is binary. For this variable, only the
comparison between the linear model and the null model is relevant. Both models
are adjusted for the current function of `cavol`, namely $\log(\texttt{cavol})$),
and for the remaining variables (`pgg45`, `cp`, `weight`, `bph`, and `age`),
which are still represented by linear functions. The test is significant
$(p = 0.0042)$, so `svi` remains in the model.

The continuous variable `pgg45` is then evaluated in the same adjustment model.
The first test, comparing the best FP2 function with the null model, is not
significant $(p = 0.3091)$. Therefore, `pgg45` is eliminated from the model
for the remainder of the first cycle.

Next, `weight` is evaluated in a model adjusted for $\log(\texttt{cavol})$,
`svi`, and the variables `cp`, `bph`, and `age`, which have not yet been visited
in this cycle. The first test is not significant $(p = 0.0521)$, so `weight`
is also eliminated for the remainder of the first cycle. In the subsequent
steps, `bph`, `cp`, and `age` are also eliminated.

**At the end of the first cycle, only two variables have been selected:
$\log(\texttt{cavol})$ and `svi`.**

These selected variables and functions form the starting model for the second
cycle. The second cycle again begins with `cavol`, but now the adjustment model
is simpler because it includes only `svi`. The deviances differ from those in
the first cycle because the variables `pgg45`, `age`, `weight`, `bph`, and `cp`
are no longer part of the adjustment model. Nevertheless, the FSP again selects
$\log(\texttt{cavol})$, although the estimated regression coefficient changes
because the adjustment model has changed.

The variable `svi` is then evaluated in a model that includes
$\log(\texttt{cavol})$, and the comparison with the null model confirms that
`svi` should remain in the model. The variable `pgg45` is also reconsidered, but
its test of inclusion remains nonsignificant, so it is again eliminated.

The variable `weight` is then reconsidered in a model adjusted for
$\log(\texttt{cavol})$ and `svi`. In contrast to the first cycle, the test for
inclusion is now significant $(p = 0.0052)$, indicating that `weight` should be
reintroduced into the model. The subsequent test for nonlinearity is not
significant, so `weight` is selected as a linear term. This change illustrates
the backfitting nature of the algorithm: once `bph`, `cp`, and `age` have been
removed, the evidence for including `weight` changes, partly because of
correlation among these variables.

In the remaining steps of the second cycle, `bph`, `cp`, and `age` are evaluated
again in models adjusted for $\log(\texttt{cavol})$, `svi`, and linear
`weight`. Their p-values remain well above 0.05, and none of these variables is
included in the model.

**The second cycle ends with a model containing $\log(\texttt{cavol})$, `svi`
(binary), and `weight` (linear).**

This model becomes the starting point for the third cycle. In the third cycle,
all seven variables are visited again, but there are no further changes in the
selected variables or their functional forms. The algorithm therefore converges,
and the final MFP model contains the three variables selected at the end of the
second cycle.

# Methods defined for `mfp2`

After fitting an MFP model with `mfp2()`, several methods can be applied to the
resulting model object to extract information or perform post-estimation tasks.
For example, an MFP model for a Gaussian outcome can be fitted with the default
settings as follows:

```{r, eval=TRUE, tidy=TRUE, tidy.opts=list(width.cutoff = 80)}
fit <- mfp2(x, y, verbose = FALSE)
class(fit)
```

The object `fit` has class `mfp2` and inherits from `glm` and `lm`. If
`family = "cox"`, the fitted object inherits from `coxph` in the **survival**
package. As a result, methods defined for `glm`, `lm`, or `coxph` objects can
also be used with fitted `mfp2` objects, where appropriate.

To display the final fitted model, enter the object name, for example `fit`, or
use `print(fit)`. The `print()` method provides a summary of the final MFP
model, including the parameters used during model selection, such as shifting
and scaling factors, initial degrees of freedom (`df_initial`), nominal
significance levels (`select` and `alpha`), and other relevant settings.

The output also reports the final degrees of freedom (`df_final`). A value of
`df_final = 0` indicates that the corresponding variable was eliminated from the
final model, as also shown in the `selected` column. The `power` column gives the
final FP powers selected for each variable; variables that were eliminated are
assigned `NA`. The `acd` column indicates whether the ACD transformation was
applied to a covariate and is displayed only when at least one covariate was
specified for ACD transformation.

```{r, eval= TRUE}
print(fit)
```

To display the regression coefficients from the final MFP model, use `coef()` or
`summary()`. In the model output, FP-transformed variables are named using dot
extensions. For example, if a variable is represented by an FP2 function, its two
transformed terms are labelled as `var.1` and `var.2`.

The displayed regression coefficients multiply the final transformed basis.
Automatic scale factors are removed from that basis before the final fit, but an
estimated shift remains part of it. For example, a coefficient may multiply
$\phi(x + s)$ rather than the raw predictor $x$. Interpret coefficients together
with `transformations` and the selected powers; prediction still expects the
original predictor values.

```{r, eval= TRUE}
# extract only regression coefficients
coef(fit)

# display regression coefficients with other statistics
summary(fit)
```

Users can generate predictions for new or existing data using the `predict()`
function. To illustrate this, consider the prostate dataset. Suppose we want to
obtain predictions for the first five rows of the covariate matrix `x`, treating
them as new observations. This can be done as follows:

```{r out.lines = 10, eval=TRUE}
# Extract the first five observations from x
new_observations <- x[1:5, ]
dim(new_observations)

# Generate predictions for these observations with their standard errors
predict(fit, newdata = new_observations, se.fit = TRUE)
```

When `newdata` is supplied, `predict()` applies the same transformations that
were used when fitting the model. In particular, any shifting required for FP
transformations is based on the shifting factors estimated from the development
data.

For models fitted with an explicit matrix-interface offset, predictions on
`newdata` must also supply the corresponding `newoffset`. Formula-interface
offset expressions are reconstructed from the variables present in `newdata`.

Users should note that, if a new observation contains values outside the range
of the development data, the original shifting factor may not be sufficient to
make all transformed covariate values positive. Prediction is undefined when a
selected logarithmic or negative-power FP term receives a non-positive value
after the fitted shift has been applied. In that situation, `predict()` stops
with an informative error identifying the affected variables and rows.

Fitted FP functions can be visualized using `plot(fit)`. The `plot()`
method is the recommended graphical interface for fitted `mfp2` objects;
the older `fracplot()` function remains available for backward
compatibility.

# Graphical presentation of FP functions

The regression coefficients $\hat{\beta}$ for FP terms provide only
limited information about the fitted function for a variable of interest.
A more informative approach is to visualize the fitted function,
especially for variables with nonlinear effects (Royston and Sauerbrei,
2008).

The recommended way to produce graphical displays from a fitted `mfp2`
object is to use the S3 `plot()` method. The method returns a list of
`ggplot2` objects, one for each selected variable requested. The `fracplot()`
function remains available for backward compatibility, but `plot()` is the
preferred interface for new code and is used throughout this vignette.

Given a fitted `mfp2` object, `plot()` produces plots of the selected
functions, together with pointwise confidence intervals. As with all
data-derived models, model-based confidence intervals may be too narrow
because they do not account for uncertainty from variable and function
selection. Bootstrap methods can be used to obtain more realistic
estimates of standard errors and pointwise confidence intervals
(Sauerbrei and Royston, 2007). The variable to be plotted can be
specified using the `terms` argument. If `terms = NULL`, which is the
default, `plot()` returns a list of plots for all variables included in
the final model.

By default, `plot()` produces a component-plus-residual plot because
`partial_only = FALSE`. For Gaussian models with constant weights and a
single covariate, this corresponds to a plot of the outcome against the
covariate, with the fitted function added. For other normal-error
models, weighted residuals are calculated and added to the fitted
values. In models with two or more covariates, the plotted function is
the partial linear predictor for the variable of interest and includes
the intercept.

For generalized linear models and Cox models, `plot()` displays fitted
values on the scale of the linear predictor. For generalized linear
models, deviance residuals are added to the partial linear predictor.
For Cox models, martingale residuals are added instead. These
component-plus-residual values are shown as points in the plot. Such
plots can reveal the amount of residual variation across the covariate
range and may indicate lack of fit or influential observations (Royston
and Sauerbrei, 2008; StataCorp, 2023).

If `partial_only = TRUE`, `plot()` displays only the partial linear
predictor, without adding residuals. Further details on
component-plus-residual plots are available in the
[Stata postestimation manual](https://www.stata.com/manuals/rmfppostestimation.pdf).

For example, the fitted function for `cavol` can be plotted as follows:

```{r, eval=TRUE, fig.cap="\\label{fig:fig2}Prostate data. Graphical presentation of results from an FP analysis", fig.width=7, fig.height=5, out.width="100%", fig.align="center"}
plots <- plot(fit)

# plots is a list
class(plots)

# The plot of cavol is the first element of the list
plots[[1]] + ggplot2::ylab("Partial predictor + residuals")
```

\newpage

The `plot()` method also allows a reference value to be included in the
plot. This is useful when the aim is to display contrasts relative to a
selected reference point. To do this, set `type = "contrasts"` and
provide the reference value using the `ref` argument. If `ref = NULL`,
which is the default, the average value is used as the reference for
continuous variables, and the minimum value is used for binary
variables. For example, to use 30 as the reference value for the
continuous variable `cavol`, use the following code:

```{r, eval=TRUE, fig.cap="\\label{fig:fig3}Prostate data. Illustration on how to use reference points (ref point = 30)", fig.width=7, fig.height=5, out.width="100%", fig.align="center"}
plots <- plot(fit, type = "contrasts", ref = list(cavol = 30))
plots[[1]] + ggplot2::ylab("Partial predictor")
```

In some cases, a fitted function plotted using the original observed
covariate values may appear very sharp near extreme values. This can
occur when only a few observations are available in the extreme part of
the covariate distribution.

The `plot()` method provides two options for the covariate values used
in the plot. The first option, `terms_seq = "data"`, uses the original
covariate values from the data used to fit the model. The second option,
`terms_seq = "equidistant"`, generates an equally spaced sequence of
values over the observed covariate range.

To illustrate these options, we use the `art` dataset included in the
`mfp2` package. This dataset is described in Royston and Sauerbrei
(2008, Chapter 10), and the outcome variable is continuous. In this
example, we fit an MFP model with a single covariate, $x5$.

```{r, eval=TRUE, fig.cap="Art data. Illustration of the `terms_seq` argument in `plot.mfp2()`. In the original data, the selected function was FP(0, 3), equivalent to $\\beta_0 + \\beta_1 \\log(x_5) + \\beta_2 x_5^3$. The function is plotted using the original observed values of `x5` (right panel) and an equally spaced sequence of values over the observed range of `x5` (left panel). The two functions are identical up to approximately x5 = 62. Beyond this value, the plot based on the original values appears nearly linear because there are no observations between 62.21 and 235. In contrast, the equally spaced sequence clearly displays the fitted FP2 curve.", fig.width=7, fig.height=5, out.width="100%", fig.align="center"}
# load art data
data("art")

# fit mfp model using art data
xx <- as.matrix(art[, "x5", drop = FALSE])
yy <- as.numeric(art$y)
fit2 <- mfp2(xx, yy, verbose = FALSE)

# generate plot using original data
plot1 <- plot(fit2)
plot1[[1]] <- plot1[[1]] +
  ggplot2::ylab("Partial predictor + residuals")

# generate plot using an equally spaced sequence
plot2 <- plot(fit2, terms_seq = "equidistant")
plot2[[1]] <- plot2[[1]] +
  ggplot2::ylab("Partial predictor + residuals")

# combine plots
patchwork::wrap_plots(plot2[[1]], plot1[[1]], ncol = 2, widths = 8, heights = 4)
```


# Handling categorical variables {#handling-categorical}

The formula interface accepts unordered and ordered factors directly. It uses
`model.matrix()` to expand each factor into its configured contrast columns, but
all columns generated from one factor are treated by `mfp2()` as one conceptual
model term. The complete block is selected, retained, tested, or removed jointly.
Naming the factor in `keep` protects the whole block from exclusion.

## Unordered factors

Unordered factors use their configured contrasts, which are treatment contrasts
by default. With a factor containing \(k\) levels, the default representation has
\(k-1\) columns relative to the reference level. `mfp2()` performs one joint test
of those columns; it does not select individual dummy variables separately.
General contrast construction is described in `?contrasts`, `?contr.treatment`,
and `?model.matrix`.

## Ordered factors

Ordered factors also enter as one joint term. Their default `contr.poly` basis
contains \(k-1\) columns. The resulting `mfp2()` test is therefore an omnibus
\(k-1\)-degree-of-freedom test of the complete ordered-factor effect. It is not a
one-degree-of-freedom linear trend test and does not impose a monotonic effect.

When subject-matter knowledge specifies a particular ordinal trend, create an
explicit numeric score and restrict it to a linear effect. For example:

```{r ordinal-score-example, eval=FALSE}
dat$severity_score <- c(0, 1, 2)[match(dat$severity,
                                      c("mild", "moderate", "severe"))]
fit <- mfp2(y ~ fp(severity_score, df = 1) + fp(age), data = dat)
```

Factor contrast blocks are fixed linear terms. Shifting, scaling, FP, ACD,
`zero`, `catzero`, and SAZ processing do not apply to them.

## Formula-interface example

The following example supplies both an ordered and an unordered factor directly.

```{r categorical-formula-example, message=TRUE, warning=TRUE}
data("art")

art$x4 <- ordered(art$x4, levels = c(1, 2, 3))
art$x9 <- factor(art$x9, levels = c(1, 2, 3))

fit_formula <- mfp2(
  y ~ fp(x1) + x2 + fp(x3) + x4 + fp(x5) + fp(x6) +
    fp(x7) + x8 + x9 + fp(x10),
  data = art,
  keep = c("x4", "x9"),
  verbose = FALSE
)

fit_formula
```

For prediction from a formula fit, supply the original factor variables in
`newdata`. The fitted levels and contrasts are reused. New or unknown levels are
rejected by the standard `model.frame()` and `model.matrix()` checks.

## Matrix-interface example

The matrix interface requires numeric columns. When several generated columns
represent one factor, identify them with `term_groups`; otherwise they would be
considered separate selection candidates.

```{r categorical-matrix-example, message=TRUE, warning=TRUE}
mm <- model.matrix(
  y ~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9 + x10,
  data = art
)

x_matrix <- mm[, colnames(mm) != "(Intercept)", drop = FALSE]
y_matrix <- art$y

x4_columns <- grep("^x4", colnames(x_matrix), value = TRUE)
x9_columns <- grep("^x9", colnames(x_matrix), value = TRUE)

fit_matrix <- mfp2(
  x = x_matrix,
  y = y_matrix,
  term_groups = list(x4 = x4_columns, x9 = x9_columns),
  keep = c("x4", "x9"),
  verbose = FALSE
)

fit_matrix
```

The subsetting rules for factors and grouped matrix terms are described in
[Subsetting and analysis populations](#subsetting-analysis-populations).

# Logistic regression

Logistic regression is used to model binary outcomes. In `mfp2()`, logistic
regression is fitted by setting `family = "binomial"`.

For illustration, we use the Pima Indians dataset included in the `mfp2`
package. The dataset was downloaded from the
[MFP website](https://mfp.imbi.uni-freiburg.de/book#dataset_tables). It
originates from a study of 768 female Pima Indians, a population known to be
particularly susceptible to diabetes. Of the 768 individuals, 268 developed
diabetes. The dataset contains a binary outcome variable indicating the presence
(1) or absence (0) of diabetes, together with eight covariates. Further details
are given in Royston and Sauerbrei (2008, Chapter 9.7).

In `mfp2()`, a binomial response may be supplied as a binary 0/1 vector, a
two-level factor, or a two-column matrix of grouped counts in the form
`cbind(successes, failures)`. Apart from the `family` argument, the remaining
arguments are nearly identical to those used for Gaussian models in both the
default and formula interfaces.

```{r out.lines = 103}
# Load pima data
data("pima")
head(pima)

# Covariate matrix
x <- as.matrix(pima[, 2:9])

# Binary outcome
y <- as.vector(pima$y)

# Fit MFP logistic regression model
fit <- mfp2(
  x, y,
  family = "binomial",
  verbose = FALSE
)

fit
```

The output shows that four continuous variables are selected: `glucose` with
power 1, `bmi` with power -2, `diabetes` with power 1, and `age` with powers
0 and 3. The selected functions can be visualized using the `plot()` method, as
shown below.

```{r out.lines = 103, fig.show='hold', fig.cap="Pima data. Graphical presentation of results from an MFP analysis", echo=FALSE, fig.width=7, fig.height=6, out.width="100%", fig.align="center"}
plots <- plot(fit)

# Rename y-label for all plots
plots <- lapply(plots, function(v) {
  v + ggplot2::ylab("Partial predictor + residuals")
})

# Combine multiple ggplot objects into a single figure
patchwork::wrap_plots(plots, ncol = 2)
```

# Survival data

We illustrate two analyses reported by Sauerbrei and Royston (1999). The example
uses the `gbsg` dataset, which contains prognostic-factor data from the German
Breast Cancer Study Group for patients with node-positive breast cancer. The
response is recurrence-free survival time (`rectime`), and the censoring
indicator is `censrec`. The dataset contains 686 patients, of whom 299
experienced the event.

We use Cox regression to model the log hazard of recurrence as a function of
several prognostic factors. Five covariates are continuous (`age`, `size`,
`nodes`, `pgr`, and `er`), one is binary (`meno`), and `grade` is ordinal. The
ordinal variable `grade` is represented using two dummy variables, `grade_1` and
`grade_2`. The treatment variable, hormonal therapy (`hormon`), is forced into
the model. As an alternative analysis, we also stratify the model by `hormon`.

We first use `mfp2()` with the default settings to build a model from the initial
set of eight covariates. The nominal significance levels for variable selection
and FP function selection are set to 0.05 for all variables, except for
`hormon`, which is forced into the model using the `keep` argument.

For survival outcomes, the response variable must be created using
`survival::Surv()`. Other arguments are used in the same way as described above
for Gaussian models.

```{r out.lines = 103}
# Load gbsg data
data("gbsg")

# Encode grade manually using cumulative threshold indicators.
# Construct the ordinal coding explicitly so that the thresholds and
# reference category are visible in the vignette.
grade_levels <- sort(unique(as.character(gbsg$grade)))
grade_index <- match(as.character(gbsg$grade), grade_levels)

if (length(grade_levels) != 3L || anyNA(grade_index)) {
  stop(
    "`grade` must contain exactly three non-missing ordered levels.",
    call. = FALSE
  )
}

# grade_1 compares grades 2 and 3 with grade 1.
gbsg$grade_1 <- as.integer(grade_index >= 2L)

# grade_2 compares grade 3 with grades 1 and 2.
gbsg$grade_2 <- as.integer(grade_index >= 3L)

# Preserve the source variable as an ordered factor for the formula example.
gbsg$grade_ord <- ordered(gbsg$grade, levels = grade_levels)

# Covariate matrix. Exclude the source grade variables and retain the manually
# created grade indicators. They are grouped below as one conceptual term.
x <- as.matrix(gbsg[, setdiff(
  names(gbsg),
  c("id", "grade", "grade_ord", "enodes", "rectime", "censrec")
), drop = FALSE])
head(x, 10)

# Create survival outcome
y <- survival::Surv(gbsg$rectime, gbsg$censrec)

# Fit MFP Cox model and force hormon into the model
fit1 <- mfp2(
  x, y,
  family = "cox",
  term_groups = list(grade = c("grade_1", "grade_2")),
  keep = "hormon",
  control = survival::coxph.control(iter.max = 50),
  verbose = FALSE
)

fit1
```

Sauerbrei and Royston (1999) noted that, on medical grounds, the risk function
for `nodes`, the number of positive lymph nodes, should be monotonic. However,
the selected FP function for `nodes`, with powers $(-2, -1)$, was not
monotonic. They therefore proposed a preliminary exponential transformation,

$$
\text{enodes} = \exp(-0.12 \times \text{nodes}),
$$

followed by fitting a degree-1 FP function for `enodes`. This produced a
monotonic risk function. The value $-0.12$ was estimated univariately using
nonlinear Cox regression. To ensure the desired direction of the transformation,
Sauerbrei and Royston (1999) restricted the candidate powers for `enodes` to
positive values. Their Model III can be fitted using the following R code, which
yields results identical to those from the
[Stata `mfp` command](https://www.stata.com/manuals13/rmfp.pdf):

```{r out.lines = 103}
# Replace nodes with enodes, retaining the grouped grade indicators.
x <- cbind(
  x[, setdiff(colnames(x), "nodes"), drop = FALSE],
  enodes = gbsg$enodes
)

# Fit MFP Cox model and force hormon into the model
fit2 <- mfp2(
  x, y,
  family = "cox",
  term_groups = list(grade = c("grade_1", "grade_2")),
  keep = "hormon",
  powers = list(enodes = c(0.5, 1, 2, 3)),
  control = survival::coxph.control(iter.max = 50),
  verbose = FALSE
)

fit2
```

See Figure 1 in Sauerbrei and Royston (1999) for the fitted functions for
`nodes`, with powers $(-2, -1)$, and for `enodes`.

## Stratified Cox model

The stratified Cox model is an extension of the Cox proportional hazards (PH)
model that allows the baseline hazard to differ across strata (Therneau et al.,
2000). This is useful when a covariate does not satisfy the PH assumption. The
model includes covariates that are assumed to satisfy the PH assumption, while
the stratification variable itself is not estimated as a regression coefficient.
Instead, it is used to define subgroups with separate baseline hazard functions.

Similar to `coxph()` in the **survival** package, `mfp2()` supports
stratification through the `strata` argument or through `strata()` in the formula
interface. The fitted MFP model assumes that the selected FP functions and their
coefficients are common across strata. Therefore, users should assess whether
this assumption is appropriate for their data.

As an illustration, we stratify the analysis by hormonal therapy (`hormon`). The
following code shows how to fit a stratified Cox model using both the default
and formula interfaces.

```{r out.lines = 103}
# Default interface
hormon_column <- match("hormon", colnames(x))
fit2 <- mfp2(
  x[, -hormon_column, drop = FALSE],
  y,
  family = "cox",
  strata = x[, hormon_column],
  term_groups = list(grade = c("grade_1", "grade_2")),
  verbose = FALSE
)

fit2

# Formula interface
fit3 <- mfp2(
  survival::Surv(rectime, censrec) ~ age + meno + size + enodes + pgr + er +
    grade_ord + survival::strata(hormon),
  family = "cox",
  data = gbsg,
  verbose = FALSE
)

fit3
```

# References

Sauerbrei, W., Royston, P. and Kipruto, E. (2025). Multivariable
fractional polynomial models and extensions. In *International
Encyclopedia of Statistical Science* (pp. 1609-1616). Berlin,
Heidelberg: Springer Berlin Heidelberg.

Huebner, M., le Cessie, S., Schmidt, C.O. and Vach, W. (2018). A
contemporary conceptual framework for initial data analysis.
*Observational Studies*, 4, 171-192.

Royston, P. and Altman, D.G. (1994). Regression using fractional
polynomials of continuous covariates: parsimonious parametric modelling.
*Journal of the Royal Statistical Society Series C: Applied Statistics*,
43(3), 429-467.

Royston, P. (2014). A smooth covariate rank transformation for use in
regression models with a sigmoid dose-response function. *Stata
Journal*, 14(2), 329-341.

Royston, P. and Sauerbrei, W. (2004). A new approach to modelling
interactions between treatment and continuous covariates in clinical
trials by using fractional polynomial submodels. *Statistics in
Medicine*, 23(16), 2509-2525.

Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A
Pragmatic Approach to Regression Analysis Based on Fractional
Polynomials for Modelling Continuous Variables*. John Wiley & Sons.

Royston, P. and Sauerbrei, W. (2016). mfpa: Extension of mfp using the
ACD covariate transformation for enhanced parametric multivariable
modeling. *Stata Journal*, 16(1), 72-87.

Sauerbrei, W. and Royston, P. (1999). Building multivariable prognostic
and diagnostic models: transformation of the predictors by using
fractional polynomials. *Journal of the Royal Statistical Society Series
A: Statistics in Society*, 162(1), 71-94.

Sauerbrei, W., Meier-Hirmer, C., Benner, A. and Royston, P. (2006).
Multivariable regression model building by using fractional polynomials:
description of SAS, STATA and R programs. *Computational Statistics and
Data Analysis*, 50, 3464-3485.

Sauerbrei, W. and Royston, P. (2007). A new strategy for meta-analysis
of continuous covariates in observational studies. *Statistics in
Medicine*, 26(7), 1424-1442.

Stamey, T.A., Kabalin, J.N., McNeal, J.E., Johnstone, I.M., Freiha, F.,
Redwine, E.A. and Yang, N. (1989). Prostate specific antigen in the
diagnosis and treatment of adenocarcinoma of the prostate. II. Radical
prostatectomy treated patients. *Journal of Urology*, 141(5), 1076-1083.

StataCorp (2023). *Stata 18 Base Reference Manual*. College Station, TX:
Stata Press.

Therneau, T.M., Grambsch, P.M. and Pankratz, V.S. (2000). Penalized
survival models and frailty. *Journal of Computational and Graphical
Statistics*, 12(1), 156-175.

```{r cleanup, include=FALSE}
options(width = oldwidth)
```
