---
title: "Spike-at-Zero Modelling with mfp2"
subtitle: 'Package version `r as.character(utils::packageVersion("mfp2"))`'
author: "Edwin Kipruto and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Spike-at-Zero Modelling with mfp2}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  out.width = "100%",
  fig.align = "center",
  dev = "png",
  dpi = 96
)

library(mfp2)
library(ggplot2)
```

# Overview

In epidemiological and clinical research, some covariates have a
semi-continuous distribution: a non-negligible proportion of observations
are concentrated exactly at zero, and the remaining observations are
distributed continuously over positive values. This pattern is commonly
referred to as a *spike at zero* (SAZ). Typical examples include
alcohol consumption (zero for teetotallers, positive for drinkers) and
cigarette smoking (zero for non-smokers, positive for smokers).

Standard approaches to modelling SAZ covariates have important
limitations. Categorising the covariate (e.g. zero, low, medium, high)
discards information, depends on arbitrary cutpoints, and can obscure
nonlinear relationships. Treating the covariate as an ordinary
continuous variable constrains zero and positive values to lie on the
same response function, which may be inappropriate when the zero group
represents a qualitatively different state.

The SAZ extension in `mfp2` addresses this by decomposing an SAZ
covariate into two components (Royston, Sauerbrei, and Becher, 2010;
Becher, Lorenz, and Engel, 2012):

- a **binary indicator** $Z$ for whether the value is zero ($Z = 1$
  when $x = 0$; $Z = 0$ when $x > 0$), and
- a **fractional polynomial function** $f^+(x)$ of the positive values
  only, with zero values contributing nothing to this component.

The two components do not contribute to the linear predictor
simultaneously: for observations with $x = 0$, the linear predictor
includes only the binary indicator's coefficient; for observations with
$x > 0$, it includes only the FP function of $x$. The model imposes no
continuity constraint at zero, allowing the zero group to have an
effect distinct from the function fitted to positive values.

The SAZ selection algorithm then determines which components the final
model requires. Under $p$-value-based selection, a two-stage procedure
is used: Stage 1 determines variable inclusion and selects the
positive-component functional form (linear, FP1, or FP2), and Stage 2
determines whether to retain both components, only the binary
indicator, or only the positive component. Under AIC or BIC, a
one-stage procedure compares all eligible models jointly. In a
multivariable model, SAZ selection for each covariate is performed
conditionally on the current forms of the other covariates during MFP
backfitting, which allows multiple SAZ covariates to be modelled
simultaneously.

For the complete algorithm, eligibility equations, candidate families,
and simulation results, see:

> Kipruto E, Sauerbrei W (YEAR). "Modelling Spike-at-Zero Covariates
> with Fractional Polynomials in R." *Journal of Statistical Software*,
> **VOL**(ISSUE), PAGES. doi:DOI.

This vignette demonstrates the practical workflow.

# When to use spike-at-zero modelling

SAZ modelling is appropriate when a nonnegative covariate has a
meaningful group of exact zeros alongside a continuous positive
component. Common examples include:

- biomarkers that are undetectable in some patients (recorded as zero),
- exposure variables where some individuals have no exposure, and
- counts or measurements where zero represents a qualitatively
  different state.

The decision to model zero separately from positive values should be
guided by subject-matter knowledge, particularly whether a distinct
effect at zero is scientifically plausible. A covariate qualifies for
SAZ modelling when both the zero group and the positive group each
contain at least the proportion of observations specified by
`min_saz_prop` (default 0.10, i.e. 10%). This threshold is a pragmatic
safeguard; in large samples, a smaller value may be appropriate. Users
should consider both the proportion and the absolute number of
observations in each component when choosing `min_saz_prop`.

SAZ covariates must be nonnegative: $x = 0$ defines the zero component
and $x > 0$ the positive component. Negative values are rejected; they
must be explicitly recoded if they should represent the zero group.
Binary covariates are ineligible because their two levels are already
fully represented by a single indicator.

# Example: prostate cancer data

We use the `prostate` dataset (Stamey et al., 1989), bundled with
`mfp2`. The outcome is `lpsa` (log prostate-specific antigen). The
covariate `pgg45` (percentage of Gleason score 4 or 5) is a natural SAZ
candidate.

```{r data-summary}
data("prostate")

cat("Number of observations:", nrow(prostate), "\n")
cat("pgg45 == 0:", sum(prostate$pgg45 == 0),
    sprintf("(%.0f%%)\n", 100 * mean(prostate$pgg45 == 0)))
cat("pgg45  > 0:", sum(prostate$pgg45 > 0),
    sprintf("(%.0f%%)\n", 100 * mean(prostate$pgg45 > 0)))
```

Both proportions exceed the default eligibility threshold of 10%, so
`pgg45` qualifies for SAZ modelling.

## Univariable SAZ analysis (p-value criterion)

Request SAZ modelling with `spike = TRUE` inside the `fp()` term:

```{r fit-univariable}
fit_pgg45 <- mfp2(
  lpsa ~ fp(pgg45, spike = TRUE),
  data = prostate,
  family = "gaussian",
  criterion = "pvalue",
  verbose = FALSE
)

fit_pgg45
```

The output reports the SAZ decision. In this example, only the binary
indicator is retained ("binary only"). The positive-component FP
function was dropped at Stage 2 of the closed-test procedure, meaning
that among patients with any Gleason 4/5 cells, the percentage does
not add predictive information beyond simply knowing that such cells
were present.

## Predictions

The `predict()` method applies the selected SAZ structure
automatically. Supply new data on the original covariate scale:

```{r predictions}
newdat <- data.frame(pgg45 = c(0, 4, 20, 45, 60, 100))
predict(fit_pgg45, newdata = newdat, type = "link")
```

For a "binary only" model, all positive values produce the same
prediction (the intercept), while `pgg45 = 0` shifts the prediction by
the binary indicator's coefficient.

## Plot

The `plot()` method for `mfp2` objects produces a ggplot2 figure. For
a "binary only" SAZ decision, it shows two point estimates (zero and
positive groups) with error bars:

```{r plot-binary-only, fig.height=4}
plots <- plot(fit_pgg45, terms = "pgg45", partial_only = TRUE)
plots[["pgg45"]]
```

## Using AIC instead of p-values

With `criterion = "aic"`, model selection uses a one-stage procedure
that compares all eligible models (null, binary-only, positive-only,
and positive-plus-binary) jointly, selecting the one with the smallest
AIC. This approach tends to be less stringent and may retain more
components:

```{r fit-aic}
fit_aic <- mfp2(
  lpsa ~ fp(pgg45, spike = TRUE),
  data = prostate,
  family = "gaussian",
  criterion = "aic",
  verbose = FALSE
)

fit_aic
```

With AIC, both the binary indicator and a linear positive-component
function are retained ("continuous + binary"). The plot now shows a
continuous curve over positive values with a separate point at zero:

```{r plot-aic, fig.height=4.5}
plots_aic <- plot(fit_aic, terms = "pgg45", partial_only = TRUE)
plots_aic[["pgg45"]]
```

BIC applies a stronger penalty and, for these data, selects the same
"binary only" model as the p-value criterion.

# Multivariable SAZ analysis

In a multivariable model, SAZ selection for each covariate is performed
conditionally on the current forms of the other covariates during MFP
backfitting. In each cycle, every SAZ covariate is re-evaluated given
the latest selected forms of the remaining predictors. In this example,
`pgg45` is the SAZ candidate while the other covariates use standard MFP
selection:

```{r fit-multivariable}
fit_multi <- mfp2(
  lpsa ~ fp(pgg45, spike = TRUE) + fp(age) +
    fp(cavol) + fp(weight) + fp(bph) + fp(cp) + svi,
  data = prostate,
  family = "gaussian",
  criterion = "pvalue",
  verbose = FALSE
)

fit_multi
```

After adjustment for the other covariates, `pgg45` is excluded
entirely ("not selected"). The retained model includes only `cavol`,
`weight`, and `svi`.

# Controlling SAZ retention

Three arguments influence how much of the SAZ machinery can be
overridden:

## `select = 1`: force inclusion, allow form selection

Setting `select = 1` inside `fp()` forces the variable into the
model (Stage 1 retention is guaranteed), but `alpha` still governs
functional-form selection and Stage 2 component removal:

```{r select-1, eval=FALSE}
mfp2(
  lpsa ~ fp(pgg45, select = 1, spike = TRUE),
  data = prostate,
  family = "gaussian",
  criterion = "pvalue",
  verbose = FALSE
)
```

## `keep`: retain a variable across criteria

The `keep` argument ensures a variable cannot be excluded, regardless
of the selection criterion:

```{r keep, eval=FALSE}
mfp2(
  lpsa ~ fp(pgg45, spike = TRUE) + fp(age) +
    fp(cavol) + fp(weight) + fp(bph) + fp(cp) + svi,
  data = prostate,
  family = "gaussian",
  criterion = "pvalue",
  keep = "pgg45",
  verbose = FALSE
)
```

## `force_max_fp = TRUE`: bypass Stage 2

Setting `force_max_fp = TRUE` retains the maximum-complexity SAZ
representation (the full FP positive component plus the binary
indicator), bypassing Stage 2 component selection entirely:

```{r force-max, eval=FALSE}
mfp2(
  lpsa ~ fp(pgg45, spike = TRUE, force_max_fp = TRUE),
  data = prostate,
  family = "gaussian",
  criterion = "pvalue",
  verbose = FALSE
)
```

# Matrix interface

The matrix interface identifies SAZ candidates through `spike_vars`
instead of `fp(..., spike = TRUE)`:

```{r matrix-interface}
x <- as.matrix(prostate[, "pgg45", drop = FALSE])
y <- prostate$lpsa

fit_mat <- mfp2(
  x, y,
  family = "gaussian",
  criterion = "pvalue",
  spike_vars = "pgg45",
  verbose = FALSE
)

fit_mat
```

Both interfaces use the same fitting procedure and produce identical
results.

# Related options: `zero` and `catzero`

The `spike` option is the most complete SAZ workflow, but two simpler
options are available when you want to fix the model structure rather
than select it:

- `fp(x, zero = TRUE)` applies the FP function to positive values only;
  zero values contribute nothing to the linear predictor. No binary
  indicator is included.
- `fp(x, catzero = TRUE)` adds a fixed binary indicator for zero
  values alongside the positive-component FP function. Unlike `spike`,
  neither the binary indicator nor the continuous component can be
  dropped — both are always in the model.

`spike = TRUE` implies both `catzero` and `zero` while SAZ remains
eligible, and then uses the SAZ selection procedure to determine which
components to retain.

# References

Kipruto E, Sauerbrei W (YEAR). "Modelling Spike-at-Zero Covariates
with Fractional Polynomials in R." *Journal of Statistical Software*,
**VOL**(ISSUE), PAGES. doi:DOI.

Becher H, Lorenz E, Engel A (2012). "Analysing Covariates with Spike
at Zero: A Modified FP Procedure and Conceptual Issues."
*Biometrical Journal*, 54(5), 686--700. doi:10.1002/bimj.201100263.

Royston P, Sauerbrei W, Becher H (2010). "Modelling Continuous
Exposures with a 'Spike' at Zero: A New Procedure Based on Fractional
Polynomials." *Statistics in Medicine*, 29, 1219--1227.
doi:10.1002/sim.3864.

Royston P, Sauerbrei W (2008). *Multivariable Model-Building: A
Pragmatic Approach to Regression Analysis based on Fractional
Polynomials for Modelling Continuous Variables*. John Wiley & Sons.

Stamey TA, Kabalin JN, McNeal JE, Johnstone IM, Freiha F, Redwine EA,
Yang N (1989). "Prostate Specific Antigen in the Diagnosis and
Treatment of Adenocarcinoma of the Prostate." *Journal of Urology*,
141(5), 1076--1083.
