---
title: "Introduction to Multivariable Fractional Polynomials"
author:
  - Edwin Kipruto
  - Michael Kammer
  - Patrick Royston
  - Willi Sauerbrei
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Introduction to Multivariable Fractional Polynomials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  out.width = "100%"
)

oldwidth <- getOption("width")
options(width = 100)

library(ggplot2)
library(patchwork)
```

# Introduction to Multivariable Fractional Polynomials (MFP)

## Overview of MFP

Multivariable regression models are widely used across various fields of science
where empirical data is analyzed. In model building, many researchers often
assume a linear function for continuous variables, sometimes after applying
“standard” transformations such as logarithms, or dividing the variable into
several categories. Assuming linearity without considering non-linear
relationships may hinder the detection of effects or cause the effects to be
mismodeled. Categorizing continuous variables, which results in modelling
implausible step functions, is a common practice but widely criticized (Royston
et al. 2006; Sauerbrei et al. 2020).

When building a descriptive model with the aim of capturing the data structure
parsimoniously, two components should be considered: first, variable selection
to identify the subset of "important" covariates that have a significant impact
on the outcome and second, whether non-linear relationships of continuous
covariates fit the data (substantially) better.

The MFP approach has been proposed as a pragmatic method that combines variable
and function selection simultaneously in multivariable  regression modelling.
This approach identifies non-linear functions for continuous variables if
sufficiently supported by the data, and eliminates covariates with no or weak
effects by backward elimination (BE). Despite its relative simplicity, the
selected models often capture the essential information from the data. The MFP
models are relatively straightforward to interpret and report, which is
important for transportability and practical usability.

In detail, the MFP procedure combines:
- variable selection through backward elimination with
- selection of fractional polynomial (FP) functions for continuous variables.

The analyst must decide on nominal significance levels $(\alpha_1, \alpha_2)$
for both components. This choice has a strong influence on the complexity of the
final model. Often, the same significance levels $(\alpha_1 = \alpha_2)$ are
used for both components, but they can also differ. The decision regarding these
significance levels strongly depends on the specific aim of the analysis.

The rest of the paper is organized as follows. Section 1.2 provides an overview
of fractional polynomial functions for a single continuous covariate in the
model, including the function selection procedure (FSP). Section 1.3 describes
the MFP approach, focusing on models involving two or more covariates.

Section 2 is an introduction to the `mfp2` package. It covers the
installation process and provides instructions for utilizing the package in
various linear regression models. The package functionality is predominantly
demonstrated using Gaussian linear models (Subsection 2.3), while other models,
such as logistic (Subsection 2.4),and Cox (Subsection
2.5), are briefly explained.

Section 3 introduces an extension of MFP using the approximate cumulative
distribution (ACD) transformation of a continuous covariate. This extension
allows for modelling a sigmoid relationship between covariates and an outcome
variable (subsection 3.1).

For more comprehensive information about MFP and its extensions, please refer to
Royston and Sauerbrei 2008 or visit [MFP website (click here)](https://mfp.imbi.uni-freiburg.de/) 
and explore the references given there.

## Fractional polynomial models for a continuous variable

Suppose that we have an outcome variable, a single continuous covariate $x$, and
a regression model relating them. A starting point is the straight-line model
$\beta_1x$ (for simplicity, we suppress the constant term, $\beta_0$). Often, a
straight line is an adequate description of the relationship, but other models
should be investigated for possible improvements in ﬁt. A simple extension of
the straight line is a power transformation model, $\beta_1x^{p}$. The latter
model has often been used by practitioners in an ad hoc way, utilizing different
choices of $p$. Royston and Altman (1994) formalized the model by calling it a
ﬁrst-degree fractional polynomial or FP1 function. The power $p$ is chosen from
a pragmatically restricted set of eight elements: $S = \{-2, -1, -0.5, 0, 0.5, 1, 2, 3\}$, 
where $x^0$ denotes the natural logarithm $\log(x)$.

As with polynomial regression, extension from one-term FP1 functions to more
complex and ﬂexible two-term FP2 functions is straightforward. The quadratic
function $\beta_1x^1 + \beta_2x^2$ is written as $\beta_1x^{p1} + \beta_2x^{p2}$
in FP terminology. The powers $p1=1$ and $p2=2$ are members of set $S$. Royston
and Altman extended the class of FP2 functions with different powers to cases
with equal powers ($p1=p2=p$) by defining them as $\beta_1x^p + \beta_2x^p \log(x)$.
These are known as repeated-powers functions. Detailed definitions of FP
functions are given in Section 4.3.1 of Royston and Sauerbrei (2008). For formal deﬁnitions, we use their notation.

FP1 functions are always monotonic and those with power $p < 0$ have an asymptote
as $x\rightarrow \infty$. FP2 functions may be monotonic or unimodal (i.e., 
have one maximum or one minimum for positive values of $x$), and they have an
asymptote as $x\rightarrow \infty$ when both $p1$ and $p2$ are negative.
For more details, see Royston and Sauerbrei (2008), Section 4.4.

Extension to FPm is straightforward but $m>2$ is hardly needed when modelling
health research data with MFP. For $m = 2$, there are 44 models available within
the set of FP powers ($S$), consisting of 8 FP1 models and 36 FP2 models. Although
the allowed class of FP functions may seem limited, it encompasses a wide range
of diverse shapes. This is illustrated in the Figure below, with the left panel
displaying eight FP1 powers and the right panel depicting a subset of FP2
powers.

```{r, eval=TRUE, echo=FALSE}
#==============================================================================
# 8 FP1 FUNCTIONS 
#===============================================================================
x <- seq(0.05, 1.05, length.out = 1000)
funx <- function(x, power){
  ifelse(power == rep(0, length(x)), log(x), x^power)
}
s <- c(-2, -1, -0.5, 0, 0.5, 1, 2, 3)
# transform x using the powers in s
outx <- lapply(s, function(s) funx(x = x, power = s))

# multiply the first 3 with -1 so that the function increase rather than decrease
# due to negative powers. these are betas i,e y = beta*x^p
k <- c(-1, -1, -1, 1, 1, 1, 1, 1)
datx <- matrix(unlist(outx), ncol = length(s))
datax <- datx %*% diag(k)
# standardize the data so that y is in the same range
dataxx <- apply(datax, 2, function(x) (x - min(x)) / (max(x) - min(x)))
colnames(dataxx) <- c(paste0("x", 1:length(s)))
dataxx <- as.data.frame(dataxx)
dataxx$x <- x
width <- 0.5
fig1 <- ggplot(dataxx, aes(x)) + 
  geom_line(aes(y = x1, colour = "x1"), linewidth = width, color = "#006400") + 
  geom_line(aes(y = x2, colour = "x2"), linewidth = width, color ="#ff0000") + 
  geom_line(aes(y = x3, colour = "x3"), linewidth = width, color ="#ffd700") +
  geom_line(aes(y = x4, colour = "x4"), linewidth = width, color ="#ff00ff") +
  geom_line(aes(y = x5, colour = "x5"), linewidth = width, color ="#ffb6c1") +
  geom_line(aes(y = x6, colour = "x6"), linewidth = width, color ="#00ff00") +
  geom_line(aes(y = x7, colour = "x7"), linewidth = width, color ="#0000ff") +
  geom_line(aes(y = x8, colour = "x8"), linewidth = width, color ="#000000") +
  geom_text(aes(x = 0.175, y = 1,size = 5, label = "-2"), color = "#006400") +
  geom_text(aes(x = 0.25, y = 0.9,size = 5, label = "-1"), color = "#ff0000") +
  geom_text(aes(x = 0.33, y = 0.83,size = 5, label = "-0.5"), color = "#ffd700") +
  geom_text(aes(x = 0.4, y = 0.725,size = 5, label = "0"), color = "#ff00ff") +
  geom_text(aes(x = 0.5, y = 0.65,size = 5, label = "0.5"), color = "#ffb6c1") +
  geom_text(aes(x = 0.59, y = 0.575,size = 5, label = "1"), color = "#00ff00") +
  geom_text(aes(x = 0.7, y = 0.475,size = 5, label = "2"), color = "#0000ff") +
  geom_text(aes(x = 0.75, y = 0.4,size = 5, label = "3"), color = "#000000") +
  labs(x="x", y="Fractional polynomial f(x)") +
  theme_bw() +
  theme(
    legend.position = "none",
    panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5),
    axis.ticks = element_blank(),
    axis.text.y = element_blank(),
    axis.title=element_text(size=18),
    axis.text.x = element_blank()) 

# Define functions
f1 <- function(x) 3 - 10*x^2 + 4*x^3
f2 <- function(x) 20 - 15.4*x^2 + 4*x^3
f3 <- function(x) -20 + 6*log(x) + 6*log(x)*log(x)
f4 <- function(x) 20 + 0.3*(x^-2) - 4*(x^-1)
f5 <- function(x) -10 + 5*(x^0.5) + 14*(x^-0.5)
f6 <- function(x) 33 + 19*log(x) - 7*(x^2)
f7 <- function(x) -10 + 10*(x - 1.5) + 10*(x - 1.5)^2

# Create data frames for each function
x <- seq(0.1, 3, by = 0.01)
df1 <- data.frame(x = x, y = f1(x))
df2 <- data.frame(x = x, y = f2(x))
df3 <- data.frame(x = x, y = f3(x))
df4 <- data.frame(x = x, y = f4(x))
df5 <- data.frame(x = x, y = f5(x))
df6 <- data.frame(x = x, y = f6(x))
df7 <- data.frame(x = x, y = f7(x))

# Plot functions
fig2 <- ggplot() +
    geom_line(data = df1, aes(x = x, y = y), color = "#e41a1c") +
    geom_line(data = df2, aes(x = x, y = y), color = "#377eb8") +
    geom_line(data = df3, aes(x = x, y = y), color = "#4daf4a") +
    geom_line(data = df4, aes(x = x, y = y), color = "#984ea3") +
    geom_line(data = df5, aes(x = x, y = y), color = "#ff7f00") +
    geom_line(data = df6, aes(x = x, y = y), color = "#FF00FF") +
    geom_line(data = df7, aes(x = x, y = y), color = "#a65628") +
    scale_x_continuous(limits = c(0, 3)) +
    scale_y_continuous(expand = c(0, 0), limits = c(-25,40)) + theme_classic() +
    geom_text(aes(x = 0.75, y = 2.4,size = 5, label = "(2, 3)"), color = "#e41a1c") +
    geom_text(aes(x = 1.6, y =1.95,size = 5, label = "(2, 3)"), color = "#377eb8")+
    geom_text(aes(x = 0.7, y = -18,size = 5, label = "(0, 0)"), color = "#4daf4a") +
    geom_text(aes(x = 1.25, y = 20,size = 5, label = "(-2, -1)"), color = "#984ea3") +
    geom_text(aes(x = 1.3, y = 12,size = 5, label = "(-0.5, 0.5)"), color = "#ff7f00") +
    geom_text(aes(x = 1.25, y = 29.5,size = 5, label = "(0, 2)"), color = "#FF00FF") +
    geom_text(aes(x = 1, y = -9.5,size = 5, label = "(1, 2)"), color = "#a65628") +
    ylab(" ") +
    theme_bw() +
    theme(
      legend.position = "none",
      panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5),
      axis.title = element_text(size=18),
      axis.ticks = element_blank(),
      axis.text.y = element_blank(),
      axis.text.x = element_blank()
    )

figure <- patchwork::wrap_plots(fig1, fig2, 
                               ncol = 2, nrow = 1, 
                               widths = 8, heights = 2)

```


```{r fig1, echo=FALSE, fig.cap="Illustration of the flexibility of the FP family. 8 FP1 functions (left panel) and a subset of FP2 functions (right panel). FPs are global functions. See section 1.3.1.4 for a proposed extension of local features", fig.height=4.5, fig.width=7, out.width="100%", fig.align="center"}
figure
```


Based on extensive experience with real data and several simulation studies, FP1
and FP2 are generally considered adequate in the context of multivariable model
building, particularly when the selection of variable and functional forms are required
(Royston and Sauerbrei 2008; Sauerbrei et al. 2007a).

## Function selection procedure (FSP)

This short description of the MFP methodology content is based on previously
published encyclopedia articles (Sauerbrei and Royston, 2011; Sauerbrei and
Royston, 2016).

Choosing the best fitting FP1, FP2, or linear function for a given dataset is
conducted by grid search, computing the deviance (minus twice the maximized
log-likelihood) and making a decision for the best function based on a suitable
algorithm (e.g. via pre-defined significance levels). 

The default function in most implementations of MFP is a linear function.
Therefore, unless the data support a more complex FP function, a straight line
model is chosen. Concerning the default, there are occasional exceptions. For example, to model
time-varying regression coefficients in the Cox model, Sauerbrei et al. (2007b)
chose a default time ($t$) transformation of $\log (t)$ rather than $t$.

It can be assumed that deviance diﬀerence between an FPm and an FP(m-1) model is
distributed approximately as central $\chi^2$ on 2 degrees of freedom (d.f.)
(Royston and Sauerbrei 2008, Chapter 4.9; Ambler and Royston (2001)). To select a
speciﬁc function, a closed test procedure was proposed (Royston and Sauerbrei
2008, Section 4.10). The complexity of the ﬁnally chosen function requires
pre-definition of the nominal significance level $(\alpha)$ and the degree ($m$)
of the most complex FP model allowed. Typical choices are $\alpha = 0.05$ and
FP2 ($m = 2$).  We illustrate the strategy for $m = 2$, which runs as follows:

 1. Use the deviance difference to test the best FP2 model for $x$ at the
 $(\alpha)$ signiﬁcance level against the null model using 4 d.f. If the test is
 not signiﬁcant, stop and conclude that the eﬀect of $x$ is "not signiﬁcant" at
 the $\alpha$ level. Otherwise continue. 
 2. Test the best FP2 for $x$ against a straight line at the $\alpha$ level
 using 3 d.f. If the test is not signiﬁcant, stop, the ﬁnal model being a
 straight line. Otherwise continue.
 3. Test the best FP2 for $x$ against the best FP1 at the $\alpha$ level using 2
 d.f. If the test is not signiﬁcant, the ﬁnal model is FP1, otherwise, the ﬁnal
 model is FP2. This marks the end of procedure.

Step 1 is a test of the overall association of the outcome with $x$. 
Step 2 examines the evidence for nonlinearity. Step 3 chooses between a simpler 
or a more complex non-linear model.
 
# MFP procedure

When developing a multivariable model with a relatively large number of
candidate covariates (say between 10 and 30, we are not envisaging the case of
high-dimensional data), an important distinction is between descriptive,
predictive and explanatory modelling (Shmueli, 2010). MFP was mainly developed
for descriptive modelling, aiming to capture the data structure parsimoniously.
Nevertheless, a suitable descriptive model often has a fit similar to a model
whose aim is good prediction (Sauerbrei et al. 2007a). In some fields, the term
explanatory modelling is used exclusively for testing causal theory.

In many areas of science, the main interest often lies in the identification of
inﬂuential variables and determination of appropriate functional forms for
continuous variables. Often, linearity is presumed without checking this
important assumption. The MFP procedure was proposed as a pragmatic strategy to
investigate whether nonlinear functions can improve the model ﬁt (Royston and
Sauerbrei, 2008; Sauerbrei and Royston, 1999). MFP combines backward elimination
for the selection of variables with a systematic search for possible
nonlinearity by the function selection procedure (FSP). The extension is
feasible with any type of regression model to which BE is applicable. The
philosophy behind MFP modeling is to create interpretable and relatively simple
models (Sauerbrei et al. 2007a). Consequently, an analyst using MFP should be
less concerned about failing to include variables with a weak eﬀect or failing
to identify minor curvature in a functional form of a continuous covariate.
Modiﬁcations that may improve MFP models are combination with post-estimation
shrinkage (Dunkler et al., 2016, R-package shrink) and a more systematic check
for overlooked local features (Binder and Sauerbrei, 2010, currently not
implemented in the `mfp2` package). Successful use of MFP requires only general
knowledge about building regression models.

Two nominal signiﬁcance level values are the main tuning parameters: $\alpha_1$
for selecting variables with BE (in the ﬁrst step of the FSP) and $\alpha_2$ for
comparing the ﬁt of functions within the FSP. Often, $\alpha_1=\alpha_2$ is a
good choice. If available, subject-matter knowledge should replace or at least
guide data-dependent model choice. Only minor modiﬁcations are required to
incorporate various types of subject-matter knowledge into MFP modeling. For a
detailed example, see Sauerbrei and Royston (1999). Recommendations for MFP
modeling in practice are given in Sauerbrei et al. (2007a) and in Royston and
Sauerbrei (2008, Section 12.2). In the `mfp2` package you may replace
$\alpha_1$ and $\alpha_2$ by AIC-P or BIC-P.

# MFP -- Key issues and approaches to handling them

Mainly focusing on the FP component, we brieﬂy mention key issues of MFP
modeling and refer to the literature for further reading. A brief overview is
also given on the MFP website (https://mfp.imbi.uni-freiburg.de/fp). Regarding
variable selection, we have summarized relevant issues and provided arguments
for backward elimination as our preferred strategy (Royston and Sauerbrei 2008,
Chapter 2). Even when a search for model improvement using a nonlinear function
is not considered, that is, all functions are assumed linear, it is infeasible
to derive a suitable and stable model for description in small datasets. Below
we provide some information about sample size needed, but implicitly we assume
that the sample size is "suﬃcient".

## The variable has to be positive

The class of FP1 and FP2 functions includes a log and other transformations
which require that the continuous variable must be positive. A preliminary
origin-shift transformation can be applied as shown in section 2.3.2.

Sometimes, there are variables that exhibit a spike at zero. This means that a
certain proportion of individuals have zero values, while among those with
positive values, the variable has a continuous distribution. In epidemiology,
typical examples are cigarette consumption or alcohol intake. There are
different approaches to modeling such variables. Royston and Sauerbrei
(2008, chapter 4.15) and Royston et al. (2010) proposed a method that involves
adding a small constant (to have positive values) and a binary indicator (for
zero or positive values) to the regression model. Additionally, they utilized
fractional polynomials to determine the functional form of the variable. In
a two-stage procedure, they assessed whether the binary variable and/or the
continuous function is required for a good fit to the data. Becher et al. (2012)
proposed a slight modification of this approach. These two approaches are currently not
implemented in the `mfp2` package.

## Sample size and influential observations

All statistical models are potentially adversely aﬀected by inﬂuential
observations or "outliers". However, compared with models that comprise only
linear functions, the situation may be more critical for FP functions because
logarithmic or negative power transformations may produce extreme functional
estimates at small values of $x$. Conversely, the same may happen with large
positive powers at large values of $x$. Such transformations may create inﬂuential
observations that may affect parts of the FSP. To mitigate the impact of
influential observations, it is important to assess the robustness of FP
functions. Some suggestions for investigating inﬂuential points (IPs) and
handling such issues in MFP modeling can be found in Royston and Sauerbrei
(2008, Chapters 5 and 10) and their paper on improving the robustness of FP
models (Royston and Sauerbrei, 2007). Using synthetic data, a more detailed
investigation of IPs is given in Sauerbrei et al. (2023). The authors conclude
that for smaller sample sizes, IPs and low power are important reasons that the
MFP approach may not be able to identify underlying functional relationships for
continuous variables and selected models might differ substantially from the
true model. However, for larger sample sizes (about 50 or more observations per
variable) a carefully conducted MFP analysis (which includes checks for IPs and
other aspects of model criticism) is often a suitable way to select a
multivariable regression model which includes continuous variables.

## Lack of fit

Based on experience we argue that FP1 or FP2 functions are sufficient to derive
a good fitting model. Occasionally, higher order FP functions (with $m = 3$ or 
$m = 4$) are useful in data analysis. Negative exponential or other types of
pre-transformations may help to improve the fit of a complex function or may
allow to incorporate subject specific knowledge. An example is discussed in
section 3.1.3 in Sauerbrei and Royston (1999).

## Local features

Unlike splines which have a local interpretation of the fitted function, FPs
provide a curve with a global interpretation. To investigate possible
"overlooked" local features of an FP function, Binder and Sauerbrei (2010)
conducted a systematic analysis of residuals from a model obtained by MFP. If
local features are detected by their MFP + L procedure, statistically signiﬁcant
local polynomials are then parsimoniously added to the FP function. This enables the
identification and incorporation of local features that may have been missed by
the global FP function. This approach is not currently implemented in the `mfp2`
package, but it is a potential area for future extensions.


# References

Ambler, G. and Royston, P. (2001). Fractional polynomial model selection procedures: investigation of type I error rate. *Journal of Statistical Computation and Simulation*, 69, 89–108.

Binder, H. and Sauerbrei, W. (2010). Adding local components to global functions for continuous covariates in multivariable regression modeling. *Statistics in Medicine*, 29, 808–817.

Binder, H., Sauerbrei, W. and Royston, P. (2013). Comparison between splines and fractional polynomials for multivariable model building with continuous covariates: a simulation study with continuous response. *Statistics in Medicine*, 32, 2262–2277.

Buchholz, A. and Sauerbrei, W. (2011). Comparison of procedures to assess non-linear and time-varying effects in multivariable models for survival data. *Biometrical Journal*, 53, 308–331. DOI: 10.1002/bimj.201000159.

Draper, D. (1995). Assessment and propagation of model selection uncertainty (with discussion). *Journal of the Royal Statistical Society: Series B*, 57, 45–97.

Dunkler, D., Sauerbrei, W. and Heinze, G. (2016). Global, parameterwise and joint post-estimation shrinkage. *Journal of Statistical Software*.

Faes, C., Aerts, M., Geys, H. and Molenberghs, G. (2007). Model averaging using fractional polynomials to estimate a safe level of exposure. *Risk Analysis*, 27, 111–123.

Huebner, M., le Cessie, S., Schmidt, C. O. and Vach, W. (2018). A contemporary conceptual framework for initial data analysis. *Observational Studies*, 4(1), 171–192.

Royston, P. (2014). A smooth covariate rank transformation for use in regression models with a sigmoid dose-response function. *The Stata Journal*, 14, 329–341.

Royston, P. and Altman, D. G. (1994). Regression using fractional polynomials of continuous covariates: parsimonious parametric modelling. *Applied Statistics*, 43(3), 429–467.

Royston, P. and Sauerbrei, W. (2004). A new approach to modelling interactions between treatment and continuous covariates in clinical trials by using fractional polynomials. *Statistics in Medicine*, 23, 2509–2525.

Royston, P. and Sauerbrei, W. (2007). Improving the robustness of fractional polynomial models by preliminary covariate transformation. *Computational Statistics and Data Analysis*, 51, 4240–4253.

Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for Modelling Continuous Variables*. John Wiley & Sons, Chichester, UK.

Royston, P. and Sauerbrei, W. (2013). Interaction of treatment with a continuous variable: simulation study of significance level for several methods of analysis. *Statistics in Medicine*, 32(22), 3788–3803.

Royston, P. and Sauerbrei, W. (2014). Interaction of treatment with a continuous variable: simulation study of power for several methods of analysis. *Statistics in Medicine*, 33, 4695–4708.

Royston, P., Altman, D. G. and Sauerbrei, W. (2006). Dichotomizing continuous predictors in multiple regression: a bad idea. *Statistics in Medicine*, 25, 127–141.

Royston, P. and Sauerbrei, W. (2016). mfpa: Extension of mfp using the ACD covariate transformation for enhanced parametric multivariable modeling. *The Stata Journal*, 16(1), 72–87.

Royston, P., Sauerbrei, W. and Becher, H. (2010). Modelling continuous exposures with a ‘spike’ at zero: a new procedure based on fractional polynomials. *Statistics in Medicine*, 29, 1219–1227.

Sauerbrei, W. (1999). The use of resampling methods to simplify regression models in medical statistics. *Applied Statistics*, 48, 313–329.

Sauerbrei, W., Abrahamowicz, M., Altman, D. G., et al. (2014). STrengthening Analytical Thinking for Observational Studies: the STRATOS initiative. *Statistics in Medicine*, 33(30), 5413–5432.

Sauerbrei, W., Kipruto, E. and Balmford, J. (2022). Effects of influential points and sample size on the selection and replicability of multivariable fractional polynomial models. *Diagnostic and Prognostic Research*, 7(1), 7.

Sauerbrei, W., Meier-Hirmer, C., Benner, A. and Royston, P. (2006). Multivariable regression model building by using fractional polynomials: description of SAS, STATA and R programs. *Computational Statistics and Data Analysis*, 50, 3464–3485.

Sauerbrei, W., Perperoglou, A., Schmid, M., Abrahamowicz, M., Becher, H., Binder, H., Dunkler, D., Harrell Jr., F. E., Royston, P. and Heinze, G. for TG2 of the STRATOS initiative (2020). State of the art in selection of variables and functional forms in multivariable analysis: outstanding issues. *Diagnostic and Prognostic Research*, 4, 3, 1–18. DOI: 10.1186/s41512-020-00074-3.

Sauerbrei, W. and Royston, P. (1999). Building multivariable prognostic and diagnostic models: transformation of the predictors using fractional polynomials. *Journal of the Royal Statistical Society: Series A*, 162, 71–94.

Sauerbrei, W. and Royston, P. (2011). Multivariable fractional polynomial models. In Lovric, M. (ed.), *International Encyclopedia of Statistical Science*. Springer, Berlin, pp. 899–902.

Sauerbrei, W. and Royston, P. (2016). Multivariable fractional polynomial models. *Wiley StatsRef: Statistics Reference Online*, 1–8. DOI: 10.1002/9781118445112.stat07861.

Sauerbrei, W., Royston, P. and Binder, H. (2007b). Selection of important variables and determination of functional form for continuous predictors in multivariable model-building. *Statistics in Medicine*, 26, 5512–5528.

Sauerbrei, W., Royston, P. and Look, M. (2007a). A new proposal for multivariable modelling of time-varying effects in survival data based on fractional polynomial time-transformation. *Biometrical Journal*, 49, 453–473.

Sauerbrei, W. and Royston, P. (2007). Modelling to extract more information from clinical trials data: on some roles for the bootstrap. *Statistics in Medicine*, 26, 4989–5001.

Shmueli, G. (2010). To explain or to predict? *Statistical Science*, 3, 289–310.

Stamey, T. A., Kabalin, J. N., McNeal, J. E., Johnstone, I. M., Freiha, F., Redwine, E. A. and Yang, N. (1989). Prostate specific antigen in the diagnosis and treatment of adenocarcinoma of the prostate. II. Radical prostatectomy treated patients. *The Journal of Urology*, 141(5), 1076–1083.

StataCorp. (2023). *Stata: Release 18. Statistical Software*. StataCorp LLC, College Station, TX.

Therneau, T. M. and Grambsch, P. M. (2000). *Modeling Survival Data: Extending the Cox Model*. Springer, New York, pp. 39–77.

```{r cleanup, include=FALSE}
options(width = oldwidth)
```