---
title: "MFPI: Multivariable Fractional Polynomial Interaction"
subtitle: 'Package version `r as.character(utils::packageVersion("mfp2"))`'
author: "Edwin Kipruto and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{MFPI: Multivariable Fractional Polynomial Interaction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-size: 15px;
  line-height: 1.5;
}

/* Justify narrative paragraphs */
p {
  text-align: justify;
  text-justify: inter-word;
}

/* Justify normal prose, but not code, tables, headings, or captions */
p, li {
  text-align: justify;
  line-height: 1.5;
}

h1, h2, h3, h4 {
  margin-top: 1.4em;
  margin-bottom: 0.6em;
  color: #1f77b4;
}

#TOC a {
  color: inherit;
  text-decoration: none;
}

pre {
  font-size: 13px;
  line-height: 1.4;
  overflow-x: auto;
}

code {
  font-size: 90%;
}

img {
  max-width: 100%;
  height: auto;
}

.figure {
  margin-top: 1.2em;
  margin-bottom: 1.2em;
}

.caption {
  font-size: 90%;
  color: #555555;
  text-align: left;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  dev = "png",
  dpi = 150
)

oldwidth <- getOption("width")
options(width = 100)

library(mfp2)
library(ggplot2)
library(patchwork)
library(survival)
```


# Introduction

Investigating whether the effect of a treatment or other grouping variable changes across the values of a continuous covariate is a common and important problem in all areas of science in which empirical data are analyzed. For many years our work concentrates on methodological research for clinical and epidemiological studies, a reason that terminology, explanations and examples come from these areas. However, the Multivariable Fractional Polynomial Interaction (MFPI) approach is based on multiple linear regression, which belongs to the most popular approaches in many other areas. To investigate for non-linear relationship MFP extended the approach (Sauerbrei and Royston 1999) and MFPI is a further extension to investigate for interactions (Royston and Sauerbrei 2004). 

In clinical and epidemiolocal studies, MFPI can be used in randomized trials to investigate for interactions of continuous variables with treatment and in epidemiological studies to improve multivariable models by checking for interactions of a continuous variable with a categorical variable. Obviously, there may also be an interaction of two continuous variables. An approach to investigate for such interactions is published in Royston and Sauerbrei (2008, MFPIgen, chap 7.11) but we are not aware of any further developments and this important topic is not included in this package. 

MFPI was developed in the context of a RCT to investigate for an interaction between a categorical (treatment) and a continuous covariate (Royston and Sauerbrei (2004). It aimed to replace the (still)  common approach to categorise the continuous variable using (often one) cutpoint(s) and check for an interaction in 2*2 or 2*k tables. Obviously, this discards information, reduces power, and can make the evidence for interaction depend strongly  on  the  number  and location of the cutpoints. For this reason, continuous variables should preferably be analysed  on  their original scale, with nonlinear relationships  evaluated where appropriate, as recommended in the PROGRESS framework for prognosis research (Hingorani et al 2013). The Instrument to assess the Credibility of Effect Modification Analyses (ICEMAN) in randomized controlled trials and meta-analyses has nine core questions, one reads ‘If the effect modifier is a continuous variable, were arbitrary cut points avoided?’ This question is one of the four which are relevant for both methods (RCT and meta-analysis) (Schandelmaier et al , 2020). Assuming a linear interaction  avoids  arbitrary cutpoints but may still be restrictive: the main effect, the interaction function, or both may be nonlinear (Royston and Sauerbrei, 2013).

The Multivariable Fractional Polynomial Interaction (MFPI) procedure (Royston and Sauerbrei,  2004) addresses these issues by modelling covariate effects using fractional polynomial (FP) functions within levels of a grouping variable. It is a parametric approach based on fractional polynomials and includes the linear interaction model as its simplest special case. Four flexibility variants (flex1 to flex4) control how FP powers are selected and  constrained across groups  (Royston and Sauerbrei,  2013, 2014).  If needed, for example often in observational studies, the procedure first constructs an adjustment model using the MFP algorithm, then tests  whether  group- specific functions (best FP1 function is the default, see xy) are parallel on the scale of the continuous  predictor.

This vignette describes `mfpi()` in the `mfp2` package. The function supports Gaussian, binomial, Poisson, and Cox regression models, and provides formula, matrix, and data-frame interfaces. It investigates interactions between a categorical grouping variable (`group_var`) and one or more continuous covariates (`cont_vars`). The grouping variable is remapped internally to consecutive integer levels for fitting, but its original labels are preserved for printing, summaries, plotting, prediction metadata, and the group-level FP power tables.

# Issues in the context of MFPI modelling
Examples published in the literature concentrate on the investigation of interactions in an RCT. There may be one specific predefined variable of interest or there may be a dozen of variables which may be checked for an interaction. Obviously, the analyst needs to decide about the significance level, interpretation of results, whether adjustment for multiple testing is needed and how to proceed if more than one variable is significant. The program helps by providing the relevant information (eg p values with and without adjustment for multiple testing (adjust-MT) and you can easily investigate for an interaction in a given (prespecified or based on earlier results) or selected adjustment model (adjust-model). 

It is well-known that single observations can have a strong influence on the selection of the best fitting function (Sauerbrei et al 2022, Infpoints paper). Already in the original MFPI paper, Royston and Sauerbrei (2004) stressed the importance to check robustness of the estimated functions by using Kaplan–Meier plots (NOT INTEGRATED IN THIS PACKAGE) and estimated treatment effects in subgroups defined by the covariate which (seems to) interact with treatment. Being new to this approach, we strongly recommend reading sections 2.4 to 2.6 in the original paper. 
We also warn against using MFPI when the (effective) sample size is small. We are not aware of any investigation about the sample size needed, but we assume that the effective sample size in each group should be (much) larger than 50. 

In the context of observational studies model adjustment plays a key role and there are many options to build a multivariable model including interactions. The main aim of the model should guide the strategy but there is no guidance on the ‘best’ procedure. The analyst may decide to derive an MFP model and check for all potential interactions of continuous by categorical variable. Alternatively, depending on the aim it may be preferable to  concentrate on a pair of variables of main interest, with or without adjustment for other variables. Anyhow, the package supports many options, but the strategy and interpretation of results is the task to be done by the analyst. Using some examples, we will illustrate how to work with this very flexible approach.

# Quick start

We begin with a worked example before describing the methodology in detail. The
`advanced_prostate_cancer` dataset in `mfp2` contains 475 observations, of
which 338 are observed deaths. It includes survival time (`survtime`),
event/censoring status (`cens`), a treatment variable (`rx`), eight
continuous covariates (`age`, `wt`, `sbp`, `dbp`, `sz`, `ap`,  `hg`, `sg`), and five categorical covariates (`pf`, `hx`, `bm`, `stage`, `ekg`). In this vignette, the original 0/1 treatment codes are recoded as the factor levels `Placebo` and `Treatment` before model fitting. This dataset was also used in Royston and 
Sauerbrei (2008) where more details are given.

```{r data}
data("advanced_prostate_cancer", package = "mfp2")

# Use descriptive treatment labels. The original
# dataset stores rx as 0/1. Converting it to a factor makes printed tables,
# summaries, and plots easier to read.
advanced_prostate_cancer$rx <- factor(
  advanced_prostate_cancer$rx,
  levels = c(0, 1),
  labels = c("Placebo", "Treatment")
)

head(advanced_prostate_cancer)
```

The following call tests whether the treatment effect on survival varies across
values of `age`, while adjusting for the other selected predictors. The
functional form for `age` is pre-specified as FP2 using `cont_var_forms`.
Here, `criterion = "pvalue"` uses p-values to select the adjustment model and
the functional forms of its continuous predictors, with `select = alpha = 0.05`.
The p-value threshold for the interaction test is set separately using
`p_interact = 0.05`. The arguments `select` and `alpha` apply only to the
adjustment-model selection step and are described in detail in the documentation
for `mfp2()`.

```{r quickstart}
fit_quick <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var         = "rx",
  cont_vars         = "age",
  cont_var_forms    = c(age = "fp2"),
  flex              = "flex1",
  family            = "cox",
  p_interact        = 0.05,
  select            = 0.05,
  alpha             = 0.05,
  criterion         = "pvalue",
  data              = advanced_prostate_cancer,
  verbose           = FALSE
)
print(fit_quick)
```

The output is organised into sequential steps. Step 1 presents the adjustment
model selected by MFP; in this example, six variables are retained as linear
adjustment terms for the interaction analysis involving age. Step 2 presents the
interaction model metrics. Directly below Step 2, an unnumbered detail table
shows the interaction-side FP powers by group level when such powers are
available. Step 3 provides the final interaction summary. For
`criterion = "pvalue"`, this summary includes both raw and adjusted p-values and
reports the p-value adjustment method.

The `summary()` method additionally prints a retained-only interaction-power
detail table, followed by the regression coefficients for retained interaction
models. The `plot()` method computes plotting data on demand through
`predict.mfpi()` and displays the group-specific fitted functions and the
fitted-function difference. The remainder of this vignette explains each
component in detail.

```{r quickstart-plot, fig.width = 10, fig.cap = "Fitted functions and fitted-function difference for the age-by-treatment interaction."}
plot(fit_quick, plot_type = "both")
```

## Grouping variables, labels, and reference levels

The grouping variable is specified by `group_var`, which must be a character
scalar naming a column. In the formula interface, the grouping variable may be a
factor, character, logical, integer, or numeric variable in the supplied data. In
the default interface, `x` may be a numeric matrix or a data frame. If `x` is a
data frame, only `group_var` may be categorical; all other variables must be
numeric, integer, or logical. This keeps the default interface explicit and
avoids silently dummy-coding adjustment variables.

Internally, `mfpi()` remaps the grouping variable to consecutive integer levels
`0, 1, ..., K - 1`. These internal levels are used in model matrices, coefficient
names, prediction-column names, and group-specific FP basis construction. The
original labels are stored separately and are used for printed output, summaries,
plots, and prediction metadata. For example, in this vignette `rx` is recoded as a factor with levels
`c("Placebo", "Treatment")`. The internal levels are `0` and `1`, but the
printed group-level tables and plot legends use `Placebo` and `Treatment`.

The reference level is determined as follows. For a factor `group_var`, the
reference level is the first element of `levels(droplevels(group_var))`, so users
can control the reference by setting factor levels before fitting. For a
character `group_var`, the reference is the first distinct value encountered in
the data. For logical, integer, numeric, and numeric-matrix input, the reference
is the smallest observed value. The mapping is stored in the fitted object as
`group_level_map` when available.

If `include_group_var = TRUE`, the grouping variable is also included in the
adjustment model. In that case, `mfpi()` creates `K - 1` group dummy variables
and forces them into the adjustment model. This can be useful when the group
main effect must be adjusted for in Step 1, but it increases the number of
forced adjustment parameters when `group_var` has many levels. Sparse group
levels can therefore make the adjustment model less stable.

# Prespecified and exploratory interaction analyses

Royston and Sauerbrei (2008) distinguished between two settings for
investigating interactions:

1. **Prespecified analysis.** This evaluates an interaction defined before
   examining the data, usually based on prior evidence or the trial protocol.
   Because the analysis focuses on a single interaction, multiplicity adjustment
   across several covariates is not required.

2. **Hypothesis generation.** This screens several covariates for possible
   interactions with the grouping variable. Because multiple interaction tests
   are performed, some small p-values may occur by chance. The interaction
   p-values may therefore need to be adjusted for multiplicity. Any interaction
   identified in such an exploratory analysis should be regarded as a candidate
   finding that requires confirmation in independent data (Royston and Sauerbrei,
   2008).

The `mfpi()` function supports both settings. For a prespecified analysis,
`cont_vars` contains a single continuous covariate. For an exploratory analysis,
`cont_vars` contains several continuous covariates, which are assessed one at a
time. For each variable under test, the adjustment covariates are the remaining
variables selected in Step 1. Specifically, the continuous variable currently
being tested and the dummy variables corresponding to the grouping variable are
excluded from the adjustment set.

When `criterion = "pvalue"`, multiplicity adjustment is controlled by the
`p_adjust_method` argument, which specifies the adjustment method, such as
`"holm"`, `"bonferroni"`, `"hochberg"`, `"BH"`, or `"BY"`. The default is
`p_adjust_method = "none"`, meaning that no multiplicity correction is applied.
When multiple variables are screened, adjustment is applied across all tested
variables, with one interaction test per variable. The adjusted p-values are
shown in Step 3 and used for the final selection decision.

## Pre-specifying the functional form: `cont_var_forms`

In both settings, the functional form of the interaction --- linear, FP1, or
FP2 --- should be pre-specified rather than selected from the data. Choosing
the functional form by comparing all three candidates on the same data inflates
the Type I error and introduces selection bias: more flexible forms tend to fit
the data better by chance, making spurious interactions more likely to be
detected.

The `cont_var_forms` argument specifies the functional form to use for each
variable in `cont_vars`. It accepts a named character vector whose names are
variable names from `cont_vars` and whose values are one of `"linear"`,
`"fp1"`, or `"fp2"`. Variables not named in `cont_var_forms` are assigned
`"linear"` by default. The argument is `NULL` by default, which assigns
`"linear"` to all variables.

For example, to test `age` as FP2 and leave `wt` as linear (the default):

```{r cont-var-forms-example, eval=FALSE}
mfpi(
  ...,
  cont_vars      = c("age", "wt"),
  cont_var_forms = c(age = "fp2")
)
```

Each variable in `cont_vars` is tested using exactly its pre-specified form.
The functional form should be chosen based on subject-matter knowledge or prior
evidence, ideally before looking at the data. For guidance on choosing a
functional form, see the [Flexibility levels](#Flex) section and the
recommendations of Royston and Sauerbrei (2014).

Unnamed entries and unnamed scalars (e.g. `cont_var_forms = "fp2"`) are not
accepted; every entry must carry the name of the target variable to ensure the
specification is unambiguous.

# The MFPI algorithm

The MFPI procedure operates in two broad stages.

First, an adjustment model is selected using the MFP algorithm. The algorithm is
run once on the full predictor matrix to select adjustment variables and their
functional forms. This initial selection also includes variables that may later
be tested for interaction. By default, the grouping variable is excluded from
the MFP step (`include_group_var = FALSE`). If `include_group_var = TRUE`, the
grouping variable is represented by dummy variables and forced into the
adjustment model via `select = 1` and the `keep` set, so that it is never
eliminated during backfitting regardless of the selection criterion. The
selected adjustment model is then used to control for confounding in the
subsequent interaction analyses.

Second, for each continuous variable listed in `cont_vars`, `mfpi()` evaluates
whether the relationship between that covariate and the outcome differs across
the levels of `group_var`. For each covariate under investigation, denoted by
$z$, the adjustment set contains the variables selected by the MFP adjustment
model, excluding $z$ itself and the dummy variables representing the grouping
variable. The selected adjustment variables remain in their selected functional
forms. Thus, different variables in `cont_vars` may be evaluated with slightly
different adjustment sets. When `verbose = TRUE`, the adjustment set used for
each variable is printed so that the user can verify which covariates are being
controlled for.

For each variable in `cont_vars`, the interaction model is fitted using the
functional form pre-specified in `cont_var_forms` (default: `"linear"`). The
selection criterion, specified by `criterion`, determines how the interaction
model is compared with the corresponding main-effects model. The available
criteria are `"pvalue"`, `"aic"`, and `"bic"`.

For `criterion = "pvalue"`, each variable yields one p-value from its
pre-specified interaction test. When `p_adjust_method != "none"`, all p-values
are collected after the per-variable loop and adjusted together before the
final selection decisions are made. During verbose model fitting, the p-value
reported within each variable is therefore provisional when adjustment is
requested. After all variables have been evaluated and p-values adjusted, the
final interaction decisions are determined and printed in Step 3.
When `p_adjust_method = "none"`, the p-value printed during fitting is the
final result and is shown immediately as `>> p_raw = ... -> Selected/Not selected`.

## The interaction model

For a binary grouping variable, let \(t = 0\) denote the reference group and
\(t = 1\) denote the comparison group. Let \(z\) be the continuous covariate
under investigation. Prior to FP transformation, \(z\) is shifted when it
contains negative or zero values to ensure positivity, and optionally scaled for
numerical stability.

If an FP2 function with powers \(p_1\) and \(p_2\) is selected, the interaction
model for the linear predictor is

\[
\eta_i =
\mathbf{x}_i^\top \boldsymbol{\gamma}
+ \alpha t_i
+ \sum_{j = 0}^{1} I(t_i = j)
\left[
\beta_{j1}\phi_1(z_i)
+
\beta_{j2}\phi_2(z_i)
\right].
\]

Here, \(\mathbf{x}_i^\top \boldsymbol{\gamma}\) denotes the linear predictor
contribution from the adjustment covariates, \(\alpha\) is the main effect of
the grouping variable, \(I(t_i = j)\) is the indicator for group \(j\), and
\(\phi_1(z_i)\) and \(\phi_2(z_i)\) are the selected FP basis functions. For
distinct FP powers,

\[
\phi_1(z_i) = z_i^{p_1},
\qquad
\phi_2(z_i) = z_i^{p_2},
\]

with \(z^0 = \log(z)\) by convention. For repeated FP2 powers,
\(p_1 = p_2 = p\), the basis functions are

\[
\phi_1(z_i) = z_i^p,
\qquad
\phi_2(z_i) = z_i^p \log(z_i).
\]

The group-specific contribution of \(z\) for an observation in group \(j\) is

\[
\hat{f}_j(z_i)
=
\hat{\beta}_{j1}\phi_1(z_i)
+
\hat{\beta}_{j2}\phi_2(z_i),
\qquad
j = 0, 1.
\]

The interaction test assesses whether the group-specific functions are parallel
on the scale of the linear predictor. For Gaussian models, this is the outcome
scale. For logistic regression, it is the log-odds scale; for Poisson
regression, it is the log-rate or log-mean scale; and for Cox models, it is the
log-hazard scale.

A useful way to display the interaction is the fitted-function difference,

\[
\hat{d}(z_i)
=
\hat{f}_1(z_i) - \hat{f}_0(z_i),
\]

together with a pointwise 95% confidence interval. If the two functions are
parallel, the difference is approximately constant over \(z\). If the difference
varies with \(z\), this suggests that the effect of the grouping variable on the
outcome depends on the value of the continuous covariate.

Pointwise standard errors for \(\hat{d}(z_i)\) and for the group-specific
functions \(\hat{f}_j(z_i)\) are estimated using the delta method applied to the
variance-covariance matrix of the fitted interaction model.

The fitted-function difference plot should be interpreted as a model-based
visual summary. The confidence interval is pointwise and depends on the selected
model.

# Flexibility levels {#Flex}

The argument `flex` controls how FP powers are selected and constrained across
the levels of the grouping variable. Four flexibility levels are available:
`"flex1"`, `"flex2"`, `"flex3"`, and `"flex4"`. These variants are available
for both FP1 and FP2 functions.

In general, `flex1` is the least flexible and `flex4` is the most flexible
option. Increasing flexibility allows more complex group-specific functional
forms, but it also increases the risk of instability, especially in small
samples or when the continuous covariate has a skewed distribution, influential
observations, or sparse data in parts of its range. The appropriate flexibility
level should therefore be chosen using subject-matter knowledge, the expected
shape of the interaction, and sample-size considerations. For further details,
see Royston and Sauerbrei (2004, 2013, 2014).

## `flex1`

`flex1` is the default and least flexible option. FP powers for the continuous
covariate are estimated once from the pooled main-effects model. This model
includes the FP terms for the continuous covariate and the group indicators, but
no group-by-covariate interaction, while adjusting for the selected covariates
in the adjustment model. The FP powers selected from this pooled main-effects
model are then held fixed and used at each level of the grouping variable. Thus,
the same FP basis functions are used in all groups, and only the regression
coefficients of these basis functions are allowed to differ between groups.

Because the main-effects and interaction models use the same FP basis functions,
the main-effects model is nested within the interaction model. The
interaction test is therefore a standard likelihood-ratio test comparing the two
models, with $(K - 1) \times m$ degrees of freedom, where $K$ is the number of
groups and $m$ is the number of FP terms. Thus, $m = 1$ for FP1 and $m = 2$ for
FP2. For the common case of two groups, this gives 1 df for FP1 and 2 df for
FP2.

`flex1` corresponds to the original MFPI procedure proposed by Royston and
Sauerbrei (2004). In the significance-level simulation study, Royston and
Sauerbrei (2013) found that FP2 with `flex1` maintained significance levels
close to nominal, whereas the other FP2 flexibility variants showed larger
departures from the nominal level. When subject-matter knowledge gives good
reasons to expect a non-monotone treatment-effect function, Royston and
Sauerbrei (2014) recommended FP2 with `flex1`.

## `flex2`

`flex2` estimates FP powers from an interaction model rather than from the
pooled main-effects model used by `flex1`. The powers are optimized over the
levels of the grouping variable, but are constrained to be the same across
groups. These same powers are then also used in the corresponding main-effects
model.

Thus, `flex2` may select powers different from those selected by `flex1`,
because power selection is informed by group-specific effects. However, once the
powers have been selected, the same FP powers are used in both the
main-effects and interaction models. The main-effects model is therefore nested
within the interaction model. The interaction test uses the same nominal degrees
of freedom as `flex1`, namely $(K - 1) \times m$.

In the simulation study by Royston and Sauerbrei (2013, 2014), FP1 with `flex2`
had high power but slightly inflated significance levels. FP2 with `flex2` 
showed clearly inflated type I error and was therefore not recommended as a 
routine choice.

## `flex3`

`flex3` estimates FP powers separately for the main-effects and interaction
models. The main-effects powers are selected from the pooled main-effects model,
as in `flex1`, whereas the interaction powers are selected from a model allowing
group-by-covariate interaction, as in `flex2`. These interaction powers are
constrained to be the same across groups, so group-specific functions differ
through their regression coefficients rather than through group-specific FP
powers.

Because the main-effects and interaction models may use different FP power sets,
the two fitted models may be non-nested. If the selected powers differ, the
main-effects and interaction models are based on different transformations of
the continuous covariate. In that case, the main-effects model is not nested
within the interaction model, and cannot be obtained simply by setting the interaction regression coefficients to zero. If the selected powers are the same, the usual
nested-model interpretation applies. In all cases, the likelihood-ratio
statistic is computed and compared with a chi-square distribution using the
nominal degrees of freedom, $(K - 1) \times m$.

In the significance-level simulation study of Royston and Sauerbrei (2013), FP1
with `flex3` had significance levels close to nominal. In the power study, FP1
with `flex3` performed well. Royston and Sauerbrei (2014) considered FP1 with
`flex1`--`flex3` to have the largest power among methods with acceptable
significance levels and recommended FP1 with `flex3` as the preferred practical
choice when there is no strong subject-matter reason to expect a non-monotone
treatment-effect function. They noted that the FP1 class includes the linear
function and tends to yield functions that are simple, understandable, and transferable. 

## `flex4`

`flex4` is the most flexible option. Like `flex3`, it allows the main-effects
and interaction models to use separately selected FP powers. Its distinguishing
feature is that the FP powers in the interaction model are also allowed to differ
across the levels of the grouping variable. Thus, each group may have its own FP
power set and its own regression coefficients.

Because `flex4` allows separate power selection for the main-effects and
interaction models, and may also allow group-specific powers in the interaction
model, the comparison may be non-nested. When the FP basis functions used in the
main-effects and interaction models differ, the main-effects model is not a
simple restriction of the interaction model. If the same basis functions are estimated
in the fitted main-effects and interaction models, the comparison is the usual 
nested-model setting. The interaction test uses $2(K - 1) \times m$ degrees of 
freedom, which is twice as many as `flex1`--`flex3` to account for the additional
flexibility from allowing group-specific powers.

In the simulation studies of Royston and Sauerbrei (2013, 2014), FP2 with
`flex4` was conservative, with empirical significance levels around 2--3%
rather than the nominal 5%, but had low power.

`flex4` may be useful when there is a strong subject-matter reason to believe
that the functional form of the covariate effect differs between groups.
However, this may require sufficiently large samples to detect interaction
effects.

## Summary table

The four MFPI variants are summarized in the table below.

```{r flex-table, echo=FALSE}
flex_table <- data.frame(
  Variant = c("flex1", "flex2", "flex3", "flex4"),
  Description = c(
    "FP powers determined for the main effect of z; use these same powers at each level of t",
    "FP optimized over levels of t; use these same powers at each level of t",
    "As flex2, but re-estimate powers for the main effect of z",
    "Optimize FP for the main effect and separately at each level of t"
  ),
  `d.f. FP1` = c(1, 1, 1, 2),
  `d.f. FP2` = c(2, 2, 2, 4),
  check.names = FALSE
)

knitr::kable(
  flex_table,
  align = c("l", "l", "c", "c"),
  caption = "Variants of MFPI with degrees of freedom for the tests of
  interaction from Royston and Sauerbrei (2013)."
)
```

The choice of functional form --- linear, FP1, or FP2 --- is made via
`cont_var_forms` and should be based on subject-matter knowledge before the
data are examined. Each variable in `cont_vars` is then tested using exactly
its pre-specified form. For example, a user may choose FP2 with `flex3` when
there is subject-matter reason to allow a more flexible nonlinear interaction
function while allowing the main-effects and interaction models to select
separate FP powers:

```{r cont-var-forms-flex3, eval=FALSE}
mfpi(
  ...,
  cont_vars      = "age",
  cont_var_forms = c(age = "fp2"),
  flex           = "flex3"
)
```

## Note on nested and non-nested comparisons {#note-on-non-nested-comparisons}

Whether the main-effects and interaction models are nested depends on the FP
basis functions used in the two models. If the same FP powers, and therefore the
same transformed covariate columns, are used in both models, the main-effects
model is nested within the interaction model.

If the FP powers differ between the main-effects and interaction models, the two
models no longer use the same covariate basis. In that case, the main-effects
model is not nested within the interaction model. The likelihood-ratio statistic
may still be computed from the fitted deviances and compared with a chi-square
distribution using the nominal degrees of freedom, but the resulting p-values
lack a firm theoretical underpinning. They should therefore be interpreted
cautiously, because the significance levels may be liberal or conservative
(Royston and Sauerbrei, 2013).

This issue is most relevant for flexibility variants that allow separate power
selection for the main-effects and interaction models, such as `flex3` and
`flex4`. However, the nestedness of a particular fitted comparison should be
judged from the selected FP basis functions, not only from the flexibility label.
For example, `flex3` and `flex4` fits may still give a nested comparison if the
selected main-effects and interaction powers are the same.

As an alternative, AIC or BIC may be used to compare non-nested models, since
these criteria do not require the main-effects model to be nested within the
interaction model. See the
[Alternative selection criteria: AIC and BIC](#aic-bic) section for details.

# Practical considerations

## Influential observations and Winsorisation

FP models can be sensitive to outliers or influential observations in the
continuous predictor. A small number of influential observations in the tails of
a variable in `cont_vars` may strongly affect the selected functional form and
may produce spuriously significant interactions. Therefore, it is important to
inspect the data before fitting an MFPI model, for example, using univariable
plots.

In `mfpi()`, Winsorisation is provided as an optional pre-processing step to
reduce the influence of extreme values in variables listed in `cont_vars`.
Values below the lower percentile and above the upper percentile specified by
`winsorize_probs` are replaced by the corresponding percentile cutoffs. The
default is `winsorize_probs = c(0.01, 0.99)`.

Note that variables in `cont_vars` cannot be listed in `spike_vars` or
`catzero_vars` --- these options are only available for adjustment variables,
because the interaction design matrix does not include group-specific
$I(x = 0)$ indicators. If a variable appears in both `cont_vars` and
`spike_vars` or `catzero_vars`, the spike or catzero treatment is dropped
with a warning.

Variables in `cont_vars` that are also listed in `zero_vars` are excluded from
Winsorisation, because the zero structure of these variables carries substantive
modelling meaning, and truncating the lower tail could distort it. More
generally, variables flagged as `catzero_vars` or `spike_vars` are also excluded
from Winsorisation, though these options cannot currently be used with variables
in `cont_vars`. Excluded variables are recorded with `NA` limits in
`winsorize_limits` for transparency.

Winsorisation is applied after shifting and scaling, so the percentile cutoffs
and the limits stored in `winsorize_limits` are on the transformed scale. The
actual cutoffs applied are returned in the fitted object and can be inspected
via `fit$winsorize_limits`.

Winsorisation is disabled by default (`winsorize = FALSE`). To apply
Winsorisation, set `winsorize = TRUE`. More aggressive truncation can be
requested, for example:

```{r winsorize-prob, eval=FALSE}
winsorize_probs = c(0.05, 0.95)
```

Any Winsorisation or trimming applied should be reported transparently as part
of the initial data analysis.

## Shifting and scaling

Fractional polynomials require strictly positive input values. As in `mfp2()`,
`mfpi()` estimates shift terms to ensure all values of the continuous covariates
are positive before FP transformation. After shifting, `mfpi()` estimates a
scale factor for each variable, placing the shifted covariates on a convenient
numerical range for FP power selection. This improves numerical stability
during the MFP backfitting cycles. Scaling does not affect which FP powers are
selected, only the numerical precision of the fitting algorithm.

After FP powers have been selected, the variables are backscaled before the
final model is fitted. This means the regression coefficients in the returned
model are on the $\phi(z + \text{shift})$ scale, matching what `mfp2()` would
produce. The user does not need to undo the scaling manually.

## Centering

When `center = TRUE`, centering is applied in both the adjustment model and the
interaction model, but the centering constants may differ between the two
models.

**Adjustment model.** Standard `mfp2()` centering is applied. Each
FP-transformed column is centered by subtracting its column mean, that is,
`f(x_j) - mean(f(x_j))`. For binary covariates, centering uses the lower of the
two distinct values rather than the mean, following the same convention as
`mfp2()`.

**Interaction model.** The argument `center_type` controls how the
group-specific FP columns are centered:

- `center_type = "grand"` (default): the grand mean of the FP-transformed
  variable, computed across all observations regardless of group membership, is
  subtracted from each group's columns. Thus, all groups share the same
  centering constant.
- `center_type = "group"`: each group's columns are centered using the mean
  computed from observations in that group only. Thus, different groups may use
  different centering constants.

In both cases, centering affects the intercept and, when an interaction is
present, the group dummy coefficients. It does not affect the interaction test,
the fitted-function difference $\hat{f}_1(z) - \hat{f}_0(z)$, or the fitted
values.

## Handling non-positive values

When a continuous predictor contains a mixture of non-positive and positive
values, standard shifting and FP transformation may not be appropriate. The
`mfpi()` function provides three options, matching those in `mfp2()`.

The `zero_vars` option specifies variables for which FP transformations are
applied only to strictly positive values. Non-positive values are set to zero in
the transformed variable, and no additional binary indicator is added.

The `catzero_vars` option extends this approach by adding a binary indicator for
whether the variable is zero. This indicator is included in the adjustment model
alongside the FP terms for the positive part. The `catzero_vars` option implies
`zero_vars`.

The `spike_vars` option triggers the spike-at-zero (SAZ) algorithm. This
evaluates whether the binary zero indicator and the FP function each contribute
explanatory value. The `spike_vars` option implies both `catzero_vars` and
`zero_vars`. For more details about SAZ, see the corresponding vignette in the
`mfp2` package.

Variables listed in `cont_vars` cannot currently use the `catzero_vars` or
`spike_vars` options, because the methodology for combining these options with
interaction testing is not yet available. If a variable appears in both
`cont_vars` and either `catzero_vars` or `spike_vars`, it will be automatically
removed from `catzero_vars` or `spike_vars`, with a warning. However,
`zero_vars` can be used with variables in `cont_vars`: only the positive part
undergoes FP transformation, while the remaining values are set to zero.
Variables listed in `catzero_vars` or `spike_vars` that are not in `cont_vars`
are handled in the adjustment model as described in the `mfp2()` documentation.

In the formula interface, ACD handling should be requested term by term through
`fp(..., acd = TRUE)` or `fp2(..., acd = TRUE)`. Argument-level `acd_vars` is not
part of the formula interface; this avoids ambiguity between global arguments
and the term-specific settings embedded in the formula.

## F-test for Gaussian models

When `family = "gaussian"` and the sample size is small, the chi-square
approximation to the likelihood-ratio test may be unreliable. Setting
`ftest = TRUE` replaces the chi-square test with an F-test for both the
adjustment-variable selection step and the interaction test. This option is
ignored for non-Gaussian families and when `criterion` is not `"pvalue"`.

## Forcing variables and functional forms

Two arguments provide additional control over the adjustment model. The `keep`
argument accepts a character vector of variable names that are forced into the
adjustment model regardless of the selection criteria. This is useful when some
variables must always be included as adjustment variables. If such variables
are required only as linear adjustment terms, they can be specified with
`df = 1` and their names passed to `keep`.

The `force_max_fp_vars` argument accepts a character vector of variable names
for which the algorithm should select the most complex FP functional form at
the degree specified by `df`. This behaviour is consistent across all selection
criteria. When `criterion = "pvalue"`, `force_max_fp_vars` sets `alpha = 1` for
the named variables, so that the significance test always accepts the most
complex form. When `criterion = "aic"` or `criterion = "bic"`, it bypasses
information-criterion-based simplification. In both cases, the best power
combination within the requested FP degree is still selected according to the
chosen criterion.

In the formula interface, the same per-variable control is available through
`fp(force_max_fp = TRUE)`.

# Worked examples

## Data: Advanced prostate cancer

All survival model examples use the `advanced_prostate_cancer` dataset included
in the `mfp2` package. The outcome is time to death (overall survival), defined
by `survtime` and the censoring indicator `cens`. The treatment-allocation
variable `rx` is the grouping variable. In this vignette it is converted from
the original 0/1 coding to a factor with levels `Placebo` and `Treatment`. For a full description of the
data, see Royston and Sauerbrei (2008). The dataset contains 475 observations
with eight continuous measurements (`age`, `wt`, `sbp`, `dbp`, `sz`, `ap`,
`hg`, `sg`) and five categorical covariates (`pf`, `hx`, `bm`, `stage`, `ekg`).

## Prespecified interaction analysis

Suppose that, based on prior evidence, we have identified `age` as the single
covariate to investigate for interaction with treatment, and that subject-matter
knowledge suggests allowing a nonlinear treatment-by-age function, motivating an
FP2 functional form.
Because only one interaction is tested, the resulting p-value can be interpreted
directly without multiplicity adjustment.

The formula interface is the most convenient way to call `mfpi()`. Continuous
variables can be wrapped in `fp()` to control FP-specific settings, such as the
degrees of freedom or the candidate power set. For example,
`fp(age, df = 2, select = 1)` restricts `age` to an FP1 function and forces it
to remain in the adjustment model. In the interaction analysis below, however,
we use the defaults for the adjustment step and pre-specify the interaction
function for `age` as FP2 via `cont_var_forms`:

```{r prespecified}
fit_pre <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var         = "rx",
  cont_vars         = "age",
  cont_var_forms    = c(age = "fp2"),
  flex              = "flex1",
  family            = "cox",
  select            = 0.05,
  alpha             = 0.05,
  data              = advanced_prostate_cancer,
  verbose           = FALSE
)
```

### Output: prespecified analysis

Calling `print()` on the fitted object displays a structured summary. With
`criterion = "pvalue"`, the printed model-building output has three numbered
steps.

```{r print-pre}
print(fit_pre)
```

**Step 1: Adjustment model (MFP).** This section shows the results of applying
the MFP algorithm to the full set of candidate adjustment predictors. Variables
retained after selection at `select = 0.05` are shown with their final FP powers
and degrees of freedom. Variables eliminated by MFP are listed under
`"Dropped"`. By default, the grouping variable `rx` is not included in the MFP
adjustment model; it enters only in Step 2 as the variable defining the
interaction. To include `rx` in the adjustment model as well, set
`include_group_var = TRUE`, which forces it into the model so that it cannot be
eliminated during backfitting. The retained covariates, in their selected
transformed or untransformed forms, are then used for adjustment in the
interaction models for `age`.

**Step 2: All interaction models.** For `age`, one interaction model is fitted
using the pre-specified form from `cont_var_forms`, here FP2. Each row shows the
functional form, the deviance of the interaction model, the likelihood-ratio
statistic or information-criterion improvement, the raw p-value when applicable,
and the main-effect FP powers. Directly below this table, the unnumbered
`FP powers by group level` table expands the interaction-side power list into
one row per group level. This keeps the main Step 2 table compact
while still showing which FP powers were used for each group-specific
interaction function.

**Step 3: Interaction summary.** This compact table shows one row per tested
variable. It gives the functional form, the raw p-value, and the adjusted
p-value used for the final decision. The p-value adjustment method is printed
below the table. Variables whose final p-value is below `p_interact` are marked
with `*`.

### Regression output {- .unlisted}

The `summary()` method appends an unnumbered retained-only interaction-power
detail table and then adds Step 4 with the full regression coefficients of
retained interaction models.

```{r summary-pre}
summary(fit_pre)
```

The retained interaction-power detail table printed before the regression
output is restricted to selected interaction models. It is unnumbered because it
explains the retained model specification rather than adding another
model-selection step.

**Regression output.** For each variable with a significant interaction,
the regression table from the fitted `coxph()` model is displayed. This table
includes coefficient estimates, standard errors, z-values, and p-values for all
terms in the model. The interaction model still uses internal numeric group codes in coefficient
names, even though printed tables and plots use the original labels. For the
labelled `rx` factor used in this vignette, `Placebo` is the reference group and
is mapped internally to group code 0, while `Treatment` is mapped to group code
1. The mapping can be inspected with `fit_pre$group_level_map`.

The group-specific FP coefficients use the naming convention
`<varname><internal_group_code><power_index>`. For example, if `age` is modelled
as FP1, `age01` represents the FP term for `Placebo` and `age11` represents the
corresponding FP term for `Treatment`. For FP2, two terms per group are shown,
for example `age01` and `age02` for `Placebo`, and `age11` and `age12` for
`Treatment`.

The group dummy coefficient, for example `rx1`, represents the main effect of
the internally coded comparison group. In this vignette that comparison group is
`Treatment`, relative to the reference group `Placebo`. The p-values in the regression output are the standard
Wald-test p-values reported by `summary.coxph()`. These p-values do not account
for the degrees of freedom used during FP power selection. Therefore, the
appropriate p-value for testing the interaction is the one reported in the
interaction summary.

### Visualizing the interaction

The `plot()` method displays fitted interactions for continuous variables in an
`mfpi` model. By default (`terms = NULL`), it plots the selected variables,
that is, variables whose interaction passed the selection threshold. Any
variable in `cont_vars` can also be plotted, regardless of selection status, by
passing its name to the `terms` argument.

The main argument is `plot_type`, which controls the display:

- `"fitted"` overlays the group-specific fitted functions.
- `"difference"` plots the pointwise fitted-function difference,
  $\hat f_1(x) - \hat f_0(x)$.
- `"both"` places the fitted-function and difference plots side by side; this
  requires the `patchwork` package.

Two additional arguments control the graphical output. `show_ci_fitted` adds
95% pointwise confidence bands to the group-specific fitted curves, whereas
`show_ci_diff` controls the confidence band for the fitted-function difference
curve. By default, fitted-function confidence bands are suppressed and the
difference-curve confidence band is shown. The argument `show_rug = TRUE` adds
rug marks showing the observed values of the continuous variable.

Each plot includes a compact subtitle summarising the functional form, the group
contrast, and the selection metric. For example:
`type = FP1; rx: Treatment vs Placebo; p_raw = 0.031 (p_adj = 0.047)` when
original group labels are available.
For AIC or BIC, the subtitle shows `dAIC` or `dBIC`, respectively.

The fitted-function plot distinguishes groups using the legend and line
aesthetics. Original group labels are used in the legend and plot annotations,
rather than the internal recoded levels. By default, the legend is placed inside
the fitted-function panel near the top-left. Its position can be changed using
`legend_position`, `legend_inside`, and `legend_justification`.

The `line_types` argument controls the linetypes for the group-specific fitted
curves. If `line_types = NULL`, solid lines are used by default. Users may pass
an unnamed vector matched to the group order, for example
`line_types = c("solid", "dashed")`, or a named vector using the original group
labels, for example
`line_types = c("Placebo" = "solid", "Treatment" = "dashed")`.

#### Fitted functions

The fitted-function plot overlays the estimated linear predictor for the two
treatment groups as a function of `age`. Confidence bands and rug marks are
added here to show uncertainty and the distribution of the observed age values.
If the two fitted curves are approximately parallel, this suggests little
evidence that the treatment effect varies with age.

```{r plot-fitted-pre, fig.cap = "Group-specific fitted functions for the age-by-treatment interaction."}
plot(fit_pre,
     plot_type = "fitted",
     show_ci_fitted = TRUE,
     show_rug = TRUE)
```

#### Fitted-function difference

The difference plot shows the fitted-function difference between the comparison
group and the reference group, using the original group labels. For example, the
y-axis may be labelled `f(rx = Treatment) - f(rx = Placebo)` when `rx` was
supplied with those factor labels. A horizontal reference line
at zero is added. The confidence band is shown by default. Positive values
indicate that the fitted function is higher in the comparison group than in the
reference group at that covariate value.

```{r plot-diff-pre, fig.cap = "Fitted-function difference for the age-by-treatment interaction."}
plot(fit_pre, plot_type = "difference")
```

#### Side-by-side display

The `"both"` option combines the fitted-function plot and the fitted-function
difference plot in a single display. This side-by-side view makes it easier to
see how differences between the group-specific fitted functions change across
the range of the continuous variable.

```{r plot-both-pre, fig.width = 10, fig.cap = "Fitted functions and fitted-function difference for the age-by-treatment interaction."}
plot(fit_pre, plot_type = "both")
```

## Using `flex3` with a prespecified FP2 function

In `flex3`, the main-effects model and the interaction model may use different
FP power sets. This makes `flex3` more flexible than `flex1`, while still
requiring the same interaction powers to be used across treatment groups.

We now refit the prespecified analysis for `age` using `flex3`, while keeping
`age` prespecified as FP2. This makes the example comparable with the earlier
`flex1` analysis in terms of the requested interaction functional form; only the
MFPI flexibility level changes.

```{r flex3-pre}
fit_flex3 <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var      = "rx",
  cont_vars      = "age",
  cont_var_forms = c(age = "fp2"),
  flex           = "flex3",
  family         = "cox",
  select         = 0.05,
  alpha          = 0.05,
  data           = advanced_prostate_cancer,
  verbose        = FALSE
)

print(fit_flex3)
```

Compared with the earlier `flex1` analysis, the adjustment model selected in
Step 1 is determined by the MFP selection step, not by the chosen MFPI
flexibility level. The Step 2 interaction model may differ because `flex3`
allows the main-effects and interaction models to use different FP power sets.
The interaction form remains FP2, so this example isolates the effect of
changing the flexibility variant from `flex1` to `flex3`.

```{r flex3-plot, fig.width = 10, fig.cap = "Fitted functions and fitted-function difference using flex3."}
plot(fit_flex3, plot_type = "both")
```

Note that `flex3` may involve a non-nested comparison, so the reported p-values
should generally be interpreted as approximate. In this example, however, the
main-effects and interaction models use the same FP powers and adjustment set,
so the main-effects model is nested within the interaction model.


## Exploratory interaction analysis

Suppose the analyst wants to screen several continuous covariates for possible
interactions with the grouping variable. Since multiple interaction tests are
performed, some small p-values may occur by chance. The interaction p-values may
therefore need to be adjusted for multiplicity.

In this example, seven continuous covariates are screened for possible
interactions with treatment. The model also includes `sg` as an adjustment
covariate, but `sg` is not included in `cont_vars` and is therefore not tested
for interaction. Assume that, based on subject-matter knowledge, `age` is
pre-specified as FP2, while all other variables are left as linear, which is the
default. Because multiple variables are screened, the Holm step-down correction
is applied via `p_adjust_method`. The adjustment is applied across all tested
variables, with one p-value per variable.

```{r exploratory}
fit_exp <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var         = "rx",
  cont_vars         = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg"),
  cont_var_forms    = c(age = "fp2"),
  flex              = "flex1",
  family            = "cox",
  select            = 0.05,
  alpha             = 0.05,
  p_adjust_method   = "holm",
  data              = advanced_prostate_cancer
)
```

### Output: exploratory analysis

```{r print-exp}
print(fit_exp)
```

**Step 1** is identical to the prespecified analysis because the same predictors
and the same MFP settings are used.

**Step 2** shows the fitted interaction model for each variable in `cont_vars`,
using the pre-specified form from `cont_var_forms`. Each row shows the functional
form, the interaction test statistics, and the raw p-value. The unnumbered
interaction-power detail table printed directly below Step 2 shows the
group-level interaction FP powers for each candidate model.

**Step 3: Interaction summary.** This compact table shows one row per tested
variable, giving the functional form, the raw p-value, and the adjusted
p-value used for the final decision. With `p_adjust_method = "holm"`, adjustment
is applied across all seven tested variables, and the adjustment method is
printed below the table. Variables whose adjusted p-value falls below
`p_interact = 0.05` are marked with `*`.

In an exploratory analysis, the adjusted p-value is the one used for the
selection decision. Variables marked with `*` are candidates for genuine
treatment-by-covariate interaction after accounting for multiplicity. Variables
without `*` do not provide sufficient evidence of interaction after adjustment.

### Regression output for selected variables

```{r summary-exp}
summary(fit_exp)
```

For p-value-based fits, `summary()` first repeats the model-building output,
then prints an unnumbered retained-only interaction-power detail table, and then
prints Step 4 with the regression output only for variables whose interaction was
selected, that is, variables marked with `*` in Step 3. As before, the correct
interaction test p-values are those from the interaction summary, not the Wald
p-values in the regression table.

### Plotting selected variables

By default, `plot()` shows results only for variables with a significant interaction.

```{r plot-exp-fitted, fig.width=10, fig.height=5, fig.cap="Fitted functions for variables with significant interactions."}
plot(fit_exp, plot_type = "both")
```

### Plotting non-selected variables

Any variable in `cont_vars` can be plotted regardless of whether its interaction
was selected, by passing its name via `terms`. The fitted curve uses the
functional form pre-specified in `cont_var_forms` for that variable. For
example, to inspect the fitted interaction for `wt` (which was not selected):

```{r plot-nonsig, fig.width=10, fig.height=5, fig.cap = "Fitted interaction for the wt-by-treatment interaction (not selected)."}
plot(fit_exp, terms = "wt", plot_type = "both", legend_inside = c(0.60, 0.98))
```

This is useful for diagnostic inspection: it shows what the fitted interaction
looked like and helps assess whether the non-selection decision is supported by
the data.

## Alternative selection criteria: AIC and BIC {#aic-bic}

In the examples above, the default `criterion = "pvalue"` was used. With this
criterion, for each variable in `cont_vars`, `mfpi()` fits the interaction model
using the pre-specified form from `cont_var_forms` and tests it against the
corresponding main-effects model. The resulting p-value is adjusted according to
`p_adjust_method`, and the interaction is retained if the final p-value is below
`p_interact`. For `flex3` and `flex4`, the main-effects and interaction models
may be non-nested, so the likelihood-ratio-test p-values lack a firm theoretical
underpinning; see the
[Note on non-nested comparisons](#note-on-non-nested-comparisons) section.

As an alternative, `mfpi()` supports model selection based on information
criteria via `criterion = "aic"` or `criterion = "bic"`. Under these criteria,
information criteria are used in both stages of the analysis: first to select
the MFP adjustment model, and then to compare the interaction model with the
main-effects model for each variable in `cont_vars`.

For AIC and BIC, the interaction improvement for each tested variable is

$$
\mathrm{dAIC} = \mathrm{AIC}_{\mathrm{main}} - \mathrm{AIC}_{\mathrm{interaction}},
\qquad
\mathrm{dBIC} = \mathrm{BIC}_{\mathrm{main}} - \mathrm{BIC}_{\mathrm{interaction}},
$$

where both models use the functional form pre-specified in `cont_var_forms`.
The interaction is retained if this improvement exceeds `min_improvement`. A
positive value means that the interaction model has a lower information
criterion than the main-effects model after penalising for model complexity.

The `criterion` argument therefore affects two stages of the analysis. In Step
1, it controls selection of the MFP adjustment model, including both variable
selection and functional-form selection. In the interaction analysis, it
determines how the interaction model is compared with the main-effects model and
whether the interaction is retained.

BIC usually penalises model complexity more strongly than AIC. AIC uses a fixed
penalty of 2, whereas BIC uses a penalty of $\log(n)$, where $n$ is the
effective sample size. Because $\log(n)$ is greater than 2 when $n > e^2$, BIC
often favours simpler models than AIC. In this context, BIC is therefore less
likely than AIC to retain interaction terms unless they provide a sufficiently
large improvement in model fit.

When `criterion = "aic"` or `criterion = "bic"`, the `p_adjust_method` argument
does not affect the selection decision, because multiplicity adjustment applies
to p-values, not to information criteria.

### Example: BIC-based selection

We repeat the exploratory analysis using BIC as the selection criterion. The
`min_improvement` threshold is set to 2, meaning that the BIC for the
interaction model must be at least 2 units lower than the BIC for the
main-effects model for the interaction to be retained.

```{r bic-exploratory}
fit_bic <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var       = "rx",
  cont_vars       = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg"),
  cont_var_forms  = c(age = "fp2"),
  flex            = "flex1",
  family          = "cox",
  criterion       = "bic",
  min_improvement = 2,
  data            = advanced_prostate_cancer
)
```

### Output: BIC-based selection

```{r print-bic}
print(fit_bic)
```

**Step 1** may differ from the p-value-based analyses because the MFP algorithm
now uses BIC for variable selection and FP degree choice. It may retain a
different set of adjustment variables or select different functional forms.

**Step 2** shows one row per tested variable with the interaction model metrics:
the main-effects BIC, the interaction BIC, and the dBIC improvement. Each
variable is tested using the form specified in `cont_var_forms`. The unnumbered
interaction-power detail table printed directly below Step 2 shows the
group-level interaction FP powers for each candidate model.

**Step 3: Interaction summary.** This table shows one row per tested variable
with the functional form, the dBIC improvement, and the selection decision.
Variables whose dBIC exceeds `min_improvement = 2` are marked with `*`. 

### Regression output and plots

```{r summary-bic}
summary(fit_bic)
```

The regression output shows fitted models for variables selected under the BIC
criterion. Plot subtitles show `dBIC` for selected variables.

```{r plot-bic-fitted, fig.cap = "Fitted functions for variables selected by BIC."}
plot(fit_bic, plot_type = "both")
```

## Logistic regression example

`mfpi()` is not limited to survival models. The following example demonstrates
its use with logistic regression (`family = "binomial"`). We create a binary
outcome from the prostate cancer data by defining death within 36 months as
the event:

```{r logistic-data}
prostate <- advanced_prostate_cancer
prostate$death36 <- as.integer(prostate$survtime <= 36 & prostate$cens == 1)
```

We test whether the effect of treatment on 36-month mortality depends on `age`,
using `flex1` and pre-specifying `age` as FP2:

```{r logistic-fit}
fit_logistic <- mfpi(
  death36 ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var         = "rx",
  cont_vars         = "age",
  cont_var_forms    = c(age = "fp2"),
  data              = prostate,
  flex              = "flex1",
  family            = "binomial",
  select            = 0.05,
  alpha             = 0.05,
  verbose           = FALSE
)
print(fit_logistic)
```

The output follows the same structure as the Cox model. With
`criterion = "pvalue"`, the interaction summary includes both raw and adjusted
p-values and reports the adjustment method. The interaction test and
fitted-function difference are now on the log-odds scale.

```{r logistic-plot, fig.width = 10, fig.cap = "Logistic regression: fitted functions and fitted-function difference for age by treatment."}
plot(fit_logistic, plot_type = "both")
```

# Accessing model components

The object returned by `mfpi()` is a list of class `"mfpi"`. Its components
can be accessed with `$` for programmatic post-processing.

```{r components}
names(fit_pre)
```

Key components include:

- `adjustment_model`: the full `mfp2` object from the MFP adjustment step.
- `all_model_metrics`: a data frame of evaluation metrics for the interaction
  model fitted for each variable in `cont_vars`, using the form specified in
  `cont_var_forms`. For p-value-based analyses it includes raw p-values and,
  when `p_adjust_method != "none"`, adjusted p-values. The FP power columns are
  stored as list-columns so that group-level interaction powers can be printed
  and reused programmatically.
- `best_model_metrics`: one row per selected variable, containing the functional
  form and the metric used for the final interaction decision, such as
  `p_adjusted`, `dAIC`, or `dBIC`.
- `best_interaction_model`: a named list of fitted model objects for selected
  variables.
- `all_interaction_models`: a named list containing the fitted interaction model
  for every variable in `cont_vars`, regardless of selection.
- Plotting data are computed on demand by `predict.mfpi()` when `plot()` is
  called; fitted-function matrices are not precomputed or required in the fitted
  object.
- `winsorize_limits`: the actual Winsorisation cutoffs applied.
- `group_levels_original` and `group_levels_new`: the user-facing labels of
  `group_var` and the internal consecutive group codes used for fitting.
- `group_level_map`: a data frame linking user-facing group labels, input
  levels, and internal levels. In this vignette it maps `Placebo` and
  `Treatment` to the internal codes used in coefficient and prediction-column
  names.
- `cont_var_forms`: the named character vector of pre-specified functional forms,
  one entry per variable in `cont_vars`.
- `var_winners`: a named list of per-variable fitted results, including those
  that did not reach significance.

For example, to extract the metrics table for all tested variables:

```{r all-metrics}
fit_pre$all_model_metrics
```

To extract the fitted model object for a selected variable:

```{r extract-model, eval=FALSE}
fit_pre$best_interaction_model[["age"]]
```


```{r restore-options, include=FALSE}
options(width = oldwidth)
```

# References

- Royston, P. and Sauerbrei, W. (2004). A new approach to modelling
  interactions between treatment and continuous covariates in clinical trials by
  using fractional polynomial submodels. *Statistics in Medicine*, 23(16),
  2509--2525. doi:
  [10.1002/sim.1815](https://doi.org/10.1002/sim.1815)

- Sauerbrei, W., Royston, P. and Zapien, K. (2007). Detecting an interaction
  between treatment and a continuous covariate: a comparison of two approaches.
  *Computational Statistics & Data Analysis*, 51(8), 4054--4063. doi:
  [10.1016/j.csda.2006.12.041](https://doi.org/10.1016/j.csda.2006.12.041)

- Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A
  Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for
  Modelling Continuous Variables.* Wiley. doi:
  [10.1002/9780470770771](https://doi.org/10.1002/9780470770771)

- Royston, P. and Sauerbrei, W. (2013). Interaction of treatment with a
  continuous variable: simulation study of significance level for several methods
  of analysis. *Statistics in Medicine*, 32(22), 3788--3803. doi:
  [10.1002/sim.5813](https://doi.org/10.1002/sim.5813)

- Royston, P. and Sauerbrei, W. (2014). Interaction of treatment with a
  continuous variable: simulation study of power for several methods of analysis.
  *Statistics in Medicine*, 33(27), 4695--4708. doi:
  [10.1002/sim.6308]( https://doi.org/10.1002/sim.6308)

```{r cleanup, include=FALSE}
options(width = oldwidth)
```
