---
title: "MFPI: Multivariable Fractional Polynomial Interaction"
author: "Edwin Kipruto and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
vignette: >
  %\VignetteIndexEntry{MFPI: Multivariable Fractional Polynomial Interaction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center"
)

oldwidth <- getOption("width")
options(width = 100)

library(mfp2)
```

# Introduction

Investigating whether the effect of a treatment or other grouping variable
changes across the values of a continuous covariate is a common and important
problem in clinical and epidemiological studies. In randomised clinical trials,
the question is often whether the treatment effect depends on the value of a
prognostic factor. Such dependence is represented statistically as an interaction
between treatment and the covariate; when present, it may help identify patients
who benefit more or less from treatment (Royston and Sauerbrei, 2004).

When both variables are binary or categorical, interaction modelling is
relatively straightforward. The problem is more difficult when one variable is
continuous. A common approach is to categorise the continuous variable using one
or more cutpoints and then fit a model containing the resulting categories, the
grouping variable, and their interaction terms. However, this strategy has
important limitations: it discards information, may reduce power, and can make
the evidence for interaction depend strongly on the number and location of the
cutpoints. Depending on the chosen cutpoints, categorisation may even
erroneously indicate an interaction with treatment, illustrating the danger of
cutpoint-based methods (Royston and Sauerbrei, 2013).

An alternative is to keep the covariate continuous but assume that its effect,
and its interaction with the grouping variable, is linear. This avoids arbitrary
cutpoints, but it may still be restrictive. The main effect of the continuous
covariate, the interaction function, or both may be nonlinear. Royston and
Sauerbrei (2013) note that misspecifying a nonlinear main effect can generate a
spurious interaction. For this reason, continuous variables should preferably be
analysed on their continuous scale, with nonlinear relationships evaluated where
appropriate.

The Multivariable Fractional Polynomial Interaction (MFPI) procedure was
proposed by Royston and Sauerbrei (2004) to address these issues. In the
original procedure, corresponding to what is now called `flex1`, the FP powers
are selected from the main-effects model and then held fixed when fitting the
interaction model. The interaction test assesses whether the resulting
group-specific functions are parallel on the scale of the linear predictor. If
the functions are parallel, the difference between groups is constant across
covariate values, indicating no interaction. If the difference varies with the
covariate, this suggests that the effect of the grouping variable depends on the
covariate. Royston and Sauerbrei (2013) later proposed additional MFPI variants,
now implemented as `flex2`, `flex3`, and `flex4`. These variants relax different
constraints on how FP powers are selected and constrained across groups, allowing
progressively more flexible interaction functions. The ordinary linear
interaction model remains the simplest special case, while nonlinear FP
functions can be selected when supported by the data.

MFPI was developed primarily for treatment-by-covariate interactions in
randomised trials, but the same approach can also be used to investigate
interactions between a categorical grouping variable and a continuous covariate
in other regression settings. In observational studies, adjustment for
confounders and other prognostic variables is usually important. The MFPI
procedure first constructs an adjustment model using the MFP algorithm, which can
select both variables and FP transformations for continuous covariates. These
selected variables and functional forms are then kept fixed while the interaction
between the grouping variable and each covariate of interest is investigated in
turn. Therefore, if the MFP algorithm selects a non-linear function for an
adjustment variable, that variable enters the interaction model as a transformed
FP function rather than in its original linear form (Royston and Sauerbrei,
2004, 2013).

This vignette describes the implementation of MFPI in the `mfp2` package for R.
The `mfpi()` function investigates interactions between a categorical grouping
variable, specified by `group_var`, and one or more continuous covariates,
specified by `cont_vars`. The grouping variable is treated as a factor
internally, with the lowest value used as the reference group.

The function supports Gaussian, binomial, Poisson, and Cox regression models,
and provides both formula and matrix interfaces. It implements the four MFPI
flexibility levels, `flex1` to `flex4`, which differ in how FP powers are
selected and constrained across groups. These levels are described in detail in
the [Flexibility levels](#Flex) section.

# Prespecified and exploratory interaction analyses

Royston and Sauerbrei (2008) distinguished between two settings for investigating
interactions:

1. **Prespecified analysis.** This evaluates an interaction defined before
   looking at the data, usually based on prior evidence or the trial protocol. 
   Because the analysis focuses on a single interaction, multiplicity adjustment
   across several covariates is not required.

2. **Hypothesis generation.** This screens several covariates for possible
   interaction with the grouping variable. Because multiple interaction tests
   are performed, some small p-values may occur by chance. The interaction
   p-values may therefore need to be adjusted for multiplicity. Any interaction
   identified in such an exploratory analysis should be regarded as a candidate
   finding that requires confirmation in independent data.
   

The `mfpi()` function supports both settings. For a prespecified analysis,
`cont_vars` contains a single continuous covariate. For an exploratory analysis,
`cont_vars` lists several continuous covariates, each assessed in turn. For each
variable under test, the adjustment covariates consist of the remaining selected
variables, i.e., the variable being tested is excluded from the adjustment set.

When several covariates are screened and `criterion = "pvalue"`, interaction
p-values can be adjusted across `cont_vars` using the `p_adjust_method`
argument. For example, `p_adjust_method = "holm"` applies Holm's step-down
correction. The default is `p_adjust_method = "none"`, meaning that no
multiplicity adjustment is applied. This correction applies across the variables
screened in `cont_vars`, not across the internal candidate models fitted by MFPI,
such as linear, FP1, and FP2. Those internal comparisons are part of the MFPI
selection procedure.

# The MFPI approach

The MFPI procedure operates in two broad stages.

First, an adjustment model is selected using the MFP algorithm. The algorithm is
run once on the full predictor matrix to select adjustment variables and their
functional forms. This step may include variables that will later be tested for
interaction. The selected adjustment model is then used to control for
confounding in the interaction analyses.

Second, for each continuous variable listed in `cont_vars`, `mfpi()` evaluates
whether the relationship between that covariate and the outcome differs across
the levels of `group_var`. For each covariate under investigation, denoted by
$z$, the adjustment set contains the variables selected by the MFP adjustment
model, excluding $z$ itself and the dummy variables representing the grouping
variable. The selected adjustment variables remain in their selected functional
forms.

Thus, different variables in `cont_vars` may be evaluated with slightly
different adjustment sets. When `verbose = TRUE`, the adjustment set used for
each variable is printed so that the user can verify which covariates are being
controlled for.

For each variable in `cont_vars`, candidate interaction models are fitted using
linear, FP1, and FP2 functions. The selection criterion, specified by
`criterion`, determines how these candidate models are compared. The available
criteria are `"pvalue"`, `"aic"`, and `"bic"`.

For a binary grouping variable, let $t = 0$ denote the reference group and 
$t = 1$ denote the comparison group. Let $z$ be the continuous covariate under
investigation. Prior to FP transformation, $z$ is shifted when it contains 
negative or zero values to ensure positivity, and optionally scaled for 
numerical stability.

If an FP2 function with powers $p_1$ and $p_2$ is selected, the group-specific
contribution of $z$ to the linear predictor for observation $i$ is
$$
\hat{f}_j(z_i)
=
\hat{\beta}_{j1}\, \phi_1(z_i)
+
\hat{\beta}_{j2}\, \phi_2(z_i),
\qquad j = 0, 1,
$$
where $\phi_1(z_i) = z_i^{p_1}$ and $\phi_2(z_i) = z_i^{p_2}$ are the FP
basis functions (with $z_i^0 = \log z_i$ by convention). For repeated powers
($p_1 = p_2$), the second term becomes $\phi_2(z_i) = z_i^{p_1} \log z_i$.
The interaction test assesses whether the group-specific functions are parallel
on the scale of the linear predictor. For Gaussian models, this is the outcome
scale. For logistic regression, it is the log-odds scale; for Poisson
regression, it is the log-rate or log-mean scale; and for Cox models, it is the
log hazard scale.

A useful way to display the interaction is the treatment-effect function,
$$
\hat{d}(z_i)
=
\hat{f}_1(z_i) - \hat{f}_0(z_i),
$$
together with a pointwise 95% confidence interval. If the two functions are
parallel, the difference is approximately constant over $z$. If the difference
varies with $z$, this suggests that the effect of the grouping variable on the 
outcome depends on the value of the continuous covariate. Pointwise standard 
errors for $\hat{d}(z_i)$ and for the group-specific functions $\hat{f}_j(z_i)$ 
are estimated using the delta method applied to the variance-covariance matrix 
of the fitted interaction model.

The treatment-effect plot should be interpreted as a model-based visual summary.
The confidence interval is pointwise and depends on the selected model.

# Flexibility levels {#Flex}

The argument `flex` controls how FP powers are selected and constrained across 
the levels of the grouping variable. Four flexibility levels are available: 
`"flex1"`, `"flex2"`, `"flex3"`, and `"flex4"`. These variants are available 
for both FP1 and FP2 functions.

The flexibility levels differ in where the FP powers are selected and whether 
the same powers are imposed across the levels of the grouping variable. In 
general, `flex1` is the least flexible and `flex4` is the most flexible option. 
Increasing flexibility allows more complex group-specific functional forms, but 
it also increases the risk of instability, especially in small samples or when 
the continuous covariate has a skewed distribution, influential observations, 
or sparse data in parts of its range.

The appropriate flexibility level should therefore be chosen using subject matter
knowledge, the expected shape of the interaction, and sample-size considerations. 
More flexible models may be useful when there is a strong scientific rationale 
for different functional forms across groups, but simpler models are often 
preferable when the goal is stable estimation and an interpretable treatment effect
function. For further details, see Royston and Sauerbrei (2004), 
Royston and Sauerbrei (2013), and Royston and Sauerbrei (2014).

## `flex1`

`flex1` is the default and least flexible option. FP powers for the continuous 
covariate are estimated once from the pooled main-effects model. This model 
includes the FP terms for the continuous covariate and the group indicators, but
no group-by-covariate interaction, while adjusting for the selected covariates 
in the adjustment model. The FP powers selected from this pooled main effects 
model are then held fixed and used at each level of the grouping variable. Thus,
the same FP basis functions are used in all groups, and only the regression 
coefficients of these basis functions are allowed to differ between groups.

Because the main-effects and interaction models share the same FP powers, they 
are nested. The interaction test is therefore a standard likelihood-ratio test
comparing the two models, with $(K - 1) \times m$ degrees of freedom, where $K$ 
is the number of groups and $m$ is the number of FP terms. Thus, $m = 1$ for FP1
and $m = 2$ for FP2. For the common case of two groups, this gives 1 df for FP1 
and 2 df for FP2.

`flex1` corresponds to the original MFPI procedure proposed by Royston and 
Sauerbrei (2004). In the significance-level simulation study, Royston and 
Sauerbrei (2013) found that FP2 with `flex1` maintained significance levels 
close to nominal, whereas the other FP2 flexibility variants showed larger 
departures from the nominal level. In the later power study, Royston and 
Sauerbrei (2014) recommended FP1 with `flex3` as the general practical choice 
when there is no strong subject-matter reason to expect a non-monotone 
treatment-effect function. When subject-matter knowledge gives good reasons to
expect a non-monotone treatment-effect function, they instead proposed using 
FP2 with `flex1`.

## `flex2`

`flex2` estimates FP powers from an interaction model rather than from the 
pooled main-effects model used by `flex1`. The powers are optimized over the 
levels of the grouping variable, but are constrained to be the same across 
groups. These same powers are then also used in the main-effects model.

Thus, `flex2` may select powers different from those selected by `flex1`, 
because power selection is informed by group-specific effects. However, the 
fitted interaction model still uses a common FP power set across groups; the 
group-specific functions differ through their regression coefficients.

The models are non-nested, and the interaction test has the same nominal degrees
of freedom as `flex1`. The resulting chi-square-based P-values should therefore
be regarded as approximate.

In the simulations, FP1 with `flex2` had high power but slightly inflated 
significance levels. FP2 with `flex2` showed clearly inflated type I error and
was therefore not recommended as a routine choice.


## `flex3`

`flex3` estimates FP powers separately for the main-effects model and the 
interaction model. The main-effects powers are selected from the pooled 
main-effects model, as in `flex1`, while the interaction powers are selected 
from a model allowing group-by-covariate interaction, as in `flex2`. These 
interaction powers are still constrained to be the same across groups, so 
group-specific functions differ through their regression coefficients rather 
than through group-specific FP powers.

Because the main-effects and interaction models can use different FP power sets,
the models are generally non-nested. The likelihood-ratio statistic is still 
computed and compared with a chi-square distribution using the nominal degrees
of freedom: $(K - 1) \times m$. The resulting chi-square-based P-value should 
therefore be regarded as approximate.

In the significance-level simulation study of Royston and Sauerbrei (2013), 
FP1 with `flex3` had significance levels close to nominal in the well-behaved
case and in scenarios 1–6 of the badly behaved case, but FP1 models showed 
inflated significance levels in scenarios 7–9. FP2 with `flex3` showed some 
inflation: less than FP2 with `flex2`, but still large enough for Royston and 
Sauerbrei (2014) judged it unsuitable as a candidate method because of 
inflated type I error.

In the power study, FP1 with `flex3` performed well. Royston and Sauerbrei (2014) 
considered FP1 with `flex1`–`flex3` to have the largest power among methods with
acceptable significance levels and recommended FP1 with `flex3` as the preferred
practical choice. They recommended this option when there is no strong 
subject-matter reason to expect a non-monotone treatment-effect function, 
noting that the FP1 class includes the linear function and tends to yield 
functions that are simple, understandable, and transferable.

## `flex4`

`flex4` is the most flexible option. Like `flex3`, it allows the main effects 
and interaction models to use separately selected FP powers. Its distinguishing
feature is that the FP powers in the interaction model are also allowed to differ
across the levels of the grouping variable. Thus, each group may have its own FP
power set and its own regression coefficients. The models are non-nested. The 
interaction test uses $2(K - 1) \times m$ degrees of freedom, which is twice as
many as `flex1`–`flex3` to account for the additional flexibility from allowing 
group-specific powers. The chi-square-based P-value is approximate.

In the significance-level simulation study of Royston and Sauerbrei (2013), FP2
with `flex4` was conservative, with empirical significance levels around 2–3% 
rather than the nominal 5%. In the subsequent power study, Royston and 
Sauerbrei (2014) reported that FP2 with `flex4` had a rejection rate below the
target significance level and low power. They therefore did not recommend it as
a routine choice.

In the significance-level simulation study of Royston and Sauerbrei (2013), FP2 
with `flex4` was conservative: the empirical significance level was about 2% in
the well-behaved case and about 3% in the badly behaved case, rather than the 
nominal 5%. In the subsequent power study, Royston and Sauerbrei (2014) reported
that FP2 with `flex4` had significance levels below the target level and low 
power. For this reason, FP2 with `flex4` was not considered a suitable candidate
method in their recommendations.

`flex4` may be useful when there is a strong subject matter reason to believe that
the functional form of the covariate effect differs between groups. However, 
this may require sufficiently large samples.

The four MFPI variants are summarized in the table below. The table shows how
the FP powers are selected and constrained, and gives the degrees of freedom
used for the interaction test.

```{r flex-table, echo=FALSE}
flex_table <- data.frame(
  Variant = c("flex1", "flex2", "flex3", "flex4"),
  Description = c(
    "FP powers determined for the main effect of z; use these same powers at each level of t",
    "FP optimized over levels of t; use these same powers at each level of t",
    "As flex2, but re-estimate powers for the main effect of z",
    "Optimize FP for the main effect and separately at each level of t"
  ),
  `d.f. FP1` = c(1, 1, 1, 2),
  `d.f. FP2` = c(2, 2, 2, 4),
  check.names = FALSE
)

knitr::kable(
  flex_table,
  align = c("l", "l", "c", "c"),
caption = "Variants of MFPI with degrees of freedom for the tests of interaction 
from Royston and Sauerbrei (2013)."
)
```

## Summary of flexibility levels

In the original MFPI papers, the analyst typically specifies the FP degree, FP1
or FP2, and the flexibility level (`flex`) in advance. The four flexibility
levels control how FP powers are selected and constrained across the levels of
the grouping variable. Our implementation of `mfpi()` evaluates three 
functional-form candidates for each variable in `cont_vars`: linear, FP1, and 
FP2. Each candidate is evaluated using its corresponding interaction test, and 
the `criterion` parameter determines which candidate is selected. In the 
candidate table, the selected candidate is marked with `+`.

This allows users to prespecify a preferred degree based on subject matter 
reasoning while still using the other candidates for comparison and sensitivity 
assessment. For example, a user following the recommendation of Royston and 
Sauerbrei (2014) might focus on FP1 with `flex3` when there is no strong reason 
to expect a non-monotone treatment-effect function, while treating the linear 
and FP2 results as comparison or sensitivity checks. When subject-matter knowledge
gives good reasons to expect a non-monotone treatment-effect function, Royston 
and Sauerbrei (2014) recommended FP2 with `flex1`.

## Note on non-nested comparisons

For `flex2`, `flex3`, and `flex4`, significance tests for interaction are based
on chi-square distributions with the stated degrees of freedom. However, these
tests are non-nested, and the resulting p-values lack a firm theoretical
underpinning. The significance levels may therefore be liberal
or conservative (Royston and Sauerbrei, 2013).

As an alternative, AIC or BIC may be used to compare the fitted main-effects and
interaction models. Smaller values indicate a better balance between model fit
and complexity. Small differences in AIC or BIC provide little evidence for 
preferring one model over another; values around 2 are often used as a rough 
guide rather than a strict cutoff.

# Influential observations and Winsorisation

FP models can be sensitive to outliers or influential observations in the
continuous predictor. A small number of influential observations in the tails of
a variable in `cont_vars` may strongly affect the selected functional form and
may produce spuriously significant interactions. Therefore, it is important to
inspect the data before fitting an MFPI model, for example, using univariable
plots.

In `mfpi()`, Winsorisation is provided as an optional pre-processing step to
reduce the influence of extreme values in variables listed in `cont_vars`.
Values below the lower percentile and above the upper percentile specified by
`winsorize_probs` are replaced by the corresponding percentile cutoffs. The
default is
$$
\texttt{winsorize_probs} = (0.01, 0.99).
$$
Winsorisation preserves the number of observations because no rows are removed.
Only the extreme values are truncated, while the main part of the distribution
remains unchanged.

Note that variables in `cont_vars` cannot be listed in `spike_vars` or
`catzero_vars` — these options are only available for adjustment variables,
because the interaction design matrix does not include group-specific
$I(x = 0)$ indicators. If a variable appears in both `cont_vars` and
`spike_vars` or `catzero_vars`, the spike or catzero treatment is dropped
with a warning.

Variables in `cont_vars` that are also listed in `zero_vars` are excluded from
Winsorisation, because the zero structure of these variables carries substantive
modelling meaning, and truncating the lower tail could distort it. Excluded
variables are recorded with `NA` limits in `winsorize_limits` for transparency.

Winsorisation is applied after shifting and scaling, so the percentile cutoffs
and the limits stored in `winsorize_limits` are on the transformed scale. The
actual cutoffs applied are returned in the fitted object and can be inspected
via `fit$winsorize_limits`.

To disable Winsorisation, use

```{r winsorize-disable, eval=FALSE}
winsorize = FALSE
```

More aggressive truncation can be requested, for example, by using

```{r winsorize-prob, eval=FALSE}
winsorize_probs = (0.05, 0.95)
```

Any Winsorisation or trimming applied should be reported transparently as part of
the initial data analysis.

# Checking selected interactions

A selected MFPI interaction should not be accepted solely based on a
p-value. The fitted curves should be checked graphically and interpreted in the
light of subject-matter knowledge.

For survival outcomes, Royston and Sauerbrei recommended checking selected
interactions using Kaplan--Meier curves in subgroups defined by the continuous
covariate. The subgroup curves should provide a rough visual confirmation of the
pattern suggested by the continuous treatment-effect plot. For example, if the
treatment-effect function suggests that treatment benefit decreases as `z`
increases, subgroup estimates should show a similar trend.

For other regression models, analogous checks can be made using grouped summaries
or subgroup-specific estimates. These checks are not substitutes for the MFPI
analysis, because they require categorising the continuous covariate, but they
can help detect obvious artefacts from flexible modelling.

A useful workflow is:

1. inspect the distribution of each variable in `cont_vars`;
2. fit the MFPI model;
3. inspect the fitted group-specific functions;
4. inspect the treatment-effect or difference function;
5. check whether the pattern is plausible and supported by subgroup summaries;
6. consider sensitivity analyses if the result depends strongly on a few
   observations or on the adjustment model.



# Shifting, scaling, and centering

## Shifting {-}

Fractional polynomials require strictly positive input values. As in `mfp2()`,
`mfpi()` estimates shift terms to ensure all values of the continuous covariates
are positive before FP transformation. For a variable $z$ with minimum value
$z_{\min}$, the shift is chosen so that $z + \text{shift} > 0$ for all
observations.

## Scaling {-}

After shifting, `mfpi()` estimates a scale factor for each variable. Scaling
places the shifted covariates on a convenient numerical range for FP power
selection. This improves numerical stability during the MFP backfitting cycles.
Scaling does not affect which FP powers are selected, only the numerical
precision of the fitting algorithm.

After FP powers have been selected, the variables are backscaled before the
final model is fitted. This means the regression coefficients in the returned
model are on the $\phi(z + \text{shift})$ scale, matching what `mfp2()` would 
produce. The user does not need to undo the scaling manually.

## Centering {-}

When `center = TRUE`, centering is applied in both the adjustment model and
the interaction model, but the two models are centered differently.

**Adjustment model.** Standard `mfp2()` centering is applied: each
FP-transformed column is centered by subtracting its column mean. For binary
covariates, centering uses the lower of the two distinct values rather than
the mean, the same idea of `mfp2()`.

**Interaction model.** The argument `center_type` controls how the
group-specific FP columns are centered:

- `center_type = "grand"` (default): the grand mean of the FP-transformed 
  variable, computed across all observations regardless of group membership, is
  subtracted from each group's columns. All groups share the same centering
  constant. 
- `center_type = "group"`: each group's columns are centered by the mean computed
   from in-group observations only. Different groups may use different centering
  constants.

In both cases, Centering affects the intercept and, when an interaction is 
present, the group dummy coefficients. It does not affect the interaction test,
the treatment-effect function $\hat{f}_1(z) - \hat{f}_0(z)$, or the fitted
values.

# Handling non-positive values

When a continuous predictor contains a mixture of non-positive and positive
values, standard shifting and FP transformation may not be appropriate. The
`mfpi()` function provides three options, matching those in `mfp2()`.

The `zero_vars` option specifies variables for which FP transformations are
applied only to strictly positive values. Non-positive values are set to zero in
the transformed variable, and no additional binary indicator is added.

The `catzero_vars` option extends this approach by adding a binary indicator for
whether the variable is zero. This indicator is included in the adjustment model
alongside the FP terms for the positive part. The `catzero_vars` option implies
`zero_vars`.

The `spike_vars` option triggers the spike-at-zero (SAZ) algorithm. This
evaluates whether the binary zero indicator and the FP function each contribute
explanatory value. The `spike_vars` option implies both `catzero_vars` and
`zero_vars`. For more details about SAZ, see the corresponding vignette in the
`mfp2` package.

## Interaction-tested variables cannot use `catzero_vars` or `spike_vars` {-}

Variables listed in `cont_vars` are tested for interaction with the grouping
variable. These variables cannot currently use the `catzero_vars` or
`spike_vars` options, because the methodology for combining these options with
interaction testing is not yet available.

Therefore, if a variable appears in both `cont_vars` and either `catzero_vars`
or `spike_vars`, it will be automatically removed from `catzero_vars` or
`spike_vars`, with a warning.

However, `zero_vars` can be used with variables in `cont_vars`. In this case,
only the positive part undergoes FP transformation, while the remaining values
are set to zero.

Variables listed in `catzero_vars` or `spike_vars` that are not in `cont_vars`
are handled in the adjustment model as described in the `mfp2()` documentation.

# Worked example: Advanced prostate cancer data

We illustrate `mfpi()` using the `advanced_prostate_cancer` dataset included in
the `mfp2` package. The outcome is time to death (overall survival), defined by
`survtime` and the censoring indicator `cens`. The binary variable `rx`
(treatment allocation) is the grouping variable. For a full description of the
data, see Royston and Sauerbrei (2008). The dataset contains 475 observations 
with eight continuous measurements (`age`, `wt`, `sbp`, `dbp`, `sz`, `ap`, `hg`,
`sg`) and five categorical covariates (`pf`, `hx`, `bm`, `stage`, `ekg`).

```{r data}
data("advanced_prostate_cancer", package = "mfp2")
head(advanced_prostate_cancer)
```



## Formula interface

The formula interface is the most convenient way to call `mfpi()`. Continuous
variables can be wrapped in `fp()` to control FP-specific settings such as
degrees of freedom or candidate power sets. The key arguments are:

- `group_var`: the name of the grouping (treatment) variable.
- `cont_vars`: a character vector naming the continuous variables to be assessed
  for interaction with `group_var`.
- `flex`: the flexibility level for FP power selection across groups.
- `family`: the regression family (`"cox"` for survival data).
- `select` and `alpha`: significance levels for variable selection and FP degree
  selection in the adjustment model.

## Prespecified interaction analysis

Suppose that, based on prior evidence, we have identified `age` as the single
covariate to investigate for interaction with treatment. Because only one
interaction is tested, the resulting p-value can be interpreted directly without
multiplicity adjustment.

```{r prespecified}
fit_pre <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var  = "rx",
  cont_vars  = "age",
  data       = advanced_prostate_cancer,
  flex       = "flex1",
  family     = "cox",
  select     = 0.05,
  alpha      = 0.05,
  verbose = FALSE
)
```

### Understanding the output

Calling `print()` on the fitted object displays a structured three-step summary.

```{r print-pre}
print(fit_pre)
```

**Step 1: Adjustment model (MFP).** This section shows the results of running
the MFP algorithm on the full set of predictors. Variables that remain after
selection at `select = 0.05` are shown with their final FP powers and degrees of
freedom. Variables eliminated by MFP are listed under `"Dropped"`. Because
`include_group_var = TRUE` by default, the grouping variable `rx` is included in
the MFP algorithm but forced into the model: its nominal significance level is
set to 1, so it is never eliminated during backfitting. The retained covariates,
in their selected transformed or untransformed forms, are then used for adjustment
in the interaction models for `age`.

**Step 2: All interaction candidates.** For `age`, three candidate interaction
models are fitted: linear, FP1, and FP2. Each row shows the candidate type, the
deviance of the main-effects model, the deviance of the interaction model, the
likelihood ratio test statistic, the p-value, and the information-criterion
differences. AIC and BIC differences are computed as main-effects minus
interaction, so positive values favour the interaction model. The best
candidate, defined as the one with the smallest p-value, is flagged with `+`.

**Step 3: Interaction summary.** This compact table shows the selected
functional form for `age` and its p-value. If the p-value is below
`p_interact = 0.05`, the variable is marked with `*`, indicating a statistically
significant interaction.

### Regression output {-}

The `summary()` method adds a fourth section with the full regression
coefficients of the winning interaction model.

```{r summary-pre}
summary(fit_pre)
```

**Step 4: Regression Output.** For each variable with a significant interaction,
the regression table from the fitted `coxph()` model is displayed. This table 
includes coefficient estimates, standard errors, z-values, and p-values for all
terms in the model. The interaction model uses the naming convention
`<varname><group><power_index>` for the group-specific FP terms. For example,
if `age` is modelled as FP1 with groups 0 and 1, the coefficients `age01` and
`age11` represent the FP term for group 0 and group 1, respectively. For FP2,
two terms per group are shown, for example `age01` and `age02` for group 0, and
`age11` and `age12` for group 1.

The group dummy coefficient, for example `rx1`, represents the main effect of
the grouping variable. The p-values in Step 4 are the standard Wald-test p-values
reported by `summary.coxph()`. These p-values do not account for the degrees of 
freedom used during FP power selection. Therefore, the appropriate p-value for
testing the interaction is the one reported in Step 3.

### Visualising the interaction

The `plot()` method produces graphical displays of the fitted interaction. Its
main arguments are:

- `terms`: a character vector of variable names to plot. By default (`NULL`),
  all variables with a significant interaction are plotted. Specifying a
  variable name restricts the output to that variable.
- `type`: controls which functional form is shown. The default (`NULL`) plots
  the selected candidate for each variable. Setting `type` to `"linear"`,
  `"fp1"`, or `"fp2"` plots that specific candidate from the full set of fitted
  models, regardless of whether it was selected. This is useful for inspecting
  non-selected candidates or comparing functional forms.
- `plot_type`: one of `"fitted"`, `"difference"`, or `"both"`. `"fitted"`
  overlays the group-specific fitted curves; `"difference"` shows the pointwise
  treatment-effect function; and `"both"` places both plots side by side
  (requires the `patchwork` package).
- `show_title`: logical, default `TRUE`. Each plot is annotated with a title
  describing the group comparison and a subtitle showing the variable name, the
  functional form, and the selection metric. Set this to `FALSE` to produce
  cleaner plots for reports or manuscripts.

The subtitle adapts to the selection criterion used when fitting the model. When
`criterion = "pvalue"`, the subtitle shows `p = <value>`. When
`criterion = "aic"` or `criterion = "bic"`, it shows `dAIC = <value>` or
`dBIC = <value>`, respectively.

When multiplicity adjustment is active (`p_adjust_method != "none"`), the
subtitle shows both the raw and adjusted p-values for selected variables, for
example: `age | FP1 | p = 0.016 (p_adj = 0.031)`. When `type` is set explicitly
to inspect a specific candidate, the subtitle shows `p_raw = <value>` to signal
that the displayed value is the unadjusted per-candidate p-value, not the
adjusted p-value used for the selection decision.

#### Fitted functions
The fitted plot overlays the estimated linear predictor as a function of `age`
for each treatment group, with 95% pointwise confidence bands and rug marks
showing the observed values. If the two curves are roughly parallel, there is
little evidence for interaction.

```{r plot-fitted-pre, fig.cap = "Group-specific fitted functions for the age × rx interaction."}
plot(fit_pre, plot_type = "fitted", show_ci_fitted = TRUE, show_rug = TRUE)
```



#### Treatment-effect difference
The difference plot shows the pointwise treatment-effect function
$\hat{d}(\text{age}) = \hat{f}_1(\text{age}) - \hat{f}_0(\text{age})$ with a
95% confidence band and a horizontal reference line at zero. Regions where the
band excludes zero suggest that the treatment effect differs from the overall
average at those values of `age`. If the confidence band is approximately
horizontal and includes zero throughout, this supports the conclusion that
treatment benefit does not depend on `age`.

```{r plot-diff-pre, fig.cap = "Treatment-effect difference for the age × rx interaction."}
plot(fit_pre, plot_type = "difference")
```



#### Side-by-side display
The `"both"` option places the fitted and difference plots side by side using
the `patchwork` package.

```{r plot-both-pre, fig.width = 10, fig.cap = "Fitted functions and treatment-effect difference for the age × rx interaction."}
plot(fit_pre, plot_type = "both")
```

## Exploratory interaction analysis

In the second setting, we screen all eight continuous covariates for possible
interaction with treatment. Because multiple interactions are tested, we apply
the Holm step-down correction to control the family-wise error rate across the
tested variables.

```{r exploratory}
fit_exp <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var       = "rx",
  cont_vars       = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg", "sg"),
  data            = advanced_prostate_cancer,
  flex            = "flex1",
  family          = "cox",
  select          = 0.05,
  alpha           = 0.05,
  p_adjust_method = "holm"
)
```

### Understanding the output

```{r print-exp}
print(fit_exp)
```

**Step 1** is identical to the prespecified analysis because the same predictors
and the same MFP settings are used.

**Step 2** now shows three candidate rows (linear, FP1, FP2) for each of the
eight continuous variables, with the best candidate per variable marked by `+`.

**Step 3** is the key table for the exploratory analysis. It shows one row per
variable with the winning functional form, the raw p-value from the
per-variable interaction test, and the Holm-adjusted p-value. Variables whose
adjusted p-value falls below `p_interact = 0.05` are marked with `*`. In an
exploratory analysis, the adjusted p-value is the one that should be used for
the selection decision. Variables marked `*` are candidates for genuine
treatment-by-covariate interaction; variables without `*` do not provide
sufficient evidence of interaction after accounting for the number of tests
performed.

### Regression output for selected variables

```{r summary-exp}
summary(fit_exp)
```

Step 4 prints the regression output only for variables whose interaction was
selected (those marked with `*` in Step 3). As before, the correct interaction
test p-values are those from Step 3, not the Wald p-values in the regression
table.

### Plotting selected variables

By default, `plot()` shows results only for variables with a significant
interaction.

```{r plot-exp-fitted, fig.cap = "Fitted functions for variables with significant interactions."}
plot(fit_exp, plot_type = "fitted")
```

```{r plot-exp-diff, fig.cap = "Treatment-effect differences for variables with significant interactions."}
plot(fit_exp, plot_type = "difference")
```

### Plotting non-selected variables

To inspect a candidate fit for a variable that was not selected, specify the
`terms` and `type` arguments. For example, to see the FP1 candidate for `wt`:

```{r plot-nonsig, fig.cap = "FP1 candidate for the wt × rx interaction (not selected)."}
plot(fit_exp, terms = "wt", type = "fp1", plot_type = "both")
```

This is useful for diagnostic inspection: it shows what the candidate functional
form looked like before selection and helps assess whether the non-selection
decision is supported by the data.

## Alternative selection criteria: AIC and BIC

In all examples above, the default `criterion = "pvalue"` was used. This
selects the candidate interaction model (linear, FP1, or FP2) with the smallest
p-value, and retains the interaction only if the p-value is below
`p_interact`. For `flex2`, `flex3`, and `flex4`, however, the main-effects and
interaction models are non-nested, and the p-values from the likelihood ratio
test lack a firm theoretical underpinning (see the
[Note on non-nested comparisons](#note-on-non-nested-comparisons) section).

As an alternative, `mfpi()` supports model selection based on information
criteria via `criterion = "aic"` or `criterion = "bic"`. Under these criteria,
models are compared by the difference
$$
\Delta\text{IC}
= \text{IC}_{\text{main}} - \text{IC}_{\text{interaction}},
$$
where IC is AIC or BIC. A positive value means the interaction model has a
smaller information criterion and is therefore preferred. The candidate with the
largest $\Delta\text{IC}$ is selected, and the interaction is retained only if
this difference exceeds `min_improvement` (default 2). Small differences in AIC
or BIC provide little evidence for preferring one model over the other; a
threshold of 2 is a commonly used rough guide.

The `criterion` argument affects two stages of the analysis. In Step 1, it
governs the MFP adjustment model: variable elimination and FP degree selection
are based on the chosen criterion rather than on significance levels. In Step 2,
the same criterion determines which candidate interaction model is preferred
and whether the interaction is retained.

BIC penalises model complexity more heavily than AIC (the penalty per parameter
is $\log n$ rather than 2), so it tends to favour simpler functional forms and
is more conservative in declaring interactions.

Note that when `criterion = "aic"` or `"bic"`, the `p_adjust_method` argument
has no effect, because multiplicity adjustment applies to p-values, not to
information criteria.

### Example: BIC-based selection

We repeat the exploratory analysis using BIC as the selection criterion. The
`min_improvement` threshold is set to 2, meaning that the interaction model must
improve BIC by at least 2 units relative to the main-effects model.

```{r bic-exploratory}
fit_bic <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var       = "rx",
  cont_vars       = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg", "sg"),
  data            = advanced_prostate_cancer,
  flex            = "flex1",
  family          = "cox",
  criterion       = "bic",
  min_improvement = 2
)
```

### Understanding the output

```{r print-bic}
print(fit_bic)
```

**Step 1** may differ from the p-value-based analyses. Because the MFP algorithm
now uses BIC for variable selection and FP degree choice, it may retain a
different set of adjustment variables or select different functional forms. BIC's
heavier complexity penalty means that simpler models (linear terms, fewer
retained variables) are favoured compared to the p-value-based analysis.

**Step 2** shows the same three candidates (linear, FP1, FP2) per variable, but
the decisive column is now `BIC_main_minus_int` rather than `pvalue`. The `+`
marker flags the candidate with the largest BIC improvement for each variable.
The p-values and AIC differences are still shown for comparison, but BIC
governs the selection.

**Step 3** now displays one row per variable with the column `dBIC` instead of
`pvalue`. Variables whose dBIC exceeds `min_improvement = 2` are marked with
`*`. A positive dBIC means the interaction model is preferred by BIC; a negative
or small positive value means the main-effects model is adequate. Because BIC
penalises complexity more heavily, it is common for BIC-based selection to retain
fewer interactions than p-value-based selection.

### Regression output and plots

```{r summary-bic}
summary(fit_bic)
```

Step 4 shows the regression output for variables selected under the BIC
criterion. The plot subtitles adapt automatically to show `dBIC` rather than
p-values.

```{r plot-bic-fitted, fig.cap = "Fitted functions for variables selected by BIC."}
plot(fit_bic, plot_type = "fitted")
```

```{r plot-bic-diff, fig.cap = "Treatment-effect differences for variables selected by BIC."}
plot(fit_bic, plot_type = "difference")
```

### Comparing p-value and BIC results

It can be instructive to compare the results from the p-value-based and
BIC-based analyses. A variable that is selected under both criteria provides
stronger evidence for a genuine interaction than one that is selected under only
one. Conversely, a variable selected by p-value but not by BIC may reflect a
statistically detectable but small interaction that does not substantially
improve model fit. Variables selected by neither criterion can be set aside with
greater confidence.
```{r restore-options, include=FALSE}
options(width = oldwidth)
```

# References

- Royston, P. and Sauerbrei, W. (2004). A new approach to modelling interactions
  between treatment and continuous covariates in clinical trials by using
  fractional polynomial submodels. *Statistics in Medicine*, 23(16), 2509–2525.

- Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A
  Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for
  Modelling Continuous Variables.* Wiley.

- Royston, P. and Sauerbrei, W. (2013). Interaction of treatment with a
  continuous variable: simulation study of significance level for several methods
  of analysis. *Statistics in Medicine*, 32(22), 3788–3803.

- Royston, P. and Sauerbrei, W. (2014). Interaction of treatment with a
  continuous variable: simulation study of power for several methods of analysis.
  *Statistics in Medicine*, 33(27), 4695–4708.

- Sauerbrei, W., Royston, P. and Zapien, K. (2007). Detecting an interaction
  between treatment and a continuous covariate: a comparison of two approaches.
  *Computational Statistics & Data Analysis*, 51(8), 4054–4063.

```{r cleanup, include=FALSE}
options(width = oldwidth)
```
