---
title: "MFPI: Multivariable Fractional Polynomial Interaction"
author: "Edwin Kipruto and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{MFPI: Multivariable Fractional Polynomial Interaction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-size: 15px;
  line-height: 1.5;
}

/* Justify normal prose, but not code, tables, headings, or captions */
p, li {
  text-align: justify;
  line-height: 1.5;
}

h1, h2, h3, h4 {
  margin-top: 1.4em;
  margin-bottom: 0.6em;
  color: #1f77b4;
}

#TOC a {
  color: inherit;
  text-decoration: none;
}

pre {
  font-size: 13px;
  line-height: 1.4;
  overflow-x: auto;
}

code {
  font-size: 90%;
}

img {
  max-width: 100%;
  height: auto;
}

.figure {
  margin-top: 1.2em;
  margin-bottom: 1.2em;
}

.caption {
  font-size: 90%;
  color: #555555;
  text-align: left;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  dev = "png",
  dpi = 150
)

oldwidth <- getOption("width")
options(width = 100)

library(mfp2)
library(ggplot2)
library(patchwork)
library(survival)
```


# Introduction

Investigating whether the effect of a treatment or other grouping variable
changes across the values of a continuous covariate is a common and important
problem in clinical and epidemiological studies. A common approach is to
categorise the continuous variable using cutpoints, but this discards
information, may reduce power, and can make the evidence for interaction depend
strongly on the number and location of the cutpoints (Royston and Sauerbrei,
2013). Assuming a linear interaction avoids arbitrary cutpoints but may still be
restrictive: the main effect, the interaction function, or both may be
nonlinear. Misspecifying a nonlinear main effect can even generate a spurious
interaction (Royston and Sauerbrei, 2013).

The Multivariable Fractional Polynomial Interaction (MFPI) procedure (Royston
and Sauerbrei, 2004) addresses these issues by modelling covariate effects
using fractional polynomial (FP) functions within levels of a grouping
variable. Four flexibility variants (`flex1` to `flex4`) control how FP powers
are selected and constrained across groups (Royston and Sauerbrei, 2013, 2014).
The procedure first constructs an adjustment model using the MFP algorithm,
then tests whether group-specific functions are parallel on the scale of the
linear predictor.

This vignette describes `mfpi()` in the `mfp2` package. The function supports
Gaussian, binomial, Poisson, and Cox regression models, and provides both
formula and matrix interfaces (see `?mfpi` for the matrix interface). It
investigates interactions between a categorical grouping variable (`group_var`)
and one or more continuous covariates (`cont_vars`). The grouping variable is
treated as a factor internally, with the lowest value used as the reference
group.

# Quick start

We begin with a complete working example before describing the methodology in
detail. The `advanced_prostate_cancer` dataset from `mfp2` contains 475
observations with survival outcome (`survtime`, `cens`), a binary treatment
variable (`rx`), eight continuous covariates, and five categorical covariates.

```{r data}
data("advanced_prostate_cancer", package = "mfp2")
head(advanced_prostate_cancer)
```

The following call tests whether the treatment effect on survival differs
across values of `age`, adjusting for all other predictors:

```{r quickstart}
fit_quick <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var  = "rx",
  cont_vars  = "age",
  data       = advanced_prostate_cancer,
  flex       = "flex1",
  family     = "cox",
  p_interact = 0.05,
  criterion = "pvalue",
  verbose    = FALSE
)
print(fit_quick)
```

```{r quickstart-plot, fig.width = 10, fig.cap = "Fitted functions and treatment-effect difference for the age × rx interaction."}
plot(fit_quick, plot_type = "both")
```

The output is organised in three steps: the adjustment model selected by MFP
(Step 1), the candidate interaction models (Step 2), and the interaction
summary (Step 3). The `summary()` method adds a fourth step with the
regression coefficients. The `plot()` method displays the group-specific fitted
functions and the treatment-effect difference. The rest of this vignette
explains each component in detail.

# Prespecified and exploratory interaction analyses

Royston and Sauerbrei (2008) distinguished between two settings for
investigating interactions:

1. **Prespecified analysis.** This evaluates an interaction defined before
   looking at the data, usually based on prior evidence or the trial protocol.
   Because the analysis focuses on a single interaction, multiplicity adjustment
   across several covariates is not required.

2. **Hypothesis generation.** This screens several covariates for possible
   interaction with the grouping variable. Because multiple interaction tests
   are performed, some small p-values may occur by chance. The interaction
   p-values may therefore need to be adjusted for multiplicity. Any interaction
   identified in such an exploratory analysis should be regarded as a candidate
   finding that requires confirmation in independent data.

The `mfpi()` function supports both settings. For a prespecified analysis,
`cont_vars` contains a single continuous covariate. For an exploratory analysis,
`cont_vars` lists several continuous covariates, each assessed in turn. For each
variable under test, the adjustment covariates consist of the remaining selected
variables, i.e., the variable being tested is excluded from the adjustment set.

When several covariates are screened and `criterion = "pvalue"`, interaction
p-values can be adjusted across `cont_vars` using the `p_adjust_method`
argument. Available methods include `"holm"` (Bonferroni--Holm step-down), `"bonferroni"`, `"hochberg"`, `"BH"` (Benjamini--Hochberg), and `"BY"`. The default is `p_adjust_method = "none"`. This correction applies across the variables screened in `cont_vars`, not across the internal candidate models (linear, FP1, FP2) fitted by MFPI, which are part of the selection procedure.

# The MFPI approach

The MFPI procedure operates in two broad stages.

First, an adjustment model is selected using the MFP algorithm. The algorithm is
run once on the full predictor matrix to select adjustment variables and their
functional forms. This step may include variables that will later be tested for
interaction. The selected adjustment model is then used to control for
confounding in the interaction analyses.

Second, for each continuous variable listed in `cont_vars`, `mfpi()` evaluates
whether the relationship between that covariate and the outcome differs across
the levels of `group_var`. For each covariate under investigation, denoted by
$z$, the adjustment set contains the variables selected by the MFP adjustment
model, excluding $z$ itself and the dummy variables representing the grouping
variable. The selected adjustment variables remain in their selected functional
forms.

Thus, different variables in `cont_vars` may be evaluated with slightly
different adjustment sets. When `verbose = TRUE`, the adjustment set used for
each variable is printed so that the user can verify which covariates are being
controlled for.

For each variable in `cont_vars`, candidate interaction models are fitted using
linear, FP1, and FP2 functions. The selection criterion, specified by
`criterion`, determines how these candidate models are compared. The available
criteria are `"pvalue"`, `"aic"`, and `"bic"`.

For a binary grouping variable, let $t = 0$ denote the reference group and
$t = 1$ denote the comparison group. Let $z$ be the continuous covariate under
investigation. Prior to FP transformation, $z$ is shifted when it contains
negative or zero values to ensure positivity, and optionally scaled for
numerical stability.

If an FP2 function with powers $p_1$ and $p_2$ is selected, the group-specific
contribution of $z$ to the linear predictor for observation $i$ is
$$
\hat{f}_j(z_i)
=
\hat{\beta}_{j1}\, \phi_1(z_i)
+
\hat{\beta}_{j2}\, \phi_2(z_i),
\qquad j = 0, 1,
$$
where $\phi_1(z_i) = z_i^{p_1}$ and $\phi_2(z_i) = z_i^{p_2}$ are the FP
basis functions (with $z_i^0 = \log z_i$ by convention). For repeated powers
($p_1 = p_2$), the second term becomes $\phi_2(z_i) = z_i^{p_1} \log z_i$.
The interaction test assesses whether the group-specific functions are parallel
on the scale of the linear predictor. For Gaussian models, this is the outcome
scale. For logistic regression, it is the log-odds scale; for Poisson
regression, it is the log-rate or log-mean scale; and for Cox models, it is the
log hazard scale.

A useful way to display the interaction is the treatment-effect function,
$$
\hat{d}(z_i)
=
\hat{f}_1(z_i) - \hat{f}_0(z_i),
$$
together with a pointwise 95% confidence interval. If the two functions are
parallel, the difference is approximately constant over $z$. If the difference
varies with $z$, this suggests that the effect of the grouping variable on the
outcome depends on the value of the continuous covariate. Pointwise standard
errors for $\hat{d}(z_i)$ and for the group-specific functions $\hat{f}_j(z_i)$
are estimated using the delta method applied to the variance-covariance matrix
of the fitted interaction model.

The treatment-effect plot should be interpreted as a model-based visual summary.
The confidence interval is pointwise and depends on the selected model.

# Flexibility levels {#Flex}

The argument `flex` controls how FP powers are selected and constrained across
the levels of the grouping variable. Four flexibility levels are available:
`"flex1"`, `"flex2"`, `"flex3"`, and `"flex4"`. These variants are available
for both FP1 and FP2 functions.

In general, `flex1` is the least flexible and `flex4` is the most flexible
option. Increasing flexibility allows more complex group-specific functional
forms, but it also increases the risk of instability, especially in small
samples or when the continuous covariate has a skewed distribution, influential
observations, or sparse data in parts of its range. The appropriate flexibility
level should therefore be chosen using subject matter knowledge, the expected
shape of the interaction, and sample-size considerations. For further details,
see Royston and Sauerbrei (2004, 2013, 2014).

## `flex1`

`flex1` is the default and least flexible option. FP powers for the continuous
covariate are estimated once from the pooled main-effects model. This model
includes the FP terms for the continuous covariate and the group indicators, but
no group-by-covariate interaction, while adjusting for the selected covariates
in the adjustment model. The FP powers selected from this pooled main-effects
model are then held fixed and used at each level of the grouping variable. Thus,
the same FP basis functions are used in all groups, and only the regression
coefficients of these basis functions are allowed to differ between groups.

Because the main-effects and interaction models share the same FP powers, they
are nested. The interaction test is therefore a standard likelihood-ratio test
comparing the two models, with $(K - 1) \times m$ degrees of freedom, where $K$
is the number of groups and $m$ is the number of FP terms. Thus, $m = 1$ for FP1
and $m = 2$ for FP2. For the common case of two groups, this gives 1 df for FP1
and 2 df for FP2.

`flex1` corresponds to the original MFPI procedure proposed by Royston and
Sauerbrei (2004). In the significance-level simulation study, Royston and
Sauerbrei (2013) found that FP2 with `flex1` maintained significance levels
close to nominal, whereas the other FP2 flexibility variants showed larger
departures from the nominal level. When subject-matter knowledge gives good
reasons to expect a non-monotone treatment-effect function, Royston and
Sauerbrei (2014) recommended FP2 with `flex1`.

## `flex2`

`flex2` estimates FP powers from an interaction model rather than from the
pooled main-effects model used by `flex1`. The powers are optimized over the
levels of the grouping variable, but are constrained to be the same across
groups. These same powers are then also used in the main-effects model.

Thus, `flex2` may select powers different from those selected by `flex1`,
because power selection is informed by group-specific effects. However, the
fitted interaction model still uses a common FP power set across groups; the
group-specific functions differ through their regression coefficients.

The models are non-nested (see [Note on non-nested comparisons](#note-on-non-nested-comparisons)),
and the interaction test has the same nominal degrees of freedom as `flex1`.

In the simulations, FP1 with `flex2` had high power but slightly inflated
significance levels. FP2 with `flex2` showed clearly inflated type I error and
was therefore not recommended as a routine choice.

## `flex3`

`flex3` estimates FP powers separately for the main-effects model and the
interaction model. The main-effects powers are selected from the pooled
main-effects model, as in `flex1`, while the interaction powers are selected
from a model allowing group-by-covariate interaction, as in `flex2`. These
interaction powers are still constrained to be the same across groups, so
group-specific functions differ through their regression coefficients rather
than through group-specific FP powers.

Because the main-effects and interaction models can use different FP power sets,
the models are generally non-nested (see
[Note on non-nested comparisons](#note-on-non-nested-comparisons)).
The likelihood-ratio statistic is still computed and compared with a chi-square
distribution using the nominal degrees of freedom: $(K - 1) \times m$.

In the significance-level simulation study of Royston and Sauerbrei (2013),
FP1 with `flex3` had significance levels close to nominal. In the power study,
FP1 with `flex3` performed well. Royston and Sauerbrei (2014) considered FP1
with `flex1`--`flex3` to have the largest power among methods with acceptable
significance levels and recommended FP1 with `flex3` as the preferred practical
choice when there is no strong subject-matter reason to expect a non-monotone
treatment-effect function, noting that the FP1 class includes the linear
function and tends to yield functions that are simple, understandable, and
transferable. FP2 with `flex3` showed some type I error inflation and was not
recommended as a routine choice.

## `flex4`

`flex4` is the most flexible option. Like `flex3`, it allows the main-effects
and interaction models to use separately selected FP powers. Its distinguishing
feature is that the FP powers in the interaction model are also allowed to differ
across the levels of the grouping variable. Thus, each group may have its own FP
power set and its own regression coefficients. The models are non-nested (see
[Note on non-nested comparisons](#note-on-non-nested-comparisons)).
The interaction test uses $2(K - 1) \times m$ degrees of freedom, which is
twice as many as `flex1`--`flex3` to account for the additional flexibility from
allowing group-specific powers.

In the simulation studies of Royston and Sauerbrei (2013, 2014), FP2 with
`flex4` was conservative, with empirical significance levels around 2--3%
rather than the nominal 5%, and had low power. It was therefore not recommended
as a routine choice.

`flex4` may be useful when there is a strong subject-matter reason to believe
that the functional form of the covariate effect differs between groups.
However, this may require sufficiently large samples.

## Summary table

The four MFPI variants are summarized in the table below.

```{r flex-table, echo=FALSE}
flex_table <- data.frame(
  Variant = c("flex1", "flex2", "flex3", "flex4"),
  Description = c(
    "FP powers determined for the main effect of z; use these same powers at each level of t",
    "FP optimized over levels of t; use these same powers at each level of t",
    "As flex2, but re-estimate powers for the main effect of z",
    "Optimize FP for the main effect and separately at each level of t"
  ),
  `d.f. FP1` = c(1, 1, 1, 2),
  `d.f. FP2` = c(2, 2, 2, 4),
  check.names = FALSE
)

knitr::kable(
  flex_table,
  align = c("l", "l", "c", "c"),
  caption = "Variants of MFPI with degrees of freedom for the tests of
  interaction from Royston and Sauerbrei (2013)."
)
```

Our implementation of `mfpi()` evaluates three functional-form candidates for
each variable in `cont_vars`: linear, FP1, and FP2. Each candidate is evaluated
using its corresponding interaction test, and the `criterion` parameter
determines which candidate is selected. In the candidate table, the selected
candidate is marked with `+`. This allows users to prespecify a preferred degree
based on subject-matter reasoning while still using the other candidates for
comparison and sensitivity assessment. For example, a user following the
recommendation of Royston and Sauerbrei (2014) might focus on FP1 with `flex3`
when there is no strong reason to expect a non-monotone treatment-effect
function, while treating the linear and FP2 results as sensitivity checks.

## Note on non-nested comparisons {#note-on-non-nested-comparisons}

For `flex2`, `flex3`, and `flex4`, the main-effects and interaction models may
use different FP power sets and are therefore non-nested. The likelihood-ratio
test statistic is still computed and compared with a chi-square distribution
using the nominal degrees of freedom, but the resulting p-values lack a firm
theoretical underpinning. The significance levels may therefore be liberal or
conservative (Royston and Sauerbrei, 2013).

As an alternative, AIC or BIC may be used to compare models without requiring
nestedness. See the [Alternative selection criteria: AIC and BIC](#aic-bic)
section for details.

# Practical considerations

## Influential observations and Winsorisation

FP models can be sensitive to outliers or influential observations in the
continuous predictor. A small number of influential observations in the tails of
a variable in `cont_vars` may strongly affect the selected functional form and
may produce spuriously significant interactions. Therefore, it is important to
inspect the data before fitting an MFPI model, for example, using univariable
plots.

In `mfpi()`, Winsorisation is provided as an optional pre-processing step to
reduce the influence of extreme values in variables listed in `cont_vars`.
Values below the lower percentile and above the upper percentile specified by
`winsorize_probs` are replaced by the corresponding percentile cutoffs. The
default is `winsorize_probs = c(0.01, 0.99)`. Winsorisation preserves the
number of observations because no rows are removed. Only the extreme values are
truncated, while the main part of the distribution remains unchanged.

Note that variables in `cont_vars` cannot be listed in `spike_vars` or
`catzero_vars` --- these options are only available for adjustment variables,
because the interaction design matrix does not include group-specific
$I(x = 0)$ indicators. If a variable appears in both `cont_vars` and
`spike_vars` or `catzero_vars`, the spike or catzero treatment is dropped
with a warning.

Variables in `cont_vars` that are also listed in `zero_vars` are excluded from
Winsorisation, because the zero structure of these variables carries substantive
modelling meaning, and truncating the lower tail could distort it. Excluded
variables are recorded with `NA` limits in `winsorize_limits` for transparency.

Winsorisation is applied after shifting and scaling, so the percentile cutoffs
and the limits stored in `winsorize_limits` are on the transformed scale. The
actual cutoffs applied are returned in the fitted object and can be inspected
via `fit$winsorize_limits`.

To disable Winsorisation, set `winsorize = FALSE`. More aggressive truncation
can be requested, for example:

```{r winsorize-prob, eval=FALSE}
winsorize_probs = c(0.05, 0.95)
```

Any Winsorisation or trimming applied should be reported transparently as part
of the initial data analysis.

## Shifting and scaling {-}

Fractional polynomials require strictly positive input values. As in `mfp2()`,
`mfpi()` estimates shift terms to ensure all values of the continuous covariates
are positive before FP transformation. After shifting, `mfpi()` estimates a
scale factor for each variable, placing the shifted covariates on a convenient
numerical range for FP power selection. This improves numerical stability
during the MFP backfitting cycles. Scaling does not affect which FP powers are
selected, only the numerical precision of the fitting algorithm.

After FP powers have been selected, the variables are backscaled before the
final model is fitted. This means the regression coefficients in the returned
model are on the $\phi(z + \text{shift})$ scale, matching what `mfp2()` would
produce. The user does not need to undo the scaling manually.

## Centering {-}

When `center = TRUE`, centering is applied in both the adjustment model and
the interaction model, but the two models are centered differently.

**Adjustment model.** Standard `mfp2()` centering is applied: each
FP-transformed column is centered by subtracting its column mean. For binary
covariates, centering uses the lower of the two distinct values rather than
the mean, following the same convention as `mfp2()`.

**Interaction model.** The argument `center_type` controls how the
group-specific FP columns are centered:

- `center_type = "grand"` (default): the grand mean of the FP-transformed
  variable, computed across all observations regardless of group membership, is
  subtracted from each group's columns. All groups share the same centering
  constant.
- `center_type = "group"`: each group's columns are centered by the mean
  computed from in-group observations only. Different groups may use different
  centering constants.

In both cases, centering affects the intercept and, when an interaction is
present, the group dummy coefficients. It does not affect the interaction test,
the treatment-effect function $\hat{f}_1(z) - \hat{f}_0(z)$, or the fitted
values.

## Handling non-positive values

When a continuous predictor contains a mixture of non-positive and positive
values, standard shifting and FP transformation may not be appropriate. The
`mfpi()` function provides three options, matching those in `mfp2()`.

The `zero_vars` option specifies variables for which FP transformations are
applied only to strictly positive values. Non-positive values are set to zero in
the transformed variable, and no additional binary indicator is added.

The `catzero_vars` option extends this approach by adding a binary indicator for
whether the variable is zero. This indicator is included in the adjustment model
alongside the FP terms for the positive part. The `catzero_vars` option implies
`zero_vars`.

The `spike_vars` option triggers the spike-at-zero (SAZ) algorithm. This
evaluates whether the binary zero indicator and the FP function each contribute
explanatory value. The `spike_vars` option implies both `catzero_vars` and
`zero_vars`. For more details about SAZ, see the corresponding vignette in the
`mfp2` package.

Variables listed in `cont_vars` cannot currently use the `catzero_vars` or
`spike_vars` options, because the methodology for combining these options with
interaction testing is not yet available. If a variable appears in both
`cont_vars` and either `catzero_vars` or `spike_vars`, it will be automatically
removed from `catzero_vars` or `spike_vars`, with a warning. However,
`zero_vars` can be used with variables in `cont_vars`: only the positive part
undergoes FP transformation, while the remaining values are set to zero.
Variables listed in `catzero_vars` or `spike_vars` that are not in `cont_vars`
are handled in the adjustment model as described in the `mfp2()` documentation.

## F-test for Gaussian models

When `family = "gaussian"` and the sample size is small, the chi-square
approximation to the likelihood-ratio test may be unreliable. Setting
`use_ftest = TRUE` replaces the chi-square test with an F-test for both the
adjustment-variable selection step and the interaction test. This option is
ignored for non-Gaussian families and when `criterion` is not `"pvalue"`.

## Forcing variables and functional forms

Two arguments provide additional control over the adjustment model. The
`force_keep` argument accepts a character vector of variable names that are
forced into the adjustment model regardless of selection criteria. The
`force_max_fp` argument, when set to `TRUE` for a variable, forces the
algorithm to select the most complex FP functional form at the degree specified
by `df`, bypassing AIC/BIC simplification.

# Worked examples

## Data: Advanced prostate cancer

All survival-model examples use the `advanced_prostate_cancer` dataset included
in the `mfp2` package. The outcome is time to death (overall survival), defined
by `survtime` and the censoring indicator `cens`. The binary variable `rx`
(treatment allocation) is the grouping variable. For a full description of the
data, see Royston and Sauerbrei (2008). The dataset contains 475 observations
with eight continuous measurements (`age`, `wt`, `sbp`, `dbp`, `sz`, `ap`,
`hg`, `sg`) and five categorical covariates (`pf`, `hx`, `bm`, `stage`, `ekg`).

## Prespecified interaction analysis

Suppose that, based on prior evidence, we have identified `age` as the single
covariate to investigate for interaction with treatment. Because only one
interaction is tested, the resulting p-value can be interpreted directly without
multiplicity adjustment.

The formula interface is the most convenient way to call `mfpi()`. Continuous
variables can be wrapped in `fp()` to control FP-specific settings such as
degrees of freedom or candidate power sets. For example, `fp(age, df = 2)`
would restrict `age` to FP1, while `fp(age, select = 1)` would force `age`
into the adjustment model. Here we use the defaults for all variables:

```{r prespecified}
fit_pre <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var  = "rx",
  cont_vars  = "age",
  data       = advanced_prostate_cancer,
  flex       = "flex1",
  family     = "cox",
  select     = 0.05,
  alpha      = 0.05,
  verbose    = FALSE
)
```

### Output: prespecified analysis

Calling `print()` on the fitted object displays a structured three-step summary.

```{r print-pre}
print(fit_pre)
```

**Step 1: Adjustment model (MFP).** This section shows the results of running
the MFP algorithm on the full set of predictors. Variables that remain after
selection at `select = 0.05` are shown with their final FP powers and degrees of
freedom. Variables eliminated by MFP are listed under `"Dropped"`. Because
`include_group_var = TRUE` by default, the grouping variable `rx` is included in
the MFP algorithm but forced into the model: its nominal significance level is
set to 1, so it is never eliminated during backfitting. The retained covariates,
in their selected transformed or untransformed forms, are then used for
adjustment in the interaction models for `age`.

**Step 2: All interaction candidates.** For `age`, three candidate interaction
models are fitted: linear, FP1, and FP2. Each row shows the candidate type, the
deviance of the main-effects model, the deviance of the interaction model, the
likelihood ratio test statistic, the p-value, and the information-criterion
differences. AIC and BIC differences are computed as main-effects minus
interaction, so positive values favour the interaction model. The best
candidate, defined as the one with the smallest p-value, is flagged with `+`.

**Step 3: Interaction summary.** This compact table shows the selected
functional form for `age` and its p-value. If the p-value is below
`p_interact = 0.05`, the variable is marked with `*`, indicating a statistically
significant interaction.

### Regression output {-}

The `summary()` method adds a fourth section with the full regression
coefficients of the winning interaction model.

```{r summary-pre}
summary(fit_pre)
```

**Step 4: Regression output.** For each variable with a significant interaction,
the regression table from the fitted `coxph()` model is displayed. This table
includes coefficient estimates, standard errors, z-values, and p-values for all
terms in the model. The interaction model uses the naming convention
`<varname><group><power_index>` for the group-specific FP terms. For example,
if `age` is modelled as FP1 with groups 0 and 1, the coefficients `age01` and
`age11` represent the FP term for group 0 and group 1, respectively. For FP2,
two terms per group are shown, for example `age01` and `age02` for group 0, and
`age11` and `age12` for group 1.

The group dummy coefficient, for example `rx1`, represents the main effect of
the grouping variable. The p-values in Step 4 are the standard Wald-test
p-values reported by `summary.coxph()`. These p-values do not account for the
degrees of freedom used during FP power selection. Therefore, the appropriate
p-value for testing the interaction is the one reported in Step 3.

### Visualising the interaction

The `plot()` method displays the fitted interaction for continuous variables in
an `mfpi` model. By default, it plots the selected functional form for each
variable retained by the MFPI procedure.

The main argument is `plot_type`, which controls the display:

- `"fitted"` overlays the group-specific fitted functions.
- `"difference"` plots the pointwise treatment-effect difference,
  $\hat f_1(x) - \hat f_0(x)$.
- `"both"` places the fitted and difference plots side by side; this requires
  the `patchwork` package.

Two additional arguments control the graphical output. `show_ci_fitted` adds
95% pointwise confidence bands to the group-specific fitted curves, while
`show_ci_diff` controls the confidence band for the treatment-effect difference
curve. By default, fitted-function confidence bands are suppressed and the
difference-curve confidence band is shown. The argument `show_rug = TRUE` adds
rug marks showing the observed values of the continuous variable.

Each plot includes a title describing the group comparison and a subtitle giving
the variable name, selected functional form, and selection metric. For models
selected by p-value, the subtitle shows the interaction-test p-value. For models
selected by AIC or BIC, it shows `dAIC` or `dBIC`, respectively, where positive
values favour the interaction model. If multiplicity adjustment was used, plots
of selected variables show both the raw and adjusted p-values.

#### Fitted functions

The fitted-function plot overlays the estimated linear predictor for the two
treatment groups as a function of `age`. Confidence bands and rug marks are
added here to show uncertainty and the distribution of the observed age values.
If the two fitted curves are approximately parallel, this suggests little
evidence that the treatment effect varies with age.

```{r plot-fitted-pre, fig.cap = "Group-specific fitted functions for the age × rx interaction."}
plot(fit_pre,
     plot_type = "fitted",
     show_ci_fitted = TRUE,
     show_rug = TRUE)
```

#### Treatment-effect difference

The difference plot shows the pointwise treatment-effect function
$\hat d(\text{age}) = \hat f_1(\text{age}) - \hat f_0(\text{age})$ with a
horizontal reference line at zero. The confidence band is shown by default.
Values above or below zero indicate how the fitted treatment contrast changes
with age. Regions where the confidence band excludes zero suggest age ranges in
which the fitted treatment effect differs from zero.

```{r plot-diff-pre, fig.cap = "Treatment-effect difference for the age × rx interaction."}
plot(fit_pre, plot_type = "difference")
```

#### Side-by-side display

The `"both"` option combines the fitted-function plot and the treatment-effect
difference plot in a single display. This side-by-side view makes it easier to
see how differences between the group-specific fitted functions change across
the range of the continuous variable.

```{r plot-both-pre, fig.width = 10, fig.cap = "Fitted functions and treatment-effect difference for the age × rx interaction."}
plot(fit_pre, plot_type = "both")
```

#### Inspecting alternative candidate forms

By default, `plot()` displays the selected functional form for each retained
variable-specific interaction. The fitted candidate forms for the same variable
can also be inspected by setting `type` explicitly to one of `"linear"`,
`"fp1"`, or `"fp2"`, provided that the requested candidate form was fitted and
stored.

This does not compare different variables, nor does it imply that the displayed
candidate form was non-significant. Rather, it compares alternative functional
forms for the same variable-by-group interaction. For example, the interaction
involving `age` may be selected with a linear or FP1 form, while the FP2
candidate may also show evidence of interaction. In this example, the FP2
candidate has a raw p-value of 0.027, but it was not the selected functional
form.

Inspecting alternative candidate forms can be useful for assessing how sensitive
the fitted interaction is to the chosen functional form, whether different forms
lead to similar substantive conclusions, or whether a more flexible form appears
unstable. Such comparisons should be interpreted using subject-matter knowledge,
ideally with the set of candidate functional forms specified before the analysis.

The following example displays the FP2 candidate fitted functions and
treatment-effect difference for `age`, regardless of whether FP2 was the
selected functional form:

```{r plot-candidate-fp2-pre, fig.width = 10, fig.cap = "FP2 candidate fitted functions and treatment-effect difference for the age × rx interaction."}
plot(fit_pre,
     terms = "age",
     type = "fp2",
     plot_type = "both")
```

When `type` is specified explicitly, the subtitle refers to the displayed
candidate form. If p-value selection with multiplicity adjustment is used, the
subtitle shows the raw candidate p-value, labelled as `p_raw`; the adjusted
p-value is not shown because it applies to the selected winner, not necessarily
to the displayed candidate form. In this example, multiplicity adjustment was
not used, so the reported candidate p-value is unchanged by adjustment:
`p_raw = p_adj = 0.027`.

## Using `flex3`: the recommended practical choice

Royston and Sauerbrei (2014) recommended FP1 with `flex3` as the preferred
practical choice when there is no strong subject-matter reason to expect a
non-monotone treatment-effect function. In `flex3`, the main-effects model and
the interaction model may use different FP power sets, allowing more flexibility
than `flex1` while still constraining the interaction powers to be the same
across groups.

We refit the prespecified analysis for `age` using `flex3`:

```{r flex3-pre}
fit_flex3 <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var  = "rx",
  cont_vars  = "age",
  data       = advanced_prostate_cancer,
  flex       = "flex3",
  family     = "cox",
  select     = 0.05,
  alpha      = 0.05,
  verbose    = FALSE
)
print(fit_flex3)
```

Comparing with the `flex1` results, the adjustment model in Step 1 is the same
because it depends only on MFP selection, not on the flex level. However, the
interaction candidates in Step 2 may differ because `flex3` allows separately
estimated FP powers for the main-effects and interaction models.

```{r flex3-plot, fig.width = 10, fig.cap = "Fitted functions and treatment-effect difference using flex3."}
plot(fit_flex3, plot_type = "both")
```

Note that `flex3` uses a non-nested comparison, so the reported p-values should
be regarded as approximate (see
[Note on non-nested comparisons](#note-on-non-nested-comparisons)).

## Exploratory interaction analysis

Suppose the analyst wants to screen several continuous covariates for possible
interaction with the grouping variable. Because multiple interaction tests are
performed, some small p-values may occur by chance. The interaction p-values may
therefore need to be adjusted for multiplicity.

In this example, we screen seven continuous covariates for possible interaction
with treatment. The model also includes `sg` as an adjustment covariate, but
`sg` is not included in `cont_vars` and is therefore not tested for interaction.
Because multiple interactions are tested, we apply the Holm step-down correction
to control the family-wise error rate across the tested interaction variables.

```{r exploratory}
fit_exp <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var       = "rx",
  cont_vars       = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg"),
  data            = advanced_prostate_cancer,
  flex            = "flex1",
  family          = "cox",
  select          = 0.05,
  alpha           = 0.05,
  p_adjust_method = "holm"
)
```

### Output: exploratory analysis

```{r print-exp}
print(fit_exp)
```

**Step 1** is identical to the prespecified analysis because the same predictors
and the same MFP settings are used.

**Step 2** shows the fitted interaction candidates for each variable in
`cont_vars`. The best candidate form for each variable is marked by `+`.

**Step 3** is the key table for the exploratory analysis. It shows one row per
tested interaction variable, giving the winning functional form, the raw
interaction p-value, and the Holm-adjusted p-value. Variables whose adjusted
p-value falls below `p_interact = 0.05` are marked with `*`.

In an exploratory analysis, the adjusted p-value is the one used for the
selection decision. Variables marked with `*` are candidates for genuine
treatment-by-covariate interaction after accounting for the number of
interaction tests performed. Variables without `*` do not provide sufficient
evidence of interaction after multiplicity adjustment.

### Regression output for selected variables

```{r summary-exp}
summary(fit_exp)
```

Step 4 prints the regression output only for variables whose interaction was
selected (those marked with `*` in Step 3). As before, the correct interaction
test p-values are those from Step 3, not the Wald p-values in the regression
table.

### Plotting selected variables

By default, `plot()` shows results only for variables with a significant
interaction.

```{r plot-exp-fitted, fig.cap = "Fitted functions for variables with significant interactions."}
plot(fit_exp, plot_type = "both")
```

### Plotting non-selected variables

To inspect a candidate fit for a variable that was not selected, specify the
`terms` and `type` arguments. For example, to see the FP1 candidate for `wt`:

```{r plot-nonsig, fig.cap = "FP1 candidate for the wt × rx interaction (not selected)."}
plot(fit_exp, terms = "wt", type = "fp1", plot_type = "both")
```

This is useful for diagnostic inspection: it shows what the candidate functional
form looked like before selection and helps assess whether the non-selection
decision is supported by the data.

## Alternative selection criteria: AIC and BIC {#aic-bic}

In all examples above, the default `criterion = "pvalue"` was used. This
selects the candidate interaction model (linear, FP1, or FP2) with the smallest
p-value, and retains the interaction only if the p-value is below
`p_interact`. For `flex2`, `flex3`, and `flex4`, however, the main-effects and
interaction models are non-nested, and the p-values from the likelihood ratio
test lack a firm theoretical underpinning (see the
[Note on non-nested comparisons](#note-on-non-nested-comparisons) section).

As an alternative, `mfpi()` supports model selection based on information
criteria via `criterion = "aic"` or `criterion = "bic"`. Under these criteria,
models are compared by the difference
$$
\Delta\text{IC}
= \text{IC}_{\text{main}} - \text{IC}_{\text{interaction}},
$$
where IC is AIC or BIC. A positive value means the interaction model has a
smaller information criterion and is therefore preferred. The candidate with the
largest $\Delta\text{IC}$ is selected, and the interaction is retained only if
this difference exceeds `min_improvement` (default 2). Small differences in AIC
or BIC provide little evidence for preferring one model over the other; a
threshold of 2 is a commonly used rough guide.

The `criterion` argument affects two stages of the analysis. In Step 1, it
governs the MFP adjustment model: variable elimination and FP degree selection
are based on the chosen criterion rather than on significance levels. In Step 2,
the same criterion determines which candidate interaction model is preferred
and whether the interaction is retained.

BIC penalizes model complexity more heavily than AIC (the penalty is
$\log(n)$ rather than 2), so it tends to favour simpler functional forms and
is more conservative in declaring interactions.

Note that when `criterion = "aic"` or `"bic"`, the `p_adjust_method` argument
has no effect on the selection decision, because multiplicity adjustment applies
to p-values, not to information criteria.

### Example: BIC-based selection

We repeat the exploratory analysis using BIC as the selection criterion. The
`min_improvement` threshold is set to 2, meaning that the interaction model must
improve BIC by at least 2 units relative to the main-effects model.

```{r bic-exploratory}
fit_bic <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var       = "rx",
  cont_vars       = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg"),
  data            = advanced_prostate_cancer,
  flex            = "flex1",
  family          = "cox",
  criterion       = "bic",
  min_improvement = 2
)
```

### Output: BIC-based selection

```{r print-bic}
print(fit_bic)
```

**Step 1** may differ from the p-value-based analyses. Because the MFP algorithm
now uses BIC for variable selection and FP degree choice, it may retain a
different set of adjustment variables or select different functional forms.

**Step 2** shows the same three candidates (linear, FP1, FP2) per variable, but
the decisive column is now `BIC_main_minus_int` rather than `pvalue`. The `+`
marker flags the candidate with the largest BIC improvement for each variable.
The p-values and AIC differences are still shown for comparison, but BIC
governs the selection.

**Step 3** now displays one row per variable with the column `dBIC` instead of
`pvalue`. Variables whose dBIC exceeds `min_improvement = 2` are marked with
`*`. A positive dBIC means the interaction model is preferred by BIC; a negative
or small positive value means the main-effects model is adequate. Because BIC
penalizes complexity more heavily, it is common for BIC-based selection to retain
fewer interactions than AIC-based selection.

### Regression output and plots

```{r summary-bic}
summary(fit_bic)
```

Step 4 shows the regression output for variables selected under the BIC
criterion. The plot subtitles adapt automatically to show `dBIC` rather than
p-values.

```{r plot-bic-fitted, fig.cap = "Fitted functions for variables selected by BIC."}
plot(fit_bic, plot_type = "both")
```

## Logistic regression example

`mfpi()` is not limited to survival models. The following example demonstrates
its use with logistic regression (`family = "binomial"`). We create a binary
outcome from the prostate cancer data by defining death within 36 months as
the event:

```{r logistic-data}
prostate <- advanced_prostate_cancer
prostate$death36 <- as.integer(prostate$survtime <= 36 & prostate$cens == 1)
```

We test whether the effect of treatment on 36-month mortality depends on `age`,
using `flex1`:

```{r logistic-fit}
fit_logistic <- mfpi(
  death36 ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var  = "rx",
  cont_vars  = "age",
  data       = prostate,
  flex       = "flex1",
  family     = "binomial",
  select     = 0.05,
  alpha      = 0.05,
  verbose    = FALSE
)
print(fit_logistic)
```

The output follows the same three-step structure as the Cox model. The
interaction test and treatment-effect function are now on the log-odds scale.
Plotting works identically:

```{r logistic-plot, fig.width = 10, fig.cap = "Logistic regression: fitted functions and treatment-effect difference for age × rx."}
plot(fit_logistic, plot_type = "both")
```

# Accessing model components

The object returned by `mfpi()` is a list of class `"mfpi"`. Its components
can be accessed with `$` for programmatic post-processing.

```{r components}
names(fit_pre)
```

Key components include:

- `adjustment_model`: the full `mfp2` object from the MFP adjustment step.
- `all_model_metrics`: a data frame of evaluation metrics for all candidates
  (linear, FP1, FP2) for every variable in `cont_vars`.
- `best_model_metrics`: metrics for the selected (winning) interaction model
  per variable.
- `best_interaction_model`: a named list of fitted model objects for selected
  variables.
- `all_interaction_models`: a named list of lists containing all three
  candidate model objects for every variable, regardless of selection.
- `best_fitted_functions` and `all_fitted_functions`: group-specific fitted
  values used by `plot()`.
- `winsorize_limits`: the actual Winsorisation cutoffs applied.
- `group_levels_original` and `group_levels_new`: the original and remapped
  factor levels of `group_var`.
- `var_winners`: a named list of per-variable best candidates, including those
  that did not reach significance.

For example, to extract the metrics table for all candidates:

```{r all-metrics}
fit_pre$all_model_metrics
```

To extract the fitted model object for a selected variable:

```{r extract-model, eval=FALSE}
fit_pre$best_interaction_model[["age"]]
```


```{r restore-options, include=FALSE}
options(width = oldwidth)
```

# References

- Royston, P. and Sauerbrei, W. (2004). A new approach to modelling
  interactions between treatment and continuous covariates in clinical trials by
  using fractional polynomial submodels. *Statistics in Medicine*, 23(16),
  2509--2525. doi:
  [10.1002/sim.1815](https://doi.org/10.1002/sim.1815)

- Sauerbrei, W., Royston, P. and Zapien, K. (2007). Detecting an interaction
  between treatment and a continuous covariate: a comparison of two approaches.
  *Computational Statistics & Data Analysis*, 51(8), 4054--4063. doi:
  [10.1016/j.csda.2006.12.041](https://doi.org/10.1016/j.csda.2006.12.041)

- Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A
  Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for
  Modelling Continuous Variables.* Wiley.

- Royston, P. and Sauerbrei, W. (2013). Interaction of treatment with a
  continuous variable: simulation study of significance level for several methods
  of analysis. *Statistics in Medicine*, 32(22), 3788--3803. doi:
  [10.1002/sim.5813](https://doi.org/10.1002/sim.5813)

- Royston, P. and Sauerbrei, W. (2014). Interaction of treatment with a
  continuous variable: simulation study of power for several methods of analysis.
  *Statistics in Medicine*, 33(27), 4695--4708. doi:
  [10.1002/sim.6245](https://doi.org/10.1002/sim.6245)

```{r cleanup, include=FALSE}
options(width = oldwidth)
```
