---
title: "MFPI: Multivariable Fractional Polynomial Interaction"
author: "Edwin Kipruto and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{MFPI: Multivariable Fractional Polynomial Interaction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-size: 15px;
  line-height: 1.5;
}

/* Justify normal prose, but not code, tables, headings, or captions */
p, li {
  text-align: justify;
  line-height: 1.5;
}

h1, h2, h3, h4 {
  margin-top: 1.4em;
  margin-bottom: 0.6em;
  color: #1f77b4;
}

#TOC a {
  color: inherit;
  text-decoration: none;
}

pre {
  font-size: 13px;
  line-height: 1.4;
  overflow-x: auto;
}

code {
  font-size: 90%;
}

img {
  max-width: 100%;
  height: auto;
}

.figure {
  margin-top: 1.2em;
  margin-bottom: 1.2em;
}

.caption {
  font-size: 90%;
  color: #555555;
  text-align: left;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  dev = "png",
  dpi = 150
)

oldwidth <- getOption("width")
options(width = 100)

library(mfp2)
library(ggplot2)
library(patchwork)
library(survival)
```


# Introduction

Investigating whether the effect of a treatment or other grouping variable
changes across the values of a continuous covariate is a common and important
problem in clinical and epidemiological studies. A common approach is to
categorise the continuous variable using cutpoints, but this discards
information, may reduce power, and can make the evidence for interaction depend
strongly on the number and location of the cutpoints (Royston and Sauerbrei,
2013). Assuming a linear interaction avoids arbitrary cutpoints but may still be
restrictive: the main effect, the interaction function, or both may be
nonlinear. Misspecifying a nonlinear main effect can even generate a spurious
interaction (Royston and Sauerbrei, 2013).

The Multivariable Fractional Polynomial Interaction (MFPI) procedure (Royston
and Sauerbrei, 2004) addresses these issues by modelling covariate effects
using fractional polynomial (FP) functions within levels of a grouping
variable. Four flexibility variants (`flex1` to `flex4`) control how FP powers
are selected and constrained across groups (Royston and Sauerbrei, 2013, 2014).
The procedure first constructs an adjustment model using the MFP algorithm,
then tests whether group-specific functions are parallel on the scale of the
linear predictor.

This vignette describes `mfpi()` in the `mfp2` package. The function supports
Gaussian, binomial, Poisson, and Cox regression models, and provides both
formula and matrix interfaces (see `?mfpi` for the matrix interface). It
investigates interactions between a categorical grouping variable (`group_var`)
and one or more continuous covariates (`cont_vars`). The grouping variable is
treated as a factor internally, with the lowest value used as the reference
group.

# Quick start

We begin with a worked example before describing the methodology in detail. The
`advanced_prostate_cancer` dataset from `mfp2` contains 475 observations, of
which 338 are observed deaths. It includes survival time (`survtime`),
event/censoring status (`cens`), a binary treatment variable (`rx`), eight
continuous covariates, and five categorical covariates.

```{r data}
data("advanced_prostate_cancer", package = "mfp2")
head(advanced_prostate_cancer)
```

The following call tests whether the treatment effect on survival varies across
values of `age`, while adjusting for the other selected predictors:

```{r quickstart}
fit_quick <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var         = "rx",
  cont_vars         = "age",
  data              = advanced_prostate_cancer,
  flex              = "flex1",
  family            = "cox",
  p_interact        = 0.05,
  criterion         = "pvalue",
  verbose           = FALSE
)
print(fit_quick)
```
The output is organised in steps. For p-value-based analyses, Step 1 shows
the adjustment model selected by MFP, Step 2 shows the candidate interaction
models, Step 3 shows the candidate p-value table, and Step 4 shows the final
interaction summary. For AIC- and BIC-based analyses, the output contains the
adjustment model, the candidate interaction models, and the interaction
summary. The `summary()` method additionally prints the regression coefficients
for retained interactions. The `plot()` method displays the group-specific
fitted functions and the fitted-function difference. The rest of this vignette
explains each component in detail.

```{r quickstart-plot, fig.width = 10, fig.cap = "Fitted functions and fitted-function difference for the age × rx interaction."}
plot(fit_quick, plot_type = "both")
```

# Prespecified and exploratory interaction analyses

Royston and Sauerbrei (2008) distinguished between two settings for
investigating interactions:

1. **Prespecified analysis.** This evaluates an interaction defined before
   looking at the data, usually based on prior evidence or the trial protocol.
   Since the analysis focuses on a single interaction, multiplicity adjustment
   across several covariates is not required.

2. **Hypothesis generation.** This screens several covariates for possible
   interaction with the grouping variable. Because multiple interaction tests
   are performed, some small p-values may occur by chance. The interaction
   p-values may therefore need to be adjusted for multiplicity. Any interaction
   identified in such an exploratory analysis should be regarded as a candidate
   finding that requires confirmation in independent data (Royston and Sauerbrei,
   2008).

The `mfpi()` function supports both settings. For a prespecified analysis,
`cont_vars` contains a single continuous covariate. For an exploratory analysis,
`cont_vars` contains several continuous covariates, which are assessed one at a
time. For each variable under test, the adjustment covariates are the remaining
selected variables. Specifically, the continuous variable currently being tested
and the dummy variables corresponding to the grouping variable are excluded from
the adjustment set.

When `criterion = "pvalue"`, multiplicity adjustment is controlled by two
arguments. The argument `p_adjust_method` specifies the adjustment method, such
as `"holm"`, `"bonferroni"`, `"hochberg"`, `"BH"`, or `"BY"`. The default is
`p_adjust_method = "none"`, meaning that no multiplicity correction is applied.
The argument `p_adjust_scope` defines the family of p-values to which that
adjustment is applied.

## P-value adjustment scope

The argument `p_adjust_scope` specifies which p-values are included in the
adjustment set. With the default `p_adjust_scope = "candidates"`, all candidate
tests are adjusted together. This means that the raw p-values from every tested
variable and every fitted functional form are put into one adjustment set.

For example, suppose `age` has raw p-values 0.08, 0.03, and 0.06 for the
linear, FP1, and FP2 candidates, respectively. Suppose also that `wt` has raw
p-values 0.04, 0.20, and 0.11 for the linear, FP1, and FP2 candidates. With
`p_adjust_scope = "candidates"`, the multiplicity adjustment is applied to all
six candidate p-values: 0.08, 0.03, 0.06, 0.04, 0.20, and 0.11.

The adjusted p-values are then placed back into the candidate table. For each
variable, the candidate with the smallest adjusted p-value is selected as the
variable-level candidate. The interaction is retained only if this selected
candidate's adjusted p-value is below `p_interact`. If seven variables are
screened and all three candidate forms are fitted for each variable, the
adjustment is applied to up to 21 p-values. This option gives stronger
protection against false positives because it accounts for searching across both
variables and functional forms.

With `p_adjust_scope = "variables"`, adjustment is done in two steps. First,
within each variable, `mfpi()` chooses the candidate form with the smallest raw
p-value. This gives one p-value per variable. Second, `p_adjust_method` is
applied only to these selected variable-level p-values.

Using the same example, the smallest raw p-value for `age` is 0.03 from the FP1
candidate, and the smallest raw p-value for `wt` is 0.04 from the linear
candidate. Therefore, the adjustment is applied only to these two selected
p-values: 0.03 and 0.04. The non-selected candidate forms are still shown in the candidate table with
their raw p-values, but they do not receive adjusted p-values because they were
not included in the adjustment set. This option is less conservative, but it does not adjust for the internal
functional-form search. It can be ideal when the functional
form is prespecified.


# The MFPI algorithm

The MFPI procedure operates in two broad stages.

First, an adjustment model is selected using the MFP algorithm. The algorithm is
run once on the full predictor matrix to select adjustment variables and their
functional forms. This initial selection also includes variables that may later
be tested for interaction. By default, the grouping variable is excluded from 
the MFP step (`include_group_var = FALSE`). If `include_group_var = TRUE`, the 
grouping variable is represented by dummy variables and forced into the 
adjustment model. When `criterion = "pvalue"`, these dummy variables are assigned
`select = 1`, meaning that they are always retained in the model regardless of 
their statistical significance. When `criterion = "aic"` or `criterion = "bic"`,
they are added to the `keep` set and are therefore not eligible for elimination
during backfitting. The selected adjustment model is then used to control for 
confounding in the subsequent interaction analyses.

Second, for each continuous variable listed in `cont_vars`, `mfpi()` evaluates
whether the relationship between that covariate and the outcome differs across
the levels of `group_var`. For each covariate under investigation, denoted by
$z$, the adjustment set contains the variables selected by the MFP adjustment
model, excluding $z$ itself and the dummy variables representing the grouping
variable. The selected adjustment variables remain in their selected functional
forms. Thus, different variables in `cont_vars` may be evaluated with slightly
different adjustment sets. When `verbose = TRUE`, the adjustment set used for
each variable is printed so that the user can verify which covariates are being
controlled for.

For each variable in `cont_vars`, candidate interaction models are fitted using
linear, FP1, and FP2 functions. The selection criterion, specified by
`criterion`, determines how these candidate models are compared. The available
criteria are `"pvalue"`, `"aic"`, and `"bic"`.

For `criterion = "pvalue"`, all candidate p-values are collected before the
final p-value adjustment and variable-level decisions are made. During verbose
model fitting, the best candidate reported within each variable is therefore
only provisional and is based on the raw p-value. After all variables and
candidate forms have been evaluated, p-values are adjusted according to
`p_adjust_method` and `p_adjust_scope`, and the final selected candidate and
interaction decision are determined.

## The interaction model

For a binary grouping variable, let $t = 0$ denote the reference group and
$t = 1$ denote the comparison group. Let $z$ be the continuous covariate under
investigation. Prior to FP transformation, $z$ is shifted when it contains
negative or zero values to ensure positivity, and optionally scaled for
numerical stability.

If an FP2 function with powers $p_1$ and $p_2$ is selected, the interaction
model for the linear predictor is
$$
\eta_i
=
\mathbf{x}_i^\top \boldsymbol{\gamma}
+ \alpha \, t_i
+ \sum_{j=0}^{1} I(t_i = j)
  \bigl[
    \beta_{j1}\, \phi_1(z_i)
    + \beta_{j2}\, \phi_2(z_i)
  \bigr],
$$

where $\mathbf{x}_i^\top \boldsymbol{\gamma}$ denotes the linear predictor 
contribution from the adjustment covariates, $\alpha$ is the main effect of the
grouping variable, $I(t_i = j)$ is the indicator for group $j$, and 
$\phi_1(z_i) = z_i^{p_1}$, $\phi_2(z_i) = z_i^{p_2}$ are the FP basis functions 
(with $z^0 = \log z$ by convention).


The group-specific contribution of $z$ for an observation in group $j$ is then

$$
\hat{f}_j(z_i)
=
\hat{\beta}_{j1}\, \phi_1(z_i)
+
\hat{\beta}_{j2}\, \phi_2(z_i),
\qquad j = 0, 1.
$$

The interaction test assesses whether the group-specific functions are parallel
on the scale of the linear predictor. For Gaussian models, this is the outcome
scale. For logistic regression, it is the log-odds scale; for Poisson
regression, it is the log-rate or log-mean scale; and for Cox models, it is the
log-hazard scale.

A useful way to display the interaction is the fitted-function difference,

$$
\hat{d}(z_i)
=
\hat{f}_1(z_i) - \hat{f}_0(z_i),
$$

together with a pointwise 95% confidence interval. If the two functions are
parallel, the difference is approximately constant over $z$. If the difference
varies with $z$, this suggests that the effect of the grouping variable on the
outcome depends on the value of the continuous covariate. Pointwise standard
errors for $\hat{d}(z_i)$ and for the group-specific functions $\hat{f}_j(z_i)$
are estimated using the delta method applied to the variance–covariance matrix
of the fitted interaction model.

The fitted-function difference plot should be interpreted as a model-based
visual summary. The confidence interval is pointwise and depends on the selected
model.

# Flexibility levels {#Flex}

The argument `flex` controls how FP powers are selected and constrained across
the levels of the grouping variable. Four flexibility levels are available:
`"flex1"`, `"flex2"`, `"flex3"`, and `"flex4"`. These variants are available
for both FP1 and FP2 functions.

In general, `flex1` is the least flexible and `flex4` is the most flexible
option. Increasing flexibility allows more complex group-specific functional
forms, but it also increases the risk of instability, especially in small
samples or when the continuous covariate has a skewed distribution, influential
observations, or sparse data in parts of its range. The appropriate flexibility
level should therefore be chosen using subject-matter knowledge, the expected
shape of the interaction, and sample-size considerations. For further details,
see Royston and Sauerbrei (2004, 2013, 2014).

## `flex1`

`flex1` is the default and least flexible option. FP powers for the continuous
covariate are estimated once from the pooled main-effects model. This model
includes the FP terms for the continuous covariate and the group indicators, but
no group-by-covariate interaction, while adjusting for the selected covariates
in the adjustment model. The FP powers selected from this pooled main-effects
model are then held fixed and used at each level of the grouping variable. Thus,
the same FP basis functions are used in all groups, and only the regression
coefficients of these basis functions are allowed to differ between groups.

Because the main-effects and interaction models use the same FP basis functions,
the main-effects model is a restricted version of the interaction model. The
interaction test is therefore a standard likelihood-ratio test comparing the two
models, with $(K - 1) \times m$ degrees of freedom, where $K$ is the number of
groups and $m$ is the number of FP terms. Thus, $m = 1$ for FP1 and $m = 2$ for
FP2. For the common case of two groups, this gives 1 df for FP1 and 2 df for
FP2.

`flex1` corresponds to the original MFPI procedure proposed by Royston and
Sauerbrei (2004). In the significance-level simulation study, Royston and
Sauerbrei (2013) found that FP2 with `flex1` maintained significance levels
close to nominal, whereas the other FP2 flexibility variants showed larger
departures from the nominal level. When subject-matter knowledge gives good
reasons to expect a non-monotone treatment-effect function, Royston and
Sauerbrei (2014) recommended FP2 with `flex1`.

## `flex2`

`flex2` estimates FP powers from an interaction model rather than from the
pooled main-effects model used by `flex1`. The powers are optimized over the
levels of the grouping variable, but are constrained to be the same across
groups. These same powers are then also used in the corresponding main-effects
model.

Thus, `flex2` may select powers different from those selected by `flex1`,
because power selection is informed by group-specific effects. However, once the
powers have been selected, the same FP powers are used in both the
main-effects and interaction models. The main-effects model is therefore nested
within the interaction model.

The interaction test uses the same nominal degrees of freedom as `flex1`,
namely $(K - 1) \times m$, where $K$ is the number of groups and $m$ is the
number of FP terms.

In the simulations, FP1 with `flex2` had high power but slightly inflated
significance levels. FP2 with `flex2` showed clearly inflated type I error and
was therefore not recommended as a routine choice.

## `flex3`

`flex3` estimates FP powers separately for the main-effects model and the
interaction model. The main-effects powers are selected from the pooled
main-effects model, as in `flex1`, while the interaction powers are selected
from a model allowing group-by-covariate interaction, as in `flex2`. These
interaction powers are constrained to be the same across groups, so
group-specific functions differ through their regression coefficients rather
than through group-specific FP powers.

Because the main-effects and interaction models are allowed to use different FP
power sets, the two fitted models may be non-nested. If the selected powers
differ, the main-effects model is not simply obtained by imposing equality
constraints on the interaction model, because the two models use different
transformed covariate columns. If the selected powers happen to be the same, the
usual nested-model interpretation applies. In all cases, the likelihood-ratio
statistic is computed and compared with a chi-square distribution using the
nominal degrees of freedom, $(K - 1) \times m$.

In the significance-level simulation study of Royston and Sauerbrei (2013),
FP1 with `flex3` had significance levels close to nominal. In the power study,
FP1 with `flex3` performed well. Royston and Sauerbrei (2014) considered FP1
with `flex1`--`flex3` to have the largest power among methods with acceptable
significance levels and recommended FP1 with `flex3` as the preferred practical
choice when there is no strong subject-matter reason to expect a non-monotone
treatment-effect function, noting that the FP1 class includes the linear
function and tends to yield functions that are simple, understandable, and
transferable. FP2 with `flex3` showed some type I error inflation and was not
recommended as a routine choice.

## `flex4`

`flex4` is the most flexible option. Like `flex3`, it allows the main-effects
and interaction models to use separately selected FP powers. Its distinguishing
feature is that the FP powers in the interaction model are also allowed to differ
across the levels of the grouping variable. Thus, each group may have its own FP
power set and its own regression coefficients.

Because `flex4` allows separate power selection for the main-effects and
interaction models, and may also allow group-specific powers in the interaction
model, the comparison may be non-nested. When the FP basis functions used in the
main-effects and interaction models differ, the main-effects model is not a
simple restriction of the interaction model. If the same basis functions are used
in the fitted main-effects and interaction models, the comparison is closer to
the usual nested-model setting. The interaction test uses
$2(K - 1) \times m$ degrees of freedom, which is twice as many as
`flex1`--`flex3` to account for the additional flexibility from allowing
group-specific powers.

In the simulation studies of Royston and Sauerbrei (2013, 2014), FP2 with
`flex4` was conservative, with empirical significance levels around 2--3%
rather than the nominal 5%, and had low power.

`flex4` may be useful when there is a strong subject-matter reason to believe
that the functional form of the covariate effect differs between groups.
However, this may require sufficiently large samples.

## Summary table

The four MFPI variants are summarized in the table below.

```{r flex-table, echo=FALSE}
flex_table <- data.frame(
  Variant = c("flex1", "flex2", "flex3", "flex4"),
  Description = c(
    "FP powers determined for the main effect of z; use these same powers at each level of t",
    "FP optimized over levels of t; use these same powers at each level of t",
    "As flex2, but re-estimate powers for the main effect of z",
    "Optimize FP for the main effect and separately at each level of t"
  ),
  `d.f. FP1` = c(1, 1, 1, 2),
  `d.f. FP2` = c(2, 2, 2, 4),
  check.names = FALSE
)

knitr::kable(
  flex_table,
  align = c("l", "l", "c", "c"),
  caption = "Variants of MFPI with degrees of freedom for the tests of
  interaction from Royston and Sauerbrei (2013)."
)
```

Our implementation of `mfpi()` evaluates three functional-form candidates for
each variable in `cont_vars`: linear, FP1, and FP2. Each candidate is evaluated
using its corresponding interaction test, and the `criterion` parameter
determines how candidate forms are compared. Candidate tables use `+` to mark
the candidate selected for the variable-level decision. This allows users to
prespecify a preferred degree based on subject-matter reasoning while still
using the other candidates for comparison and sensitivity assessment. For
example, a user following the recommendation of Royston and Sauerbrei (2014)
might focus on FP1 with `flex3` when there is no strong reason to expect a
non-monotone treatment-effect function, while treating the linear and FP2
results as sensitivity checks.

## Note on nested and non-nested comparisons {#note-on-non-nested-comparisons}

Whether the main-effects and interaction models are nested depends on the FP
basis functions used in the two models. If the same FP powers, and therefore the
same transformed covariate columns, are used in both models, the main-effects
model is nested within the interaction model.

If the FP powers differ between the main-effects and interaction models, the two
models no longer use the same covariate basis. In that case, the main-effects
model is not nested within the interaction model. The likelihood-ratio statistic
may still be computed from the fitted deviances and compared with a chi-square
distribution using the nominal degrees of freedom, but the resulting p-values
lack a firm theoretical underpinning. They should therefore be interpreted
cautiously, because the significance levels may be liberal or conservative
(Royston and Sauerbrei, 2013).

This issue is most relevant for flexibility variants that allow separate power
selection for the main-effects and interaction models, such as `flex3` and
`flex4`. However, the nestedness of a particular fitted comparison should be
judged from the selected FP basis functions, not only from the flexibility label.
For example, a `flex3` fit may still give a nested comparison if the selected
main-effects and interaction powers are the same.

As an alternative, AIC or BIC may be used to compare non-nested models, since
these criteria do not require the main-effects model to be nested within the
interaction model. See the
[Alternative selection criteria: AIC and BIC](#aic-bic) section for details.

# Practical considerations

## Influential observations and Winsorisation

FP models can be sensitive to outliers or influential observations in the
continuous predictor. A small number of influential observations in the tails of
a variable in `cont_vars` may strongly affect the selected functional form and
may produce spuriously significant interactions. Therefore, it is important to
inspect the data before fitting an MFPI model, for example, using univariable
plots.

In `mfpi()`, Winsorisation is provided as an optional pre-processing step to
reduce the influence of extreme values in variables listed in `cont_vars`.
Values below the lower percentile and above the upper percentile specified by
`winsorize_probs` are replaced by the corresponding percentile cutoffs. The
default is `winsorize_probs = c(0.01, 0.99)`. Winsorisation preserves the
number of observations because no rows are removed. Only the extreme values are
truncated, while the main part of the distribution remains unchanged.

Note that variables in `cont_vars` cannot be listed in `spike_vars` or
`catzero_vars` --- these options are only available for adjustment variables,
because the interaction design matrix does not include group-specific
$I(x = 0)$ indicators. If a variable appears in both `cont_vars` and
`spike_vars` or `catzero_vars`, the spike or catzero treatment is dropped
with a warning.

Variables in `cont_vars` that are also listed in `zero_vars` are excluded from
Winsorisation, because the zero structure of these variables carries substantive
modelling meaning, and truncating the lower tail could distort it. Excluded
variables are recorded with `NA` limits in `winsorize_limits` for transparency.

Winsorisation is applied after shifting and scaling, so the percentile cutoffs
and the limits stored in `winsorize_limits` are on the transformed scale. The
actual cutoffs applied are returned in the fitted object and can be inspected
via `fit$winsorize_limits`.

To disable Winsorisation, set `winsorize = FALSE`. More aggressive truncation
can be requested, for example:

```{r winsorize-prob, eval=FALSE}
winsorize_probs = c(0.05, 0.95)
```

Any Winsorisation or trimming applied should be reported transparently as part
of the initial data analysis.

## Shifting and scaling {-}

Fractional polynomials require strictly positive input values. As in `mfp2()`,
`mfpi()` estimates shift terms to ensure all values of the continuous covariates
are positive before FP transformation. After shifting, `mfpi()` estimates a
scale factor for each variable, placing the shifted covariates on a convenient
numerical range for FP power selection. This improves numerical stability
during the MFP backfitting cycles. Scaling does not affect which FP powers are
selected, only the numerical precision of the fitting algorithm.

After FP powers have been selected, the variables are backscaled before the
final model is fitted. This means the regression coefficients in the returned
model are on the $\phi(z + \text{shift})$ scale, matching what `mfp2()` would
produce. The user does not need to undo the scaling manually.

## Centering {-}

When `center = TRUE`, centering is applied in both the adjustment model and
the interaction model, but the two models can be centered differently.

**Adjustment model.** Standard `mfp2()` centering is applied: each
FP-transformed column is centered by subtracting its column mean. For binary
covariates, centering uses the lower of the two distinct values rather than
the mean, following the same convention as `mfp2()`.

**Interaction model.** The argument `center_type` controls how the
group-specific FP columns are centered:

- `center_type = "grand"` (default): the grand mean of the FP-transformed
  variable, computed across all observations regardless of group membership, is
  subtracted from each group's columns. All groups share the same centering
  constant.
- `center_type = "group"`: each group's columns are centered by the mean
  computed from in-group observations only. Different groups may use different
  centering constants.

In both cases, centering affects the intercept and, when an interaction is
present, the group dummy coefficients. It does not affect the interaction test,
the fitted-function difference $\hat{f}_1(z) - \hat{f}_0(z)$, or the fitted
values.

## Handling non-positive values

When a continuous predictor contains a mixture of non-positive and positive
values, standard shifting and FP transformation may not be appropriate. The
`mfpi()` function provides three options, matching those in `mfp2()`.

The `zero_vars` option specifies variables for which FP transformations are
applied only to strictly positive values. Non-positive values are set to zero in
the transformed variable, and no additional binary indicator is added.

The `catzero_vars` option extends this approach by adding a binary indicator for
whether the variable is zero. This indicator is included in the adjustment model
alongside the FP terms for the positive part. The `catzero_vars` option implies
`zero_vars`.

The `spike_vars` option triggers the spike-at-zero (SAZ) algorithm. This
evaluates whether the binary zero indicator and the FP function each contribute
explanatory value. The `spike_vars` option implies both `catzero_vars` and
`zero_vars`. For more details about SAZ, see the corresponding vignette in the
`mfp2` package.

Variables listed in `cont_vars` cannot currently use the `catzero_vars` or
`spike_vars` options, because the methodology for combining these options with
interaction testing is not yet available. If a variable appears in both
`cont_vars` and either `catzero_vars` or `spike_vars`, it will be automatically
removed from `catzero_vars` or `spike_vars`, with a warning. However,
`zero_vars` can be used with variables in `cont_vars`: only the positive part
undergoes FP transformation, while the remaining values are set to zero.
Variables listed in `catzero_vars` or `spike_vars` that are not in `cont_vars`
are handled in the adjustment model as described in the `mfp2()` documentation.

## F-test for Gaussian models

When `family = "gaussian"` and the sample size is small, the chi-square
approximation to the likelihood-ratio test may be unreliable. Setting
`ftest = TRUE` replaces the chi-square test with an F-test for both the
adjustment-variable selection step and the interaction test. This option is
ignored for non-Gaussian families and when `criterion` is not `"pvalue"`.

## Forcing variables and functional forms

Two arguments provide additional control over the adjustment model. The `keep`
argument accepts a character vector of variable names that are forced into the
adjustment model regardless of selection criteria. The `force_max_fp_vars`
argument accepts a character vector of variable names for which the algorithm
should select the most complex FP functional form at the degree specified by
`df`. This works consistently across all selection criteria: under
`criterion = "pvalue"`, it sets `alpha = 1` for the named variables so that the
significance test always accepts the most complex form; under
`criterion = "aic"` or `"bic"`, it bypasses information-criterion-based
simplification. In both cases, the best power combination within the requested
degree is still selected by the criterion. In the formula interface,
per-variable control is also available via `fp(force_max_fp = TRUE)`.

# Worked examples

## Data: Advanced prostate cancer

All survival-model examples use the `advanced_prostate_cancer` dataset included
in the `mfp2` package. The outcome is time to death (overall survival), defined
by `survtime` and the censoring indicator `cens`. The binary variable `rx`
(treatment allocation) is the grouping variable. For a full description of the
data, see Royston and Sauerbrei (2008). The dataset contains 475 observations
with eight continuous measurements (`age`, `wt`, `sbp`, `dbp`, `sz`, `ap`,
`hg`, `sg`) and five categorical covariates (`pf`, `hx`, `bm`, `stage`, `ekg`).

## Prespecified interaction analysis

Suppose that, based on prior evidence, we have identified `age` as the single
covariate to investigate for interaction with treatment. Because only one
interaction is tested, the resulting p-value can be interpreted directly without
multiplicity adjustment.

The formula interface is the most convenient way to call `mfpi()`. Continuous
variables can be wrapped in `fp()` to control FP-specific settings, such as the
degrees of freedom or the candidate power set. For example,
`fp(age, df = 2, select = 1)` restricts `age` to an FP1 function and forces it
to remain in the adjustment model. Here we use the defaults for all variables:

```{r prespecified}
fit_pre <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var         = "rx",
  cont_vars         = "age",
  data              = advanced_prostate_cancer,
  flex              = "flex1",
  family            = "cox",
  select            = 0.05,
  alpha             = 0.05,
  verbose           = FALSE
)
```

### Output: prespecified analysis

Calling `print()` on the fitted object displays a structured summary. With
`criterion = "pvalue"`, the output has four steps.

```{r print-pre}
print(fit_pre)
```

**Step 1: Adjustment model (MFP).** This section shows the results of running
the MFP algorithm on the full set of predictors. Variables that remain after
selection at `select = 0.05` are shown with their final FP powers and degrees of
freedom. Variables eliminated by MFP are listed under `"Dropped"`. By default,
the grouping variable `rx` is not included in the MFP adjustment model; it
enters only in Step 2 as the variable defining the interaction. To include `rx`
in the adjustment model as well, set `include_group_var = TRUE`, which forces it
into the model with its nominal significance level set to 1 so that it is never
eliminated during backfitting. The retained covariates, in their selected
transformed or untransformed forms, are then used for adjustment in the
interaction models for `age`.

**Step 2: All interaction candidates.** For `age`, three candidate interaction
models are fitted: linear, FP1, and FP2. Each row shows the candidate type, the
deviance of the main-effects model, the deviance of the interaction model, the
likelihood-ratio statistic, the raw p-value, and the information-criterion
differences. During model fitting, the `+` marker denotes the provisional best
candidate by raw p-value.

**Step 3: Candidate p-value table.** For p-value-based analyses, this table
shows the raw p-values for all candidate functional forms. If
`p_adjust_scope = "candidates"`, adjusted p-values are also shown for all
candidate rows because all candidate tests are included in the adjustment. If
`p_adjust_scope = "variables"`, only raw p-values are shown in the candidate
table; adjusted p-values are shown only for the selected candidate in the final
interaction summary. The `+` marker identifies the selected candidate for each
variable after applying the relevant p-value selection rule.

**Step 4: Interaction summary.** This compact table shows one row per tested
variable. It gives the selected functional form, the raw p-value, and the
adjusted p-value used for the final decision. Variables whose final p-value is
below `p_interact` are marked with `*`.

### Regression output {-}

For p-value-based fits, the `summary()` method adds a fifth section with the
full regression coefficients of retained interaction models. For AIC- and
BIC-based fits, the regression output follows the interaction summary.

```{r summary-pre}
summary(fit_pre)
```

**Regression output.** For each variable with a significant interaction,
the regression table from the fitted `coxph()` model is displayed. This table
includes coefficient estimates, standard errors, z-values, and p-values for all
terms in the model. The interaction model uses the naming convention
`<varname><group><power_index>` for the group-specific FP terms. For example,
if `age` is modelled as FP1 with groups 0 and 1, the coefficients `age01` and
`age11` represent the FP term for group 0 and group 1, respectively. For FP2,
two terms per group are shown, for example `age01` and `age02` for group 0, and
`age11` and `age12` for group 1.

The group dummy coefficient, for example `rx1`, represents the main effect of
the grouping variable. The p-values in the regression output are the standard
Wald-test p-values reported by `summary.coxph()`. These p-values do not account
for the degrees of freedom used during FP power selection. Therefore, the
appropriate p-value for testing the interaction is the one reported in the
interaction summary.

### Visualising the interaction

The `plot()` method displays the fitted interaction for continuous variables in
an `mfpi` model. By default, it plots the selected functional form for each
variable retained by the MFPI procedure.

The main argument is `plot_type`, which controls the display:

- `"fitted"` overlays the group-specific fitted functions.
- `"difference"` plots the pointwise fitted-function difference,
  $\hat f_1(x) - \hat f_0(x)$.
- `"both"` places the fitted and difference plots side by side; this requires
  the `patchwork` package.

Two additional arguments control the graphical output. `show_ci_fitted` adds
95% pointwise confidence bands to the group-specific fitted curves, while
`show_ci_diff` controls the confidence band for the fitted-function difference
curve. By default, fitted-function confidence bands are suppressed and the
difference-curve confidence band is shown. The argument `show_rug = TRUE` adds
rug marks showing the observed values of the continuous variable.

Each plot includes a compact title and subtitle. The fitted-function plot is
titled `"Group-specific fitted functions"`, and the difference plot is titled
`"Difference in fitted functions"`. The subtitle gives the selected or displayed
functional form, the group contrast using the original group labels, and the
relevant selection metric. For example, a p-value subtitle may read
`type = FP1; rx: 1 vs 0; p_raw = 0.031 (p_adj = 0.047)`. For AIC or BIC,
selected plots show the global improvement used for the final decision, such as
`global dAIC`, while explicit candidate plots show the within-form candidate
difference, such as `dAIC`.

The fitted-function plot distinguishes groups using the legend and line
aesthetics. Original group labels are used in the legend and plot annotations,
rather than the internal recoded levels. By default, the legend is placed inside
the fitted-function panel near the top-left. Its position can be changed using
`legend_position`, `legend_inside`, and `legend_justification`.

The `line_types` argument controls the linetypes for the group-specific fitted
curves. If `line_types = NULL`, solid lines are used by default. Users may pass
an unnamed vector matched to the group order, for example
`line_types = c("solid", "dashed")`, or a named vector using the original group
labels, for example
`line_types = c("placebo" = "solid", "standard" = "dashed")`.

#### Fitted functions

The fitted-function plot overlays the estimated linear predictor for the two
treatment groups as a function of `age`. Confidence bands and rug marks are
added here to show uncertainty and the distribution of the observed age values.
If the two fitted curves are approximately parallel, this suggests little
evidence that the treatment effect varies with age.

```{r plot-fitted-pre, fig.cap = "Group-specific fitted functions for the age × rx interaction."}
plot(fit_pre,
     plot_type = "fitted",
     show_ci_fitted = TRUE,
     show_rug = TRUE)
```

#### Fitted-function difference

The difference plot shows the fitted-function difference between the comparison
group and the reference group, using the original group labels. For example, the
y-axis may be labelled `f(rx = 1) - f(rx = 0)` or
`f(treatment = standard) - f(treatment = placebo)`. A horizontal reference line
at zero is added. The confidence band is shown by default. Positive values
indicate that the fitted function is higher in the comparison group than in the
reference group at that covariate value.

```{r plot-diff-pre, fig.cap = "Fitted-function difference for the age × rx interaction."}
plot(fit_pre, plot_type = "difference")
```

#### Side-by-side display

The `"both"` option combines the fitted-function plot and the fitted-function
difference plot in a single display. This side-by-side view makes it easier to
see how differences between the group-specific fitted functions change across
the range of the continuous variable.

```{r plot-both-pre, fig.width = 10, fig.cap = "Fitted functions and fitted-function difference for the age × rx interaction."}
plot(fit_pre, plot_type = "both")
```

#### Inspecting alternative candidate forms

By default, `plot()` displays the selected functional form for each retained
variable-specific interaction. The fitted candidate forms for the same variable
can also be inspected by setting `type` explicitly to one of `"linear"`,
`"fp1"`, or `"fp2"`, provided that the requested candidate form was fitted and
stored.

This does not compare different variables, nor does it imply that the displayed
candidate form was non-significant. Rather, it compares alternative functional
forms for the same variable-by-group interaction. For example, the interaction
involving `age` may be selected with a linear or FP1 form, while the FP2
candidate may also show evidence of interaction. In this example, the FP2
candidate has a raw p-value of 0.027, but it was not the selected functional
form.

Inspecting alternative candidate forms can be useful for assessing how sensitive
the fitted interaction is to the chosen functional form, whether different forms
lead to similar substantive conclusions, or whether a more flexible form appears
unstable. Such comparisons should be interpreted using subject-matter knowledge,
ideally with the set of candidate functional forms specified before the analysis.

The following example displays the FP2 candidate fitted functions and
fitted-function difference for `age`, regardless of whether FP2 was the
selected functional form:

```{r plot-candidate-fp2-pre, fig.width = 10, fig.cap = "FP2 candidate fitted functions and fitted-function difference for the age × rx interaction."}
plot(fit_pre,
     terms = "age",
     type = "fp2",
     plot_type = "both")
```

When `type` is specified explicitly, the subtitle refers to the displayed
candidate form. For `criterion = "pvalue"`, the subtitle shows the raw candidate
p-value. If `p_adjust_scope = "candidates"` and candidate-level adjusted
p-values are available, the candidate adjusted p-value is also shown as
`p_raw = ... (p_adj = ...)`. If `p_adjust_scope = "variables"`, adjusted
p-values are shown only for the selected variable-level candidate in the
interaction summary, because non-selected candidate forms are not part of the
variable-level adjustment set.

## Using `flex3`: the recommended practical choice

Royston and Sauerbrei (2014) recommended FP1 with `flex3` as a practical
default when there is no strong subject-matter reason to expect a non-monotone
treatment-effect function. In `flex3`, the main-effects model and the
interaction model may use different FP power sets. This makes `flex3` more
flexible than `flex1`, while still requiring the same interaction powers to be
used across treatment groups.

We now refit the prespecified analysis for `age` using `flex3`:

```{r flex3-pre}
fit_flex3 <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var = "rx",
  cont_vars = "age",
  data      = advanced_prostate_cancer,
  flex      = "flex3",
  family    = "cox",
  select    = 0.05,
  alpha     = 0.05,
  verbose   = FALSE
)

print(fit_flex3)
```
Compared with the `flex1` analysis, the adjustment model selected in Step 1 is
unchanged because it is determined by the MFP selection step, not by the chosen
flexibility level. The Step 2 interaction candidates may differ because `flex3`
allows the main effects and interaction models to use different FP power sets. 
In this example, the `flex1` and `flex3` analyses lead to the same result.

```{r flex3-plot, fig.width = 10, fig.cap = "Fitted functions and fitted-function difference using flex3."}
plot(fit_flex3, plot_type = "both")
```

Note that `flex3` may involve a non-nested comparison, so the reported p-values
should generally be regarded as approximate. In this example, however, the
main-effects and interaction models use the same FP powers and adjustment set,
so the main-effects model is nested in the interaction model.

## Exploratory interaction analysis

Suppose the analyst wants to screen several continuous covariates for possible
interaction with the grouping variable. Because multiple interaction tests are
performed, some small p-values may occur by chance. The interaction p-values may
therefore need to be adjusted for multiplicity.

In this example, we screen seven continuous covariates for possible interaction
with treatment. The model also includes `sg` as an adjustment covariate, but
`sg` is not included in `cont_vars` and is therefore not tested for interaction.
Since multiple variables and candidate functional forms are screened, we apply
the Holm step-down correction. With the default
`p_adjust_scope = "candidates"`, the adjustment is applied across all fitted
candidate interaction tests, that is, across all tested variables and candidate
functional forms.

```{r exploratory}
fit_exp <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var         = "rx",
  cont_vars         = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg"),
  data              = advanced_prostate_cancer,
  flex              = "flex1",
  family            = "cox",
  select            = 0.05,
  alpha             = 0.05,
  p_adjust_method   = "holm",
  p_adjust_scope    = "candidates"
)
```

### Output: exploratory analysis

```{r print-exp}
print(fit_exp)
```

**Step 1** is identical to the prespecified analysis because the same predictors
and the same MFP settings are used.

**Step 2** shows the fitted interaction candidates for each variable in
`cont_vars`. During fitting, the best candidate form by raw p-value is marked by
`+` only provisionally.

**Step 3** shows the candidate p-value table. With
`p_adjust_scope = "candidates"`, each candidate row has both a raw and an
adjusted p-value. The `+` marker identifies the selected candidate form for each
variable after candidate-level p-value adjustment.

**Step 4** is the final interaction summary. It shows one row per tested
variable, giving the selected functional form, the raw p-value, and the adjusted
p-value used for the final decision. Variables whose adjusted p-value falls
below `p_interact = 0.05` are marked with `*`.

In an exploratory analysis, the adjusted p-value is the one used for the
selection decision. Variables marked with `*` are candidates for genuine
treatment-by-covariate interaction after accounting for the specified p-value
adjustment family. Variables without `*` do not provide sufficient evidence of
interaction after multiplicity adjustment.

### Regression output for selected variables

```{r summary-exp}
summary(fit_exp)
```

For p-value-based fits, Step 5 prints the regression output only for variables
whose interaction was selected, that is, variables marked with `*` in Step 4. As
before, the correct interaction test p-values are those from the interaction
summary, not the Wald p-values in the regression table.

### Plotting selected variables

By default, `plot()` shows results only for variables with a significant
interaction.

```{r plot-exp-fitted, fig.cap = "Fitted functions for variables with significant interactions."}
plot(fit_exp, plot_type = "both")
```

### Plotting non-selected variables

To inspect a candidate fit for a variable that was not selected, specify the
`terms` and `type` arguments. For example, to see the FP1 candidate for `wt`:

```{r plot-nonsig, fig.cap = "FP1 candidate for the wt × rx interaction (not selected)."}
plot(fit_exp, terms = "wt", type = "fp1", plot_type = "both")
```

This is useful for diagnostic inspection: it shows what the candidate functional
form looked like before selection and helps assess whether the non-selection
decision is supported by the data.

## Alternative selection criteria: AIC and BIC {#aic-bic}

In the examples above, the default `criterion = "pvalue"` was used. With this
criterion, `mfpi()` evaluates the linear, FP1, and FP2 interaction models for
each variable under test. Raw candidate p-values are then adjusted according to
`p_adjust_method` and `p_adjust_scope`. For each variable, the selected
candidate is the form with the smallest final p-value under the chosen
adjustment scope. The interaction is retained if that final p-value is below
`p_interact`. For `flex3` and `flex4`, the main-effects and interaction models
may be non-nested, so the likelihood-ratio-test p-values lack a firm theoretical
underpinning; see the
[Note on non-nested comparisons](#note-on-non-nested-comparisons) section.

As an alternative, `mfpi()` supports model selection based on information
criteria via `criterion = "aic"` or `criterion = "bic"`. Under these criteria,
information criteria are used in both stages of the analysis: first to select
the MFP adjustment model, and then to compare the candidate interaction models.

For AIC and BIC, the final decision is based on a model-class comparison. For
each tested variable, `mfpi()` first identifies the best no-interaction model
among the linear, FP1, and FP2 main-effects candidates, and the best interaction
model among the corresponding interaction candidates. The global improvement is

\[
\Delta \mathrm{IC}_{\mathrm{global}}
=
\min_f \mathrm{IC}_{\mathrm{main}, f}
-
\min_f \mathrm{IC}_{\mathrm{interaction}, f},
\]

where \(f\) indexes the candidate functional forms. Interaction is retained if
this global improvement exceeds `min_improvement`.

If interaction is retained, the selected interaction form is the form with the
smallest absolute interaction-model IC. If interaction is not retained, the best
no-interaction form is retained. The within-form differences
`AIC_main_minus_int` and `BIC_main_minus_int` remain useful diagnostics for each
candidate form but are not used as the final AIC/BIC selection rule.

The `criterion` argument therefore affects two stages of the analysis. In Step
1, it controls selection of the MFP adjustment model, including both variable
selection and functional-form selection. In the interaction analysis, it
determines how the candidate interaction forms are compared and whether the
interaction is retained.

BIC usually penalizes model complexity more strongly than AIC. AIC adds a fixed
penalty of 2 for each model parameter, whereas BIC adds a penalty of
$\log(n)$ for each parameter, where $n$ is the sample size. Because
$\log(n)$ is greater than 2 for samples larger than about 7 observations, BIC
often favours simpler models than AIC. In this context, BIC is therefore
less likely than AIC to select more complex FP functions or retain interaction
terms unless they provide a sufficiently large improvement in model fit.

When `criterion = "aic"` or `criterion = "bic"`, the `p_adjust_method` argument
does not affect the selection decision, because multiplicity adjustment applies
to p-values, not to information criteria.

### Example: BIC-based selection

We repeat the exploratory analysis using BIC as the selection criterion. The
`min_improvement` threshold is set to 2, meaning that the best interaction
model class must improve BIC by at least 2 units relative to the best
main-effects model class.

```{r bic-exploratory}
fit_bic <- mfpi(
  Surv(survtime, cens) ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var       = "rx",
  cont_vars       = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg"),
  data            = advanced_prostate_cancer,
  flex            = "flex1",
  family          = "cox",
  criterion       = "bic",
  min_improvement = 2
)
```

### Output: BIC-based selection

```{r print-bic}
print(fit_bic)
```

**Step 1** may differ from the p-value-based analyses. Because the MFP algorithm
now uses BIC for variable selection and FP degree choice, it may retain a
different set of adjustment variables or select different functional forms.

**Step 2** shows the same three candidates (linear, FP1, FP2) per variable. For
BIC selection, the candidate table still reports within-form
`BIC_main_minus_int` values as diagnostics. The final interaction candidate is
chosen using the absolute `BIC_interaction` among interaction candidates, while
the final retain/drop decision is based on the global BIC improvement comparing
the best interaction model class with the best no-interaction model class.

**Step 3** displays one row per variable with the selected interaction form, the
best main-effects form, the best main-effects IC, the best interaction IC, and
the global dBIC. Variables whose global dBIC exceeds `min_improvement = 2` are
marked with `*`. Because BIC penalizes model complexity more heavily than AIC,
it is common for BIC-based selection to retain fewer interactions than
AIC-based selection.

### Regression output and plots

```{r summary-bic}
summary(fit_bic)
```

The regression output shows fitted models for variables selected under the BIC
criterion. Plot subtitles show `global dBIC` for selected BIC plots. When a
specific candidate form is requested with `type = "linear"`, `"fp1"`, or
`"fp2"`, the subtitle shows the within-form candidate `dBIC` for that displayed
form.

```{r plot-bic-fitted, fig.cap = "Fitted functions for variables selected by BIC."}
plot(fit_bic, plot_type = "both")
```

## Logistic regression example

`mfpi()` is not limited to survival models. The following example demonstrates
its use with logistic regression (`family = "binomial"`). We create a binary
outcome from the prostate cancer data by defining death within 36 months as
the event:

```{r logistic-data}
prostate <- advanced_prostate_cancer
prostate$death36 <- as.integer(prostate$survtime <= 36 & prostate$cens == 1)
```

We test whether the effect of treatment on 36-month mortality depends on `age`,
using `flex1`:

```{r logistic-fit}
fit_logistic <- mfpi(
  death36 ~ fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg + rx,
  group_var         = "rx",
  cont_vars         = "age",
  data              = prostate,
  flex              = "flex1",
  family            = "binomial",
  select            = 0.05,
  alpha             = 0.05,
  verbose           = FALSE
)
print(fit_logistic)
```

The output follows the same structure as the Cox model. With
`criterion = "pvalue"`, the printed output includes the candidate p-value table
before the final interaction summary. The interaction test and fitted-function
difference are now on the log-odds scale.
Plotting works identically:

```{r logistic-plot, fig.width = 10, fig.cap = "Logistic regression: fitted functions and fitted-function difference for age × rx."}
plot(fit_logistic, plot_type = "both")
```

# Accessing model components

The object returned by `mfpi()` is a list of class `"mfpi"`. Its components
can be accessed with `$` for programmatic post-processing.

```{r components}
names(fit_pre)
```

Key components include:

- `adjustment_model`: the full `mfp2` object from the MFP adjustment step.
- `all_model_metrics`: a data frame of evaluation metrics for all candidate
  models (`linear`, `fp1`, `fp2`) for every variable in `cont_vars`. For
  p-value-based analyses it includes raw p-values and, when applicable,
  candidate-level adjusted p-values.
- `best_model_metrics`: one row per tested variable, containing the selected
  candidate form and the metric used for the final interaction decision, such as
  `p_adjusted`, `AIC_global_improvement`, or `BIC_global_improvement`.
- `best_interaction_model`: a named list of fitted model objects for selected
  variables.
- `all_interaction_models`: a named list of lists containing all three
  candidate model objects for every variable, regardless of selection.
- `best_fitted_functions` and `all_fitted_functions`: group-specific fitted
  values used by `plot()`.
- `winsorize_limits`: the actual Winsorisation cutoffs applied.
- `group_levels_original` and `group_levels_new`: the original and remapped
  factor levels of `group_var`.
- `p_adjust_scope`: the p-value adjustment scope used for p-value-based
  analyses, either `"candidates"` or `"variables"`.
- `var_winners`: a named list of per-variable best candidates, including those
  that did not reach significance.

For example, to extract the metrics table for all candidates:

```{r all-metrics}
fit_pre$all_model_metrics
```

To extract the fitted model object for a selected variable:

```{r extract-model, eval=FALSE}
fit_pre$best_interaction_model[["age"]]
```


```{r restore-options, include=FALSE}
options(width = oldwidth)
```

# References

- Royston, P. and Sauerbrei, W. (2004). A new approach to modelling
  interactions between treatment and continuous covariates in clinical trials by
  using fractional polynomial submodels. *Statistics in Medicine*, 23(16),
  2509--2525. doi:
  [10.1002/sim.1815](https://doi.org/10.1002/sim.1815)

- Sauerbrei, W., Royston, P. and Zapien, K. (2007). Detecting an interaction
  between treatment and a continuous covariate: a comparison of two approaches.
  *Computational Statistics & Data Analysis*, 51(8), 4054--4063. doi:
  [10.1016/j.csda.2006.12.041](https://doi.org/10.1016/j.csda.2006.12.041)

- Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A
  Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for
  Modelling Continuous Variables.* Wiley.

- Royston, P. and Sauerbrei, W. (2013). Interaction of treatment with a
  continuous variable: simulation study of significance level for several methods
  of analysis. *Statistics in Medicine*, 32(22), 3788--3803. doi:
  [10.1002/sim.5813](https://doi.org/10.1002/sim.5813)

- Royston, P. and Sauerbrei, W. (2014). Interaction of treatment with a
  continuous variable: simulation study of power for several methods of analysis.
  *Statistics in Medicine*, 33(27), 4695--4708. doi:
  [10.1002/sim.6245](https://doi.org/10.1002/sim.6245)

```{r cleanup, include=FALSE}
options(width = oldwidth)
```
