---
title: "MFPI: Multivariable Fractional Polynomial Interaction"
subtitle: 'Package version `r as.character(utils::packageVersion("mfp2"))`'
author: "Edwin Kipruto and Willi Sauerbrei"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
    highlight: tango
vignette: >
  %\VignetteIndexEntry{MFPI: Multivariable Fractional Polynomial Interaction}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
body {
  font-size: 15px;
  line-height: 1.5;
}

/* Justify narrative paragraphs */
p {
  text-align: justify;
  text-justify: inter-word;
}

/* Justify normal prose, but not code, tables, headings, or captions */
p, li {
  text-align: justify;
  line-height: 1.5;
}

h1, h2, h3, h4 {
  margin-top: 1.4em;
  margin-bottom: 0.6em;
  color: #1f77b4;
}

#TOC a {
  color: inherit;
  text-decoration: none;
}

pre {
  font-size: 13px;
  line-height: 1.4;
  overflow-x: auto;
}

code {
  font-size: 90%;
}

img {
  max-width: 100%;
  height: auto;
}

.figure {
  margin-top: 1.2em;
  margin-bottom: 1.2em;
}

.caption {
  font-size: 90%;
  color: #555555;
  text-align: left;
}

.note-box {
  background-color: #fff3cd;
  border-left: 4px solid #ffc107;
  padding: 0.8em 1em;
  margin: 1.2em 0;
  font-size: 95%;
}
</style>
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  out.width = "100%",
  fig.align = "center",
  dev = "png",
  dpi = 96
)

oldwidth <- getOption("width")
options(width = 100)

library(mfp2)
library(ggplot2)
library(patchwork)
library(survival)
```

# Introduction

Investigating whether the effect of a treatment or other grouping variable changes across the values of a continuous covariate is a common and important problem in many areas of science in which empirical data are analysed. Much of our methodological research has focused on clinical and epidemiological studies, which explains the terminology and examples used in this vignette. However, the underlying methodology is based on regression modelling with fractional polynomials and is applicable beyond clinical and epidemiological research.

## Background and motivation

Multivariable fractional polynomial (MFP) models (Sauerbrei and Royston, 1999) extend standard regression by using fractional polynomials (FPs) to represent nonlinear covariate effects. The Multivariable Fractional Polynomial Interaction (MFPI) procedure (Royston and Sauerbrei, 2004) builds on MFP to investigate whether the relationship between a continuous covariate and the outcome differs across the levels of a categorical grouping variable, such as a treatment indicator. In randomised controlled trials (RCTs), MFPI can be used to investigate whether a treatment effect varies across values of a continuous covariate, such as age. In observational studies, it can be used to assess interactions between a continuous and a categorical variable, typically after adjustment for relevant confounders. An extension to interactions of two continuous variables has been proposed (Royston and Sauerbrei, 2008, MFPIgen, chap 7.11) but is not included in this package.

MFPI was developed to address the limitations of categorising a continuous variable using one or more cutpoints and then testing its interaction with the grouping variable (Royston and Sauerbrei, 2004). Such categorisation can introduce multiplicity, reduce statistical power, obscure linear or nonlinear patterns of effect modification, and complicate comparisons across studies. Moreover, selecting cutpoints on the basis of the observed data can further increase the risk of false-positive findings. These concerns are reflected in the Instrument for assessing the Credibility of Effect Modification Analyses (ICEMAN), a structured tool for evaluating claims of effect modification in RCTs and their meta-analyses, which considers analyses that avoid cutpoints and make use of the full spectrum of values to be the most credible (Schandelmaier et al., 2020). Accordingly, continuous variables should preferably be analysed on their original scale, with nonlinear relationships evaluated where appropriate, as recommended in the PROGRESS framework for prognosis research (Hingorani et al., 2013). Assuming a linear interaction avoids arbitrary cutpoints but may still be restrictive: the main effect, the interaction function, or both may be nonlinear (Royston and Sauerbrei, 2013).

## The MFPI procedure in brief

The MFPI procedure addresses these issues by modelling covariate effects using FP functions within levels of a grouping variable. It is a parametric approach that includes the linear interaction model as its simplest special case. Four flexibility variants (`flex1` to `flex4`) control how FP powers are selected and constrained across groups; see the [Flexibility levels](#Flex) section for details (Royston and Sauerbrei, 2013, 2014). When needed, for example in observational studies, the procedure first constructs an adjustment model using the MFP algorithm and then tests whether group-specific FP functions, by default of degree one (FP1), are parallel on the scale of the linear predictor.

## Scope of this vignette

This vignette describes `mfpi()` in the `mfp2` package. The function supports Gaussian, binomial, Poisson, and Cox regression models and provides formula and matrix/data-frame interfaces. It investigates interactions between a categorical grouping variable (`group_var`) and one or more continuous covariates (`cont_vars`). The grouping variable is remapped internally to consecutive integer levels for fitting, but its original labels are preserved for printing, summaries, plotting, prediction metadata, and the group-level FP power tables.

# The MFPI algorithm

The MFPI procedure operates in two broad stages, implemented as three numbered
steps in the `mfpi()` output. First, an adjustment model is selected (Step 1).
Then, for each continuous variable listed in `cont_vars`, an interaction model
is fitted and compared with a main-effects model (Step 2), and the specified
interaction-selection rule is applied (Step 3). The mathematical formulation of
the models compared in Step 2 is given in the
[Interaction model formulation](#the-interaction-model) section; the
[Quick start](#quick-start) section demonstrates the procedure on data.

## Step 1: Select the adjustment model {#step1}

The MFP algorithm is run once on the candidate predictors to select adjustment variables and, for continuous variables, their functional forms. The resulting model provides the pool of adjustment terms used in the variable-specific interaction analyses.

A variable selected in Step 1 may subsequently be evaluated for interaction. In that analysis, its complete Step 1 representation is removed from the corresponding adjustment set so that it is not included twice. The variable is then reintroduced using the functional form specified in `cont_var_forms`. Further details on Step 1, including the role of `include_group_var`, are provided
in [Specifying the adjustment model](#specifying-the-adjustment-model).

## Step 2: Fit and compare the main-effects and interaction models {#step2}

For each variable in `cont_vars`, `mfpi()` constructs a variable-specific adjustment set from the terms selected in Step 1. The complete Step 1 representation of the variable being evaluated is excluded from this set, while the remaining selected adjustment terms retain their Step 1 functional forms. The same adjustment set is included in both models fitted for that variable. Its construction is described in [Construct variable-specific adjustment sets](#variable-specific-adjustment-sets).

The variable being evaluated is then reintroduced using the functional form specified in `cont_var_forms`. The **main-effects model** includes the adjustment terms, the group main effect, and a single function of the continuous variable that is shared across groups. It contains no group-by-variable interaction, so the group-specific functions are parallel on the scale of the linear predictor.

The **interaction model** includes the same adjustment terms and group main effect but represents the continuous variable using group-specific functions. Depending on `flex`, these functions may differ across groups in their coefficients, their selected FP powers, or both. The function selected for the main-effects model may also differ from the functions selected for the interaction model.

Step 2 compares the main-effects and interaction model specifications to assess whether the group-specific functions are parallel on the scale of the linear predictor. Non-parallel functions indicate effect modification. The `flex` argument determines how FP powers are selected and shared between the models and across groups. The `criterion` argument determines how the fitted models are compared. The precise interpretation of the comparison depends on the selected `flex` and `criterion`, as described in [Flexibility levels](#Flex) and [Selection criteria](#selection-criteria).

Candidate interactions are evaluated separately rather than fitted jointly. With `verbose = TRUE`, the adjustment set used for each variable is printed for inspection. For each variable, Step 2 records the selected powers, model dimensions, and relevant comparison statistics, which are described further in [Interpreting fitted models](#interpreting-fitted-models).

## Step 3: Select the interactions {#step3}

With `criterion = "pvalue"`, each tested variable provides one interaction
p-value. When `p_adjust_method` is not `"none"`, the p-values are adjusted
together after all variables have been tested. The final decision is based on
the adjusted p-value, which equals the raw p-value when no adjustment is
requested.

With `criterion = "aic"` or `criterion = "bic"`, selection is based on the
improvement in the corresponding information criterion achieved by the fitted interaction model relative to the fitted main-effects model. Step 3 records one
final `Yes` or `No` decision for each tested variable. `selected = "Yes"`
indicates evidence that the group effect varies with the tested continuous
covariate under the specified threshold. `selected = "No"` indicates insufficient
evidence of such variation; it does not imply that the covariate has no
prognostic association with the outcome. The decision rules are described in
[Selection criteria](#selection-criteria).

# Quick start {#quick-start}

This quick start addresses one clinically relevant question:

> Does the effect of treatment on survival vary continuously with age?

The example deliberately starts with one treatment variable and one continuous effect modifier, and no adjustment variables. One figure is shown which displays both the
group-specific functions and their pointwise difference.
Multivariable adjustment is introduced in the next section.

## Load and prepare the data

The `advanced_prostate_cancer` dataset in `mfp2` contains 475 observations, of
which 338 are observed deaths. It includes survival time (`survtime`),
event/censoring status (`cens`), a treatment variable (`rx`), eight continuous
covariates (`age`, `wt`, `sbp`, `dbp`, `sz`, `ap`, `hg`, `sg`), and five
categorical covariates (`pf`, `hx`, `bm`, `stage`, `ekg`). This dataset was also
used in Royston and Sauerbrei (2008), where further details are provided.

The original 0/1 treatment codes are converted to descriptive factor levels.
The first factor level, `Placebo`, is the reference group.

```{r data}
data("advanced_prostate_cancer", package = "mfp2")

advanced_prostate_cancer$rx <- factor(
  advanced_prostate_cancer$rx,
  levels = c(0, 1),
  labels = c("Placebo", "Treatment")
)

head(advanced_prostate_cancer)
```

## Fit a Cox MFPI model

The following model investigates whether the treatment effect on survival
varies with age. It uses an FP2 function for the interaction, allowing a more flexible nonlinear
age-by-treatment relationship, with the most restrictive MFPI flexibility level (`flex1`).

```{r quickstart}
fit_quick <- mfpi(
  Surv(survtime, cens) ~ rx + fp(age),
  data = advanced_prostate_cancer,
  family = "cox",
  group_var = "rx",
  cont_vars = "age",
  cont_var_forms = c(age = "fp2"),
  flex = "flex1",
  verbose = FALSE
)
```

The essential arguments are:

- `Surv(survtime, cens)` defines the right-censored survival outcome, where
  `cens = 1` denotes an observed death and `cens = 0` denotes censoring.
- `group_var = "rx"` identifies the groups to be compared.
- `cont_vars = "age"` identifies age as the continuous effect modifier to be
  investigated.
- `fp(age)` provides the variable-specific FP and preprocessing specification
  for age.
- `cont_var_forms = c(age = "fp2")` specifies an FP2 age-by-treatment
  interaction, represented by two fractional-polynomial basis functions.
- `flex = "flex1"` selects the FP2 power pair from the pooled model without
  an interaction and uses that same pair in both treatment groups.

> **Important:** `fp(age)` and `cont_var_forms` have different roles.
> `fp(age)` controls the FP and preprocessing settings available for age,
> whereas `cont_var_forms` determines whether the age-by-treatment interaction
> is fitted as linear, FP1, or FP2.

The call omits `p_interact`, `select`, `alpha`, and `criterion` because their
default values are used. These controls are introduced later, after the basic
workflow is established.

## Inspect the interaction result

```{r quickstart-print}
print(fit_quick)
```

The printed output follows the three algorithmic steps described above. Step 1
reports the selected adjustment model. Step 2 reports the model comparison for
`age`, including the selected powers and the interaction statistic. Step 3
reports the final interaction decision.

This minimal model contains no additional prognostic variables, so the
age-specific comparison is unadjusted for other prognostic factors. The
interaction is fitted as FP2, as specified by `cont_var_forms`, using the
power-selection constraints imposed by `flex1`.

Under the default `criterion = "pvalue"`, `p_interact = 0.05`, and
`p_adjust_method = "none"`, the interaction is selected when its p-value is
below 0.05. A non-selected interaction indicates insufficient evidence that the
treatment effect varies with age; it does not imply that age has no prognostic
association with survival.

## Plot the fitted functions and group difference

```{r quickstart-plot, fig.width = 10, fig.cap = "Group-specific partial fitted treatment functions and the treatment-minus-placebo difference across age on the Cox model's log-hazard scale."}
plot(
  fit_quick,
  terms = "age",
  plot_type = "both"
)
```

The left panel displays the partial fitted functions for `Placebo` and `Treatment`, showing how the log hazard varies with age in each treatment group. The right panel displays the estimated treatment effect on the log-hazard scale. A horizontal difference curve indicates that the estimated treatment effect is constant across age, whereas a changing curve indicates potential treatment-effect modification. The pointwise confidence bands do not account for uncertainty arising from variable, functional form, or interaction selection.

Because `plot_type = "both"` combines the two panels, it requires the `patchwork` package.

# Practical considerations

Before fitting an MFPI model, the analyst should consider the choice of significance level and any adjustment for multiple testing, the robustness of estimated functions to influential observations, sample size requirements, and, in observational studies, the strategy for model adjustment. This section discusses each of these issues. Arguments and settings specific to `mfpi()` are described in the subsequent methodology sections.

## Significance levels and multiplicity

In an RCT there may be a single prespecified variable of interest, or a dozen variables may be screened for interactions. The analyst must decide about the significance level, whether adjustment for multiple testing is needed, and how to proceed if more than one variable is significant. The `mfpi()` function provides both raw and adjusted p-values (controlled by the `p_adjust_method` argument; see the [Prespecified and exploratory interaction analyses](#prespecified-and-exploratory-interaction-analyses) section).

## Robustness and influential observations

Individual observations can strongly influence the selection of the best-fitting FP function (Sauerbrei et al., 2022). When an MFPI analysis indicates an interaction, Royston and Sauerbrei (2004, Section 2.6) recommend checking whether the estimated interaction pattern is robust and not an artefact of flexible modelling. In their survival-analysis examples, they used Kaplan–Meier curves and estimated treatment effects in several subgroups defined by the interacting covariate. The subgroup analysis was used as a diagnostic check rather than as the primary method for detecting an interaction, with subgroup boundaries guided by the estimated treatment-effect function and the distribution of the covariate. Kaplan–Meier curves are applicable only to time-to-event outcomes and are not currently produced by `mfpi()`; for other outcome types, corresponding outcome-specific graphical checks and estimates within covariate-defined subgroups should be considered. We strongly recommend reading this section before interpreting MFPI results for the first time. See also the [Influential observations and Winsorisation](#influential-observations-and-winsorisation) section for the Winsorisation option available in `mfpi()`.

## Sample size considerations

MFPI should be used with caution when the effective sample size is small. Although no formal investigation of the required sample size has been published, the effective sample size in each group should be large enough to allow reliable estimation of group-specific FP functions. In practice, higher [flexibility levels](#Flex) and FP2 functions require more observations than lower flexibility levels and FP1 functions
because they estimate more parameters.

## Observational studies

In observational studies, model adjustment plays a key role, and there are many options for building a multivariable model that includes interactions. The research aim should guide the modelling strategy, but there is no consensus on the optimal procedure. An analyst may choose to derive an MFP adjustment model and then evaluate all continuous-by-categorical interactions, or may prefer to focus on a single pair of variables, with or without adjustment for other covariates. The `mfpi()` function supports both strategies. Practical specification is described in the [Specifying the adjustment model](#specifying-the-adjustment-model) and [Prespecified and exploratory interaction analyses](#prespecified-and-exploratory-interaction-analyses) sections.

## Missing data

`mfpi()` does not automatically perform complete-case analysis. Missing values in variables used by the model are not silently omitted; instead, the function stops with an error. In the formula interface, the model frame is constructed using `na.fail`, while the default matrix interface explicitly rejects predictor data containing `NA` values. Analysts should therefore examine missing-data patterns and remove or appropriately impute missing values before fitting an MFPI model.


## Interpreting confidence intervals {#ci-caveat}

All pointwise confidence intervals and confidence bands reported by
`predict.mfpi()` and displayed by `plot.mfpi()` are conditional on the
selected model. They do not account for uncertainty arising from variable
selection in Step 1, functional-form selection (FP power choice), or
interaction selection in Step 3. As a result, the intervals may understate the
true uncertainty. 

# Specifying the adjustment model {#specifying-the-adjustment-model}

The Quick Start used a minimal model to isolate the interaction workflow. In
most applications, the interaction should be evaluated while controlling for
other prognostic or confounding variables. This section is describes how 
`mfpi()` specifies, selects, and applies those adjustment terms.

## Fit an MFPI model with adjustment variables

The following model asks the same age-by-treatment question while allowing MFP
to select adjustment terms from the remaining continuous and categorical
predictors.

```{r adjusted-fit}
fit_adjusted <- mfpi(
  Surv(survtime, cens) ~
    rx + fp(age) +
    fp(wt) + fp(sbp) + fp(dbp) + fp(sz) +
    fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg,
  data = advanced_prostate_cancer,
  family = "cox",
  group_var = "rx",
  cont_vars = "age",
  cont_var_forms = c(age = "fp2"),
  flex = "flex1",
  verbose = FALSE
)
```

## Variable roles in the MFPI analysis

A variable’s role depends on where it is specified. The model formula defines the variables available for adjustment, while `group_var`, `cont_vars`, and `cont_var_forms` identify the variables involved in the interaction analysis.

| Where the variable is specified | What this means                                                                               | Example in `fit_adjusted`                                                   |
| ------------------------------- | --------------------------------------------------------------------------------------------- | --------------------------------------------------------------------------- |
| Model formula                   | The variable is considered as a candidate predictor during Step 1 adjustment-model selection. | `age`, `wt`, `sbp`, `pf`, `stage`, and the other formula variables          |
| `group_var`                     | The variable defines the groups whose effects are compared.                                   | `rx`                                                                        |
| `cont_vars`                     | The continuous variable is tested for an interaction with the grouping variable in Step 2.    | `age`                                                                       |
| `cont_var_forms`                | Specifies the functional form used for the interaction test.                                  | `age` is tested using an FP2 interaction                                    |
| Formula, but not `cont_vars`    | The variable can be selected as an adjustment term, but it is not tested for interaction.     | `wt`, `sbp`, `dbp`, `pf`, `hx`, `stage`, and the other remaining predictors |

In `fit_adjusted`, `age` has two roles: it is considered during Step 1 adjustment-model selection and is then tested for interaction with `rx` in Step 2. All other predictors in the formula are adjustment-only candidates. Because `age` is the only variable listed in `cont_vars`, it is the only continuous variable tested for interaction.


## Adjustment-model selection in Step 1

Step 1 fits a single MFP adjustment model to the candidate predictor set. Within this fit, MFP may remove candidate terms that do not meet the variable-selection criterion. Under `p-value` selection, a term is removed when its contribution is not significant at the specified `select` level, unless it is forced into the model. The fitted model is stored in `adjustment_model`. In Step 2, its selected terms and fitted transformations are used to construct the adjustment set for each interaction test, except that the Step 1 representation of the continuous variable currently being tested is excluded from its own adjustment set.

The Step 1 form of a continuous variable describes how that variable is used
when it acts as an adjustment variable. It does not determine the variable's
interaction form. For example, the Step 1 representation of `age` is relevant
when another continuous variable is tested, whereas the age interaction in
`fit_adjusted` is FP2 because `cont_var_forms = c(age = "fp2")`.

Ordinary categorical predictors can be included directly in the formula. Their
encoded model-matrix columns enter Step 1 as adjustment terms and are carried
forward when retained.

## Construct variable-specific adjustment sets {#variable-specific-adjustment-sets}

For each continuous variable (z) tested for interaction, `mfpi()` constructs a separate Step 2 adjustment set:

1. Begin with the terms retained by the Step 1 MFP model.
2. Exclude the entire Step 1 term block associated with (z), including all columns used to represent its selected functional form.
3. Exclude the model-matrix columns for the grouping variable, because the group main effect is added separately in Step 2.
4. Retain all other selected adjustment terms in the functional forms chosen during Step 1.
5. Add (z) back using the interaction form specified in `cont_var_forms`, with power selection governed by `flex`.
6. Include the group main effect in both models used for the interaction comparison.

This means that each variable in `cont_vars` may be tested using a different effective adjustment set. When `verbose = TRUE`, `mfpi()` prints the adjustment set used for each tested variable.


## The role of the grouping variable

The group main effect is included in both models used for every Step 2 interaction comparison, regardless of the value of `include_group_var`.

By default, the grouping variable is excluded from the Step 1 adjustment model. When `include_group_var = TRUE`, a grouping variable with (K) observed levels is represented by (K-1) dummy variables. These terms are forced into the Step 1 model and cannot be removed during adjustment-model selection. Consequently, selection of the other adjustment terms is performed conditional on group.

When the grouping variable has many levels, or when some levels contain few observations or events, estimating the additional group effects may increase coefficient variance or make convergence less reliable. Therefore, `include_group_var = TRUE` should be used when adequately supported by the available data. 

## Prespecify the Step 1 adjustment strategy

In an RCT, investigators may wish to limit the extent to which the Step 1 adjustment model is determined by the observed data. This can be done by prespecifying the variables available to the model, requiring selected terms to remain, and constraining the functional forms of continuous covariates.

All variables used by `mfpi()` must appear in the model formula. In particular, the variable named by `group_var` and every variable listed in `cont_vars` must be included. The formula should therefore contain:

* the grouping variable;
* all continuous variables to be tested for interaction; and
* any additional variables considered for adjustment.

Variables omitted from the formula are not evaluated by `mfpi()`.

Within this formula-defined set, the Step 1 adjustment strategy can be constrained in several ways:

| Objective                                                     | Specification                                                                      | Important qualification                                                                                            |
| ------------------------------------------------------------- | ---------------------------------------------------------------------------------- | ------------------------------------------------------------------------------------------------------------------ |
| Require a term to remain in Step 1                            | Include its name in `keep`                                                         | For a factor, all associated contrast columns are retained together                                                |
| Require a continuous term to be linear                        | Set `df = 1` for that term and include it in `keep`                                | In the formula interface, use `fp(variable, df = 1)` for a variable-specific constraint                            |
| Require inclusion but allow MFP to choose the functional form | Include the variable in `keep` and permit `df > 1`                                 | The variable cannot be eliminated, but its FP degree and powers may still be selected                              |
| Limit the maximum FP degree                                   | Set `df` globally or within `fp()`                                                 | This restricts the search but does not require the maximum degree                                                  |
| Require the maximum permitted FP degree                       | Use `force_max_fp_vars` or `fp(..., force_max_fp = TRUE)`                          | The degree is fixed, but the best powers within that degree are still selected                                     |
| Restrict the candidate powers                                 | Supply `powers` globally or within `fp()`                                          | For a tested variable, the candidate powers may also affect its Step 2 interaction models                          |
| Prespecify essential terms but allow additional selection     | Put the essential terms in `keep` and include additional candidates in the formula | MFP may eliminate only candidates that are not retained through `keep`                                             |
| Fit Step 1 conditional on treatment group                     | Set `include_group_var = TRUE`                                                     | This forces the group indicators into Step 1; the group main effect is already included in every Step 2 comparison |

A fully prespecified linear Step 1 model can be obtained by retaining every intended adjustment term and restricting each continuous term to one degree of freedom:

```r
fit_prespecified <- mfpi(
  outcome ~
    treatment +
    fp(modifier, df = 1) +
    fp(baseline_score, df = 1) +
    sex + centre,
  data = trial_data,
  family = "gaussian",
  group_var = "treatment",
  cont_vars = "modifier",
  cont_var_forms = c(modifier = "linear"),
  keep = c("modifier", "baseline_score", "sex", "centre"),
  verbose = FALSE
)
```

In this example, Step 1 cannot eliminate the listed terms or replace the continuous terms with nonlinear FP functions. The Step 1 variable set and functional forms are therefore fixed. The usual Step 2 construction still applies: when `modifier` is tested, its Step 1 term is removed from the adjustment set and reintroduced using the interaction form specified in `cont_var_forms`.

Partial prespecification is also possible. For example, investigators may retain clinically essential covariates while allowing MFP to select additional prognostic covariates or their functional forms.

For a prespecified nonlinear adjustment function that cannot be represented directly through the FP controls, the transformation can be created before fitting and entered as a retained linear term. For example, a logarithmic adjustment term could be created as `log_baseline = log(baseline_score)` and included with `df = 1` and `keep`. This approach is most straightforward for adjustment-only variables.

The `select` and `alpha` settings require criterion-specific interpretation. Under `criterion = "pvalue"`, `select = 1` prevents elimination and `alpha = 1` requires the maximum permitted FP degree. Under AIC or BIC selection, these thresholds do not serve the same roles; `keep` should be used to require inclusion, and `force_max_fp_vars` should be used when the maximum FP degree must be retained.

`mfpi()` does not accept a separately fitted adjustment model as input. A prespecified Step 1 model must therefore be represented through the formula and the available inclusion and functional-form constraints.


## Inspect the adjusted analysis

```{r adjusted-print}
print(fit_adjusted)
```

Compared with the minimal analysis, Step 1 can now retain several adjustment
terms. Step 2 still reports one age-specific comparison, and Step 3 applies the
same interaction-selection rule. Further details on `summary()` and `plot()`
are given in [Interpreting fitted models](#interpreting-fitted-models).

Setting `verbose = TRUE` prints the adjustment set used for each tested
variable, which is helpful for confirming that the correct terms are included:

```{r adjusted-verbose}
fit_verbose <- mfpi(
  Surv(survtime, cens) ~
    rx + fp(age) +
    fp(wt) + fp(sbp) + fp(dbp) + fp(sz) +
    fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg,
  data = advanced_prostate_cancer,
  family = "cox",
  group_var = "rx",
  cont_vars = "age",
  cont_var_forms = c(age = "fp2"),
  flex = "flex1",
  verbose = TRUE
)
```

## Exploratory interaction analysis {#exploratory-interaction-analyses}

An exploratory analysis screens several continuous variables for possible
interactions with the grouping variable. The same variable-specific procedure
is applied to every entry in `cont_vars`; adjustment-set construction is
described in [Specifying the adjustment model](#specifying-the-adjustment-model).
Exploratory findings should be treated as candidates for confirmation in
independent data (Royston and Sauerbrei, 2008).

The following extension screens seven continuous covariates. `sg` remains in
the formula as an adjustment candidate but is omitted from `cont_vars`, so it is
not tested for interaction. `age` is prespecified as FP2 and the remaining
tested variables use the default FP1 form.

```{r exploratory-fit}
fit_exp <- mfpi(
  Surv(survtime, cens) ~
    rx + fp(age) + fp(wt) + fp(sbp) + fp(dbp) +
    fp(sz) + fp(ap) + fp(hg) + fp(sg) +
    pf + hx + bm + stage + ekg,
  group_var = "rx",
  cont_vars = c("age", "wt", "sbp", "dbp", "sz", "ap", "hg"),
  cont_var_forms = c(age = "fp2"),
  flex = "flex1",
  family = "cox",
  select = 0.05,
  alpha = 0.05,
  p_adjust_method = "holm",
  data = advanced_prostate_cancer,
  verbose = FALSE
)
```

```{r exploratory-print}
print(fit_exp)
```

Step 2 now contains one comparison per variable in `cont_vars`. Step 3 reports
both the raw and Holm-adjusted p-values, with the adjusted values used for the
final decisions.

## Multiplicity adjustment

With `criterion = "pvalue"`, `p_adjust_method` accepts the methods supported by
`stats::p.adjust()`, including `"holm"`, `"bonferroni"`, `"hochberg"`,
`"BH"`, and `"BY"`. The default, `"none"`, leaves the p-values unchanged.
When several variables are screened, adjustment is applied across all tested
variables, with one p-value per variable. The significance level and multiplicity adjustment method should be specified in advance. 

## Prespecifying interaction forms with `cont_var_forms`

The interaction form—linear, FP1, or FP2—should be prespecified rather than
selected by comparing all three forms on the same data. Data-driven selection
among increasingly flexible forms inflates the Type I error and introduces
selection bias.

`cont_var_forms` accepts a named character vector whose names are variables in
`cont_vars` and whose values are `"linear"`, `"fp1"`, or `"fp2"`. Variables
not named in the vector use `"fp1"`; `NULL` therefore assigns FP1 to every
tested variable.

For example, the following specification tests `age` as FP2 and leaves `wt` at
the FP1 default:

```{r cont-var-forms-example, eval=FALSE}
mfpi(
  ...,
  cont_vars = c("age", "wt"),
  cont_var_forms = c(age = "fp2")
)
```

Every supplied entry must be named. Unnamed scalars such as
`cont_var_forms = "fp2"` are not accepted because the target variable would be
ambiguous. The form should be chosen from subject-matter knowledge or prior
evidence, ideally before inspecting the data.

# Flexibility levels {#Flex}

## Overview

The argument `flex` controls how FP powers are selected and constrained across
the levels of the grouping variable. Four flexibility levels are available:
`"flex1"`, `"flex2"`, `"flex3"`, and `"flex4"`. These variants are available
for both FP1 and FP2 functions.

In general, `flex1` is the least flexible and `flex4` is the most flexible
option. The package default is `flex3`, following the recommendation of Royston
and Sauerbrei (2014) for FP1 interactions. Increasing flexibility allows more
complex group-specific functional forms, but it also increases the risk of
instability, especially in small samples or when the continuous covariate has a
skewed distribution, influential observations, or sparse data in parts of its
range. The appropriate flexibility level should therefore be chosen using
subject-matter knowledge, the expected shape of the interaction, and
sample-size considerations. For further details, see Royston and Sauerbrei
(2004, 2013, 2014).

## `flex1`

`flex1` is the least flexible option. FP powers for the continuous
covariate are estimated once from the pooled main-effects model. This model
includes the FP terms for the continuous covariate and the group indicators, but
no group-by-covariate interaction, while adjusting for the selected covariates
in the adjustment model. The FP powers selected from this pooled main-effects
model are then held fixed and used at each level of the grouping variable. Thus,
the same FP basis functions are used in all groups, and only the regression
coefficients of these basis functions are allowed to differ between groups.

Because the main-effects and interaction models use the same FP basis functions,
the main-effects model is nested within the interaction model. The
interaction test is therefore a standard likelihood-ratio test comparing the two
models, with $(K - 1) \times m$ degrees of freedom, where $K$ is the number of
groups and $m$ is the number of FP terms. Thus, $m = 1$ for FP1 and $m = 2$ for
FP2. For the common case of two groups, this gives 1 df for FP1 and 2 df for
FP2.

`flex1` corresponds to the original MFPI procedure proposed by Royston and
Sauerbrei (2004). In the significance-level simulation study, Royston and
Sauerbrei (2013) found that FP2 with `flex1` maintained significance levels
close to nominal, whereas the other FP2 flexibility variants showed larger
departures from the nominal level. When subject-matter knowledge gives good
reasons to expect a non-monotone treatment-effect function, Royston and
Sauerbrei (2014) recommended FP2 with `flex1`.

## `flex2`

`flex2` estimates FP powers from an interaction model rather than from the
pooled main-effects model used by `flex1`. The powers are optimized over the
levels of the grouping variable, but are constrained to be the same across
groups. These same powers are then also used in the corresponding main-effects
model.

Thus, `flex2` may select powers different from those selected by `flex1`,
because power selection is informed by group-specific effects. However, once the
powers have been selected, the same FP powers are used in both the
main-effects and interaction models. The main-effects model is therefore nested
within the interaction model. The interaction test uses the same nominal degrees
of freedom as `flex1`, namely $(K - 1) \times m$.

In the simulation study by Royston and Sauerbrei (2013, 2014), FP1 with `flex2`
had high power but slightly inflated significance levels. FP2 with `flex2` 
showed clearly inflated type I error and was therefore not recommended as a 
routine choice.

## `flex3`

`flex3` is the package default. It estimates FP powers separately for the
main-effects and interaction models. The main-effects powers are selected from
the pooled main-effects model,
as in `flex1`, whereas the interaction powers are selected from a model allowing
group-by-covariate interaction, as in `flex2`. These interaction powers are
constrained to be the same across groups, so group-specific functions differ
through their regression coefficients rather than through group-specific FP
powers.

Because the main-effects and interaction models may use different FP power sets,
the two fitted models may be non-nested. If the selected powers differ, the
main-effects and interaction models are based on different transformations of
the continuous covariate. In that case, the main-effects model is not nested
within the interaction model, and cannot be obtained simply by setting the interaction regression coefficients to zero. If the selected powers are the same, the usual
nested-model interpretation applies. In all cases, the likelihood-ratio
statistic is computed and compared with a chi-square distribution using the
nominal degrees of freedom, $(K - 1) \times m$.

```{=html}
<div class="note-box">
<strong>Note:</strong> Because <code>flex3</code> is the package default, users
should be aware that the main-effects and interaction models may be non-nested
in some applications. When the selected FP powers differ between the two
models, the resulting p-values lack a firm theoretical underpinning and should
be interpreted cautiously. As an alternative, AIC or BIC may be used because
they do not require nested models; see
<a href="#aic-bic">AIC- and BIC-based selection</a> and
<a href="#note-on-non-nested-comparisons">Note on nested and non-nested comparisons</a>.
</div>
```

In the significance-level simulation study of Royston and Sauerbrei (2013), FP1
with `flex3` had significance levels close to nominal. In the power study, FP1
with `flex3` performed well. Royston and Sauerbrei (2014) considered FP1 with
`flex1`--`flex3` to have the largest power among methods with acceptable
significance levels and recommended FP1 with `flex3` as the preferred practical
choice when there is no strong subject-matter reason to expect a non-monotone
treatment-effect function. They noted that the FP1 class includes the linear
function and tends to yield functions that are simple, understandable, and transferable. 

## `flex4`

`flex4` is the most flexible option. Like `flex3`, it allows the main-effects
and interaction models to use separately selected FP powers. Its distinguishing
feature is that the FP powers in the interaction model are also allowed to differ
across the levels of the grouping variable. Thus, each group may have its own FP
power set and its own regression coefficients.

Because `flex4` allows separate power selection for the main-effects and
interaction models, and may also allow group-specific powers in the interaction
model, the comparison may be non-nested. When the FP basis functions used in the
main-effects and interaction models differ, the main-effects model is not a
simple restriction of the interaction model. If the same basis functions are estimated
in the fitted main-effects and interaction models, the comparison is the usual 
nested-model setting. The interaction test uses $2(K - 1) \times m$ degrees of 
freedom, which is twice the degrees of freedom used by `flex1`--`flex3`. The
doubling accounts for the additional parameters introduced by allowing
group-specific FP powers in the interaction model.

In the simulation studies of Royston and Sauerbrei (2013, 2014), FP2 with
`flex4` was conservative, with empirical significance levels around 2--3%
rather than the nominal 5%, but had low power.

`flex4` may be useful when there is a strong subject-matter reason to believe
that the functional form of the covariate effect differs between groups.
However, this may require sufficiently large samples to detect interaction
effects.

## Comparison of the flexibility levels

The four MFPI variants are summarized in the table below.

```{r flex-table, echo=FALSE}
flex_table <- data.frame(
  Variant = c("flex1", "flex2", "flex3", "flex4"),
  Description = c(
    "FP powers determined for the main effect of z; use these same powers at each level of t",
    "FP optimized over levels of t; use these same powers at each level of t",
    "As flex2, but re-estimate powers for the main effect of z",
    "Optimize FP for the main effect and separately at each level of t"
  ),
  `d.f. FP1` = c(1, 1, 1, 2),
  `d.f. FP2` = c(2, 2, 2, 4),
  check.names = FALSE
)

knitr::kable(
  flex_table,
  align = c("l", "l", "c", "c"),
  caption = "Variants of MFPI with degrees of freedom for the tests of
  interaction from Royston and Sauerbrei (2013)."
)
```

The choice of functional form --- linear, FP1, or FP2 --- is made via
`cont_var_forms` and should be based on subject-matter knowledge before the
data are examined. Each variable in `cont_vars` is then tested using exactly
its prespecified form. For example, a user may choose FP2 with `flex3` when
there is subject-matter reason to allow a more flexible nonlinear interaction
function while allowing the main-effects and interaction models to select
separate FP powers:

```{r cont-var-forms-flex3, eval=FALSE}
mfpi(
  ...,
  cont_vars      = "age",
  cont_var_forms = c(age = "fp2"),
  flex           = "flex3"
)
```

## Note on nested and non-nested comparisons {#note-on-non-nested-comparisons}

Whether the main-effects and interaction models are nested depends on the FP
basis functions used in the two models. If the same FP powers, and therefore the
same transformed covariate columns, are used in both models, the main-effects
model is nested within the interaction model.

If the FP powers differ between the main-effects and interaction models, the two
models no longer use the same covariate basis. In that case, the main-effects
model is not nested within the interaction model. The likelihood-ratio statistic
may still be computed from the fitted deviances and compared with a chi-square
distribution using the nominal degrees of freedom, but the resulting p-values
lack a firm theoretical underpinning. They should therefore be interpreted
cautiously, because the significance levels may be liberal or conservative
(Royston and Sauerbrei, 2013).

This issue is most relevant for flexibility variants that allow separate power
selection for the main-effects and interaction models, such as `flex3` and
`flex4`. However, the nestedness of a particular fitted comparison should be
judged from the selected FP basis functions, not only from the flexibility label.
For example, `flex3` and `flex4` fits may still give a nested comparison if the
selected main-effects and interaction powers are the same.

As an alternative, AIC or BIC may be used to compare non-nested models, since
these criteria do not require the main-effects model to be nested within the
interaction model. See the
[AIC- and BIC-based selection](#aic-bic) section for details.

# Selection criteria {#selection-criteria}

The `criterion` argument controls both Step 1 adjustment-model selection and the
Step 2 comparison of the main-effects and interaction models.

## P-value-based selection

The default `criterion = "pvalue"` compares the relevant models using a
test statistic and selects an interaction when its final p-value is below
`p_interact`. If multiplicity adjustment is requested, the final decision uses
`p_adjusted`; otherwise, `p_adjusted` equals the raw p-value. For Gaussian
models, `ftest = TRUE` uses the corresponding F-test for nested comparisons
instead of the chi-square approximation.

The reported interaction p-value is the joint test for the prespecified
interaction form. It, rather than an individual coefficient-level Wald p-value,
should be used for the interaction decision.

## AIC- and BIC-based selection {#aic-bic}

As alternatives, `criterion = "aic"` and `criterion = "bic"` use information
criteria in both Step 1 and Step 2. They do not require the main-effects model
to be nested within the interaction model and are therefore useful when
`flex3` or `flex4` produces different FP bases.

For each tested variable, the improvement is

$$
\mathrm{dAIC} = \mathrm{AIC}_{\mathrm{main}} - \mathrm{AIC}_{\mathrm{interaction}},
\qquad
\mathrm{dBIC} = \mathrm{BIC}_{\mathrm{main}} - \mathrm{BIC}_{\mathrm{interaction}}.
$$

A positive value favours the interaction model. The interaction is retained
when the improvement exceeds `min_improvement`. BIC generally penalises model
complexity more strongly than AIC and therefore tends to favour simpler models.
Because these decisions are not based on p-values, `p_adjust_method` does not
affect AIC- or BIC-based selection.

# Grouping variables, labels, and reference levels

## Specifying the grouping variable

The grouping variable is specified by `group_var`, which must be a character
scalar naming a column. In the formula interface, it may be a factor, character,
logical, integer, or numeric variable in the supplied data.

In the default interface, `x` may be a numeric matrix or a data frame. If `x`
is a data frame, only `group_var` may be categorical; the other variables must
be numeric, integer, or logical. This keeps the default interface explicit and
avoids silently dummy-coding adjustment variables.

## Binary and multigroup variables

Binary and multigroup variables are supported. A grouping variable with \(K\)
levels is represented internally by consecutive codes and the required group
indicators.

## Internal coding and preserved labels

`mfpi()` remaps the grouping variable to consecutive integer levels
`0, 1, ..., K - 1`. These codes are used in model matrices, coefficient names,
prediction-column names, and group-specific FP basis construction. Original
labels are stored separately and used in printed output, summaries, plots, and
prediction metadata.

In this vignette, `rx` has factor levels `Placebo` and `Treatment`. They are
mapped internally to 0 and 1, while the original labels appear in group-level
power tables and plot legends. The mapping is available in
`group_level_map`.

## Choosing the reference group

For a factor, the reference is the first element of
`levels(droplevels(group_var))`; users can therefore choose it by setting the
factor levels before fitting. For a character variable, the reference is the
first distinct value encountered in the data. For logical, integer, numeric,
and numeric-matrix input, the reference is the smallest observed value.

# Data handling and transformation options

## Winsorisation {#influential-observations-and-winsorisation}

FP models can be sensitive to outliers or influential observations in the
continuous predictor. A small number of influential observations in the tails of
a variable in `cont_vars` may strongly affect the selected functional form and
may produce spuriously significant interactions. Therefore, it is important to
inspect the data before fitting an MFPI model, for example, using univariable
plots.

In `mfpi()`, Winsorisation is provided as an optional pre-processing step to
reduce the influence of extreme values in variables listed in `cont_vars`.
Values below the lower percentile and above the upper percentile specified by
`winsorize_probs` are replaced by the corresponding percentile cutoffs. The
default is `winsorize_probs = c(0.01, 0.99)`.

The following table summarizes which structural-zero options are available for
variables in `cont_vars` versus adjustment-only variables:

```{r zero-option-table, echo=FALSE}
zero_table <- data.frame(
  Option = c("zero_vars", "catzero_vars", "spike_vars", "Winsorisation"),
  `cont_vars` = c(
    "Yes (positive part FP-transformed; non-positive values set to zero)",
    "No (dropped with warning)",
    "No (dropped with warning)",
    "Yes, unless also in zero_vars"
  ),
  `Adjustment only` = c("Yes", "Yes", "Yes", "Yes, unless in zero/catzero/spike_vars"),
  check.names = FALSE
)

knitr::kable(
  zero_table,
  align = c("l", "l", "l"),
  caption = "Availability of structural-zero and Winsorisation options by variable role."
)
```

Variables in `cont_vars` cannot be listed in `spike_vars` or
`catzero_vars` --- these options are only available for adjustment variables,
because the interaction design matrix does not include group-specific
$I(x = 0)$ indicators. If a variable appears in both `cont_vars` and
`spike_vars` or `catzero_vars`, the spike or catzero treatment is dropped
with a warning.

Variables in `cont_vars` that are also listed in `zero_vars` are excluded from
Winsorisation, because the zero structure of these variables carries substantive
modelling meaning, and truncating the lower tail could distort it. More
generally, variables flagged as `catzero_vars` or `spike_vars` are also excluded
from Winsorisation, though these options cannot currently be used with variables
in `cont_vars`. Excluded variables are recorded with `NA` limits in
`winsorize_limits` for transparency.

Winsorisation is applied after shifting and scaling, so the percentile cutoffs
and the limits stored in `winsorize_limits` are on the transformed scale. The
actual cutoffs applied are returned in the fitted object and can be inspected
via `fit$winsorize_limits`.

Winsorisation is disabled by default (`winsorize = FALSE`). To apply
Winsorisation, set `winsorize = TRUE`. More aggressive truncation can be
requested, for example:

```{r winsorize-prob, eval=FALSE}
winsorize_probs = c(0.05, 0.95)
```

Any Winsorisation or trimming applied should be reported transparently as part
of the initial data analysis.

## Shifting and scaling

Fractional polynomials require strictly positive input values. As in `mfp2()`,
`mfpi()` estimates shift terms to ensure all values of the continuous covariates
are positive before FP transformation. After shifting, `mfpi()` estimates a
scale factor for each variable, placing the shifted covariates on a convenient
numerical range for FP power selection. This improves numerical stability
during the MFP backfitting cycles. Scaling does not affect which FP powers are
selected, only the numerical precision of the fitting algorithm.

After FP powers have been selected, the variables are backscaled before the
final model is fitted. This means the regression coefficients in the returned
model are on the $\phi(z + \text{shift})$ scale, matching what `mfp2()` would
produce. The user does not need to undo the scaling manually.

## Centering

When `center = TRUE`, centering is applied in both the adjustment model and the
interaction model, but the centering constants may differ between the two
models.

**Adjustment model.** Standard `mfp2()` centering is applied. Each
FP-transformed column is centered by subtracting its column mean, that is,
`f(x_j) - mean(f(x_j))`. For binary covariates, centering uses the lower of the
two distinct values rather than the mean, following the same convention as
`mfp2()`.

**Interaction model.** The argument `center_type` controls how the
group-specific FP columns are centered:

- `center_type = "grand"` (default): the grand mean of the FP-transformed
  variable, computed across all observations regardless of group membership, is
  subtracted from each group's columns. Thus, all groups share the same
  centering constant.
- `center_type = "group"`: each group's columns are centered using the mean
  computed from observations in that group only. Thus, different groups may use
  different centering constants.

In both cases, centering changes the decomposition of the fitted model among
the intercept, the group dummy coefficients, and the centered FP components.
It can therefore change the separate numerical values of
$\hat{\alpha}$ and $\hat{f}_1(z)-\hat{f}_0(z)$ while leaving their sum unchanged.
Centering does not affect the interaction test, the displayed fitted-curve
difference
$\hat{\Delta}(z)=\hat{\alpha}+\hat{f}_1(z)-\hat{f}_0(z)$, or the fitted values.

## Non-positive values

Fractional polynomial transformations require positive inputs. For ordinary
continuous variables containing zero or negative values, `mfpi()` estimates a
shift that makes the values positive before FP transformation. The shift and
scale are then applied consistently during fitting and prediction.

Automatic shifting is appropriate when zero and negative values are ordinary
points on the measurement scale. When zero has a distinct scientific meaning,
the structural-zero options below may be more appropriate.

## Structural zeros

The `zero_vars` option specifies variables for which FP transformations are
applied only to strictly positive values. Non-positive values are set to zero in
the transformed variable, and no additional binary indicator is added.

The `catzero_vars` option adds an indicator for whether the variable is zero,
alongside the FP terms for its positive part. It implies `zero_vars`. The
`spike_vars` option invokes the spike-at-zero procedure, which evaluates the
contribution of the zero indicator and the continuous FP component; it implies
both `catzero_vars` and `zero_vars`.

Variables in `cont_vars` cannot currently use `catzero_vars` or `spike_vars`,
because the interaction design does not include group-specific zero indicators.
If a tested variable is supplied in either option, it is removed with a warning.
`zero_vars` can be used for tested variables: only the positive part is FP
transformed, while the remaining values are set to zero.

Variables treated through `zero_vars`, `catzero_vars`, or `spike_vars` are
excluded from Winsorisation where necessary to preserve their zero structure.
For selected adjustment variables, the Step 1 output uses the same
reader-facing labels as `print.mfp2()`, including `catzero_final` and
`saz_decision`.

In the formula interface, approximate cumulative distribution handling is
requested term by term through `fp(..., acdx = TRUE)` or
`fp2(..., acdx = TRUE)`. The argument-level `acd_vars` option is not used in the
formula interface, avoiding ambiguity with term-specific settings.

# Interaction model formulation {#the-interaction-model}

This section provides the mathematical formulation of the models compared in
Step 2 of the MFPI algorithm. It serves as a reference for interpreting fitted
models and plotted curves. Readers primarily interested in practical usage may
wish to consult this section selectively rather than reading it end to end.

For a binary grouping variable, let \(t=0\) denote the reference group and
\(t=1\) the comparison group. Let \(z\) denote the continuous variable being
evaluated for interaction, and let
\(\mathbf{x}_i^\top\boldsymbol{\gamma}\) denote the contribution from its
variable-specific adjustment set. Before FP transformation, \(z\) is shifted
when it contains zero or negative values to ensure positivity and may
optionally be scaled for numerical stability.

The formulation below is written for an FP2 interaction, specified for \(z\)
through `cont_var_forms`, for example:

```r
cont_var_forms = c(z = "fp2")
```

An FP2 function contains two basis terms. Depending on `flex`, the powers
defining these terms may be shared or selected separately between the
main-effects and interaction models and, within the interaction model, between
groups. The same general model structure applies when `cont_var_forms`
specifies `"linear"` or `"fp1"`, but with one functional term rather than two.

Throughout, superscripts \((M)\) and \((I)\) distinguish quantities that are
estimated separately in the main-effects and interaction models, respectively.
That is, \((M)\) denotes the main-effects model and \((I)\) denotes the
interaction model. The intercept, the group main effect, and the adjustment
coefficients are re-fitted in each model and generally take different numerical
values.

## Main-effects model

Let \(\phi_1^{(M)}(z)\) and \(\phi_2^{(M)}(z)\) denote the two FP basis
functions used for \(z\) in the main-effects model. For models with an
intercept, its linear predictor is

\[
\eta_i^{(M)} =
\begin{cases}
\beta_0^{(M)}
+ \mathbf{x}_i^\top\boldsymbol{\gamma}^{(M)}
+ \beta_1^{(M)}\phi_1^{(M)}(z_i)
+ \beta_2^{(M)}\phi_2^{(M)}(z_i),
& \text{if } t_i=0, \\[6pt]
\beta_0^{(M)}
+ \mathbf{x}_i^\top\boldsymbol{\gamma}^{(M)}
+ \alpha^{(M)}
+ \beta_1^{(M)}\phi_1^{(M)}(z_i)
+ \beta_2^{(M)}\phi_2^{(M)}(z_i),
& \text{if } t_i=1.
\end{cases}
\]

Here, \(\beta_0^{(M)}\) is the intercept and \(\alpha^{(M)}\) is the main effect
of the grouping variable. Both groups share the same FP2 function of \(z\) and
differ only by the constant \(\alpha^{(M)}\). Their functions are therefore
parallel on the scale of the linear predictor, and the group difference does
not vary with \(z\).

## Interaction model

In the interaction model, \(z\) is represented by group-specific FP2
functions. Let \(\phi_{j1}^{(I)}(z)\) and \(\phi_{j2}^{(I)}(z)\) denote the two
FP basis functions used for group \(j\) (where \(j\) takes the role of the
group indicator \(t\) in subscript position). The interaction-model linear
predictor is

\[
\eta_i^{(I)} =
\begin{cases}
\beta_0^{(I)}
+ \mathbf{x}_i^\top\boldsymbol{\gamma}^{(I)}
+ \beta_{01}\phi_{01}^{(I)}(z_i)
+ \beta_{02}\phi_{02}^{(I)}(z_i),
& \text{if } t_i=0, \\[6pt]
\beta_0^{(I)}
+ \mathbf{x}_i^\top\boldsymbol{\gamma}^{(I)}
+ \alpha^{(I)}
+ \beta_{11}\phi_{11}^{(I)}(z_i)
+ \beta_{12}\phi_{12}^{(I)}(z_i),
& \text{if } t_i=1.
\end{cases}
\]

## Relationship between `flex` and the FP basis

How the group-specific FP2 functions differ depends on `flex`:

- Under `flex1`, `flex2`, and `flex3`, both groups share the same FP2 basis
  within the interaction model, so \(\phi_{01}^{(I)}=\phi_{11}^{(I)}\) and
  \(\phi_{02}^{(I)}=\phi_{12}^{(I)}\). The two group-specific FP2 functions
  therefore differ only in their coefficients
  \((\beta_{01},\beta_{02})\) versus \((\beta_{11},\beta_{12})\).
- Under `flex4`, the FP2 powers may also differ between groups, so
  \(\phi_{01}^{(I)}\) and \(\phi_{11}^{(I)}\) (and likewise \(\phi_{02}^{(I)}\)
  and \(\phi_{12}^{(I)}\)) need not coincide. The two group-specific FP2
  functions may then differ in both their basis functions and their
  coefficients.

Whether the FP2 basis of the interaction model matches that of the main-effects
model also depends on `flex`:

- Under `flex1` and `flex2`, the same FP2 basis is used in both models, so
  \(\phi_1^{(M)}=\phi_1^{(I)}\) and \(\phi_2^{(M)}=\phi_2^{(I)}\). The
  difference is only in *how* the shared powers are selected: `flex1` selects
  them from the pooled main-effects model, whereas `flex2` selects them from
  an interaction model and then reuses those same powers in the main-effects
  model.
- Under `flex3` and `flex4`, the powers for the main-effects and interaction
  models are selected separately, so the interaction-model basis may differ
  from the main-effects basis.

When the interaction model uses a common FP2 basis in both groups, the
group-specific basis functions coincide:

\[
\phi_{01}^{(I)}(z)
=
\phi_{11}^{(I)}(z)
=
\phi_1(z),
\qquad
\phi_{02}^{(I)}(z)
=
\phi_{12}^{(I)}(z)
=
\phi_2(z).
\]

For any FP2 function with distinct powers \(p_1\) and \(p_2\), the two basis
functions are

\[
\phi_1(z_i)=z_i^{p_1},
\qquad
\phi_2(z_i)=z_i^{p_2},
\]

where \(z^0=\log(z)\) by convention. When the two FP2 powers are repeated,
\(p_1=p_2=p\), the basis functions are

\[
\phi_1(z_i)=z_i^p,
\qquad
\phi_2(z_i)=z_i^p\log(z_i).
\]

The precise rules governing how the FP2 powers are selected and shared are
described in [Flexibility levels](#Flex).

## Fitted curves and displayed differences

For the interaction model, the fitted FP2 contribution of \(z\) is

\[
\hat f^{(I)}(z_i \mid t_i)
=
\begin{cases}
\hat\beta_{01}\phi_{01}^{(I)}(z_i)
+
\hat\beta_{02}\phi_{02}^{(I)}(z_i),
& \text{if } t_i=0, \\[6pt]
\hat\beta_{11}\phi_{11}^{(I)}(z_i)
+
\hat\beta_{12}\phi_{12}^{(I)}(z_i),
& \text{if } t_i=1.
\end{cases}
\]

For brevity, denote these group-specific FP2 contributions by
\(\hat f_0(z_i)\) and \(\hat f_1(z_i)\), respectively.

The FP2 contribution is not, by itself, the complete curve returned by
`predict.mfpi()` and displayed by `plot.mfpi()`. For a model with an
intercept, the group-specific partial fitted curve is

\[
\hat g(z_i \mid t_i)
=
\begin{cases}
\hat\beta_0^{(I)}+\hat f_0(z_i),
& \text{if } t_i=0, \\[6pt]
\hat\beta_0^{(I)}+\hat\alpha^{(I)}+\hat f_1(z_i),
& \text{if } t_i=1.
\end{cases}
\]

Thus, the displayed curve for the comparison group includes the estimated
group main effect \(\hat\alpha^{(I)}\). Cox models have no ordinary regression
intercept, so the \(\hat\beta_0^{(I)}\) term is absent. The adjustment
contribution \(\mathbf{x}_i^\top\hat{\boldsymbol{\gamma}}^{(I)}\) is not added
to the displayed partial curves. If the same adjustment values were added to
both groups, the common contribution would cancel from their difference.

The Step 2 comparison assesses whether the group-specific functions are
parallel on the scale of the linear predictor. Parallel functions imply that
the group difference is constant across \(z\), whereas non-parallel functions
indicate effect modification. The exact model comparison and its interpretation
depend on the selected `flex` and `criterion`.

For Gaussian models, the linear-predictor scale is the outcome scale. For
logistic models, it is the log-odds scale; for Poisson models, it is the
log-rate or log-mean scale; and for Cox models, it is the log-hazard scale.

The fitted-curve difference returned by `predict.mfpi()` and displayed by
`plot.mfpi()` is

\[
\begin{aligned}
\hat\Delta(z_i)
&=
\hat g(z_i \mid t_i=1)
-
\hat g(z_i \mid t_i=0) \\[4pt]
&=
\hat\alpha^{(I)}
+
\hat f_1(z_i)
-
\hat f_0(z_i).
\end{aligned}
\]

The group main effect is therefore included in the displayed difference. A
constant nonzero value of \(\hat\Delta(z)\) represents a group difference that
does not vary with \(z\) and is compatible with parallel functions. Variation
in \(\hat\Delta(z)\) over \(z\) indicates non-parallel functions and therefore
effect modification. The pointwise confidence band around \(\hat\Delta(z)\) is a useful visual aid when judging these patterns.

## Pointwise standard errors

For a fixed value of \(z\), both \(\hat g(z\mid t)\) and \(\hat\Delta(z)\) are
linear combinations of the fitted interaction-model coefficients, of the form
\(a(z)^\top\hat{\boldsymbol{\beta}}\). Their pointwise standard errors are
therefore obtained directly from
\(\operatorname{Var}\!\bigl(a(z)^\top\hat{\boldsymbol{\beta}}\bigr)
= a(z)^\top\hat{\mathbf V}\, a(z)\),
using the variance-covariance matrix \(\hat{\mathbf V}\) of the fitted
interaction model. The resulting confidence intervals are pointwise and
conditional on the selected model (see the [Interpreting confidence intervals](#ci-caveat) section).

# Interpreting fitted models {#interpreting-fitted-models}

## Printed and summary output

`print()` reports the three model-building steps. `summary()` retains those
results and adds the fitted regression output for selected interaction models.
For the adjusted prespecified analysis:

```{r adjusted-summary}
summary(fit_adjusted)
```

The output should be read as a sequence: the selected adjustment model, the
variable-specific comparisons, the final selection decisions, and—when an
interaction is retained—the coefficients of the fitted interaction model.

## The selected adjustment model

Step 1 displays the variables retained by MFP and the selected FP powers or
other applicable settings. Eliminated variables are omitted. The fitted `mfp2`
object is also available as `fit_adjusted$adjustment_model`.

This table describes the global Step 1 model. The exact adjustment set used for
a tested variable is obtained after removing that variable's selected block and
the group dummy columns, as described under [Construct variable-specific adjustment sets](#variable-specific-adjustment-sets).

## Variable-specific interaction models

Step 2 contains one row per variable in `cont_vars`. Depending on `criterion`,
it reports the interaction-test statistic and p-value or the AIC/BIC
improvement. The associated group-level power table records the FP powers used
for each group. Step 3 then gives the final selected/non-selected decision.

For an exploratory analysis, raw and adjusted p-values are shown together when
a multiplicity method is used. The adjusted value determines the final decision.

## Regression coefficients

For each selected interaction, `summary()` prints the regression table from the
underlying fitted model. Coefficient names use internal group codes even though
printed summaries and plots use the original group labels. For an FP2 age
interaction, for example, `age01` and `age02` refer to the two FP terms in
internal group 0, while `age11` and `age12` refer to the corresponding terms in
group 1. A coefficient such as `rx1` represents the comparison-group main
effect relative to the reference group.

```{=html}
<div class="note-box">
<strong>Important:</strong> The coefficient-level Wald p-values do not account
for the degrees of freedom used in FP power selection. The interaction decision
should use the joint test reported in Step 3, not individual Wald p-values.
</div>
```

## Group-specific fitted functions

`plot_type = "fitted"` displays the group-specific partial fitted curves defined
in the [Interaction model formulation](#the-interaction-model). Each curve
includes the model intercept when present, the applicable group main effect,
and the group-specific FP contribution for the plotted variable. Contributions
from other adjustment terms are not added.

```{r adjusted-plot-fitted, fig.cap = "Group-specific partial fitted curves for the adjusted age-by-treatment analysis."}
plot(
  fit_adjusted,
  terms = "age",
  plot_type = "fitted",
  show_ci_fitted = TRUE,
  show_rug = TRUE
)
```

`show_ci_fitted` controls pointwise confidence bands and `show_rug` adds the
observed covariate values.

## Fitted-curve differences

`plot_type = "difference"` displays the comparison-group curve minus the
reference-group curve on the linear-predictor scale. A horizontal nonzero curve
represents a constant group difference without interaction; variation over the
continuous covariate indicates effect modification.

```{r adjusted-plot-difference, fig.cap = "Fitted-curve difference for the adjusted age-by-treatment analysis."}
plot(fit_adjusted, terms = "age", plot_type = "difference")
```

For Cox models the vertical scale is log hazard; for binomial models it is log
odds; for Poisson models it is log rate or log mean; and for Gaussian models it
is the outcome scale. `show_ci_diff` controls the pointwise confidence band.

## Plotting selected and non-selected interactions

With `terms = NULL`, `plot()` displays interactions selected in Step 3. This is
useful for the exploratory model:

```{r exploratory-plot-selected, fig.width = 10, fig.height = 5, fig.cap = "Fitted functions and differences for selected exploratory interactions."}
plot(fit_exp, plot_type = "both")
```

Any tested variable can be inspected explicitly, even when its interaction was
not selected:

```{r exploratory-plot-nonselected, fig.width = 10, fig.height = 5, fig.cap = "Fitted exploratory interaction for weight, regardless of its selection status."}
plot(
  fit_exp,
  terms = "wt",
  plot_type = "both",
  legend_inside = c(0.60, 0.98)
)
```

`plot_type = "both"` combines fitted and difference panels and requires the
`patchwork` package. Original group labels are used in legends and annotations.
The `line_types`, `legend_position`, `legend_inside`, and
`legend_justification` arguments provide additional display control.

# Accessing model components

The object returned by `mfpi()` is a list of class `"mfpi"`. Its components can
be accessed with `$` for programmatic post-processing.

## Adjustment-model components

```{r components-names}
names(fit_adjusted)
```

`adjustment_model` is the complete `mfp2` object from Step 1.
`winsorize_limits` records any applied Winsorisation cutoffs.

```{r components-adjustment, eval=FALSE}
class(fit_adjusted$adjustment_model)
```

## Variable-specific model comparisons

```{r components-metrics, eval=FALSE}
fit_exp$all_model_metrics
```

`all_model_metrics` contains one row per tested variable, including the
prespecified form, selected powers, and the metric used for comparison. FP power
columns are stored as list-columns so the group-level specifications can be
reused programmatically.

## Selected interactions

```{r components-selected, eval=FALSE}
fit_exp$best_model_metrics
```

`best_model_metrics` contains one row per selected interaction. The fitted
object also stores `cont_var_forms` and the per-variable results in
`var_winners`.

## Fitted models

`best_interaction_model` contains models for selected variables, whereas
`all_interaction_models` contains the fitted interaction model for every tested
variable.

```{r components-model, eval=FALSE}
fit_adjusted$best_interaction_model[["age"]]
fit_exp$all_interaction_models[["wt"]]
```

## Plotting data

Plotting data are generated on demand by `predict.mfpi()` when `plot()` is
called; fitted-function matrices do not need to be stored in the fitted object.
The group mapping used for prediction and plotting is available through
`group_levels_original`, `group_levels_new`, and `group_level_map`.

## Programmatic extraction

The complete component names can be inspected with `names()`. The most useful
fields for automated reporting are usually `all_model_metrics`,
`best_model_metrics`, `best_interaction_model`, `all_interaction_models`,
`group_level_map`, `cont_var_forms`, and `var_winners`.

# References

- Hingorani, A.D., Windt, D.A., Riley, R.D., Abrams, K., Moons, K.G.M.,
  Steyerberg, E.W., Schroter, S., Sauerbrei, W., Altman, D.G. and Hemingway,
  H., for the PROGRESS Group (2013). Prognosis research strategy (PROGRESS) 4:
  Stratified medicine research. *BMJ*, 346, e5793. doi:
  [10.1136/bmj.e5793](https://doi.org/10.1136/bmj.e5793)

- Royston, P. and Sauerbrei, W. (2004). A new approach to modelling
  interactions between treatment and continuous covariates in clinical trials by
  using fractional polynomial submodels. *Statistics in Medicine*, 23(16),
  2509--2525. doi:
  [10.1002/sim.1815](https://doi.org/10.1002/sim.1815)

- Sauerbrei, W. and Royston, P. (1999). Building multivariable prognostic
  and diagnostic models: transformation of the predictors by using fractional
  polynomials. *Journal of the Royal Statistical Society: Series A (Statistics
  in Society)*, 162(1), 71--94. doi:
  [10.1111/1467-985X.00122](https://doi.org/10.1111/1467-985X.00122)

- Sauerbrei, W., Royston, P. and Zapien, K. (2007). Detecting an interaction
  between treatment and a continuous covariate: a comparison of two approaches.
  *Computational Statistics & Data Analysis*, 51(8), 4054--4063. doi:
  [10.1016/j.csda.2006.12.041](https://doi.org/10.1016/j.csda.2006.12.041)

- Sauerbrei, W., Royston, P. and Binder, H. (2022). Influence of
  observations on the choice of functional form when using FP-based global
  model tests. *Statistics in Medicine*, 41(15), 2854--2877. doi:
  [10.1002/sim.9390](https://doi.org/10.1002/sim.9390)

- Royston, P. and Sauerbrei, W. (2008). *Multivariable Model-Building: A
  Pragmatic Approach to Regression Analysis Based on Fractional Polynomials for
  Modelling Continuous Variables.* Wiley. doi:
  [10.1002/9780470770771](https://doi.org/10.1002/9780470770771)

- Royston, P. and Sauerbrei, W. (2013). Interaction of treatment with a
  continuous variable: simulation study of significance level for several methods
  of analysis. *Statistics in Medicine*, 32(22), 3788--3803. doi:
  [10.1002/sim.5813](https://doi.org/10.1002/sim.5813)

- Royston, P. and Sauerbrei, W. (2014). Interaction of treatment with a
  continuous variable: simulation study of power for several methods of analysis.
  *Statistics in Medicine*, 33(27), 4695--4708. doi:
  [10.1002/sim.6308](https://doi.org/10.1002/sim.6308)

- Schandelmaier, S., Briel, M., Varadhan, R., Schmid, C.H., Devasenapathy,
  N., Hayward, R.A., Gagnier, J., Borenstein, M., van der Heijden, G.J.M.G.,
  Dahabreh, I.J., Sun, X., Sauerbrei, W., Walsh, M., Ioannidis, J.P.A.,
  Thabane, L., Guyatt, G.H. (2020). Development of the Instrument to assess
  the Credibility of Effect Modification Analyses (ICEMAN) in randomized
  controlled trials and meta-analyses. *CMAJ*, 192(32), E901--E906. doi:
  [10.1503/cmaj.200077](https://doi.org/10.1503/cmaj.200077)

```{r cleanup, include=FALSE}
options(width = oldwidth)
```
